### 2013-04-17 pkg/inst/doc/msm-manual.Rnw
Doc updates.  Forgot to add manual sources to enable rebuilding of the PDF manual on r-forge.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@6 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%%\VignetteIndexEntry{User guide to msm with worked examples}
%\VignettePackage{msm}
\documentclass{article}

%% Need to modify Sweave.sty to pass pdftex option to graphicx.
\usepackage{Sweave-local}
\usepackage{times}
\usepackage{url}
\addtolength{\textwidth}{2cm}

\newcommand{\Exp}{\mathop{\mathrm{Exp}}}
\newcommand{\etal}{{\textit{et al.}}}
\newcommand{\Rfunction}[1]{{\texttt{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textit{#1}}}
\newcommand{\Rmethod}[1]{{\texttt{#1}}}
\newcommand{\Rfunarg}[1]{{\texttt{#1}}}
\newcommand{\Rclass}[1]{{\textit{#1}}}

<<echo=FALSE>>=
version <- gsub("Version: +", "", packageDescription("msm", lib.loc="~/work/msm")$Version)
@

\title{Multi-state modelling with R: the {\tt msm} package \vskip 0.2in
  \large{Version %
<<echo=FALSE,results=tex>>=
cat(version)
@
\vskip 0.1in
<<echo=FALSE,results=tex>>=
cat(format(Sys.time(), "%d %B, %Y"))
@
}}
\author{Christopher Jackson\\Medical Research Council Biostatistics Unit\\Cambridge, U.K.\\ \texttt{chris.jackson@mrc-bsu.cam.ac.uk}}
\date{} 
%% label with date when Rnw is compiled, not when tex is compiled. should be release date of package

\bibliographystyle{unsrt}

\begin{document}

\maketitle

\begin{abstract}

  The multi-state Markov model is a useful way of describing a process
  in which an individual moves through a series of states in
  continuous time.  The \Rpackage{msm} package for R allows a general
  multi-state model to be fitted to longitudinal data. Data often
  consist of observations of the process at arbitrary times, so that
  the exact times when the state changes are unobserved.  For example,
  the progression of chronic diseases is often described by stages of
  severity, and the state of the patient may only be known at doctor
  or hospital visits.  Features of \Rpackage{msm} include the ability
  to model transition rates and hidden Markov output models in terms
  of covariates, and the ability to model data with a variety of
  observation schemes, including censored states.

  Hidden Markov models, in which the true path through states is only
  observed through some error-prone marker, can also be fitted.  The
  observation is generated, conditionally on the underlying states,
  via some distribution.  An example is a screening misclassification
  model in which states are observed with error.  More generally,
  hidden Markov models can have a continuous response, with some
  arbitrary distribution, conditionally on the underlying state.

  This manual introduces the theory behind multi-state Markov and
  hidden Markov models, and gives a tutorial in the typical use of the
  \Rpackage{msm} package, illustrated by some typical applications to
  modelling chronic diseases.

  Much of the material in this manual is published, in a more concise
  form, in Journal of Statistical Software (2011) 38(8):1-29,
  \url{http://www.jstatsoft.org/v38/i08/}

\end{abstract}

\tableofcontents
\clearpage

\section{Multi-state models}
\label{sec:multistate}

\subsection{Introduction}
\label{sec:intro}

Figure~\ref{fig:multi} illustrates a multi-state model in continuous
time. Its four states are labelled {\bf 1, 2, 3, 4}.  At a time $t$
the individual is in state $S(t)$.  The arrows show which transitions
are possible between states. The next state to which the individual
moves, and the time of the change, are governed by a set of {\em
  transition intensities} $q_{rs}(t, z(t))$ for each pair of states
$r$ and $s$.  The intensities may also depend on the time of the
process $t$, or more generally a set of individual-specific or
time-varying explanatory variables $z(t)$.  The intensity represents
the instantaneous risk of moving from state $r$ to state $s$:
\begin{equation}
  \label{eq:multi:intensity}
  q_{rs}(t, z(t))  =  \lim_{\delta t \rightarrow 0} P (S(t+\delta t) = s | S(t) = r) / \delta t
\end{equation}
The intensities form a matrix $Q$ whose rows sum to zero, so that the
diagonal entries are defined by $q_{rr} = - \sum_{s \neq r} q_{rs}$.

To fit a multi-state model to data, we estimate this transition
intensity matrix.  We concentrate on {\em Markov} models here. The
Markov assumption is that future evolution only depends on the current
state.  That is, $q_{rs}(t, z(t), \mathcal{F}_t)$ is independent of
$\mathcal{F}_t$, the observation history $\mathcal{F}_t$ of the
process up to the time preceding $t$.  See, for example, Cox and
Miller\cite{cox:miller} for a thorough
introduction to the theory of continuous-time Markov chains.

\begin{figure}[p]
  \begin{center}
    \scalebox{0.4}{\includegraphics{figures/multistate}}
    \[
    Q = \left(
      \begin{array}{llll}
        q_{11} & q_{12} & q_{13} & q_{14}\\
        q_{21} & q_{22} & q_{23} & q_{24}\\
        q_{31} & q_{32} & q_{33} & q_{34}\\
        q_{41} & q_{42} & q_{43} & q_{44}\\
      \end{array}
    \right )
    \]
    \caption{\label{fig:multi}General multi-state model.}
  \end{center}
\end{figure}

In a continuous-time Markov model, a single period of occupancy (or
\emph{sojourn time}) in state $r$ has an exponential distribution,
with rate given by $-q_{rr}$, (or
mean $-1 / q_{rr}$).  The remaining elements of the $r$th row of $Q$
are \emph{proportional to} the probabilities governing the next state
after $r$ to which the individual makes a transition.  The probability
that the individual's next move from state $r$ is to state $s$ is
$-q_{rs} / q_{rr}$.

\subsection{Disease progression models}

The development of the \Rpackage{msm} package was motivated by
applications to disease modelling.  Many chronic diseases have a
natural interpretation in terms of staged progression. Multi-state
Markov models in continuous time are often used to model the course of
diseases. A commonly-used model is illustrated in
Figure~\ref{fig:disease}.  This represents a series of successively
more severe disease stages, and an `absorbing' state, often death.
The patient may advance into or recover from adjacent disease stages,
or die at any disease stage.  Observations of the state $S_i(t)$ are
made on a number of individuals $i$ at arbitrary times $t$, which may
vary between individuals.  The stages of disease may be modelled as a
homogeneous continuous-time Markov process, with a transition matrix
$Q$, pictured below Figure~\ref{fig:disease}.

A commonly-used model is the \emph{illness-death} model, with three
states representing health, illness and death
(Figure~\ref{fig:multi:illdeath}).  Transitions are permitted from
health to illness, illness to death and health to death.  Recovery
from illness to health is sometimes also considered.

A wide range of medical situations have been modelled using
multi-state methods, for example, screening for abdominal aortic
aneurysms (Jackson \etal\cite{jackson:sharples:2003}),
problems following lung transplantation (Jackson and
Sharples\cite{jackson:sharples:2002}), problems following heart
transplantation (Sharples\cite{sharp:gibbs}, Klotz and
Sharples\cite{klotz:est}), hepatic cancer (Kay\cite{kay:mark}), HIV
infection and AIDS (Longini \etal\cite{long1}, Satten and
Longini\cite{sattlong}, Guihenneuc-Jouyaux \etal\cite{rich},
Gentleman \etal\cite{gentlaw}), diabetic complications
(Marshall and Jones\cite{retino2}, Andersen\cite{andersen88}), breast
cancer screening (Duffy and Chen\cite{duffy95}, Chen \emph{et
  al.}\cite{duffy96}), cervical cancer screening (Kirby and
Spiegelhalter\cite{kirby:spiegelhalter}) and liver cirrhosis (Andersen
\etal\cite{ander:proth}).  Many of these references also
describe the mathematical theory, which will be reviewed in the
following sections.

\begin{figure}[p]
  \centering
  \vskip 1cm
  \scalebox{1.0}{\includegraphics{figures/general}}
    \[
    Q = \left(
      \begin{array}{llllll}
        q_{11} & q_{12} & 0 & 0 & \ldots & q_{1n}\\
        q_{21} & q_{22} & q_{23} & 0 & \ldots & q_{2n}\\
        0 & q_{32} & q_{33} & q_{34} & \ddots & q_{3n}\\
        0 & 0 & q_{43} & q_{44} & \ddots & q_{4n}\\
        \vdots & \vdots & \ddots & \ddots & \ddots & \vdots\\
        0 & 0 & 0 & 0 & \ldots & 0\\
      \end{array}
    \right )
    \]
  \caption{\label{fig:disease}General model for disease progression.}
  \end{figure}

\begin{figure}[p]
  \begin{center}
    \scalebox{0.4}{\includegraphics{figures/illdeath}}
    \caption{Illness-death model.}
    \label{fig:multi:illdeath}
  \end{center}
\end{figure}

\subsection{Arbitrary observation times}
\label{sec:arbitr-observ-times}

Longitudinal data from monitoring disease progression are often
incomplete in some way.  Usually patients are seen at intermittent
follow-up visits, at which monitoring information is collected, but
information from the periods between visits is not available.  Often
the exact time of disease onset is unknown. Thus, the changes of state
in a multi-state model usually occur at unknown times.  Also a subject
may only be followed up for a portion of their disease history.  A
fixed observation schedule may be specified in advance, but in
practice times of visits may vary due to patient and hospital
pressures. The states of disease progression models often include
death. Death times are commonly recorded to within a day.  Also
observations may be censored.  For example, at the end of a study, an
individual may be known only to be alive, and in an unknown state.

A typical sampling situation is illustrated in
Figure~\ref{fig:multi:sampling}. The individual is observed at four
occasions through 10 months. The final occasion is the death date
which is recorded to within a day. The only other information
available is the occupation of states 2, 2, and 1 at respective times
1.5, 3.5 and 5. The times of movement between states and the state
occupancy in between the observation times are unknown.  Although the
patient was in state 3 between 7 and 9 months this was not observed at
all.

\paragraph{Informative sampling times}
To fit a model to longitudinal data with arbitrary sampling times we
must consider the reasons why observations were made at the given
times.  This is analogous to the problem of missing data, where the
fact that a particular observation is missing may implicitly give
information about the value of that observation. Possible observation
schemes include:
\begin{itemize}
\item {\em fixed}. Each patient is observed at fixed intervals
  specified in advance.
\item {\em random}. The sampling times vary randomly, independently
  of the current state of the disease.
\item {\em doctor's care}. More severely ill patients are monitored
  more closely. The next sampling time is chosen on the basis of the
  current disease state.
\item patient {\em self-selection}. A patient may decide to visit the
  doctor on occasions when they are in a poor condition.
\end{itemize}
Gr\"uger \etal~\cite{gruger:valid} discussed conditions under
which sampling times are \emph{informative}. If a multi-state model is
fitted, ignoring the information available in the sampling times, then
inference may be biased.  Mathematically, because the sampling
times are often themselves random, they should be modelled along with
the observation process $X_t$. However the ideal situation is where
the joint likelihood for the times and the process is proportional to
the likelihood obtained if the sampling times were fixed in advance.
Then the parameters of the process can be estimated independently of
the parameters of the sampling scheme.

In particular, they showed that fixed, random and doctor's care
observation policies are not informative, whereas patient
self-selection is informative.   Note that \Rpackage{msm} does not 
deal with informative sampling times. See, e.g. \cite{sweeting:inform:msm}
for some methods in this case, which require specialised programming.

\begin{figure}[p]
  \begin{center}
    \scalebox{0.6}{\rotatebox{270}{\includegraphics{figures/sampling}}}
    \caption{Evolution of a multi-state model. The process is observed
      on four occasions.}
    \label{fig:multi:sampling}
  \end{center}
\end{figure}

\subsection{Likelihood for the multi-state model}
\label{sec:multi:likelihood}

Kalbfleisch and Lawless\cite{kalbfleisch:lawless} and later
Kay~\cite{kay:mark} described a general method for evaluating the
likelihood for a general multi-state model in continuous time,
applicable to any form of transition matrix. The only available
information is the observed state at a set of times, as in
Figure~\ref{fig:multi:sampling}.  The sampling times are assumed to be
non-informative.

\paragraph{Transition probability matrix}
The likelihood is calculated from the \emph{transition probability matrix}
$P(t)$.  For a time-homogeneous process, the $(r,s)$ entry of $P(t)$,
$p_{rs}(t)$, is the probability of being in state $s$ at a time $t+u$
in the future, given the state at time $u$ is $r$.  It does not say
anything about the time of transition from $r$ to $s$, indeed the
process may have entered other states between times $u$ and $t+u$.
$P(t)$ can be calculated by taking the matrix exponential of the
scaled transition intensity matrix (see, for example, Cox and
Miller~\cite{cox:miller}).
\begin{equation}
  \label{eq:exptq}
  P(t) = \Exp(tQ)
\end{equation}

The matrix exponential $\Exp$ is different from a scalar exponential.
The exponential of a matrix is defined by the same "power series"
$\Exp(X) = 1 + X^2/2! + X^3/3! + ...$ as the scalar exponential,
except that each term $X^k$ in the series is defined by matrix products,
not element-wise scalar multiplication.  It is notoriously difficult
to calculate reliably, as discussed by Moler and van Loan~\cite{matrixexp}.
For simpler models, it is feasible to calculate an analytic expression
for each element of $P(t)$ in terms of $Q$.  This is generally faster
and avoids the potential numerical instability of calculating the
matrix exponential.  Symbolic algebra sofware, such as Mathematica,
can be helpful for obtaining these expressions.  For example, the
three-state illness-death model with no recovery has a transition
intensity matrix of
\[
Q = \left(
  \begin{array}{llllll}
    -(q_{12} + q_{13}) & q_{12} & q_{13}\\
    0 & - q_{23} & q_{23}\\
    0 & 0 & 0\\
  \end{array}
\right )
\]
The corresponding time $t$ transition probabilities are
\begin{eqnarray*}
  p_{11}(t) & = & e^{-(q_{12} + q_{13})t}\\
  p_{12}(t) & = &
\left\{
\begin{array}{ll}
\frac{ q_{12}(-1 + e^{(q_{12}+q_{13}-q_{23})t})e^{-(q_{12}+q_{13})t} } { (q_{12} + q_{13} - q_{23}) } & (q_{12} + q_{13} \neq q_{23})\\
q_{12}te^{(-(q_{12} + q_{13})t} &  (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{13}(t) & = &
\left\{
\begin{array}{ll}
\frac{1 + (-q_{13} + q_{23})e^{-(q_{12}+q_{13})t}} {q_{12} + q_{13} - q_{23}} - \frac { q_{12}e^{-q_{23}t}} {q_{12} + q_{13} - q_{23} } & (q_{12} + q_{13} \neq q_{23})\\
 (-1 + e^{(q_{12} + q_{13})t} - q_{12}t)e^{-(q_{12} + q_{13})t} & (q_{12} + q_{13} = q_{23})
\end{array}
\right.
\\
  p_{21}(t) & = & 0\\
  p_{22}(t) & = & e^{-q_{23}t}\\
  p_{23}(t) & = & 1 - e^{-q_{23}t}\\
  p_{31}(t) & = & 0\\
  p_{32}(t) & = & 0\\
  p_{33}(t) & = & 1\\
\end{eqnarray*}
The \Rpackage{msm} package calculates $P(t)$ analytically for selected
2, 3, 4 and 5-state models. These models are illustrated in
Figures~\ref{fig:anp2}--\ref{fig:anp5}.  For other models, $P(t)$ is
determined from the matrix exponential, calculated using eigensystem
decomposition (if eigenvalues are distinct) or a method based on
Pad\'e approximants with scaling and squaring~\cite{matrixexp} (if
there are repeated eigenvalues). Notice that the states are not
labelled in these figures.  Each graph can correspond to several
different $Q$ matrices, depending on how the states are labelled.  For
example, Figure 1 a) illustrates the model defined by either \( Q =
\left(
  \begin{array}{ll}
    - q_{12} & q_{12} \\
    0 & 0
\end{array}
\right)
\)
or
\(
Q = \left( \begin{array}{ll}
0 & 0\\
q_{21} & - q_{21}
\end{array}
\right)
\).

\begin{figure}
\begin{center}
a) \includegraphics[width=3cm]{figures/p2q1} \hskip 3cm
b) \includegraphics[width=3cm]{figures/p2q12}
\end{center}
\caption{\label{fig:anp2}Two-state models fitted using analytic $P(t)$ matrices in \Rpackage{msm}. Implemented for all permutations of state labels 1, 2.}
\end{figure}
\begin{figure}
\begin{center}
a) \includegraphics[width=3cm]{figures/p3q12} \hskip 3cm
b) \includegraphics[width=5cm]{figures/p3q14} \vskip 1cm
c) \includegraphics[width=3cm]{figures/p3q16}\hskip 3cm
d) \includegraphics[width=3cm]{figures/p3q124}\vskip 1cm
e) \includegraphics[width=3cm]{figures/p3q135}\hskip 3cm
f) \includegraphics[width=3cm]{figures/p3q1246}
\end{center}
\caption{\label{fig:anp3}Three-state models fitted using analytic $P(t)$ matrices in \Rpackage{msm}. Implemented for all permutations of state labels 1, 2, 3.}
\end{figure}
\begin{figure}
\begin{center}
a) \includegraphics[width=7cm]{figures/p4q159} \vskip 1cm
b) \includegraphics[width=7cm]{figures/p4q13569}
\end{center}
\caption{\label{fig:anp4}Four-state models fitted using analytic $P(t)$ matrices in \Rpackage{msm}. Implemented for all permutations of state labels 1, 2, 3, 4.}
\end{figure}
\begin{figure}
\begin{center}
a) \includegraphics[width=9cm]{figures/p5q1_6_11_16}\vskip 1cm
b) \includegraphics[width=9cm]{figures/p5q1_4_6_8_11_12_16}\vskip 1cm
c) \includegraphics[width=5cm]{figures/p5q1_6_7_11_12}
\end{center}
\caption{\label{fig:anp5}Five-state models fitted using analytic $P(t)$ matrices in \Rpackage{msm}. Implemented for all permutations of state labels 1, 2, 3, 4, 5.}
\end{figure}

\paragraph{Likelihood for intermittently-observed processes}
Suppose $i$ indexes $M$ individuals.  The data for individual $i$
consist of a series of times $(t_{i1}, \ldots, t_{in_i})$ and
corresponding states $(S(t_{i1}), \ldots, S(t_{in_i}))$.  Consider a
general multi state model, with a pair of successive observed disease
states $S(t_j), S(t_{j+1})$ at times $t_j, t_{j+1}$. The contribution
to the likelihood from this pair of states is

\begin{equation}
  \label{eq:multi:lik:contrib}
  L_{i, j} =  p_{S(t_j)S(t_{j+1})}(t_{j+1} - t_j)
\end{equation}
This is the entry of the transition matrix $P(t)$ at the $S(t_j)$th
row and $S(t_{j+1})$th column, evaluated at $t = t_{j+1} - t_j$.

The full likelihood $L(Q)$ is the product of all such terms $L_{i,j}$
over all individuals and all transitions. It depends on the unknown
transition matrix $Q$, which was used to determine $P(t)$.

\paragraph{Exactly-observed death times}
In observational studies of chronic diseases, it is common that the
time of death is known, but the state on the previous instant before
death is unknown. If $S(t_{j+1}) = D $ is such a death state, then the
contribution to the likelihood is summed over the unknown state $m$ on
the instant before death:

\begin{equation}
  \label{eq:multi:lik:death}
  L_{i, j} =  \sum_{m \neq D} p_{S(t_j),m}(t_{j+1} - t_j) q_{m, D}
\end{equation}
The sum is taken over all possible states $m$ which can be visited
between $S(t_j)$ and $D$.

\paragraph{Exactly observed transition times}
If the times $(t_{i1}, \ldots, t_{in_i})$ had been the {\em exact}
transition times between the states, with no transitions between the
observation times, then the contributions would be of the form

\begin{equation}
  \label{eq:multi:lik:exact}
  L_{i, j} =  \exp(-q_{S(t_j)S(t_{j})}(t_{j+1} - t_j)) q_{S(t_j)S(t_{j+1})}
\end{equation}
since the state is assumed to be $S(t_j)$ throughout the interval
between $t_j$ and $t_{j+1}$ with a known transition to state
$S(t_{j+1})$ at $t_{j+1}$.  \Rpackage{msm} is restricted to Markov
models, but much richer models are possible for this type of data. For
example, Putter \etal~\cite{putter:mstate} discussed the
\Rpackage{mstate} software for semi-parametric multi-state models with
non-parametric baseline hazards and Cox regression.  The Markov
assumption is restrictive but necessary in general to compute a
likelihood for intermittently-observed processes.

\paragraph{Censored states}
A censored quantity is one whose exact value is unknown, but known to
be in a certain interval.  For example, in survival analysis, a death
time is \emph{right-censored} if the study ends and the patient is
still alive, since the death time is known to be greater than the end
time.  In multi-state models for intermittently-observed processes,
the times of changes of state are usually \emph{interval censored},
known to be within bounded intervals.  This leads to a likelihood
based on equation~\ref{eq:multi:lik:contrib}.

In some circumstances, \emph{states} may be censored as well as
\emph{event times}.  For example, at the end of some chronic disease
studies, patients are known to be alive but in an \emph{unknown
  state}.  For such a censored observation $S(t_{j+1})$, known only to
be a state in the set $C$, the equivalent contribution to the
likelihood is
\begin{equation}
  \label{eq:multi:lik:deathcens}
  L_{i, j} =  \sum_{m \in C} p_{S(t_j),m}(t_{j+1} - t_j)
\end{equation}
Note that this special likelihood is not needed if the state is known
at the end of the study.  In this case,
likelihood~\ref{eq:multi:lik:contrib} applies.  Although the
\emph{survival time} is censored, the \emph{state} at the end of the
study is not censored.

More generally, suppose every observation from a particular individual
is censored.  Observations $1, 2, \ldots n_i$ are known only to be in
the sets $C_1, C_2, \ldots, C_{n_i}$ respectively. The likelihood for
this individual is a sum of the likelihoods of all possible paths
through the unobserved states.
\begin{equation}
  \label{eq:multi:lik:cens}
  L_i = \sum_{s_{n_i} \in C_{n_i}} \ldots \sum_{s_2 \in C_2} \sum_{s_1 \in C_1} p_{s_1 s_2}(t_2 - t_1) p_{s_2 s_3} (t_3 - t_2) \ldots p_{s_{n_i-1} s_{n_i}} (t_{n_i} - t_{n_i-1})
\end{equation}
Suppose the states comprising the set $C_j$ are $c^{(j)}_1, \ldots,
c^{(j)}_{m_j}$. This likelihood can also be written as a matrix
product, say,
\begin{equation}
  \label{eq:multi:lik:cens:matrix}
  L_i = \mathbf{1}^T P^{1,2} P^{2,3} \ldots P^{n_i-1, n_i} \mathbf{1}
\end{equation}
where $P^{j-1, j}$ is a $m_{j-1} \times m_j$ matrix with $(r,s)$ entry
$p_{c^{(j-1)}_r c^{(j)}_s}(t_j - t_{j-1})$, and $\mathbf{1}$ is the
vector of ones.

The \Rpackage{msm} package allows multi-state models to be fitted to
data from processes with arbitrary observation times (panel data), exactly-observed
transition times, exact death times and censored states, or a mixture
of these schemes.

\subsection{Covariates}
\label{sec:multi:covariates}

The relation of constant or time-varying characteristics of
individuals to their transition rates is often of interest in a
multi-state model.  Explanatory variables for a particular transition
intensity can be investigated by modelling the intensity as a function
of these variables.  Marshall and Jones~\cite{retino2} described a
form of a {\em proportional hazards} model, where the transition
intensity matrix elements $q_{rs}$ which are of interest can be
replaced by
\[
q_{rs}(z(t)) = q_{rs}^{(0)} \exp(\beta_{rs}^T z(t))
\]
The new $Q$ is then used to determine the likelihood. If the
covariates $z(t)$ are time dependent, the contributions to the
likelihood of the form $p_{rs} (t - u)$ are replaced by
\[
p_{rs}(t - u, z(u))
\]
although this requires that the value of the covariate is known at
every observation time $u$. Sometimes covariates are observed at
different times to the main response, for example recurrent disease
events or other biological markers. In some of these cases it could be
assumed that the covariate is a step function which remains constant
between its observation times.  If the main response (the state of the
Markov process) is not observed at the times when the covariate
changes, it could be considered as a "censored" state (as in
Section~\ref{sec:multi:likelihood}).

The \Rpackage{msm} package allows individual-specific or time
dependent covariates to be fitted to transition intensities.  In order
to calculate transition probabilities $P(t)$ on which the likelihood
depends, time-dependent covariates are assumed to be
piecewise-constant.  Models whose intensities change with time are
called \emph{time-inhomogeneous}. An important special case handled by
\Rpackage{msm} is the model in which intensities change at a series of
times common to each individual.

Marshall and Jones~\cite{retino2} described likelihood ratio and Wald
tests for covariate selection and testing hypotheses, for example
whether the effect of a covariate is the same for all forward
transitions in a disease progression model, or whether the effect on
backward transitions is equal to minus the effect on forward
transitions.

\subsection{Hidden Markov models}
\label{sec:hmm}

In a {\em hidden Markov model} (HMM) the states of the Markov chain
are not observed. The observed data are governed by some probability
distribution (the \emph{emission} distribution) conditionally on the
unobserved state.  The evolution of the underlying Markov chain is
governed by a transition intensity matrix $Q$ as before.
(Figure~\ref{fig:multi:hidden}).  Hidden Markov models are mixture
models, where observations are generated from a certain number of
unknown distributions.  However the distribution changes through time
according to states of a hidden Markov chain.  This class of model is
commonly used in areas such as speech and signal processing
\cite{juang:rabiner} and the analysis of biological sequence data
\cite{biolog:seq}. In engineering and biological sequencing
applications, the Markov process usually evolves over an
equally-spaced, discrete `time' space.  Therefore most of the theory
of HMM estimation was developed for discrete-time models.

HMMs have less frequently been used in medicine, where continuous time
processes are often more suitable.  A disease process evolves in
continuous time, and patients are often monitored at irregular and
differing intervals.  These models are suitable for estimating
population quantities for chronic diseases which have a natural staged
interpretation, but which can only be diagnosed by an error-prone
marker.  The \Rpackage{msm} package can fit continuous-time hidden
Markov models with a variety of emission distributions.

\begin{figure}[htbp]
  \begin{center}
    \scalebox{0.6}{\rotatebox{270}{\includegraphics{figures/hidden}}}
    \caption{A hidden Markov model in continuous time. Observed states
      are generated conditionally on an underlying Markov process. }
    \label{fig:multi:hidden}
  \end{center}
\end{figure}

\subsubsection{Misclassification models}

An example of a hidden Markov model is a multi-state model with
misclassification.  Here the observed data are states, assumed to be
misclassifications of the true, underlying states.

For example, consider a disease progression model with at least a
disease-free and a disease state. When screening for the presence of
the disease, the screening process can sometimes be subject to error.
Then the Markov disease process $S_i(t)$ for individual $i$ is not
observed directly, but through realisations $O_i(t)$.  The quality of
a diagnostic test is often measured by the probabilities that the true
and observed states are equal, $Pr(O_i(t) = r | S_i(t) = r)$. Where
$r$ represents a `positive' disease state, this is the {\em
  sensitivity}, or the probability that a true positive is detected by
the test.  Where $r$ represents a `negative' or disease-free state,
this represents the {\em specificity}, or the probability that, given
the condition of interest is absent, the test produces a negative
result.

As an extension to the simple multi-state model described in
section~\ref{sec:multistate}, the \Rpackage{msm} package can fit a
general multi-state model with misclassification.  For patient $i$,
observation time $t_{ij}$, observed states $O_{ij}$ are generated
conditionally on true states $S_{ij}$ according to a {\em
  misclassification matrix} $E$. This is a $n \times n$ matrix, whose
$(r,s)$ entry is
\begin{equation}
  \label{eq:misc}
  e_{rs} = Pr(O(t_{ij}) = s | S(t_{ij}) = r),
\end{equation}
which we first assume to be independent of time $t$. Analogously to
the entries of $Q$, some of the $e_{rs}$ may be fixed to reflect
knowledge of the diagnosis process. For example, the probability of
misclassification may be negligibly small for non-adjacent states of
disease.  Thus the progression through underlying states is governed
by the transition intensity matrix $Q$, while the observation process
of the underlying states is governed by the misclassification matrix
$E$.

To investigate explanatory variables $w(t)$ for the probability $e_{rs}$ of
misclassification as state $s$ given underlying state $r$, a
multinomial logistic regression model can be used:
\begin{equation}
  \label{eq:misccovs}
  \log \frac{e_{rs}(t)}{e_{rs_0}(t)}  =   \gamma_{rs}^T w(t).
\end{equation}
where $s_0$ is some baseline state, usually chosen as the underlying
state, or the state with the highest probability (for numerical
stability).

\subsubsection{General hidden Markov model}
\label{sec:hmm:general}

Consider now a general hidden Markov model in continuous time.  The
true state of the model $S_{ij}$ evolves as an unobserved Markov
process.  Observed data $y_{ij}$ are generated conditionally true
states $S_{ij} = 1, 2, \ldots, n$ according to a set of distributions
$f_1(y | \theta_1, \gamma_1)$, $f_2(y | \theta_2, \gamma_2)$,
$\ldots$, $f_n(y | \theta_n, \gamma_n)$, respectively.  $\theta_r$ is
a vector of parameters for the state $r$ distribution.  One or more of
these parameters may depend on explanatory variables through a
link-transformed linear model with coefficients $\gamma_r$.

\subsubsection{Likelihood for general hidden Markov models}

A type of EM algorithm known as the {\em Baum-Welch} or {\em
  forward-backward} algorithm \cite{baum:petrie66, baum:petrie70}, is
commonly used for hidden Markov model estimation in discrete-time
applications.  See, for example, Durbin \etal
\cite{biolog:seq}, Albert \cite{albert99}.  A generalisation of this
algorithm to continuous time was described by Bureau \etal
\cite{bureau:hughes:shiboski:00}.

The \Rpackage{msm} package uses a direct method of calculating
likelihoods in discrete or continuous time based on matrix products.
This type of method has been described by Macdonald and Zucchini
\cite[pp.  77--79]{macdonald:zucchini}, Lindsey
\cite[p.73]{lindsey:rm} and Guttorp \cite{guttorp}.  Satten and
Longini \cite{sattlong} used this method to calculate likelihoods for
a hidden Markov model in continuous time with observations of a
continuous marker generated conditionally on underlying discrete
states.

Patient $i$'s contribution to the likelihood is
\begin{eqnarray}
  \label{eq:multi:hiddencontrib}
  L_i     & = & Pr(y_{i1}, \ldots, y_{in_i})\\
  & = & \sum Pr(y_{i1}, \ldots, y_{in_i} | S_{i1}, \ldots, S_{in_i})
  Pr(S_{i1}, \ldots, S_{in_i}) \nonumber
\end{eqnarray}
where the sum is taken over all possible paths of underlying states
$S_{i1}, \ldots, S_{in_i}$.  Assume that the observed states are
conditionally independent given the values of the underlying states.
Also assume the Markov property, $Pr(S_{ij}|S_{i,j-1}, \ldots, S_{i1})
= Pr(S_{ij}|S_{i,j-1})$.  Then the contribution $L_i$ can be written
as a product of matrices, as follows. To derive this matrix product,
decompose the overall sum in equation~\ref{eq:multi:hiddencontrib}
into sums over each underlying state. The sum is accumulated over the
unknown first state, the unknown second state, and so on until the
unknown final state:
\begin{eqnarray}
  \label{eq:multi:hiddenlik}
  L_i & = & \sum_{S_{i1}} Pr(y_{i1}|S_{i1})Pr(S_{i1}) \sum_{S_{i2}} Pr(y_{i2}|S_{i2})Pr(S_{i2}|S_{i1}) \sum_{S_{i3}} Pr(y_{i3}|S_{i3}) Pr(S_{i3}|S_{i2})
  \nonumber \\
  &   & \ldots \sum_{S_{in_i}} Pr(y_{in_i}|S_{in_i})
  Pr(S_{in_i}|S_{in_{i-1}})
\end{eqnarray}
where $Pr(y_{ij}|S_{ij})$ is the emission probability density.  For
misclassification models, this is the misclassification probability
$e_{S_{ij} O_{ij}}$.  For general hidden Markov models, this is the
probability density
$f_{S_{ij}}(y_{ij}|\theta_{S_{ij}},\gamma_{S_{ij}})$.
$Pr(S_{i,j+1}|S_{ij})$ is the $(S_{ij}, S_{i,j+1})$ entry of the
Markov chain transition matrix $P(t)$ evaluated at $t = t_{i,j+1} -
t_{ij}$.  Let $f$ be the vector with $r$ element the product of the
initial state occupation probability $Pr(S_{i1}=r)$ and $Pr(y_{i1}|
r)$, and let $\mathbf 1$ be a column vector consisting of ones. For $j
= 2, \ldots, n_i$ let $T_{ij}$ be the $R \times R$ matrix (where $R$
is the number of states) with $(r,s)$ entry
\[
Pr(y_{ij}| s) p_{rs} (t_{ij} - t_{i,j-1})
\]
Then subject $i$'s likelihood contribution is
\begin{equation}
  L_i =  f T_{i2} T_{i3}, \ldots T_{in_i} \mathbf 1
  \label{eq:multi:hidden:matprod}
\end{equation}

If $S(t_{j}) = D$ is an absorbing state such as death, measured
without error, whose entry time is known exactly, then the
contribution to the likelihood is summed over the unknown state at the
previous instant before death.  The $(r,s)$ entry of $T_{ij}$ is then
\[
p_{rs}(t_{j} - t_{j-1}) q_{s, D}
\]
Section~\ref{sec:fitting:hmm:misc} describes how to fit multi-state
models with misclassification using the \Rpackage{msm} package, and
Section~\ref{sec:fitting:hmm:general} describes how to apply general
hidden Markov models.

\subsubsection{Example of a general hidden Markov model}
\label{sec:hmm:example:fev}

Jackson and Sharples~\cite{jackson:sharples:2002} described a model
for FEV$_1$ (forced expiratory volume in 1 second) in recipients of
lung transplants.  These patients were at risk of BOS (bronchiolitis
obliterans syndrome), a progressive, chronic deterioration in lung
function.  In this example, BOS was modelled as a discrete, staged
process, a model of the form illustrated in Figure~\ref{fig:disease},
with 4 states.  State 1 represents absence of BOS. State 1 BOS is
roughly defined as a sustained drop below 80\% below baseline FEV$_1$,
while state 2 BOS is a sustained drop below 65\% baseline.  FEV$_1$ is
measured as a percentage of a baseline value for each individual,
determined in the first six months after transplant, and assumed to be
100\% baseline at six months.

As FEV$_1$ is subject to high short-term variability due to acute
events and natural fluctuations, the exact BOS state at each
observation time is difficult to determine.  Therefore, a hidden
Markov model for FEV$_1$, conditionally on underlying BOS states, was
used to model the natural history of the disease.  Discrete states are
appropriate as onset is often sudden.

\paragraph{Model 1}
Jackson~\cite{my:phd} considered models for these data where
FEV$_1$ were Normally distributed, with an unknown mean and variance
conditionally each state~(\ref{eq:fev:normal}).  This model seeks the
most likely location for the within-state FEV$_1$ means.

\begin{equation}
  \label{eq:fev:normal}
  y_{ij} | \{ S_{ij} = k\}  \sim N(\mu_k + \beta x_{ij}, \sigma^2_k)
\end{equation}

\paragraph{Model 2}
Jackson and Sharples~\cite{jackson:sharples:2002} used a more complex
two-level model for FEV$_1$ measurements. Level
1~(\ref{eq:fev:level1}) represents the short-term fluctuation error of
the marker around its underlying continuous value $y^{hid}_{ij}$.
Level 2~(\ref{eq:fev:level2}) represents the distribution of
$y^{hid}_{ij}$ conditionally on each underlying state, as follows.
\begin{equation}
  \label{eq:fev:level1}
  y_{ij} | y^{hid}_{ij} \qquad \sim  N ( y^{hid}_{ij} + \beta x_{ij} ,
  \sigma^2_\epsilon)
\end{equation}
\begin{equation}
  \label{eq:fev:level2}
  y^{hid}_{ij} | S_{ij} \quad \sim \quad
  \left\{
    \begin{array}{cll}
      \mbox{State}& \mbox{Three state model}               & \mbox{Four state model} \\
      S_{ij} = 0  & N(\mu_0, \sigma^2_0)I_{[80, \infty)} & N(\mu_0, \sigma^2_0)I_{[80, \infty)} \\
      S_{ij} = 1  & N(\mu_1, \sigma^2_1)I_{(0, 80)}      & Uniform(65, 80)                        \\
      S_{ij} = 2  & \mbox{(death)}                         & N(\mu_2, \sigma^2_2)I_{(0, 65)}      \\
      S_{ij} = 3  &                                        & \mbox{(death)}
    \end{array}
  \right .
\end{equation}

Integrating over $y^{hid}_{ij}$ gives an explicit distribution for
$y_{ij}$ conditionally on each underlying state (given in
Section~\ref{sec:fitting:hmm:general}, Table~\ref{tab:hmm:dists}).
Similar distributions were originally applied by Satten and
Longini~\cite{sattlong} to modelling the progression through discrete,
unobserved HIV disease states using continuous CD4 cell counts.  The
\Rpackage{msm} package includes density, quantile, cumulative density
and random number generation functions for these distributions.

In both models 1 and 2, the term $\beta x_{ij}$ models the short-term
fluctuation of the marker in terms of acute events. $x_{ij}$ is an
indicator for the occurrence of an acute rejection or infection
episode within 14 days of the observation of FEV$_1$.

Section~\ref{sec:fitting:hmm:general} describes how these and more
general hidden Markov models can be fitted with the \Rpackage{msm}
package.

\clearpage

\section{Fitting multi-state models with {\tt msm}}

<<echo=FALSE>>=
options(width = 60)
@

\Rpackage{msm} is a package of functions for multi-state modelling using
the R statistical software.  The \Rfunction{msm} function itself
implements maximum-likelihood estimation for general multi-state
Markov or hidden Markov models in continuous time. We illustrate
its use with a set of data from monitoring heart transplant patients.
Throughout this section ``\textsl{\texttt{>}}'' indicates the R
command prompt, \textsl{\texttt{slanted typewriter}} text indicates R
commands, and \texttt{typewriter} text R output.

\subsection{Installing \tt{msm}}
\label{sec:installing}

The easiest way to install the \Rpackage{msm} package on a computer
connected to the Internet is to run the R command:

\begin{Scode}
  install.packages("msm")
\end{Scode}

This downloads \Rpackage{msm} from the CRAN archive of contributed R
packages (\texttt{cran.r-project.org} or one of its mirrors) and
installs it to the default R system library.  To install to a
different location, for example if you are a normal user with no
administrative privileges, create a directory in which R packages are
to be stored, say, \texttt{/your/library/dir}, and run

\begin{Scode}
  install.packages("msm", lib='/your/library/dir')
\end{Scode}

After \Rpackage{msm} has been installed, its functions can be made
visible in an R session by
<<>>=
library(msm)
@
or, if it has been installed into a non-default library,

\begin{Scode}
  library(msm, lib.loc='/your/library/dir')
\end{Scode}

\subsection{Getting the data in}
\label{sec:datain}

The data are specified as a series of observations, grouped by
patient. At minimum there should be a data frame with variables
indicating
\begin{itemize}
\item the time of the observation,
\item the observed state of the process.
\end{itemize}
If the data do not also contain
\begin{itemize}
\item the subject identification number,
\end{itemize}
then all the observations are assumed to be from the same subject.
The subject ID does not need to be numeric, but data must be grouped
by subject, and observations must be ordered by time within subjects.
If the model includes variables with missing values, then the corresponding 
observations  are omitted by \Rfunction{msm} with a warning.  If you have missing data, 
as in any statistical model, it is recommended to ensure these do not 
result in biases.

An example data set, taken from monitoring a set of heart transplant
recipients, is provided with \Rpackage{msm}.  Sharples
\etal~\cite{my:cav} studied the progression of coronary allograft
vasculopathy (CAV), a post-transplant deterioration of the arterial
walls, using these data.  Risk factors and the accuracy of the
screening test were investigated using multi-state Markov and hidden
Markov models.

This data set can be made available to the current R session by
issuing the command

<<>>=
data(cav)
@

The first three patient histories are shown below. There are 622
patients in all.  \Robject{PTNUM} is the subject identifier.
Approximately each year after transplant, each patient has an
angiogram, at which CAV can be diagnosed.  The result of the test is
in the variable \Robject{state}, with possible values 1, 2, 3
representing CAV-free, mild CAV and moderate or severe CAV
respectively.  A value of 4 is recorded at the date of death.
\Robject{years} gives the time of the test in years since the heart
transplant. Other variables include \Robject{age} (age at screen),
\Robject{dage} (donor age), \Robject{sex} (0=male, 1=female),
\Robject{pdiag} (primary diagnosis, or reason for transplant - IHD
represents ischaemic heart disease, IDC represents idiopathic dilated
cardiomyopathy), \Robject{cumrej} (cumulative number of rejection
episodes), and \Robject{firstobs}, an indicator which is 1 when the
observation corresponds to the patient's transplant (the first
observation), and 0 when the observation corresponds to a later
angiogram.

<<>>=
cav[1:21,]
@

A useful way to summarise multi-state data is as a frequency table of
pairs of consecutive states.  This counts over all individuals, for
each state $r$ and $s$, the number of times an individual had an
observation of state $r$ followed by an observation of state $s$. The
function \Rfunction{statetable.msm} can be used to produce such a table,
as follows,

<<>>=
statetable.msm(state, PTNUM, data=cav)
@
Thus there were 148 CAV-free deaths, 48 deaths from state 2, and
55 deaths from state 3.  On only four occasions was there an
observation of severe CAV followed by an observation of no CAV.

\subsection{Specifying a model}
\label{sec:specifying:model}

We now specify the multi-state model to be fitted to the data.  A
model is governed by a transition intensity matrix $Q$. For the heart
transplant example, there are four possible states through which the
patient can move, corresponding to CAV-free, mild/moderate CAV, severe
CAV and death.  We assume that the patient can advance or recover from
consecutive states while alive, and die from any state. Thus the model
is illustrated by Figure~\ref{fig:disease} with four states, and we
have

\[
Q = \left(
  \begin{array}{llll}
    -(q_{12} + q_{14}) & q_{12} &  0     & q_{14}\\
    q_{21} & -(q_{21}+q_{23}+q_{24}) & q_{23} & q_{24}\\
      0    & q_{32} & -(q_{32}+q_{34}) & q_{34}\\
      0    &   0    &   0    &   0   \\
  \end{array}
\right )
\]

It is important to remember that this defines which
\emph{instantaneous} transitions can occur in the Markov process, and
that the data are \emph{snapshots} of the process (see
section~\ref{sec:arbitr-observ-times}).  Although there were 44
occasions on which a patient was observed in state 1 followed by state
3, the underlying model specifies that the patient must have passed
through state 2 in between.  If your data represent the exact and
complete transition times of the process, then you must specify
\Rfunarg{exacttimes=TRUE} or \Rfunarg{obstype=2} in the call to
\Rfunction{msm}.

To tell \Rfunction{msm} what the allowed transitions of our model are,
we define a matrix of the same size as $Q$, containing zeroes in the
positions where the entries of $Q$ are zero.  All other positions
contain an initial value for the corresponding transition intensity.
The diagonal entries supplied in this matrix do not matter, as the
diagonal entries of $Q$ are defined as minus the sum of all the other
entries in the row.  This matrix will eventually be used as the
\Rfunarg{qmatrix} argument to the \Rfunction{msm} function. For
example,

<<>>=
twoway4.q  <-  rbind ( c(0, 0.25, 0, 0.25),
                       c(0.166, 0, 0.166, 0.166),
                       c(0, 0.25, 0, 0.25),
                       c(0, 0, 0, 0) )
@

Fitting the model is a process of finding values of the seven unknown
transition intensities: $q_{12}$, $q_{14}$, $q_{21}$, $q_{23}$,
$q_{24}$, $q_{32}$, $q_{34}$, which maximise the likelihood.

\subsection{Specifying initial values}
\label{sec:inits}

The likelihood is maximised by numerical methods, which need a set of
initial values to start the search for the maximum.  For reassurance
that the true maximum likelihood estimates have been found, models
should be run repeatedly starting from different initial
values. However a sensible choice of initial values can be important
for unstable models with flat or multi-modal likelihoods.  For
example, the transition rates for a model with misclassification could
be initialised at the corresponding estimates for an approximating
model without misclassification. Initial values for a model without
misclassification could be set by supposing that transitions between
states take place only at the observation times.  If we observe
$n_{rs}$ transitions from state $r$ to state $s$, and a total of $n_r$
transitions from state $r$, then $q_{rs} / q_{rr}$ can be estimated by
$n_{rs} / n_r$. Then, given a total of $T_r$ years spent in state $r$,
the mean sojourn time $1 / q_{rr}$ can be estimated as $T_r / n_r$.
Thus, $n_{rs} / T_r$ is a crude estimate of $q_{rs}$. The \Rpackage{msm}
package provides a function \Rfunction{crudeinits.msm} for calculating
initial values in this way.

<<>>=
twoway4.crude.q  <- crudeinits.msm(state ~ years, PTNUM, data=cav, qmatrix=twoway4.q)
@

However, if there are are many changes of state in between the
observation times, then this crude approach may fail to give sensible
initial values.  For the heart transplant example we could also guess
that the mean period in each state before moving to the next is about
2 years, and there is an equal probability of progression, recovery or
death.  This gives $q_{rr} = - 0.5$ for $r = 1, 2, 3$, and $q_{12} =
q_{14} = 0.25$, $q_{21} = q_{23} = q_{24} = 0.166$, $q_{32} = q_{34} =
0.25$, and the initial value matrix \Robject{twoway4.q} shown above, 
which we now use to fit the model.

%% TODO explain about gen.inits=TRUE here

\subsection{Running \Rfunction{msm}}
\label{sec:running}

To fit the model, call the \Rfunction{msm} function with the appropriate
arguments.  For our running example, we have defined a data set
\Robject{cav}, a matrix \Robject{twoway4.q} indicating the allowed
transitions, and initial values. We are ready to run \Rfunction{msm}.

\paragraph{Model 1: simple bidirectional model}

<<>>=
cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                    qmatrix = twoway4.q, death = 4 )
@

In this example the day of death is assumed to be recorded exactly, as
is usual in studies of chronic diseases.  At the previous instant
before death the state of the patient is unknown.  Thus we specify
\Rfunarg{death = 4}, to indicate to \Rfunction{msm} that the 
entry times into state 4 are observed in this manner. 
If the model had
five states, and states 4 and 5 were two competing causes of death with
times recorded exactly in this way, then we would specify \Rfunarg{death =
  c(4,5)}.

By default, the data are assumed to represent snapshots of the process
at arbitrary times.  However, observations can also represent exact
times of transition, ``exact death times'', or a mixture of these. See the
\Rfunarg{obstype} argument to \Rfunction{msm}.

While the \Rfunction{msm} function runs, it searches for the maximum
of the likelihood of the unknown parameters.  Internally, it uses the
R function \Rfunction{optim} to minimise the minus log-likelihood.
When the data set, the model, or both, are large, then this may take a
long time.  It can then be useful to see the progress of the
optimisation algorithm. To do this, we can specify a \Rfunarg{control}
argument to \Rfunction{msm}, which is passed internally to the
\Rfunction{optim} function.  For example \texttt{control = list(trace=1,
  REPORT=1)}. See the help page for \Rfunction{optim},

<<eval=FALSE>>=
help(optim)
@
for more options to control the optimisation.  When completed, the
\Rfunction{msm} function returns a value.  This value is a list of the
important results of the model fitting, including the parameter
estimates and their covariances.  To keep these results for
post-processing, we store them in an R object, here called
\Robject{cav.msm}.  When running several similar \Rfunction{msm} models,
it is recommended to store the respective results in
informatively-named objects.

\subsection{Showing results}

To show the maximum likelihood estimates and 95\% confidence
intervals, type the name of the fitted model object at the R command
prompt.  \footnote{This is equivalent to typing
  \texttt{print.msm(cav.msm)}.  The function \Rfunction{print.msm}
  formats the important information in the model object for printing
  on the screen.}  The confidence level can be changed using the
\Rfunarg{cl} argument to \Rfunction{msm}.

<<>>=
cav.msm
@

From the estimated intensity matrix, we see patients are three times
as likely to develop symptoms than die without symptoms (first row).
After disease onset (state 2), progression to severe symptoms (state
3) is 50\% more likely than recovery.  Once in the severe state, death
is more likely than recovery, and a mean of 1 / -0.44 = 2.3 years is
spent in state 3 before death or recovery.

Section \ref{sec:extractor} describes various functions that can be
used to obtain summary information from the fitted model.

\subsection{Covariates on the transition rates}
\label{sec:msm:covariates}

We now model the effect of explanatory variables on the rates of
transition, using a proportional intensities model. Now we have an
intensity matrix $Q(z)$ which depends on a covariate vector $z$. For
each entry of $Q(z)$, the transition intensity for patient $i$ at
observation time $j$ is $q_{rs}(z_{ij}) = q_{rs}^{(0)}
\exp(\beta_{rs}^T z_{ij})$.  The covariates $z$ are specified through
the \Rfunarg{covariates} argument to \Rfunction{msm}. If $z_{ij}$ is
time-dependent, we assume it is constant in between the observation
times of the Markov process.  \Rfunction{msm} calculates the
probability for a state transition from times $t_{i,j-1}$ to $t_{ij}$
using the covariate value at time $t_{i,j-1}$.

We consider a model with just one covariate, female sex.  Out of the
622 transplant recipients, 535 are male and 87 are female.  By
default, all linear covariate effects $\beta_{rs}$ are initialised to
zero. To specify different initial values, use a \Rfunarg{covinits}
argument, as described in \Rfunction{help(msm)}.  Initial values given
in the \Rfunarg{qmatrix} represent the intensities with covariate values
set to their means in the data.

\paragraph{Model 2: sex as a covariate}
<<>>=
cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                     qmatrix = twoway4.q, death = 4, covariates = ~ sex)
@

The \Robject{msm} object will now include the estimated covariate
effects and their confidence intervals.

<<>>=
cavsex.msm
@
Comparing the estimated log-linear effects of sex and their standard
errors, we see that the disease onset rate is smaller for females,
whereas none of the other effects are significant. The first matrix shown in the output of printing
\Robject{cavsex.msm} is the estimated transition intensity matrix
$q_{rs}(z) = q_{rs}^{(0)} \exp(\beta_{rs}^T z)$ with the covariate $z$
set to its mean value in the data.  This represents an average
intensity matrix for the population of 535 male and 87 female
patients. To extract separate intensity matrices for male and female
patients ($z = 0$ and $1$ respectively), use the function
\Rfunction{qmatrix.msm}, as shown below.  This and similar summary
functions will be described in more detail in
section~\ref{sec:extractor}.

<<>>=
qmatrix.msm(cavsex.msm, covariates=list(sex=0)) # Male
qmatrix.msm(cavsex.msm, covariates=list(sex=1)) # Female
@

We may also want to constrain the effect of a covariate to be equal for
certain transition rates, to reduce the number of parameters in the
model, or to investigate hypotheses on the covariate effects.  A
\Rfunarg{constraint} argument can be used to indicate which of the
transition rates have common covariate effects.

\paragraph{Model 3: constrained covariate effects}

<<eval=FALSE>>=
cav3.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                   qmatrix = twoway4.q, death = 4,
                   covariates = ~ sex, constraint = list(sex=c(1,2,3,1,2,3,2)) )
@

This constrains the effect of sex to be equal for the progression
rates $q_{12}, q_{23}$, equal for the death rates $q_{14}, q_{24},
q_{34}$, and equal for the recovery rates $q_{21}, q_{32}$. The
intensity parameters are assumed to be ordered by reading across the
rows of the transition matrix, starting at the first row: ($q_{12},
q_{14}, q_{21}, q_{23}, q_{24}, q_{32}, q_{34}$), giving constraint
indicators \Rfunarg{(1,2,3,1,2,3,2)}.  Any vector of increasing
numbers can be used for the indicators.  Negative entries can be used
to indicate that some effects are equal to minus others:
\Rfunarg{(1,2,3,-1,2,3,2)} sets the fourth effect to be minus the
first.

In a similar manner, we can constrain some of the baseline transition
intensities to be equal to one another, using the
\Rfunarg{qconstraint} argument.  For example, to constrain the rates
$q_{12}$ and $q_{23}$ to be equal, and $q_{24}$ and $q_{34}$ to be
equal, specify \Rfunarg{qconstraint = c(1,2,3,1,4,5,4)}.

\subsection{Fixing parameters at their initial values}

For exploratory purposes we may want to fit a model assuming that some
parameters are fixed, and estimate the remaining parameters.  This may
be necessary in cases where there is not enough information in the
data to be able to estimate a proposed model, and we have strong prior
information about a certain transition rate. To do this, use the
\Rfunarg{fixedpars} argument to \Rfunction{msm}.  For model 1, the
following statement fixes the parameters numbered 6, 7, that is,
$q_{32}$, $q_{34}$, to their initial values (0.25 and 0.25,
respectively).

\paragraph{Model 4: fixed parameters}
<<eval=FALSE>>=
cav4.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                qmatrix = twoway4.q, death = 4,
                control = list(trace=2, REPORT=1),
                fixedpars = c(6, 7) )
@

A \Rfunarg{fixedpars} statement can also be useful for fixing covariate
effect parameters to zero, that is to assume no effect of a covariate
on a certain transition rate.

\subsection{Extractor functions}
\label{sec:extractor}

We may want to extract some of the information from the \Rfunction{msm}
model fit for post-processing, for example for plotting graphs or
generating summary tables.  A set of functions is provided for
extracting interesting features of the fitted model.

\begin{description}

\item[Intensity matrices] The function \Rfunction{qmatrix.msm}
  extracts the estimated transition intensity matrix and its confidence intervals
  for a given set of covariate values, as shown in
  section~\ref{sec:msm:covariates}.  Confidence intervals are
  calculated from the covariance matrix of the estimates by assuming
  the distribution is symmetric on the log scale.  Standard errors for
  the intensities are also available from the object returned by
  \Rfunction{qmatrix.msm}.  These are calculated by the delta method.
  The \Rpackage{msm} package provides a general-purpose function
  \Rfunction{deltamethod} for estimating the variance of a function of
  a random variable $X$ given the expectation and variance of $X$. See
  \texttt{help(deltamethod)} for further details.  Bootstrap
  confidence intervals are also available for \Rfunction{qmatrix.msm}
  and for most output functions; these are often more accurate, at the
  cost of computational time.  For more about bootstrapping in
  \Rpackage{msm}, see Section~\ref{sec:boot}.

\item[Transition probability matrices] The function
  \Rfunction{pmatrix.msm} extracts the estimated transition probability
  matrix $P(t)$ within a given time. For example, for model 1, the 10
  year transition probabilities are given by:

<<>>=
pmatrix.msm(cav.msm, t=10)
@

  Thus, a typical person in state 1, disease-free, has a probability
  of 0.5 of being dead ten years from now, a probability of 0.3 being
  still disease-free, and probabilities of 0.1 of being alive with
  mild/moderate or severe disease, respectively.

  This assumes $Q$ is constant within the desired time interval.  For
  non-homogeneous processes, where $Q$ varies with time-dependent
  covariates but can be approximated as piecewise constant, there is
  an equivalent function \Rfunction{pmatrix.piecewise.msm}.  Consult its
  help page for further details.

  If \Rfunarg{ci=''norm''} is specified, then a confidence interval is
  calculated based on drawing a random sample (default size 1000) from
  the assumed multivariate normal distribution of the maximum
  likelihood estimates and covariance matrix, and transforming.  If
  \Rfunarg{ci=''boot''} is specified, then a bootstrap confidence
  interval for the transition probability matrix is calculated (see Section~\ref{sec:boot}) .
  However, both of these are computationally intensive, particularly 
  the bootstrap method, so no confidence interval is calculated by default.

\item[Mean sojourn times] The function \Rfunction{sojourn.msm}
  extracts the estimated mean sojourn times in each transient state
  $r$, for a given set of covariate values. This is calculated as $-1
  / \hat q_{rr}$, where $\hat q_{rr}$ is the $r$th diagonal entry of
  the estimated transition intensity matrix.

<<>>=
sojourn.msm(cav.msm)
@

\item[Probability that each state is next] The function
  \Rfunction{pnext.msm} extracts the matrix of probabilities $-q_{rs}
  / q_{rr}$ that the next state after state $r$ is state $s$, for each
  $r$ and $s$.  Together with the mean sojourn times, this gives a more
  intuitive parameterisation of a continuous-time Markov model than 
  the raw transition intensities $q_{rs}$.  Note these are different from 
  the transition probabilities in a given time $t$ returned by 
  \Rfunction{pmatrix.msm}.

<<>>=
pnext.msm(cav.msm)
@

\item[Total length of stay] Mean sojourn times describe the
  average period in a single stay in a state. For processes with
  successive periods of recovery and relapse, we may want to
  forecast the total time spent healthy or diseased, before
        death.  The function \Rfunction{totlos.msm} estimates the
        forecasted total length of time spent in each transient state
        $s$ between two future time points $t_1$ and $t_2$, for a
        given set of covariate values. This defaults to the expected
        amount of time spent in each state between the start of the
        process (time 0, the present time) and death or a specified
        future time. This is obtained as
\[
L_s = \int_{t_1}^{t_2} P(t)_{r,s} dt
\]
where $r$ is the state at the start of the process, which defaults to
1. This is calculated using numerical integration.  For model 1, each
patient is forecasted to spend 8.8 years disease free, 2.2 years with
mild or moderate disease and 1.8 years with severe disease.

Bootstrap and asymptotic confidence intervals are available, as for
\Rfunction{pmatrix.msm}, but are not calculated by default.

<<>>=
totlos.msm(cav.msm)
@

\item[Expected first passage times] The function \Rfunction{efpt.msm} estimates
  the expected time until the process first enters a given state or set of states, 
  also called the ``hitting time''.  See its help page for further details.

\item[Ratio of transition intensities] The function
  \Rfunction{qratio.msm} estimates a ratio of two entries of the
  transition intensity matrix at a given set of covariate values,
  together with a confidence interval estimated assuming normality on
  the log scale and using the delta method.  For example, we may want
  to estimate the ratio of the progression rate $q_{12}$ into the
  first state of disease to the corresponding recovery rate $q_{21}$.
  For example in model 1, recovery is 1.8 times as likely as
  progression.

<<>>=
qratio.msm(cav.msm, ind1=c(2,1), ind2=c(1,2))
@

\item[Hazard ratios for transition] The function \Rfunction{hazard.msm}
  gives the estimated hazard ratios corresponding to each covariate
  effect on the transition intensities. 95\% confidence limits are
  computed by assuming normality of the log-effect. For example, for
  model 2 with female sex as a covariate, the following hazard ratios
  show more clearly that the only transition on which the effect of
  sex is significant at the 5\% level is the 1-2 transition.

<<>>=
hazard.msm(cavsex.msm)
@

\end{description}

\paragraph{Setting covariate values}
All of these extractor functions take an argument called
\Rfunarg{covariates}.  If this argument is omitted, for example,

<<eval=FALSE>>=
qmatrix.msm(cav.msm)
@
then the intensity matrix is evaluated as $Q(\bar x)$ with all
covariates set to their mean values $\bar x$ in the data.
Alternatively, set \Rfunarg{covariates} to 0 to return the result
$Q(0)$ with covariates set to zero. This will usually be
preferable for categorical covariates, where we wish to see the
result for the baseline category.

<<eval=FALSE>>=
qmatrix.msm(cavsex.msm, covariates = 0)
@
Alternatively, the desired covariate values can be specified
explicitly as a list,

<<eval=FALSE>>=
qmatrix.msm(cavsex.msm, covariates = list(sex = 1))
@

Values of categorical covariates must be quoted. For example, consider
a covariate \texttt{smoke}, representing tobacco smoking status, with
three levels, \texttt{NON, CURRENT, EX}, representing a non-smoker,
current smoker or ex-smoker.

\begin{Scode}
qmatrix.msm(example.msm, covariates = list(age = 60, smoke=''CURRENT''))
\end{Scode}

\subsection{Survival plots}

In studies of chronic disease, an important use of multi-state models
is in predicting the probability of survival for patients in
increasingly severe states of disease, for some time $t$ in the
future.  This can be obtained directly from the transition probability
matrix $P(t)$.

The \Rfunction{plot} method for \Robject{msm} objects produces a plot
of the expected probability of survival against time, from each
transient state. Survival is defined as not entering the final
absorbing state.

<<fig=TRUE>>=
plot(cav.msm, legend.pos=c(8, 1))
@

This shows that the 10-year survival probability with severe CAV is
approximately 0.1, as opposed to 0.3 with mild CAV and 0.5 without
CAV.  With severe CAV the survival probability diminishes very quickly
to around 0.3 in the first five years after transplant.  The
\Rfunarg{legend.pos} argument adjusts the position of the legend in
case of clashes with the plot lines.  A \Rfunarg{times} argument can
be supplied to indicate the time interval to forecast survival for.

A more sophisticated analysis of these data might explore competing
causes of death from causes related or unrelated to the disease under
study.

\subsection{Bootstrapping}
\label{sec:boot}
Most of \Rpackage{msm}'s output functions present confidence intervals
based on asymptotic standard errors calculated from the Hessian, or
transformations of these using the delta method.  The asymptotic
standard errors are expected to be underestimates of the true standard
errors (Cramer-Rao lower bound).  For some output functions, such as
\Rfunction{pmatrix.msm}, and functions based on
\Rfunction{pmatrix.msm} such as \Rfunction{totlos.msm} and
\Rfunction{prevalence.msm}, the delta method cannot be used at all to
obtain standard errors.  In these cases, confidence intervals can be
calculated by drawing a random sample from the assumed multivariate
normal distribution of the maximum likelihood estimates and covariance
matrix, and transforming.  However, this is still based on potentially
inaccurate asymptotic theory.  The \Rpackage{msm} package provides the
function \Rfunction{boot.msm} to enable bootstrap refitting of
\Rfunction{msm} models, an alternative way to estimate uncertainty.

For non-hidden Markov models, a bootstrap dataset is drawn by
resampling pairs of consecutive states from the full data, i.e.
\emph{transitions}.  These are assumed to be independent when
calculating the likelihood (Section~\ref{sec:multi:likelihood}).  For
hidden Markov models and models with censoring, a bootstrap dataset is
drawn by resampling complete series from independent subjects.  The
bootstrap datasets have the same number of transitions, or subjects,
respectively, as the original data.

For most output extractor functions provided with \Rpackage{msm}, the
option \Rfunarg{ci=''boot''} is available, as a wrapper around
\Rfunction{boot.msm}, to enable bootstrap confidence intervals to be calculated.
But any user-defined output statistic can be bootstrapped, as follows.
The function \Rfunction{boot.msm} is called with the fitted
\Rfunction{msm} model as first argument, and an R function specifying
the statistic to be bootstrapped as the second argument \texttt{stat}.
The return value from \Rfunction{boot.msm} is a list of \texttt{B}
replicates (by default, \texttt{B=1000}) of the desired statistic.  For
example, to bootstrap the transition intensity matrix of the heart
transplantation model \Robject{cav.msm}:
\begin{Scode}
  q.list <- boot.msm(cav.msm, stat=function(x){qmatrix.msm(x)$estimates})
\end{Scode}
Note that for \Rfunction{boot.msm} to be able to refit the original
model that produced \Robject{cav.msm}, all objects used in the
original model fit (for example, in this case, \Robject{twoway4.q})
must be in the working environment.  Otherwise, \Rfunction{boot.msm}
will give an ``object not found'' error.

The user can then summarise these replicates by calculating empirical
standard deviations or quantile-based intervals.  In this example,
\Robject{q.list} is a list of 1000 4$\times$4 matrices.  The following
code calculates the bootstrap standard error as the empirical standard
deviation of the 1000 replicates, and a similar 95\% bootstrap
confidence interval.
\begin{Scode}
  q.array <- array(unlist(q.list), dim=c(4,4,1000))
  apply(q.array, c(1,2), sd)
  apply(q.array, c(1,2), function(x)quantile(x, c(0.025, 0.975)))
\end{Scode}

Note that when bootstrapping, the refits of the model to the resampled
datasets may occasionally fail to converge (as discussed in
Section~\ref{sec:failure}) even if the original model fit did
converge.  In these cases, a warning is given, but
\Rfunction{boot.msm} simply discards the failed dataset and moves on
to the next bootstrap iteration. Unless convergence failure occurs for
a large proportion of iterations, this should not affect the accuracy
of the final bootstrap confidence intervals.

\subsection{Convergence failure}
\label{sec:failure}

Inevitably if over-complex models are applied with insufficient data
then the parameters of the model will not be identifiable.  This will
result in the optimisation algorithm failing to find the maximum of
the log-likelihood, or even failing to evaluate the likelihood. For
example, it will commonly be inadvisable to include several covariates
in a model simultaneously.

In some circumstances, the optimisation may report convergence, but
fail to calculate any standard errors.  In these cases, the Hessian of
the log-likelihood at the reported solution is not positive definite.
Thus the reported solution may be a saddle point rather than the
maximum likelihood, or it may be close to the maximum.

\begin{description}
\item[Initial values] Make sure that a sensible set of initial values
  have been chosen. The optimisation may only converge within a
  limited range of `informative' initial values.  It is also sensible
  to run the model for several different initial values to ensure that
  the estimation has converged to a global rather than a local optimum.

\item[Scaling] It is often necessary to apply a scaling factor to
  normalise the likelihood (\Rfunarg{fnscale}), or certain individual
  parameters \Rfunarg{(parscale)}.  This may prevent overflow or
  underflow problems within the optimisation.  For example, if the value
  of the -2 $\times$ log-likelihood is around 5000, then the following
  option leads to an minimisation of the -2 $\times$ log-likelihood on
  an approximate unit scale: \Rfunarg{control = list(fnscale = 5000)}

  It is also advisable to analyse all variables, including covariates
  and the time unit, on a roughly normalised scale. For example,
  working in terms of a time unit of months or years instead of days,
  when the data range over thousands of days.

\item[Convergence criteria] ``False convergence'', in which
  \Rfunction{optim()} reports convergence of the optimisation but the
  Hessian is not positive definite, can sometimes be solved by
  tightening the criteria (\Rfunarg{reltol}, defaults to
  \texttt{1e-08}) for reporting convergence. For example,
  \Rfunarg{control = list(reltol = 1e-16)}.

  Alternatively consider using smaller step sizes for the numerical
  approximation to the gradient, used in calculating the Hessian.
  This is given by the control parameter \Rfunarg{ndeps}. For example,
  for a model with 5 parameters, \Rfunarg{control = list(ndeps =
    rep(1e-6, 5))}

\item[Choice of algorithm] By default, \Rfunction{msm} uses a
  Nelder-Mead algorithm which does not use analytic derivatives of the
  objective function.   For potentially greater speed and accuracy,
  consider using a quasi-Newton algorithm such as
  the ``BFGS'' method of \Rfunction{optim}, which can make use of
  analytic derivatives, for example, \Rfunarg{method = ``BFGS'', use.deriv = TRUE}.
  Or use the \Rfunction{nlm} R function via \Rfunarg{msm(..., opt.method = "nlm" , ...)}
  Note also the Fisher scoring method available for non-hidden Markov 
  models for panel data, via  \Rfunarg{msm(..., opt.method = "fisher", ...)}, 
  is expected to be faster than the generic methods, but less robust to 
  bad initial values.

\item[Model simplification]

  If none of these numerical adjustments lead to convergence, then the
  model is probably over-complicated.  There may not be enough
  information in the data on a certain transition rate.  It is always
  recommended to count all the pairs of transitions between states in
  successive observation times, making a frequency table of previous
  state against current state (function \Rfunction{statetable.msm}).
  Although the data are a series of snapshots of a continuous-time
  process, and the actual transitions take place in between the
  observation times, this type of table may still be helpful.  If
  there are not many observed `transitions' from state 2 to state 4,
  for example, then the data may be insufficient to estimate $q_{24}$.

  For a staged disease model (Figure~\ref{fig:disease}), the number of
  disease states should be low enough that all transition rates can be
  estimated.  Consecutive states of disease severity should be merged
  if necessary.  If it is realistic, consider applying constraints on
  the intensities or the covariate effects so that the parameters are
  equal for certain transitions, or zero for certain transitions.

  Be careful to use a observation scheme and transition matrix
  appropriate to your data (see
  Section~\ref{sec:arbitr-observ-times}). By default, \Rfunction{msm}
  assumes that the data represent snapshots of the process, and the
  true state is unknown between observation times.  In such
  circumstances, it is rarely feasible to estimate an intensity matrix
  with \emph{instantaneous} transitions allowed between every pair of
  states.  This would be easier if the complete course of the process
  is known \Rfunarg{(exacttimes = TRUE)} in the call to
  \Rfunction{msm}.  Understand the difference between
  \emph{instantaneous} and \emph{interval} transitions - although
  individuals may be in state 1 at time $t_r$, and state 3 at time
  $t_{r+1}$, that doesn't mean that instantaneous transitions from 1
  to 3 should be permitted.

\end{description}

\subsection{Model assessment}
\label{sec:model-assessment}

\paragraph{Observed and expected prevalence}

To compare the relative fit of two nested models, it is easy to
compare their likelihoods.  However it is not always easy to determine
how well a fitted multi-state model describes an irregularly-observed
process.  Ideally we would like to compare observed data with fitted
or expected data under the model.  If there were times at which all
individuals were observed then the fit of the expected numbers in each
state or {\em prevalences} can be assessed directly at those times.
Otherwise, some approximations are necessary.  We could assume that an
individual's state at an arbitrary time $t$ was the same as the state
at their previous observation time.  This might be fairly accurate if
observation times are close together.  This approach is taken by the
function \Rfunction{prevalence.msm}, which constructs a table of observed
and expected numbers and percentages of individuals in each state at a
set of times.

A set of expected counts can be produced if the process begins at a
common time for all individuals.  Suppose at this time, each
individual is in state 0.  Then given $n(t)$ individuals are under
observation at time $t$, the expected number of individuals in state
$r$ at time $t$ is $n(t) P(t)_{0,r}$.

For example, we calculate the observed and expected numbers and
percentages at two-yearly intervals up to 20 years after transplant,
for the heart transplant model \Rfunction{cav.msm}.  The number of
individuals still alive and under observation decreases from 622 to
251 at year 20.  The observed and expected percentages are plotted
against time.

<<>>=
options(digits=3)
prevalence.msm(cav.msm, times=seq(0,20,2))
@
<<fig=TRUE>>=
plot.prevalence.msm(cav.msm, mintime=0, maxtime=20)
@

Comparing the observed and expected percentages in each state,
we see that the predicted number of individuals who die (State 4) is
under-estimated by the model from about year 8 onwards.  Similarly the
number of individuals sill alive and free of CAV (State 1) is
over-estimated by the model from about year 8 onwards.

Such discrepancies could have many causes.  One possibility is that
the transition rates vary with the time since the beginning of the
process, the age of the patient, or some other omitted covariate, so
that the Markov model is {\em non-homogeneous}.  This could be
accounted for by modelling the intensity as a function of age, for
example, such as a piecewise-constant function.  The \Rfunarg{pci}
argument to \Rfunction{msm} can be used to automatically construct
models with transition intensities which are piecewise-constant in
time.

In this example, the hazard of death may increase with age, so that
the model underestimates the number of deaths when forecasting far
into the future.  Another cause of poor model fit may sometimes be the
failure of the Markov assumption.  That is, the transition intensities
may depend on the time spent in the current state (a semi-Markov
process) or other characteristics of the process history.  Accounting
for the process history is difficult as the process is only observed
through a series of snapshots.  For a multi-state model with one-way
progression through states, and frequent observations, we may be able
to estimate the time spent in each state by each individual.

However, if it is known that individuals who died would not have been
followed up after a certain time, had they survived to that time, then
they should not be included in the observed prevalence of the death
state after that time.  This can be accounted for by passing a vector
of maximum potential follow-up times, one for each individual in the
same order as the original data, in the \Rfunarg{censtime} argument to
\Rfunction{prevalence.msm}.  Ignoring the potential follow-up times is
likely to have resulted in overestimates of the number of deaths
at later times in the CAV example, though these times are not
available in the data supplied with \Rpackage{msm}.

\paragraph{Pearson-type goodness-of-fit test}
\label{sec:pearson}

Suppose that the true transition times are unknown, and data consist
of observations of the process at arbitrary times which differ between
individuals (panel data).  Assessing goodness of fit by prevalence
counts then involves estimating the observed prevalence at a series of
points by some form of interpolation. This is only advisable if
observation times are close together.  An alternative method of
assessing goodness-of-fit is to construct tables of observed and
expected numbers of transitions, as described by Aguirre-Hernandez and
Farewell~\cite{ahf}.  This leads to a formal test of goodness-of-fit,
analogous to the classical Pearson $\chi^2$ test for contingency
tables. The tables are constructed as follows.  Each pair of
successive observations in the data (\emph{transition}) is classified
by
\begin{itemize}
\item the starting state $r$ and finishing state $s$,
\item time between the start of the process and the first of the pair
  of observations (indexed by $h$),
\item time interval between the observations (indexed by $l_h$, within
  categories $h$),
\item (if there are fitted covariates) the impact of covariates, as
  summarised by $q_{rr}$ (indexed by $c$),
\item any other grouping of interest for diagnosing lack of fit
  (indexed by $g$).
\end{itemize}
Groupings of continuous quantities are normally defined by quantiles,
so that there are a similar number of observed transitions in each
(one-dimensional) category.  The observed and expected numbers of
transitions in each group are then defined by
\[
o_{hl_h rscg} = \sum I(S(t_{i,j+1}) = s, S(t_{ij}) = r)
\]
\[
e_{hl_h rscg} = \sum P(S(t_{i,j+1}) = s | S(t_{ij}) = r)
\]
where $I(A)$ is the indicator function for an event $A$ and the
summation is over the set of transitions in the category defined by
$h,l_h,c,g$, over all individuals $i$. The Pearson-type test statistic
is then
\[
T = \sum_{hl_h rscg} \frac{(o_{hl_h rscg} - e_{hl_h rscg})^2}{e_{hl_h rscg}}
\]
The classical Pearson test statistic is distributed as $\chi^2_{n-p}$,
where $n$ is the number of independent cells in the table and $p$ is
the number of estimated parameters $p$. But the null distribution of
$T$ is not exactly $\chi^2$, since the time intervals are
non-identical, therefore the observed transitions are realizations
from a set of independent but non-identical multinomial distributions.
Titman~\cite{titman:asympnull} showed that the null distribution 
of $T$ is asymptotically approximated by a weighted sum of $\chi^2_1$ 
random variables.   Aguirre-Hernandez and Farewell~\cite{ahf} also showed that $\chi^2_{n-p}$
is a good approximation if there are no fitted covariates. For models
with covariates, the null mean of $T$ is higher than $n - p$, but
lower than $n$.  Therefore, upper and lower bounds for the true
$p$-value of the statistic can be obtained from the $\chi^2_{n-p}$ and
$\chi^2_n$ distributions.  Aguirre-Hernandez and Farewell~\cite{ahf}
also described a bootstrap procedure for obtaining an accurate $p$-value.

Titman and Sharples~\cite{titman:sharples} described modifications to
the test to correct for the biases introduced where in addition to the
panel-data observation scheme:
\begin{itemize}
\item Times of death are known exactly.  In this case, transitions
  ending in death are classified according to the next scheduled
  observation time after the death, which is estimated by multiple
  imputation from a Kaplan-Meier estimate of the distribution of time
  intervals between observations.

\item An individual's final observation is censored, so that they are
  only known to be alive at that point.

\item States are misclassified. 
\end{itemize}

The \Rpackage{msm} package provides the function
\Rfunction{pearson.msm} to perform the Pearson-type test.  By default,
three groups are used for each of $h$, $l_h$ and $c$.  Often the
number of groups will need to be reduced in cases where the resulting
contingency tables are sparse (thus there are several low expected
counts and the variance of $T$ is inflated).

The test is now performed on the model \Robject{cav.msm} for the heart
transplant dataset (a version of which was also analysed by
Titman and Sharples~\cite{titman:sharples}). The default three interval groups are used,
and two groups of the time since the start of the process. The
\Rfunarg{transitions} argument groups the transitions from state 3 to
each of states 1, 2 and 3 (the 9th, 10th and 11th transitions)
together in the table, since these transitions are infrequent.

<<>>=
options(digits=2)
pearson.msm(cav.msm, timegroups=2, transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
@

The first two tables in the output show the contingency tables of
observed and expected numbers of transitions.  Note that the observed
number of transitions in certain categories is not a whole number,
since these are averaged over multiple imputations of the next
scheduled observation time following deaths.  The column \texttt{Time}
is the group defined by time since the start of the process, and the
column \texttt{Interval} is the group defined by intervals between
observations.  The columns indicate the allowed transitions, or pairs
of states which can be observed in successive observations.

The third table presents the ``deviance'', the value of
$\frac{(o_{hl_h rscg} - e_{hl_h rscg})^2}{e_{hl_h rscg}}$ for each
cell, multipled by the sign of $o_{hl_h rscg} - e_{hl_h rscg}$ to
indicate whether there were more or fewer transitions than expected in
each cell.  These can indicate areas of bad fit.  For example,
systematic changes in deviance by time or time interval between
observations can indicate that a model with time-varying transition
intensities is more suitable.  Changes in deviance by covariate impact
may indicate heterogeneity between individuals which is unexplained
by the fitted covariates.  Changes in deviance with the length of the
interval between observations may also indicate failure of the Markov
assumption, and that a semi-Markov model (in which intensities depend
on the time spent in the current state) may fit better.

In this example, the test statistic is 100.  \Robject{p.upper} is an
upper bound for the $p$-value of the statistic based on an asymptotic
$\chi^2_{42}$ distribution, therefore the model does not fit well.  It
is not clear from the table of deviances which aspects of the fit are
most influental to the test statistic.  However, the two-way Markov
model itself is not biologically plausible, as discussed in
Section~\ref{sec:fitting:hmm:misc}.

For non-hidden Markov models for panel data, \Rfunction{pearson.msm}
also presents the accurate analytic p-value of
Titman \cite{titman:asympnull}.  For all models,
\Rfunction{pearson.msm} provides an option for parametric
bootstrapping to obtain an accurate p-value.

\subsection{Fitting misclassification models with \Rpackage{msm}}
\label{sec:fitting:hmm:misc}

In fact, in the heart transplant example from
section~\ref{sec:datain}, it is not medically realistic for patients
to recover from a diseased state to a healthy state.  Progression of
coronary artery vasculopathy is thought to be an irreversible process.
The angiography scan for CAV is actually subject to error, which leads
to some false measurements of CAV states and apparent recoveries.
Thus we account for misclassification by fitting a \emph{hidden Markov
model} using \Rpackage{msm}.  Firstly we replace the two-way multi-state
model by a one-way model with transition intensity matrix
\[
Q = \left(
  \begin{array}{llll}
    -(q_{12} + q_{14}) & q_{12} &  0     & q_{14}\\
      0    & -(q_{23}+q_{24}) & q_{23} & q_{24}\\
      0    &   0    & -q_{34} & q_{34}\\
      0    &   0    &   0    &   0   \\
  \end{array}
\right )
\]
We also assume that true state 1 (CAV-free) can be classified as state
1 or 2, state 2 (mild/moderate CAV) can be classified as state 1, 2 or
3, while state 3 (severe CAV) can be classified as state 2 or 3.
Recall that state 4 represents death.  Thus our matrix of misclassification
probabilities is
\[
E = \left(
  \begin{array}{llll}
    1 - e_{12} & e_{12} & 0 & 0 \\
    e_{21} & 1 - e_{21} - e_{23} & e_{23} & 0 \\
    0 & e_{32} & 1 - e_{32} & 0 \\
    0 & 0 & 0 & 0\\
  \end{array}
  \right)
\]
with underlying states as rows, and observed states as columns.

To model observed states with misclassification, we define an matrix
\Rfunarg{ematrix} indicating the states that can be misclassified.
Rows of this matrix correspond to true states, columns to observed
states.  It should contains zeroes in the positions where
misclassification is not permitted.  Non-zero entries are initial
values for the corresponding misclassification probabilities. We then
call \Rfunction{msm} as before, but with this matrix as the
\Rfunarg{ematrix} argument.  Initial values of 0.1 are assumed for
each of the four misclassification probabilities $e_{12}, e_{21},
e_{23}, e_{32}$.  Zeroes are given where the elements of $E$ are zero.
The diagonal elements supplied in \Rfunarg{ematrix} are ignored, as
rows must sum to one. The matrix \Rfunarg{qmatrix}, specifying
permitted transition intensities and their initial values, also
changes to correspond to the new $Q$ representing the progression-only
model for the underlying states.

The true state for every patient at the date of transplant is known to
be ``CAV-free'', not misclassified.  To indicate this we use the
argument \Rfunarg{obstrue} to \Rfunction{msm}.  This is set to be a
variable in the dataset, \Rfunarg{firstobs}, indicating where the
observed state equals the true state.  This takes the value of 1 at
the patient's first observation, at the transplant date, and 0
elsewhere.

We use an alternative quasi-Newton optimisation algorithm
\Rfunarg{(method="BFGS")} which can often be faster or more robust
than the default Nelder-Mead simplex-based algorithm.  An optional
argument \Rfunarg{initprobs} could also have been given here,
representing the vector of the probabilities of occupying each true
state at the initial observation
(equation~\ref{eq:multi:hidden:matprod}).  If not given, all
individuals are assumed to be in true state 1 at their initial
observation.  If \Rfunarg{est.initprobs=TRUE} is specified, then these
probabilites are estimated as part of the model fit, using
\Rfunarg{initprobs} as initial values.  Covariate effects on these
probabilities can also be estimated using a multinomial logistic
regression model, if an \Rfunarg{initcovariates} argument is
specified.  See \Rfunction{help(msm)} for further details.

\paragraph{Model 5: multi-state model with misclassification}

<<>>=
oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
                     c(0, 0, 0.202, 0.081),
                     c(0, 0, 0, 0.126),
                     c(0, 0, 0, 0))
ematrix <- rbind(c(0, 0.1, 0, 0),
                 c(0.1, 0, 0.1, 0),
                 c(0, 0.1, 0, 0),
                 c(0, 0, 0, 0))
cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                   qmatrix = oneway4.q, ematrix = ematrix, death = 4,
                   obstrue = firstobs, method="BFGS")
cavmisc.msm
@

Thus there is an estimated probability of about 0.03 that
mild/moderate CAV will be diagnosed erroneously, but a rather higher
probability of 0.17 that underlying mild/moderate CAV will be
diagnosed as CAV-free.  Between the two CAV states, the mild state
will be misdiagnosed as severe with a probability of 0.06, and the
severe state will be misdiagnosed as mild with a probability of 0.12.

The model also estimates the progression rates through underlying
states.  An average of 8 years is spent disease-free, an average of
about 3 years is spent with mild/moderate disease, and periods of severe
disease also last about 3 years on average before death.

\subsection{Effects of covariates on misclassification rates}

We can investigate how the probabilities of misclassification depend
on covariates in a similar way to the transition intensities, using a
\Rfunarg{misccovariates} argument to \Rfunction{msm}. For example, we
now include female sex as a covariate for the misclassification
probabilities.  The linear effects on the log odds of each misclassified
state relative to the true state are initialised to zero by default (but
this can be changed with the \Rfunarg{misccovinits} argument).

\paragraph{Model 6: misclassification model with misclassification probabilities modelled on sex}

<<>>=
cavmiscsex.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                          qmatrix = oneway4.q, ematrix = ematrix, death = 4,
                          misccovariates = ~sex, obstrue=firstobs, method="BFGS")
@
<<>>=
cavmiscsex.msm
@
The large confidence interval for the 1/2 misclassification suggests
that women are no more or less likely than men to have a false
positive angiography scan. On the other hand, women have slightly more
false negatives.

\subsection{Extractor functions}

As well as the functions described in section~\ref{sec:extractor} for
extracting useful information from fitted models, there are a number
of extractor functions specific to models with misclassification.

\begin{description}
\item[Misclassification matrix] The function \Rfunction{ematrix.msm}
gives the estimated misclassification probability matrix at the given
covariate values.  For illustration, the fitted misclassification
probabilities for men and women in model 6 are given by

<<>>=
ematrix.msm(cavmiscsex.msm, covariates=list(sex=0))
ematrix.msm(cavmiscsex.msm, covariates=list(sex=1))
@
The confidence intervals for the estimates for women are higher, since
there are only 87 women in this set of 622 patients.

\item[Odds ratios for misclassification] The function
  \Rfunction{odds.msm} gives the estimated odds ratios corresponding to
  each covariate effect on the misclassification probabilities.

<<>>=
odds.msm(cavmiscsex.msm)
@

\item[Observed and expected prevalences]

  The function \Rfunction{prevalence.msm} is intended to assess the
  goodness of fit of the hidden Markov model for the \emph{observed}
  states to the data.  Tables of observed prevalences of observed
  states are calculated as described in
  section~\ref{sec:model-assessment}, by assuming that observed states
  are retained between observation times.

  The expected numbers of individuals in each observed state are
  calculated similarly. Suppose the process begins at a common time
  for all individuals, and at this time, the probability of occupying
  \emph{true} state $r$ is $f_r$.  Then given $n(t)$ individuals under
  observation at time $t$, the expected number of individuals in true
  state $r$ at time $t$ is the $r$th element of the vector $n(t) f
  P(t)$. Thus the expected number of individuals in \emph{observed}
  state $r$ is the $r$th element of the vector $n(t) f P(t) E$, where
  $E$ is the misclassification probability matrix.

  The expected prevalences (not shown) for this example are similar to
  those forecasted by the model without misclassification, with
  underestimates of the rates of death from 8 years onwards.  To
  improve this model's long-term prediction ability, it is probably
  necessary to account for the natural increase in the hazard of death
  from any cause as people become older.

\item[Goodness-of-fit test] The Pearson-type goodness-of-fit test is
  performed, as in Section~\ref{sec:pearson}.  The table of deviances
  indicates that there are more 1-3 and 1-4 transitions than expected
  in short intervals, and fewer in long intervals.  This may indicate
  some time-dependence in the transition rates. Indeed, Titman
  \cite{titman:phd} found that a model with piecewise-constant
  transition intensities gave a greatly improved fit to these data.
<<>>=
pearson.msm(cavmisc.msm, timegroups=2, transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
@

\end{description}

\subsection{Recreating the path through underlying states}

In speech recognition and signal processing, {\em decoding} is the
procedure of determining the underlying states that are most likely to
have given rise to the observations. The most common method of
reconstructing the most likely state path is the {\em Viterbi}
algorithm. Originally proposed by Viterbi~\cite{viterbi}, it is also
described by Durbin \etal~\cite{biolog:seq} and Macdonald and
Zucchini~\cite{macdonald:zucchini} for discrete-time hidden Markov
chains.  For continuous-time models it proceeds as follows. Suppose
that a hidden Markov model has been fitted and a Markov transition
matrix $P(t)$ and misclassification matrix $E$ are known.  Let
$v_k(t_i)$ be the probability of the most probable path ending in state
$k$ at time $t_i$.
\begin{enumerate}
\item Estimate $v_k(t_1)$ using known or estimated initial-state
  occupation probabilities.
\item For $i = 1 \ldots N$, calculate $v_l(t_i) = e_{l,O_{t_i}} \max_k
  v_k(t_{i-1}) P_{kl}(t_{i} - t_{i-1})$. Let $K_i(l)$ be the
  maximising value of $k$.
\item At the final time point $t_N$, the most likely underlying
  state $S^*_N$ is the value of $k$ which maximises $v_k(t_N)$.
\item Retrace back through the time points, setting $S^*_{i-1} =
  K_i(S^*_i)$.
\end{enumerate}

The computations should be done in log space to prevent underflow.
The \Rpackage{msm} package provides the function \Rfunction{viterbi.msm} to
implement this method.  For example, the following is an extract from
a result of calling \Rfunction{viterbi.msm} to determine the most likely
underlying states for all patients. The results for patient 100103 are
shown, who appeared to `recover' to a less severe state of disease
while in state 3.  We assume this is not biologically possible for the
true states, so we expect that either the observation of state 3 at
time 4.98 was an erroneous observation of state 2, or their apparent
state 2 at time 5.94 was actually state 3.  According to the expected
path constructed using the Viterbi algorithm, it is the observation at
time 5.94 which is most probably misclassified.

<<>>=
vit <- viterbi.msm(cavmisc.msm)
vit[vit$subject==100103,]
@

\subsection{Fitting general hidden Markov models with \Rpackage{msm}}
\label{sec:fitting:hmm:general}

The \Rpackage{msm} package provides a framework for fitting
continuous-time hidden Markov models with general, continuous
outcomes.  As before, we use the \Rfunction{msm} function itself.

\paragraph{Specifying the hidden Markov model}

A hidden Markov model consists of two related components:
\begin{itemize}
\item the model for the evolution of the underlying Markov chain,
\item the set of models for the observed data conditionally on each
  underlying state.
\end{itemize}

The model for the transitions between underlying states is specified
as before, by supplying a \Rfunarg{qmatrix}.  The model for the
outcomes is specified using the argument \Rfunarg{hmodel} to
\Rfunction{msm}.  This is a list, with one element for each underlying
state, in order.  Each element of the list should be an object
returned by a hidden Markov model \emph{constructor function}.  The
HMM constructor functions provided with \Rpackage{msm} are listed in
Table~\ref{tab:hmm:dists}.  There is a separate constructor function
for each class of outcome distribution, such as uniform, normal or
gamma.

Consider a three-state hidden Markov model, with a transition
intensity matrix of
\[
Q = \left(
  \begin{array}{llll}
    -q_{12} & q_{12} &  0  \\
    0  & -q_{23} & q_{23}\\
    0    &   0    &   0   \\
  \end{array}
\right )
\]
Suppose the outcome distribution for state 1 is Normal$(\mu_1,
\sigma^2_1)$, the distribution for state 2 is Normal$(\mu_2,
\sigma^2_2)$, and state 3 is exactly observed. Observations of state 3
are given a label of -9 in the data.  Here our \Rfunarg{hmodel}
argument should be a list of objects returned by \Rfunction{hmmNorm} and
\Rfunction{hmmIdent} constructor functions.

We must specify initial values for the parameters as the arguments to
the constructor functions.  For example, we take initial values of
$\mu_1 = 90, \sigma_1 = 8, \mu_2 = 70, \sigma_2 = 8$.  Initial values
for $q_{12}$ and $q_{23}$ are 0.25 and 0.2. Finally suppose the
observed data are in a variable called \Robject{y}, the measurement
times are in \Robject{time}, and subject identifiers are in
\Robject{ptnum}.  The call to \Rfunction{msm} to estimate the
parameters of this hidden Markov model would then be
\begin{Scode}
  msm( y ~ time, subject=ptnum, data = example.df,
  qmatrix = rbind( c(0, 0.25, 0), c(0, 0, 0.2), c(0, 0, 0)),
  hmodel = list (hmmNorm(mean=90, sd=8), hmmNorm(mean=70, sd=8),
                 hmmIdent(-9)) )
\end{Scode}

\begin{table}[htbp]
\scriptsize
  \centering
  \begin{tabular}{lp{0.8in}p{0.6in}p{0.6in}p{0.7in}l}
    \hline
    Function & Distribution & Parameters &  &  Location (link) & Density for an observation $x$  \\
    \hline
    \Rfunction{hmmCat}   & Categorical   & \Rfunarg{prob, basecat}  & $p,c_0$ & $p$ (logit) & $p_x$, $x = 1, \ldots, n$  \\
    \Rfunction{hmmIdent}   & Identity  & \Rfunarg{x}  & $x_0$  & & $I_{x = x_0}$  \\
    \Rfunction{hmmUnif}   & Uniform  & \Rfunarg{lower, upper}  & $l,u$  & & $1 / (u - l)$, $u \leq x \leq l$  \\
    \Rfunction{hmmNorm}   & Normal  & \Rfunarg{mean, sd}  & $\mu,\sigma$  & $\mu$ (identity) & $\phi(x, \mu, \sigma) = \frac{1}{\sqrt{2 \pi \sigma^2}} \exp(-(x - \mu)^2/(2 \sigma^2) )$  \\
    \Rfunction{hmmLNorm}   & Log-normal  & \Rfunarg{meanlog, sdlog}  & $\mu,\sigma$  & $\mu$ (identity) & $\frac{1}{x \sqrt{2 \pi \sigma^2}} \exp(-(\log x - \mu)^2 / (2 \sigma^2))$   \\
    \Rfunction{hmmExp}   & Exponential  & \Rfunarg{rate}  & $\lambda$  & $\lambda$ (log) & $\lambda e^{- \lambda x}$, $x > 0$  \\
    \Rfunction{hmmGamma}   & Gamma  & \Rfunarg{shape, rate}  & $n,\lambda$  &  $\lambda$ (log) & $\frac{\lambda^n}{\Gamma(n)}x^{n-1} \exp(-\lambda x)$, $x > 0, n > 0, \lambda > 0$   \\
    \Rfunction{hmmWeibull}   & Weibull  & \Rfunarg{shape, scale}  & $a, b$  & $b$ (log) & $\frac{a}{b} (\frac{x}{b})^{a-1} \exp{(-(\frac{x}{b})^a)}$, $x > 0$  \\
    \Rfunction{hmmPois}   & Poisson  & \Rfunarg{rate}  & $\lambda$  & $\lambda$ (log) & $\lambda^x \exp(-\lambda)/x!$, $x = 0, 1, 2, \ldots$  \\
    \Rfunction{hmmBinom}   & Binomial  & \Rfunarg{size, prob}  & $n,p$  & $p$ (logit) & ${n \choose x} p^x (1-p)^{n-x}$  \\
    \Rfunction{hmmNBinom}   & Negative binomial  & \Rfunarg{disp, prob}  & $n,p$  & $p$ (logit) & $\Gamma(x+n)/(\Gamma(n)x!) p^n (1-p)^x$ \\
    \Rfunction{hmmBeta}   & Beta  & \Rfunarg{shape1,shape2}  & $a,b$  & & $ \Gamma(a+b) / (\Gamma(a)\Gamma(b))x^{a-1}(1-x)^{b-1}$ \\
    \Rfunction{hmmT}   & Student $t$  & \Rfunarg{mean, scale, df}  & $\mu,\sigma,k$  &  $\mu$ (identity) & $\frac{\Gamma\left((k+1)/2\right)}{\Gamma(k/2)}{\sqrt{\frac{1}{k\pi\sigma^2}}}\left\{1 + \frac{1}{k\sigma^2}(x - \mu)^{2} \right\}^{-(k + 1)/2}$ \\
    \Rfunction{hmmTNorm}   & Truncated normal  & \Rfunarg{mean, sd, lower, upper}  & $\mu,\sigma,l,u$  & $\mu$ (identity) &
\parbox[t]{2in}{$\phi(x, \mu, \sigma) / \\ (\Phi(u, \mu, \sigma) - \Phi(l, \mu, \sigma))$, \\ where $\Phi(x,\mu,\sigma) = \int_{-\infty}^x \phi(u,\mu,\sigma) du$}
 \\
    \Rfunction{hmmMETNorm}   & Normal with truncation and measurement error  & \Rfunarg{mean, sd, lower, upper, sderr, meanerr}  & \parbox[t]{1in} {$\mu_0,\sigma_0,l,u$, \\ $\sigma_\epsilon,\mu_\epsilon$}  &  $\mu_\epsilon$ (identity) &
\parbox[t]{2in}{$( \Phi(u, \mu_2, \sigma_3) - \Phi(l, \mu_2, \sigma_3)) / $ \\
$(\Phi(u, \mu_0, \sigma_0) - \Phi(l, \mu_0, \sigma_0)) $ \\
$\times \phi(x, \mu_0 + \mu_\epsilon, \sigma_2)$, \\
$\sigma_2^2 = \sigma_0^2 + \sigma_\epsilon^2$, \\
$\sigma_3 = \sigma_0 \sigma_\epsilon / \sigma_2$, \\
$\mu_2 = (x - \mu_\epsilon) \sigma_0^2 + \mu_0 \sigma_\epsilon^2$}
 \\
    \Rfunction{hmmMEUnif}   & Uniform with measurement error  & \Rfunarg{lower, upper, sderr, meanerr}  & $l,u,\mu_\epsilon,\sigma_\epsilon$  & $\mu_\epsilon$ (identity) &
    \parbox[t]{2in}{$(\Phi(x, \mu_\epsilon+l, \sigma_\epsilon) - \Phi(x, \mu_\epsilon+u, \sigma_\epsilon)) / \\ (u - l)$} \\
    \hline
  \end{tabular}
  \caption{Hidden Markov model distributions in \Rpackage{msm}.}
  \label{tab:hmm:dists}
\end{table}

\paragraph{Covariates on hidden Markov model parameters}

Most of the outcome distributions can be parameterised by covariates,
using a link-transformed linear model.  For example, an observation
$y_{ij}$ may have distribution $f_1$ conditionally on underlying state
1. The link-transformed parameter $\theta_1$ is a linear function of
the covariate vector $x_{ij}$ at the same observation time.
\begin{eqnarray*}
  \label{eq:hmm:covs}
  y_{ij} | S_{ij} &  \sim &   f_1 (y | \theta_1, \gamma_1)\\
  g(\theta_1) & = & \alpha + \beta^T x_{ij}
\end{eqnarray*}
Specifically, parameters named as the ``Location'' parameter in
Table~\ref{tab:hmm:dists} can be modelled in terms of covariates, with
the given link function.

The \Rfunarg{hcovariates} argument to \Rfunction{msm} specifies the
model for covariates on the hidden Markov outcome distributions.  This
is a list of the same length as the number of underlying states, and
the same length as the \Rfunarg{hmodel} list.  Each element of the
list is a formula, in standard R linear model notation, defining the
covariates on the distribution for the corresponding state.  If there
are no covariates for a certain hidden state, then insert a \texttt{NULL} in
the corresponding place in the list.  For example, in the three-state
normal-outcome example above, suppose that the normal means on states
1 and 2 are parameterised by a single covariate $x$.
\[
\mu_1 = \alpha_1 + \beta_1 x_{ij}, \qquad \mu_2 = \alpha_2 + \beta_2 x_{ij}.
\]
The equivalent call to \Rfunction{msm} would be
\begin{Scode}
  msm( state ~ time, subject=ptnum, data = example.df,
  qmatrix = rbind( c(0, 0.25, 0), c(0, 0, 0.2), c(0, 0, 0)),
  hmodel = list (hmmNorm(mean=90, sd=8), hmmNorm(mean=70, sd=8),
                 hmmIdent(-9)),
  hcovariates = list ( ~ x, ~ x, NULL)
 ).
\end{Scode}

\paragraph{Constraints on hidden Markov model parameters}

Sometimes it is realistic that parameters are shared between some of
the state-specific outcome distributions.  For example, the
Normally-distributed outcome in the previous example could have a
common variance $\sigma^2_1 = \sigma^2_2 = \sigma^2$ between states 1
and 2, but differing means.  It would also be realistic for any
covariates on the mean to have a common effect $\beta_1 = \beta_2 =
\beta$ on the state 1 and 2 outcome distributions.

The argument \Rfunarg{hconstraint} to \Rfunction{msm} specifies which
hidden Markov model parameters are constrained to be equal.  This is a
named list.  Each element is a vector of constraints on the named
hidden Markov model parameter. The vector has length equal to the
number of times that class of parameter appears in the whole model.
As for the other constraint arguments such as \Rfunarg{qconstraint},
identical values of this vector indicate parameters constrained to be
equal.

For example consider the three-state hidden Markov model described
above, with normally-distributed outcomes for states 1 and 2.
To constrain the outcome variance to be equal for states 1 and 2,
and to also constrain the effect of \Robject{x} on the outcome mean
to be equal for states 1 and 2, specify
\begin{Scode}
hconstraint = list(sd = c(1,1), x=c(1,1))
\end{Scode}

\paragraph{FEV$_1$ after lung transplants}

Now we give an example of fitting a hidden Markov model to a real
dataset.  The data on FEV$_1$ measurements from lung transplant
recipients, described in \ref{sec:hmm:example:fev}, are provided with
the \Rpackage{msm} package in a dataset called \Robject{fev}.

<<>>=
data(fev)
@

We fit models Models 1 and 2, each with three states and common $Q$
matrix.
<<>>=
three.q <- rbind(c(0, exp(-6), exp(-9)), c(0, 0, exp(-6)), c(0, 0, 0))
@

The simpler Model 1 is specified as follows. Under this model the
FEV$_1$ outcome is Normal with unknown mean and variance, and the mean
and variance are different between BOS state 1 and state 2.
\Rfunarg{hcovariates} specifies that the mean of the Normal outcome
depends linearly on acute events.  Specifically, this covariate is an
indicator for the occurrence of an acute event within 14 days of the
observation, denoted \texttt{acute} in the data.  As an initial guess,
we suppose the mean FEV$_1$ is 100\% baseline in state 1, and 54\%
baseline in state 2, with corresponding standard deviations 16 and 18,
and FEV$_1$ observations coinciding with acute events are on average
8\% baseline lower. \Rfunarg{hconstraint} specifies that the acute
event effect is equal between state 1 and state 2.

Days of death are coded as 999 in the \texttt{fev} outcome variable.

<<>>=
hmodel1 <- list(hmmNorm(mean=100, sd=16), hmmNorm(mean=54, sd=18), hmmIdent(999))

fev1.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q, death=3, hmodel=hmodel1,
                hcovariates=list(~acute, ~acute, NULL), hcovinits = list(-8, -8, NULL),
                hconstraint = list(acute = c(1,1)), method="BFGS")

fev1.msm

sojourn.msm(fev1.msm)
@

Printing the \Rclass{msm} object \Rfunarg{fev1.msm} shows estimates
and confidence intervals for the transition intensity matrix and the
hidden Markov model parameters.  The estimated within-state means of
FEV$_1$ are around 98\% and 52\% baseline respectively.  From the
estimated transition intensities, individuals spend around 1421 days
(3.9 years) before getting BOS, after which they live for an average
of 1248 days (3.4 years).  FEV$_1$ is lower by an average of 8\%
baseline within 14 days of acute events.

Model 2, where the outcome distribution is a more complex two-level
model, is specified as follows.  We use the distribution defined by
equations \ref{eq:fev:level1}--\ref{eq:fev:level2}. The
\Rfunction{hmmMETNorm} constructor defines the truncated normal
outcome with an additional normal measurement error.  The explicit
probability density for this distribution is given in
Table~\ref{tab:hmm:dists}.

Our initial values are again 100 and 54 for the means of the
within-state distribution of \emph{underlying} FEV$_1$, and 16 and 18
for the standard errors.  This time, underlying FEV$_1$ is truncated
normal. The truncation limits \Rfunarg{lower} and \Rfunarg{upper} are
not estimated.  We take an initial measurement error standard
deviation of \Rfunarg{sderr=8}. The extra shift \Rfunarg{meanerr} in
the measurement error model is fixed to zero and not estimated.

The \Rfunarg{hconstraint} specifies that the measurement error
variance $\sigma^2_\epsilon$ is equal between responses in states 1
and 2, as is the effect of short-term acute events on the FEV$_1$
response.

<<>>=
hmodel2 <- list(hmmMETNorm(mean=100, sd=16, sderr=8, lower=80, upper=Inf, meanerr=0),
                hmmMETNorm(mean=54, sd=18, sderr=8, lower=0, upper=80, meanerr=0),
                hmmIdent(999))

fev2.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q, death=3, hmodel=hmodel2,
                hcovariates=list(~acute, ~acute, NULL), hcovinits = list(-8, -8, NULL),
                hconstraint = list(sderr = c(1,1), acute = c(1,1)), method="BFGS")

fev2.msm

sojourn.msm(fev2.msm)
@

Under this model the standard deviation of FEV$_1$ measurements caused
by measurement error (more realistically, natural short-term
fluctuation) is around 9\% baseline.  The estimated effect of acute
events on FEV$_1$ and sojourn times in the BOS-free state and in BOS
before death are similar to Model 1.

The following plot illustrates a trajectory of declining FEV$_1$ from
the first lung transplant recipient in this dataset.  This is produced
by the following R code.  The Viterbi algorithm is used to locate the
most likely point at which this individual moved from BOS state 1 to
BOS state 2, according to the fitted Model 2.  This is illustrated by
the vertical dotted line.  This is the point at which the individual's
lung function started to remain consistently below 80\% baseline
FEV$_1$.

<<fig=TRUE>>=
keep <- fev$ptnum==1 & fev$fev<999
plot(fev$days[keep], fev$fev[keep], type="l",
ylab=expression(paste("% baseline ", FEV[1])), xlab="Days after transplant")
vit <- viterbi.msm(fev2.msm)[keep,]
(max1 <- max(vit$time[vit$fitted==1]))
(min2 <- min(vit$time[vit$fitted==2]))
abline(v = mean(max1,min2), lty=2)
text(max1 - 500, 50, "STATE 1")
text(min2 + 500, 50, "STATE 2")
@

\paragraph{An alternative way of specifying a misclassification model}

This general framework for specifying hidden Markov models can also be
used to specify multi-state models with misclassification.  A
misclassification model is a hidden Markov model with a categorical
outcome distribution.  So instead of an \Rfunarg{ematrix} argument to
\Rfunction{msm}, we can use a \Rfunarg{hmodel} argument with
\Rfunction{hmmCat} constructor functions.

\Rfunction{hmmCat} takes at least one argument \Rfunarg{prob}, a
vector of probabilities of observing outcomes of $1, 2, \ldots, n$
respectively, where $n$ is the length of \Rfunarg{prob}.  All outcome
probabilities with an initial value of zero are assumed to be fixed at
zero.  \Rfunarg{prob} is scaled if necessary to sum to one.

The model in section~\ref{sec:fitting:hmm:misc} specifies that an
individual occupying underlying state 1 can be observed as states 2
(and 1), underlying state 2 can be observed as states 1, 2 or 3, and
state 3 can be observed as states 2 or 3, and underlying state 4
(death) cannot be misclassified. Initial values of 0.1 are given for
the 1-2, 2-1, 2-3 and 3-2 misclassification probabilities.

This is equivalent to the model below, specified using a
\Rfunarg{hmodel} argument to \Rfunction{msm}.  The maximum likelihood
estimates should be the same as before (Model 5).

<<>>=
oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
                   c(0, 0, 0.202, 0.081),
                   c(0, 0, 0, 0.126),
                   c(0, 0, 0, 0))
cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                     hmodel = list (hmmCat(c(0.9, 0.1, 0, 0)), hmmCat(c(0.1, 0.8, 0.1, 0)), hmmCat(c(0, 0.1, 0.9, 0)), hmmIdent(4)),
                     qmatrix = oneway4.q, obstrue=firstobs, death = 4, method="BFGS")
cavmisc.msm
@

\subsubsection{Defining a new hidden Markov model distribution}

Suppose the hidden Markov model outcome distributions supplied with
\Rpackage{msm} (Table~\ref{tab:hmm:dists}) are insufficient.  We want
to define our own univariate distribution, called
\Rfunction{hmmNewDist}, taking two parameters \Robject{location} and
\Robject{scale}.  Download the source package, for example
\texttt{msm-0.7.2.tar.gz} for version 0.7.2, from CRAN and edit the
files in there, as follows.

\begin{enumerate}
\item Add an element to \Robject{.msm.HMODELPARS} in the file
  \texttt{R/constants.R}, naming the parameters of the distribution.
  For example
\begin{Scode}
  newdist = c('location', 'scale')
\end{Scode}

\item Add a corresponding element to the C variable \texttt{HMODELS}
  in the file \texttt{src/lik.c}.  This MUST be in the same position
  as in the \Robject{.msm.HMODELPARS} list.  For example,
\begin{Scode}
hmmfn HMODELS[] = {
  ...,
  hmmNewDist
};.
\end{Scode}

\item The new distribution is allowed to have one parameter which can
  be modelled in terms of covariates.  Add the name of this parameter
  to the named vector \Robject{.msm.LOCPARS} in
  \texttt{R/constants.R}.  For example \texttt{newdist = 'location'}.
  Specify \texttt{newdist = NA} if there is no such parameter.

\item Supposed we have specified a parameter with a non-standard name,
  that is, one which doesn't already appear in
  \Robject{.msm.HMODELPARS}. Standard names include, for example,
  \texttt{'mean'}, \texttt{'sd'}, \texttt{'shape'} or
  \texttt{'scale'}.  Then we should add the allowed range of the
  parameter to \Robject{.msm.PARRANGES}. In this example, we add
  \texttt{meanpars = c(-Inf, Inf)} to \Robject{.msm.PARRANGES}.  This
  ensures that the optimisation to estimate the parameter takes place
  on a suitable scale, for example, a log scale for a parameter
  constrained to be positive.  If the parameter should be fixed during
  maximum likelihood estimation (for example, the denominator of a
  binomial distribution) then add its name to \Robject{.msm.AUXPARS}.

\item Add an R constructor function for the distribution to
  \texttt{R/hmm-dists.R}.  For a simple univariate distribution, this is of the
  form
\begin{Scode}
  hmmNewDist <- function(location, scale)
  {
    hmmDIST(label = "newdist",
    link = "identity",
    r = function(n) rnewdist(n, location, scale),
    match.call())
  }
\end{Scode}

  \begin{itemize}
  \item The \texttt{'label'} must be the same as the name you
    supplied for the new element of \Robject{.msm.HMODELPARS}

  \item \texttt{link} is the link function for modelling the location
    parameter of the distribution as a linear function of covariates. This
    should be the quoted name of an R function.  A log link is
    \texttt{'log'} and a logit link is \texttt{'qlogis'}.  If
    using a new link function other than \texttt{'identity'},
    \texttt{'log'}, or \texttt{'qlogis'}, you should add its name
    to the vector \Robject{.msm.LINKFNS} in \texttt{R/constants.R},
    and add the name of the corresponding inverse link to
    \Robject{.msm.INVLINK}.  You should also add the names of these
    functions to the C array \texttt{LINKFNS} in \texttt{src/lik.c},
    and write the functions if they do not already exist.

  \item \texttt{r} is an R function, of the above format, returning a
    vector of \texttt{n} random values from the distribution.
    You should write this if it doesn't already exist.
  \end{itemize}

\item Add the name of the new constructor function to the NAMESPACE in
  the top-level directory of the source package.

\item Write a C function to compute the probability density of the
  distribution, and put this in \texttt{src/hmm.c}, with a declaration
  in \texttt{src/hmm.h}. This must be of the form
\begin{Scode}
  double hmmNewDist(double x, double *pars)
\end{Scode}

  where \texttt{*pars} is a vector of the parameters of the
  distribution, and the density is evaluated at \texttt{x}.

\item Update the documentation (\texttt{man/hmm-dists.Rd}) and the
  distribution table in\\ \texttt{inst/doc/msm-manual.Rnw}) if you like.

\item Recompile the package (see the ``Writing R Extensions'' manual)

\end{enumerate}

Your new distribution will be available to use in the \Rfunarg{hmodel}
argument to \Rfunction{msm}, as, for example
\begin{Scode}
hmodel = list(..., hmmNewDist(location = 0, scale = 1), ...)
\end{Scode}
If your distribution may be of interest to others, ask me nicely
(\texttt{chris.jackson@mrc-bsu.cam.ac.uk}) to include it in a future
release.

\clearpage

\section{\Rpackage{msm} reference guide}

The R help page for \Rfunction{msm} gives details of all the allowed
arguments and options to the \Rfunction{msm} function.  To view this
online in R, type:

<<eval=FALSE>>=
help(msm)
@

Similarly all the other functions in the package have help pages,
which should always be consulted in case of doubt about how to
call them.  The web-browser based help interface may often be
convenient - type

<<eval=FALSE>>=
help.start()
@
and navigate to \textsf{Packages} $\ldots$ \textsf{msm}, which brings
up a list of all the functions in the package with links to their
documentation, and a link to this manual in PDF format.

\appendix

\section{Changes in the msm package}

For a detailed list of the changes in recent versions of
\Rpackage{msm}, see the \texttt{NEWS} file in the top-level directory
of the installed package.  The \texttt{Changelog} file in the
top-level directory of the source package contains more technical
detail about these changes.

\section{Get in touch}

If you use \Rpackage{msm} in your work, whatever your field of
application, please let me know, for my own interest!  Suggestions for
improvement are welcome.

\clearpage

\bibliography{msm}

\end{document}

### 2013-04-18 pkg/inst/doc/msm-manual.Rnw
See if the version number code in the Rnw works this time with a relative path.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@7 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Remove NBSPs from doc. ... in summary, damp in Fisher.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@9 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 <<echo=FALSE>>=
-version <- gsub("Version: +", "", packageDescription("msm", lib.loc="~/work/msm")$Version)
+version <- gsub("Version: +", "",
+                packageDescription("msm", lib.loc="../../..")$Version)
 @
@@
 
-In a continuous-time Markov model, a single period of occupancy (or
+In a time-homogeneous continuous-time Markov model, a single period of occupancy (or
 \emph{sojourn time}) in state $r$ has an exponential distribution,
@@
 diseases. A commonly-used model is illustrated in
-Figure~\ref{fig:disease}.  This represents a series of successively
+Figure \ref{fig:disease}.  This represents a series of successively
 more severe disease stages, and an `absorbing' state, often death.
@@
 homogeneous continuous-time Markov process, with a transition matrix
-$Q$, pictured below Figure~\ref{fig:disease}.
+$Q$, pictured below Figure \ref{fig:disease}.
 
@@
 states representing health, illness and death
-(Figure~\ref{fig:multi:illdeath}).  Transitions are permitted from
+(Figure \ref{fig:multi:illdeath}).  Transitions are permitted from
 health to illness, illness to death and health to death.  Recovery
@@
 A typical sampling situation is illustrated in
-Figure~\ref{fig:multi:sampling}. The individual is observed at four
+Figure \ref{fig:multi:sampling}. The individual is observed at four
 occasions through 10 months. The final occasion is the death date
@@
 \end{itemize}
-Gr\"uger \etal~\cite{gruger:valid} discussed conditions under
+Gr\"uger \etal \cite{gruger:valid} discussed conditions under
 which sampling times are \emph{informative}. If a multi-state model is
@@
 Kalbfleisch and Lawless\cite{kalbfleisch:lawless} and later
-Kay~\cite{kay:mark} described a general method for evaluating the
+Kay \cite{kay:mark} described a general method for evaluating the
 likelihood for a general multi-state model in continuous time,
@@
 information is the observed state at a set of times, as in
-Figure~\ref{fig:multi:sampling}.  The sampling times are assumed to be
+Figure \ref{fig:multi:sampling}.  The sampling times are assumed to be
 non-informative.
@@
 scaled transition intensity matrix (see, for example, Cox and
-Miller~\cite{cox:miller}).
+Miller \cite{cox:miller}).
 \begin{equation}
@@
 not element-wise scalar multiplication.  It is notoriously difficult
-to calculate reliably, as discussed by Moler and van Loan~\cite{matrixexp}.
+to calculate reliably, as discussed by Moler and van Loan \cite{matrixexp}.
 For simpler models, it is feasible to calculate an analytic expression
@@
 2, 3, 4 and 5-state models. These models are illustrated in
-Figures~\ref{fig:anp2}--\ref{fig:anp5}.  For other models, $P(t)$ is
+Figures \ref{fig:anp2}--\ref{fig:anp5}.  For other models, $P(t)$ is
 determined from the matrix exponential, calculated using eigensystem
 decomposition (if eigenvalues are distinct) or a method based on
-Pad\'e approximants with scaling and squaring~\cite{matrixexp} (if
+Pad\'e approximants with scaling and squaring \cite{matrixexp} (if
 there are repeated eigenvalues). Notice that the states are not
@@
 models, but much richer models are possible for this type of data. For
-example, Putter \etal~\cite{putter:mstate} discussed the
+example, Putter \etal \cite{putter:mstate} discussed the
 \Rpackage{mstate} software for semi-parametric multi-state models with
@@
 known to be within bounded intervals.  This leads to a likelihood
-based on equation~\ref{eq:multi:lik:contrib}.
+based on equation \ref{eq:multi:lik:contrib}.
 
@@
 at the end of the study.  In this case,
-likelihood~\ref{eq:multi:lik:contrib} applies.  Although the
+likelihood \ref{eq:multi:lik:contrib} applies.  Although the
 \emph{survival time} is censored, the \emph{state} at the end of the
@@
 intensity can be investigated by modelling the intensity as a function
-of these variables.  Marshall and Jones~\cite{retino2} described a
+of these variables.  Marshall and Jones \cite{retino2} described a
 form of a {\em proportional hazards} model, where the transition
@@
 changes, it could be considered as a "censored" state (as in
-Section~\ref{sec:multi:likelihood}).
+Section \ref{sec:multi:likelihood}).
 
@@
 
-Marshall and Jones~\cite{retino2} described likelihood ratio and Wald
+Marshall and Jones \cite{retino2} described likelihood ratio and Wald
 tests for covariate selection and testing hypotheses, for example
@@
 governed by a transition intensity matrix $Q$ as before.
-(Figure~\ref{fig:multi:hidden}).  Hidden Markov models are mixture
+(Figure \ref{fig:multi:hidden}).  Hidden Markov models are mixture
 models, where observations are generated from a certain number of
@@
 As an extension to the simple multi-state model described in
-section~\ref{sec:multistate}, the \Rpackage{msm} package can fit a
+section \ref{sec:multistate}, the \Rpackage{msm} package can fit a
 general multi-state model with misclassification.  For patient $i$,
@@
 as a product of matrices, as follows. To derive this matrix product,
-decompose the overall sum in equation~\ref{eq:multi:hiddencontrib}
+decompose the overall sum in equation \ref{eq:multi:hiddencontrib}
 into sums over each underlying state. The sum is accumulated over the
@@
 \]
-Section~\ref{sec:fitting:hmm:misc} describes how to fit multi-state
+Section \ref{sec:fitting:hmm:misc} describes how to fit multi-state
 models with misclassification using the \Rpackage{msm} package, and
-Section~\ref{sec:fitting:hmm:general} describes how to apply general
+Section \ref{sec:fitting:hmm:general} describes how to apply general
 hidden Markov models.
@@
 
-Jackson and Sharples~\cite{jackson:sharples:2002} described a model
+Jackson and Sharples \cite{jackson:sharples:2002} described a model
 for FEV$_1$ (forced expiratory volume in 1 second) in recipients of
@@
 function.  In this example, BOS was modelled as a discrete, staged
-process, a model of the form illustrated in Figure~\ref{fig:disease},
+process, a model of the form illustrated in Figure \ref{fig:disease},
 with 4 states.  State 1 represents absence of BOS. State 1 BOS is
@@
 \paragraph{Model 1}
-Jackson~\cite{my:phd} considered models for these data where
+Jackson \cite{my:phd} considered models for these data where
 FEV$_1$ were Normally distributed, with an unknown mean and variance
-conditionally each state~(\ref{eq:fev:normal}).  This model seeks the
+conditionally each state (\ref{eq:fev:normal}).  This model seeks the
 most likely location for the within-state FEV$_1$ means.
@@
 \paragraph{Model 2}
-Jackson and Sharples~\cite{jackson:sharples:2002} used a more complex
+Jackson and Sharples \cite{jackson:sharples:2002} used a more complex
 two-level model for FEV$_1$ measurements. Level
-1~(\ref{eq:fev:level1}) represents the short-term fluctuation error of
+1 (\ref{eq:fev:level1}) represents the short-term fluctuation error of
 the marker around its underlying continuous value $y^{hid}_{ij}$.
-Level 2~(\ref{eq:fev:level2}) represents the distribution of
+Level 2 (\ref{eq:fev:level2}) represents the distribution of
 $y^{hid}_{ij}$ conditionally on each underlying state, as follows.
@@
 $y_{ij}$ conditionally on each underlying state (given in
-Section~\ref{sec:fitting:hmm:general}, Table~\ref{tab:hmm:dists}).
+Section \ref{sec:fitting:hmm:general}, Table \ref{tab:hmm:dists}).
 Similar distributions were originally applied by Satten and
-Longini~\cite{sattlong} to modelling the progression through discrete,
+Longini \cite{sattlong} to modelling the progression through discrete,
 unobserved HIV disease states using continuous CD4 cell counts.  The
@@
 
-Section~\ref{sec:fitting:hmm:general} describes how these and more
+Section \ref{sec:fitting:hmm:general} describes how these and more
 general hidden Markov models can be fitted with the \Rpackage{msm}
@@
 recipients, is provided with \Rpackage{msm}.  Sharples
-\etal~\cite{my:cav} studied the progression of coronary allograft
+\etal \cite{my:cav} studied the progression of coronary allograft
 vasculopathy (CAV), a post-transplant deterioration of the arterial
@@
 consecutive states while alive, and die from any state. Thus the model
-is illustrated by Figure~\ref{fig:disease} with four states, and we
+is illustrated by Figure \ref{fig:disease} with four states, and we
 have
@@
 that the data are \emph{snapshots} of the process (see
-section~\ref{sec:arbitr-observ-times}).  Although there were 44
+section \ref{sec:arbitr-observ-times}).  Although there were 44
 occasions on which a patient was observed in state 1 followed by state
@@
 functions will be described in more detail in
-section~\ref{sec:extractor}.
+section \ref{sec:extractor}.
 
@@
   for a given set of covariate values, as shown in
-  section~\ref{sec:msm:covariates}.  Confidence intervals are
+  section \ref{sec:msm:covariates}.  Confidence intervals are
   calculated from the covariance matrix of the estimates by assuming
@@
   cost of computational time.  For more about bootstrapping in
-  \Rpackage{msm}, see Section~\ref{sec:boot}.
+  \Rpackage{msm}, see Section \ref{sec:boot}.
 
@@
   \Rfunarg{ci=''boot''} is specified, then a bootstrap confidence
-  interval for the transition probability matrix is calculated (see Section~\ref{sec:boot}) .
+  interval for the transition probability matrix is calculated (see Section \ref{sec:boot}) .
   However, both of these are computationally intensive, particularly
@@
 \emph{transitions}.  These are assumed to be independent when
-calculating the likelihood (Section~\ref{sec:multi:likelihood}).  For
+calculating the likelihood (Section \ref{sec:multi:likelihood}).  For
 hidden Markov models and models with censoring, a bootstrap dataset is
@@
 datasets may occasionally fail to converge (as discussed in
-Section~\ref{sec:failure}) even if the original model fit did
+Section \ref{sec:failure}) even if the original model fit did
 converge.  In these cases, a warning is given, but
@@
 
-  For a staged disease model (Figure~\ref{fig:disease}), the number of
+  For a staged disease model (Figure \ref{fig:disease}), the number of
   disease states should be low enough that all transition rates can be
@@
   appropriate to your data (see
-  Section~\ref{sec:arbitr-observ-times}). By default, \Rfunction{msm}
+  Section \ref{sec:arbitr-observ-times}). By default, \Rfunction{msm}
   assumes that the data represent snapshots of the process, and the
@@
 expected numbers of transitions, as described by Aguirre-Hernandez and
-Farewell~\cite{ahf}.  This leads to a formal test of goodness-of-fit,
+Farewell \cite{ahf}.  This leads to a formal test of goodness-of-fit,
 analogous to the classical Pearson $\chi^2$ test for contingency
@@
 from a set of independent but non-identical multinomial distributions.
-Titman~\cite{titman:asympnull} showed that the null distribution 
+Titman \cite{titman:asympnull} showed that the null distribution
 of $T$ is asymptotically approximated by a weighted sum of $\chi^2_1$
-random variables.   Aguirre-Hernandez and Farewell~\cite{ahf} also showed that $\chi^2_{n-p}$
+random variables.   Aguirre-Hernandez and Farewell \cite{ahf} also showed that $\chi^2_{n-p}$
 is a good approximation if there are no fitted covariates. For models
@@
 $p$-value of the statistic can be obtained from the $\chi^2_{n-p}$ and
-$\chi^2_n$ distributions.  Aguirre-Hernandez and Farewell~\cite{ahf}
+$\chi^2_n$ distributions.  Aguirre-Hernandez and Farewell \cite{ahf}
 also described a bootstrap procedure for obtaining an accurate $p$-value.
 
-Titman and Sharples~\cite{titman:sharples} described modifications to
+Titman and Sharples \cite{titman:sharples} described modifications to
 the test to correct for the biases introduced where in addition to the
@@
 transplant dataset (a version of which was also analysed by
-Titman and Sharples~\cite{titman:sharples}). The default three interval groups are used,
+Titman and Sharples \cite{titman:sharples}). The default three interval groups are used,
 and two groups of the time since the start of the process. The
@@
 model itself is not biologically plausible, as discussed in
-Section~\ref{sec:fitting:hmm:misc}.
+Section \ref{sec:fitting:hmm:misc}.
 
@@
 In fact, in the heart transplant example from
-section~\ref{sec:datain}, it is not medically realistic for patients
+section \ref{sec:datain}, it is not medically realistic for patients
 to recover from a diseased state to a healthy state.  Progression of
@@
 state at the initial observation
-(equation~\ref{eq:multi:hidden:matprod}).  If not given, all
+(equation \ref{eq:multi:hidden:matprod}).  If not given, all
 individuals are assumed to be in true state 1 at their initial
@@
 
-As well as the functions described in section~\ref{sec:extractor} for
+As well as the functions described in section \ref{sec:extractor} for
 extracting useful information from fitted models, there are a number
@@
   states are calculated as described in
-  section~\ref{sec:model-assessment}, by assuming that observed states
+  section \ref{sec:model-assessment}, by assuming that observed states
   are retained between observation times.
@@
 \item[Goodness-of-fit test] The Pearson-type goodness-of-fit test is
-  performed, as in Section~\ref{sec:pearson}.  The table of deviances
+  performed, as in Section \ref{sec:pearson}.  The table of deviances
   indicates that there are more 1-3 and 1-4 transitions than expected
@@
 reconstructing the most likely state path is the {\em Viterbi}
-algorithm. Originally proposed by Viterbi~\cite{viterbi}, it is also
-described by Durbin \etal~\cite{biolog:seq} and Macdonald and
-Zucchini~\cite{macdonald:zucchini} for discrete-time hidden Markov
+algorithm. Originally proposed by Viterbi \cite{viterbi}, it is also
+described by Durbin \etal \cite{biolog:seq} and Macdonald and
+Zucchini \cite{macdonald:zucchini} for discrete-time hidden Markov
 chains.  For continuous-time models it proceeds as follows. Suppose
@@
 HMM constructor functions provided with \Rpackage{msm} are listed in
-Table~\ref{tab:hmm:dists}.  There is a separate constructor function
+Table \ref{tab:hmm:dists}.  There is a separate constructor function
 for each class of outcome distribution, such as uniform, normal or
@@
 Specifically, parameters named as the ``Location'' parameter in
-Table~\ref{tab:hmm:dists} can be modelled in terms of covariates, with
+Table \ref{tab:hmm:dists} can be modelled in terms of covariates, with
 the given link function.
@@
 probability density for this distribution is given in
-Table~\ref{tab:hmm:dists}.
+Table \ref{tab:hmm:dists}.
 
@@
 
-The model in section~\ref{sec:fitting:hmm:misc} specifies that an
+The model in section \ref{sec:fitting:hmm:misc} specifies that an
 individual occupying underlying state 1 can be observed as states 2
@@
 Suppose the hidden Markov model outcome distributions supplied with
-\Rpackage{msm} (Table~\ref{tab:hmm:dists}) are insufficient.  We want
+\Rpackage{msm} (Table \ref{tab:hmm:dists}) are insufficient.  We want
 to define our own univariate distribution, called

### 2013-04-18 pkg/R/msm.R

@@ -222,6 +222,7 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
             if (any(msmdata$obstype==3))
                 stop("Fisher scoring not supported with exact death times")
             if (is.null(optim.args$control$reltol)) reltol <- sqrt(.Machine$double.eps)
+            damp <- if (is.null(optim.args$control$damp)) 0 else optim.args$control$damp
             theta <- p$inits
             lik.old <- -lik.msm(theta, msmdata=msmdata, qmodel=qmodel, qcmodel=qcmodel,
                                cmodel=cmodel, hmodel=hmodel, paramdata=p)
@@ -232,7 +233,7 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
                 VI <- information.msm(theta, msmdata=msmdata, qmodel=qmodel, qcmodel=qcmodel,
                                       cmodel=cmodel, hmodel=hmodel, paramdata=p)
                 V <- -VI$deriv
-                Info <- VI$info
+                Info <- VI$info + diag(damp, nrow(VI$info))
                 theta <- theta + solve(Info, V)
                 lik.new <- -lik.msm(theta, msmdata=msmdata, qmodel=qmodel, qcmodel=qcmodel,
                                    cmodel=cmodel, hmodel=hmodel, paramdata=p)

### 2013-04-18 pkg/R/outputs.R

@@ -57,7 +57,7 @@ summary.msm <- function(object, # fitted model
                         )
 {
     if (!inherits(object, "msm")) stop("expected object to be a msm model")
-    prevalences <- prevalence.msm(object, times, timezero, initstates, covariates, misccovariates)
+    prevalences <- prevalence.msm(object, times, timezero, initstates, covariates, misccovariates, ...)
     if (object$qcmodel$ncovs > 0) {
         if (missing (hazard.scale))
             hazard.scale <- rep(1, object$data$covdata$ncovs)
@@ -1150,6 +1150,7 @@ intervaltrans.msm <- function(x=NULL, qmatrix=NULL, ematrix=NULL, exclude.absabs
 ### Estimate observed state occupancies in the data at a series of times
 ### Assume previous observed state is retained until next observation time
 ### Assumes times are sorted within patient (they are in data in msm objects)
+### TODO allow subset
 
 observed.msm <- function(x, times=NULL, interp=c("start","midpoint"), censtime=Inf)
 {
@@ -1207,6 +1208,8 @@ observed.msm <- function(x, times=NULL, interp=c("start","midpoint"), censtime=I
     obstab <- cbind(obstab, Total=rowSums(obstab))
     list(obstab=obstab, obsperc=obsperc, risk=obstab[,ncol(obstab)])
 }
+
+## TODO allow this to be integrated over covariates observed in data
 
 expected.msm <- function(x,
                          times=NULL,
@@ -1315,6 +1318,8 @@ prevalence.msm <- function(x,
                                   piecewise.times=piecewise.times, piecewise.covariates=piecewise.covariates, ...)
     res
 }
+
+## TODO plot only observed/expected curve, or add to existing plot
 
 plot.prevalence.msm <- function(x, mintime=NULL, maxtime=NULL, timezero=NULL, initstates=NULL,
                                 interp=c("start","midpoint"), censtime=Inf, covariates="mean", misccovariates="mean",

### 2013-05-17 pkg/inst/doc/msm-manual.Rnw
Bits left off last commit for matrix initprobs
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@16 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 observation at time $t$, the expected number of individuals in state
-$r$ at time $t$ is $n(t) P(t)_{0,r}$.
+$r$ at time $t$ is $n(t) P(t)_{0,r}$.   If the covariates on which 
+$P(t)$ depends vary between individuals, then this can be 
+averaged over the covariates observed in the data.
 
@@
 state at the initial observation
-(equation \ref{eq:multi:hidden:matprod}).  If not given, all
+(equation \ref{eq:multi:hidden:matprod}).  This can also be a matrix
+with number of rows equal to the number of subjects, if these 
+probabilities are subject-dependent and known. If not given, all
 individuals are assumed to be in true state 1 at their initial
@@
 probabilites are estimated as part of the model fit, using
-\Rfunarg{initprobs} as initial values.  Covariate effects on these
+a vector \Rfunarg{initprobs} as initial values.  Covariate effects on these
 probabilities can also be estimated using a multinomial logistic

### 2013-05-25 pkg/inst/doc/msm-manual.Rnw
New interface for transition-specific covariates and subject-specific likelihoods.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@18 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+\item[\Rfunction{optim} "function cannot be evaluated at initial parameters"]
+  To diagnose this problem, run \Rfunction{msm} again with
+  \Rfunarg{fixedpars=TRUE} set, to calculate the -2 log-likelihood at
+  the initial values.  This will probably be \Robject{Inf}. To show
+  the contributions of individual subjects to the overall log
+  likelihood, call \Rfunction{logLik.msm(x, by.subject=TRUE)}, where
+  \Robject{x} is the fitted model object.  If only a few subjects give
+  an infinite log-likelihood, then you can check whether their state
+  histories are particularly unusual and conflict with the model.  For
+  example, they might appear to make unusually large jumps between
+  states in short periods of time.  For models with misclassification,
+  note that the default true initial state distribution
+  \Rfunarg{initprobs} puts all individuals in true state 1 at their
+  first observation.  If someone starts in a much higher state, this
+  may result in an infinite log-likelihood, and changing
+  \Rfunarg{initprobs} would be sensible.
+
 \item[Model simplification]

### 2013-05-25 pkg/R/msm.R

@@ -11,10 +11,10 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
                 hmodel = NULL,  # list of constructors for hidden emission distributions
                 obstype = NULL, # optional, defaults to all 1 (snapshots) if not given
                 obstrue = NULL, # for hidden Markov models, which observations represent the true state
-                covariates = NULL, # formula specifying covariates on transition rates.
+                covariates = NULL, # formula or list of formulae specifying covariates on transition rates.
                 covinits = NULL,      # initial values for covariate effects
                 constraint = NULL, # which intensities have covariates on them (as in Marshall et al.)
-                misccovariates = NULL, # formula specifying covariates on misclassification probs
+                misccovariates = NULL, # formula or list of formulae specifying covariates on misclassification probs
                 misccovinits = NULL,      # initial values for misclassification covariate effects
                 miscconstraint = NULL, # which misc probs have covariates on them
                 hcovariates = NULL, # list of formulae specifying covariates model for each hidden state
@@ -132,7 +132,6 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
             if (!is.null(constraint)) warning("constraint specified but no covariates")
             list(npars=0, ncovs=0, ndpars=0)
         }
-
 ### MODEL FOR COVARIATES ON MISCLASSIFICATION PROBABILITIES
     if (!emodel$misc || is.null(misccovariates))
         ecmodel <- list(npars=0, ncovs=0)
@@ -163,7 +162,7 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
         else warning("initprobs and initcovariates ignored for non-hidden Markov models")
     }
     else if (hmodel$hidden) {
-        hmodel <- c(hmodel, list(nicovs=rep(0, hmodel$nstates-1), nicoveffs=0))
+        hmodel <- c(hmodel, list(nicovs=rep(0, hmodel$nstates-1), nicoveffs=0, cri=ecmodel$cri))
         class(hmodel) <- "hmodel"
     }
     if (hmodel$hidden && !emodel$misc) {
[…]

### 2013-05-25 pkg/R/outputs.R

@@ -1007,12 +1007,17 @@ coef.msm <- function(object, ...)
 
 ### Extract the log-likelihood
 
-logLik.msm <- function(object, ...)
+logLik.msm <- function(object, by.subject=FALSE, ...)
 {
     if (!inherits(object, "msm")) stop("expected object to be a msm model")
-    val <- - 0.5 * object$minus2loglik
-    attr(val, "df") <- object$paramdata$nopt
-    class(val) <- "logLik"
+    if (by.subject){
+        val <- -0.5*likderiv.msm(object$paramdata$params, deriv=6, object$data, object$qmodel, object$qcmodel, object$cmodel, object$hmodel, object$paramdata)
+        names(val) <- unique(object$data$subject)
+    } else { 
+        val <- - 0.5 * object$minus2loglik
+        attr(val, "df") <- object$paramdata$nopt
+        class(val) <- "logLik"
+    }
     val
 }
 

### 2013-05-25 pkg/src/lik.c

@@ -433,6 +433,35 @@ double liksimple(msmdata *d, qmodel *qm, qcmodel *qcm,
     return (-2*lik);
 }
 
+/* Likelihood for the non-hidden multi-state Markov model, by subject */
+
+double liksimple_subj(int pt, /* ordinal subject ID */
+		      msmdata *d, qmodel *qm, qcmodel *qcm,
+		      cmodel *cm, hmodel *hm)
+{
+    int i,totcovs=0, np=qm->npars, from, to;
+    double lik=0, pm=0, dt;
+    double *pmat      = Calloc((qm->nst)*(qm->nst), double);
+    double *newintens = Calloc ( qm->npars , double);
+    double *x = Calloc(qcm->ncovs[0], double);
+    for (i = d->firstobs[pt]+1; i < d->firstobs[pt+1]; ++i) {
+	GetCovData(i, d->covobs, d->whichcov, x, qcm->ncovs[0], d->n);
+	AddCovs(i, d->n, np, qcm->ncovs, qm->intens, newintens,
+		qcm->coveffect, d->covobs, d->whichcov, &totcovs, log, exp);
+	dt = d->time[i] - d->time[i-1];
+	from = fprec(d->obs[i-1] - 1, 0); /* convert state outcome to integer */
+	to = fprec(d->obs[i] - 1, 0);
+	Pmat(pmat, dt, newintens, np, qm->ivector, qm->nst, (d->obstype[i] == OBS_EXACT), qm->analyticp, qm->iso, qm->perm,  qm->qperm, 0);
+	if (d->obstype[i] == OBS_DEATH)
+	    pm = pijdeath(from, to, pmat, newintens, qm->ivector, qm->nst);
+	else
+	    pm = pmat[MI(from, to, qm->nst)];
+	lik += log(pm);
+    }
+    Free(pmat); Free(newintens); Free(x);
+    return (-2*lik);
+}
+
 /* Find zero-based index of maximum element of a vector x */
 
 void pmax(double *x, int n, int *maxi)
@@ -481,7 +510,7 @@ void Viterbi(msmdata *d, qmodel *qm, qcmodel *qcm,
 	  else lvold[k] = R_NegInf;
 	}
       }
-      else { /* use initprobs */	
+      else { /* use initprobs */
 	for (k = 0; k < qm->nst; ++k)
[…]

### 2013-05-25 pkg/src/pijt.c

@@ -719,7 +719,7 @@ void dpijdeath(int r, int s, vector x, Array3 dpmat, Matrix pmat, vector intens,
 void DPmat(Array3 dpmat, double t, vector x, vector intens, vector oldintens, ivector qvector,
 	   int n, int np, int ndp, int ndc, ivector qconstr, ivector bconstr, ivector wcov, int exacttimes)
 {
-	int i, j, p, pp, err=0;
+	int i, j, p, err=0;
     double eit, ejt;
     Matrix DQ = (Matrix) Calloc(n*n, double); /* should initialize to zero */
     vector revals = (vector) Calloc(n, double);
@@ -754,7 +754,7 @@ if (fixed[p]==1) { ++pp; }
 and replace p by pp in formdq and last multmat assignment
 need to edit dpijdeath, dmatrixexpseries, dpmatexact, derivsimple, derivsimple_subj in same way.
 */
-	    for (p=0, pp=0; p<ndp+ndc; ++p) {
+	    for (p=0; p<ndp+ndc; ++p) {
 		if (p < ndp)
 		    FormDQ(DQ, qmat, qbase, p, n, qconstr, np);
 		else FormDQCov(DQ, qmat, p-ndp, n, bconstr, wcov, np, x);

### 2013-05-25 pkg/inst/doc/msm-manual.Rnw
Manual update for new covariates interface.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@19 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 in the \Rfunarg{qmatrix} represent the intensities with covariate values
-set to their means in the data.
+set to their means in the data.  In the following model, all transition
+intensities are modelled in terms of sex.
 
@@
 qmatrix.msm(cavsex.msm, covariates=list(sex=1)) # Female
+@
+
+Different transition rates may be easily modelled on different
+covariates by specifying a named list of formulae as the
+\Rfunarg{covariates} argument.  Each element of the list has a name
+identifying the transition.  In the model below, the transition rate
+from state 1 to state 2 and the rate from state 3 to state 4 are each
+modelled on sex as a covariate, but no other intensities have
+covariates on them.
+
+\paragraph{Model 2a: transition-specific covariates}
+<<eval=FALSE>>=
+cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
+                     qmatrix = twoway4.q, death = 4,
+                     covariates = list("1-2" = ~ sex, "3-4" = ~sex) )
 @

### 2013-06-21 pkg/R/constants.R
Major internals update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@20 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -40,7 +40,7 @@
 .msm.INVLINK <- c(identity="identity", log="exp", qlogis="plogis")
 
 ### Parameters which are always fixed, never estimated
-.msm.AUXPARS <- c("lower", "upper", "which", "size", "meanerr", "ncats", "basecat", "p0", "pbase", "initp0")
+.msm.AUXPARS <- c("lower", "upper", "which", "size", "meanerr", "ncats", "basecat", "p0", "pbase", "initpbase", "initp0")
 
 ### Parameters which should be defined as integer
 .msm.INTEGERPARS <- c("size")
@@ -49,12 +49,12 @@
 .msm.PARRANGES <- list(qbase=c(0, Inf), lower=c(-Inf,Inf), upper=c(-Inf, Inf),
                        mean=c(-Inf, Inf), sd=c(0, Inf),
                        meanlog=c(-Inf,Inf), sdlog=c(0, Inf), rate=c(0, Inf), shape=c(0, Inf), shape1=c(0,Inf), shape2=c(0,Inf),
-                       prob=c(0, 1), meanerr=c(0, Inf), sderr=c(0, Inf), disp=c(0, Inf),
-                       initp=c(0, 1), initpcov=c(-Inf,Inf), df=c(0, Inf)
+                       prob=c(0, 1), meanerr=c(-Inf, Inf), sderr=c(0, Inf), disp=c(0, Inf),
+                       df=c(0, Inf)
                        )
 
 ### Transforms to optimise some parameters on a different scale
-### Univariate transforms only: doesn't include multinomial logit transform used for misclassification p.
+### Univariate transforms only: doesn't include multinomial logit transform used for misclassification p and initial state probs
 .msm.TRANSFORMS <-
   do.call("rbind",
           lapply(.msm.PARRANGES,
@@ -71,10 +71,10 @@
 ### This database is used to determine the appropriate method for calculating the analytic P matrix.
 
 ### The numbered label gives the indices into the matrix of rates (vectorised by reading across rows)
-### e.g. the model with qmatrix of the form 
-### *,1,1    
+### e.g. the model with qmatrix of the form
+### *,1,1
 ### 0,*,1
-### 0,0,*    is "1-2-4" 
+### 0,0,*    is "1-2-4"
 ### well-disease, well-death, disease-death transitions allowed.
 
 .msm.graphs <-
@@ -131,3 +131,7 @@
          )
 
        )
+
[…]

### 2013-06-21 pkg/R/hmm.R

@@ -53,7 +53,7 @@ msm.form.hmodel <- function(hmodel, hconstraint=NULL, initprobs=NULL, est.initpr
               }
           }
       }
-      nipars <- if (est.initprobs) nst - 1 else 0
+      nipars <- if (est.initprobs) nst else 0
       labels <- sapply(hmodel, function(x) x$label)
       models <- match(labels, .msm.HMODELS)
       pars <- lapply(hmodel, function(x) x$pars)

### 2013-06-21 pkg/R/msm.R

@@ -37,7 +37,7 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
                 opt.method = c("optim","nlm","fisher"),
                 hessian = TRUE,
                 use.deriv = TRUE,
-                analyticp = TRUE,
+                analyticp=TRUE,
                 ... # options to optim or nlm
                 )
 {
@@ -127,7 +127,7 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
 ### MODEL FOR COVARIATES ON INTENSITIES
     qcmodel <-
         if (msmdata$covdata$ncovs > 0)
-            msm.form.covmodel(msmdata$covdata, constraint, qmodel$npars, covinits)
+            msm.form.covmodel(msmdata$covdata, constraint, qmodel$npars, covinits, msmdata)
         else {
             if (!is.null(constraint)) warning("constraint specified but no covariates")
             list(npars=0, ncovs=0, ndpars=0)
@@ -140,7 +140,7 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
             warning("misccovariates have been specified, but misc is FALSE. Ignoring misccovariates.")
         }
         else {
-            ecmodel <- msm.form.covmodel(msmdata$misccovdata, miscconstraint, emodel$npars, misccovinits)
+            ecmodel <- msm.form.covmodel(msmdata$misccovdata, miscconstraint, emodel$npars, misccovinits, msmdata)
             hcovariates <- msm.misccov2hcov(misccovariates, emodel)
             hcovinits <- msm.misccovinits2hcovinits(misccovinits, hcovariates, emodel, ecmodel)
         }
@@ -178,10 +178,9 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
 ### CALCULATE LIKELIHOOD AT INITIAL VALUES...
     if (p$fixed) {
         p$lik <- lik.msm(p$inits, msmdata, qmodel, qcmodel, cmodel, hmodel, p)
-        p$deriv <- if (!hmodel$hidden && cmodel$ncens==0) deriv.msm(p$inits, msmdata, qmodel, qcmodel, cmodel, hmodel, p) else NULL
-        p$information <- if (!hmodel$hidden && cmodel$ncens==0) information.msm(p$inits, msmdata, qmodel, qcmodel, cmodel, hmodel, p) else NULL
[…]

### 2013-06-21 pkg/R/outputs.R

@@ -200,7 +200,7 @@ surface.msm <- function(x, params=c(1,2), np=10, type=c("contour","filled.contou
     for (i in 1:np) {
         for (j in 1:np) {
             point[i1] <- p1[i]; point[i2] <- p2[j]
-            z[i,j] <- -0.5*likderiv.msm(point, 0, x$data, x$qmodel, x$qcmodel, x$cmodel, x$hmodel, x$paramdata)
+            z[i,j] <- -0.5*Ccall.msm(point, "lik", x$data, x$qmodel, x$qcmodel, x$cmodel, x$hmodel, x$paramdata)
         }
     }
 
@@ -342,8 +342,10 @@ qematrix.msm <- function(x, covariates="mean", intmisc="intens", sojourn=FALSE, 
             0 else x$ecmodel$ncovs
         if (nc==0){
             logest <- x$Ematrices$logitbaseline
-            plabs <- x$emodel$imatrix; plabs[x$emodel$imatrix==1] <- "p"; diag(plabs)[rowSums(x$emodel$imatrix)>0] <- "pbase"
-            mat <- matrix(msm.mninvlogit.transform(t(logest), t(plabs), rep(1:nst, each=nst)), nrow=nst, byrow=TRUE)
+            plabs <- x$emodel$imatrix
+            plabs[x$emodel$imatrix==1] <- "p"
+            diag(plabs)[rowSums(x$emodel$imatrix)>0] <- "pbase"
+            mat <- matrix(msm.mninvlogit.transform(as.vector(t(logest)), as.vector(t(plabs)), rep(1:nst, each=nst)), nrow=nst, byrow=TRUE)
             diag(mat)[rowSums(x$emodel$imatrix)==0] <- 1
             mat[is.infinite(logest)] <- 1 # if any offdiagonal misc probs 1
         }
@@ -387,7 +389,7 @@ qematrix.msm <- function(x, covariates="mean", intmisc="intens", sojourn=FALSE, 
                 logest <- logest + x$Ematrices[[i+1]] * covi
             }
             plabs <- x$emodel$imatrix; plabs[x$emodel$imatrix==1] <- "p"; diag(plabs)[rowSums(x$emodel$imatrix)>0] <- "pbase"
-            mat <- matrix(msm.mninvlogit.transform(t(logest), t(plabs), rep(1:nst, each=nst)), nrow=nst, byrow=TRUE)
[…]

### 2013-06-21 pkg/R/pearson.R

@@ -7,7 +7,7 @@ pearson.msm <- function(x, transitions=NULL, timegroups=3, intervalgroups=3, cov
                         indep.cens=TRUE, # use censoring times when calculating the empirical distribution of sampling times
                         maxtimes=NULL,# upper limit for imputed next observation times
                         pval=TRUE # calculate p-values for test. false if calling from bootstrap
-                        ) {
+                        ){
 
     dat <- x$data
     ## Error handling
@@ -376,7 +376,7 @@ pearson.msm <- function(x, transitions=NULL, timegroups=3, intervalgroups=3, cov
         ## compute Psi  = Cov(score(theta), Orc), where Orc is observed counts in pearson table
         ## arranged as npars x RC
         ## bottom left and top right blocks wrong way round in paper
-        dp <- likderiv.msm(x$paramdata$opt$par, deriv=5, x$data, x$qmodel, x$qcmodel, x$cmodel, x$hmodel, x$paramdata) # ntrans x R x npars: trans in data order
+        dp <- Ccall.msm(x$paramdata$opt$par, do.what="dpmat", x$data, x$qmodel, x$qcmodel, x$cmodel, x$hmodel, x$paramdata) # ntrans x R x npars: trans in data order
         Psiarr <- apply(dp, c(2,3), function(x)tapply(x, md$group, sum))
         npars <- x$paramdata$nopt # only includes qbase,qcov, since no hmms here
         ## permute C x R x npars Psiarr to R x C x npars Psi and then collapse first two dims
@@ -426,12 +426,11 @@ pearson.msm <- function(x, transitions=NULL, timegroups=3, intervalgroups=3, cov
 
     ## Simulated observation times to use as sampling frame for bootstrapped data
     if (!is.null(imputation)) {
-      imp.times <- matrix(rep(od$time, N), nrow=length(od$time), ncol=N)
-      prevtime <- imp.times[which(od$state %in% dstates) - 1,]
-      imp.times[od$state %in% dstates, ] <- prevtime + imputation[,,"times"]
+      imp.times <- matrix(rep(x$data$time, N), nrow=length(x$data$time), ncol=N)
+      prevtime <- imp.times[which(x$data$state %in% dstates) - 1,]
+      imp.times[x$data$state %in% dstates, ] <- prevtime + imputation[,,"times"]
     }
     else imp.times <- NULL
-
     if (boot) {
[…]

### 2013-06-21 pkg/src/analyticp.c

@@ -6,36 +6,36 @@ Increases speed and stability by avoiding the numeric calculation of
 the matrix exponential.
 
 ### The numbered label gives the indices into the matrix of rates (vectorised by reading across rows)
-### e.g. the 3-state model with qmatrix of the form 
+### e.g. the 3-state model with qmatrix of the form
 ### well-disease, well-death, disease-death transitions allowed.
-### *,1,1    
+### *,1,1
 ### 0,*,1
 ### 0,0,*    uses the function p3q124()
 
-### Some models are isomorphic and use the same p?q? function. 
-### See .msm.graphs in R/constants.R for permutations. 
+### Some models are isomorphic and use the same p?q? function.
+### See .msm.graphs in R/constants.R for permutations.
 */
 
 #include "msm.h"
 
-void p2q1(Matrix pmat, double t, vector intens, int *degen);
-void p2q12(Matrix pmat, double t, vector intens, int *degen);
+void p2q1(Matrix pmat, double t, Matrix qmat, int *degen);
+void p2q12(Matrix pmat, double t, Matrix qmat, int *degen);
 
-void p3q12(Matrix pmat, double t, vector intens, int *degen);
-void p3q14(Matrix pmat, double t, vector intens, int *degen);
-void p3q16(Matrix pmat, double t, vector intens, int *degen);
-void p3q124(Matrix pmat, double t, vector intens, int *degen);
-void p3q135(Matrix pmat, double t, vector intens, int *degen);
-void p3q1246(Matrix pmat, double t, vector intens, int *degen);
+void p3q12(Matrix pmat, double t, Matrix qmat, int *degen);
+void p3q14(Matrix pmat, double t, Matrix qmat, int *degen);
+void p3q16(Matrix pmat, double t, Matrix qmat, int *degen);
+void p3q124(Matrix pmat, double t, Matrix qmat, int *degen);
[…]

### 2013-06-21 pkg/src/hmm.c

@@ -6,31 +6,20 @@
    of outcome conditionally on the hidden state).  All of the form
    double f(double x, double *pars) */
 
-/* Categorical distribution on the set (1, 2, 3, ..., pars[0]), 
-   Baseline category is given by pars[1]
-   Probabilities are defined by pars[2], ... pars[ncats+1] 
-
-   Multinomial logistic regression: 
-   On entry to this function, the baseline probabilities "oldpars"
-   may have been transformed by covariates using AddCovs, as: 
-   newpars[r] --> oldpars[r]*exp(beta x)
-   In this function, these are used to calculate the actual probabilities as: 
-   p[r] = newpars[r] / (sum of non-baseline newpars)  for r = not baseline
-   p[r] = 1 / (sum of non-baseline newpars)  for r = baseline
-      
+/* Categorical distribution on the set (1, 2, 3, ..., pars[0]),
+   Baseline category is given by pars[1] (not used any more, to delete?)
+   Probabilities are defined by pars[2], ... pars[ncats+1]
+   NEW IMPLEMENTATION in v1.3
+   pars[2],pars[3]... are absolute probs.  covariates applied in R.
+   used to be relative with covariates applied in C.
 */
 
 double hmmCat(double x, double *pars)
 {
-    /* check that data x is in the set should already have been done in R */
     int cat = fprec(x, 0);
     int ncats = fprec(pars[0], 0);
-    int basecat = fprec(pars[1], 0); /* pars[2], ... pars[ncats+1] are the probabilities */
-    double ret, *p = Calloc(ncats, double); 
-    relative2absolutep(&pars[2], p, ncats, basecat-1);
-    ret = p[cat-1];
-    Free(p);
-    return ret;	    
+    if ((cat > ncats) || (cat < 1)) return 0;
+    return pars[1 + cat];
 }
 
 double hmmIdent(double x, double *pars)
@@ -63,7 +52,7 @@ double hmmExp(double x, double *pars)
 }
 
 double hmmGamma(double x, double *pars)
-{ 
+{
     double shape = pars[0], scale = 1 / pars[1];
     return dgamma(x, shape, scale, 0);
 }
@@ -112,7 +101,7 @@ double hmmMETNorm(double x, double *pars)
     double mutmp = ((x - meanerr)*sd*sd + mean*sderr*sderr) / sumsq;
[…]

### 2013-06-21 pkg/src/lik.c

@@ -15,6 +15,7 @@
 #define NODEBUG2
 #define NOVITDEBUG
 #define NOFISHDEBUG
+#define NODERIVDEBUG
 
 linkfn LINKFNS[3][2] = {
     {identity, identity},
@@ -42,7 +43,17 @@ hmmfn HMODELS[] = {
     hmmT
 };
 
+/* MUST MATCH order of .msm.CTASKS in R/constants.R */
+#define DO_LIK 0
+#define DO_DERIV 1
+#define DO_INFO 2
+#define DO_VITERBI 3
+#define DO_LIK_SUBJ 4
+#define DO_DERIV_SUBJ 5
+#define DO_DPMAT 6
+
 #define OBS_SNAPSHOT 1
+#define OBS_PANEL 1 /* preferred term now */
 #define OBS_EXACT 2
 #define OBS_DEATH 3
 
@@ -66,49 +77,6 @@ double identity(double x)
 int all_equal(double x, double y)
 {
     return fabs (x - y) <= DBL_EPSILON * fabs(x);
-}
-
-/*
-   Add effects of covariates at an observation number obs to a set of link-transformed parameters
-   For covs on the qmatrix, we have npars q's, each with the same number of covs on them
-   For HMMs, each set is the set of parameters of the model for one state
-   Only certain of these parameters (means and rates) can have
-   covariates on them, ncovs[i] > 0
-   whichcov[j] is the column of the (vectorised) data matrix "cov" containing the jth covariate
-   totcovs is a call-by-referenced integer which counts the covariates so far, used for indexing whichcov
-   coveffect is a vectorised ncovs * npars matrix, filled by rows
-   (par1) cov1 cov2 ..  (par2) cov1 cov2
-*/
-
-void AddCovs(int obs, int nobs, int npars, int *ncovs,
-	     double *oldpars, double *newpars,
-	     double *coveffect, double *cov, int *whichcov,
-	     int *totcovs,
-	     double link(double x), double invlink(double x))
-{
-    int i, j, k=0;
-    double x;
-    for (i = 0; i < npars; ++i)
-	{
-	    newpars[i] = oldpars[i];
-	    if (ncovs[i] > 0) {
-		newpars[i] = link(newpars[i]);
-		for (j = 0; j < ncovs[i]; ++j) {
[…]

### 2013-06-21 pkg/src/msm.h

@@ -6,7 +6,8 @@
 /* index to treat a vector as a matrix. ith row, jth column. Fills columns first, as in R */
 #define MI(i, j, nrows) ( (int) ((j)*(nrows) + (i)) )
 /* index to treat a vector as a 3-dimensional array. Left-most index varies fastest, as in R */
-#define MI3(i, j, k, n1, n2) ( (int) ((k)*(n1*n2) + (j)*(n1) + (i)) )
+#define MI3(i, j, k, n1, n2) ( (int) ((k)*((n1)*(n2)) + (j)*(n1) + (i)) )
+#define MI4(i, j, k, m, n1, n2, n3) ( (int) ((m)*((n1)*(n2)*(n3)) + (k)*((n1)*(n2)) + (j)*(n1) + (i)) )
 
 /* Macros to switch quickly between C and S memory handling. Currently not used */
 
@@ -22,6 +23,7 @@
 #endif
 
 typedef double * Array3;
+typedef double * Array4;
 typedef double * Matrix;
 typedef int * iMatrix;
 typedef double * vector;
@@ -58,21 +60,15 @@ struct qmodel {
     int nst;
     int npars;
     int ndpars;
+    int ncpars;
     int *ivector;
     double *intens;
+    double *dintens;
     int analyticp;
     int iso;
     int *perm;
     int *qperm;
     int *constr;
-};
-
-struct qcmodel {
-    int *ncovs;
-    double *coveffect;
-    int *constr;
-    int ndpars;
-    int *wcov;
 };
 
 struct cmodel {
@@ -97,26 +93,17 @@ struct hmodel {
 
 typedef struct msmdata msmdata;
 typedef struct qmodel qmodel;
-typedef struct qcmodel qcmodel;
 typedef struct cmodel cmodel;
 typedef struct hmodel hmodel;
 
-double qij(int i, int j, vector intens, ivector qvector, int nstates);
-double pijdeath(int r, int s, Matrix pmat, vector intens, ivector qvector, int n);
-void Pmat(Matrix pmat, double t, vector intens, int nintens, int *qvector, int nstates, int exacttimes, int analyticp, int iso, int *perm, int *qperm, int debug);
-void DPmat(Array3 dpmat, double t, vector x, vector intens, vector oldintens, ivector qvector,
-	   int n, int np, int ndp, int ndc, ivector qconstr, ivector bconstr, ivector wcov, int exacttimes);
[…]

### 2013-06-21 pkg/src/pijt.c

@@ -25,6 +25,8 @@
 #define _MEXP_METHOD_ 1 /* 1 for Pade approximation, 2 for series. Pade is more robust. */
 #include "R_ext/Lapack.h"
 
+#define NODERIVDEBUG
+
 /* Set A to be an n x n identity matrix */
 
 void FormIdentity(Matrix A, int n)
@@ -361,50 +363,15 @@ void MatrixExp(Matrix mat, int n, Matrix expmat, double t, int debug, int degen)
     Free(work);  Free(revals);  Free(ievals); Free(evecs);  Free(evecsinv);
 }
 
-/* Fills the required entries of the intensity matrix with the current intensities */
-/* qvector supplied filled by row. intens ordered by row, qmat filled by column */
-/* TODO can probably optimise by passing this from R */
-
-void FillQmatrix(ivector qvector, vector intens, Matrix qmat, int nstates)
-{
-    int i, j, k=0;
-    for (i=0; i<nstates; ++i) {
-	qmat[MI(i, i, nstates)] = 0;
-	for (j=0; j<nstates; ++j) {
-	    if (j != i) {
-		qmat[MI(i, j, nstates)] = 0;
-		if (qvector[MI(j, i, nstates)] == 1) {
-		    qmat[MI(i, j, nstates)] += intens[k];
-		    qmat[MI(i, i, nstates)] -= intens[k];
-		    ++k;
-		}
-	    }
-	}
-    }
-}
 
 /* Returns i-j transition intensity time t given vectors of intensities and transition indicators */
-/* TODO can probably save time by passing filled Qmatrix from R */
-
-double qij(int i, int j, vector intens, ivector qvector, int nstates)
-{
-    double q;
-    Matrix qmat = Calloc( (nstates)*(nstates), double);
-    FillQmatrix(qvector, intens, qmat, nstates);
-    q = qmat[MI(i,j,nstates)];
-    Free(qmat);
-    return q;
-}
 
 /* Calculates the whole transition matrix in time t given an intensity matrix */
 
-void Pmat(Matrix pmat, double t, vector intens, int nintens, ivector qvector, int nstates, int exacttimes,
[…]

### 2013-06-21 pkg/tests/misc.R

@@ -29,14 +29,17 @@ misccov.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                 qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars = TRUE,
                 control = list(trace=1, REPORT=1), method="BFGS",
                 covariates = ~ sex, covinits=list(sex=rep(0.1, 5)))
-stopifnot(isTRUE(all.equal(4299.35653620144, misccov.msm$minus2loglik, tol=1e-06)))
+#stopifnot(isTRUE(all.equal(4299.35653620144, misccov.msm$minus2loglik, tol=1e-06))) # with last obs included in cov means, as pre 1.2.3
+stopifnot(isTRUE(all.equal(4299.38058878142, misccov.msm$minus2loglik, tol=1e-06)))
+
 
 ## Covs on misc probs, old way.
 misccov.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                    qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars=TRUE,
                    misccovariates = ~dage + sex, misccovinits = list(dage=c(0.01,0.02,0.03,0.04), sex=c(-0.013,-0.014,-0.015,-0.016)),
                    control = list(trace=1, REPORT=1), method="BFGS")
-stopifnot(isTRUE(all.equal(4304.90609473048, misccov.msm$minus2loglik, tol=1e-06)))
+#stopifnot(isTRUE(all.equal(4304.90609473048, misccov.msm$minus2loglik, tol=1e-06))) # with last obs included in cov means
+stopifnot(isTRUE(all.equal(4306.3077053482, misccov.msm$minus2loglik, tol=1e-06)))
 
 if (developer.local) {
     system.time(misccov.msm <- msm(state ~ years, subject = PTNUM, data = cav,
@@ -44,7 +47,8 @@ if (developer.local) {
                                    misccovariates = ~dage + sex, # fixedpars=c(1:5,11),
                                    control = list(trace=1, REPORT=1), method="BFGS"))
     ##    stopifnot(isTRUE(all.equal(3929.39438312539, misccov.msm$minus2loglik, tol=1e-06))) ## 0.7.1 and earlier
-    stopifnot(isTRUE(all.equal(3929.59504496140, misccov.msm$minus2loglik, tol=1e-06)))
[…]

### 2013-06-21 pkg/tests/simple.R

@@ -72,17 +72,19 @@ cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                  qmatrix = twoway4.q, death = TRUE, fixedpars=TRUE,
                  covariates = ~ sex, covinits = list(sex=rep(0.01, 7)), # , dage=rep(0, 7)),
                  method="BFGS", control=list(trace=5, REPORT=1))
-stopifnot(isTRUE(all.equal(4909.08442586298, cav.msm$minus2loglik, tol=1e-06)))
+#stopifnot(isTRUE(all.equal(4909.08442586298, cav.msm$minus2loglik, tol=1e-06))) # pre-1.2.3
+stopifnot(isTRUE(all.equal(4909.17147259115, cav.msm$minus2loglik, tol=1e-06)))
 
 if (developer.local) {
     system.time(cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                                  qmatrix = twoway4.q, death = TRUE, fixedpars=FALSE,
                                  covariates = ~ sex, method="BFGS", control=list(trace=5, REPORT=1))) # 44.13 on new, 260.14 on old
-    stopifnot(isTRUE(all.equal(3954.77699876128, cav.msm$minus2loglik, tol=1e-06)))
-    qmat <- qmatrix.msm(cav.msm)[c("estimates","SE")]
-    stopifnot(isTRUE(all.equal(c(0.226768225214069, -0.583572966218831, 0.337068668450776, 0.0197360725539862), as.numeric(qmat$estimates[2,]), tol=1e-06)))
-    stopifnot(isTRUE(all.equal(c(0.0341912476474469, 0.177757314782152, 0.0383135782775986, 0.171176619065207), as.numeric(qmat$SE[2,]), tol=1e-02)))   ### SEs slightly different on different machines.
-    stopifnot(isTRUE(all.equal(5.29003443121721, sojourn.msm(cav.msm)[1,3], tol=1e-06)))
+    stopifnot(isTRUE(all.equal(3954.77700438268, cav.msm$minus2loglik, tol=1e-06)))
+    qe <- qmatrix.msm(cav.msm)$estimates
+    qs <- qmatrix.msm(cav.msm)$SE
[…]

### 2013-07-23 pkg/inst/doc/msm-manual.Rnw
Major internals update
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@20 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Range constraints for HMM parameters
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@22 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
   \label{eq:multi:lik:exact}
-  L_{i, j} =  \exp(-q_{S(t_j)S(t_{j})}(t_{j+1} - t_j)) q_{S(t_j)S(t_{j+1})}
+  L_{i, j} =  \exp(q_{S(t_j)S(t_{j})}(t_{j+1} - t_j)) q_{S(t_j)S(t_{j+1})}
 \end{equation}
@@
 
-For example consider the three-state hidden Markov model described
+For example, consider the three-state hidden Markov model described
 above, with normally-distributed outcomes for states 1 and 2.
@@
 \end{Scode}
+
+Parameters of the 
+outcome distributions may also be constrained within specific ranges. 
+If chosen carefully, this may improve identifiability of hidden Markov states. 
+For example to constrain the mean for state 1 to be between 80 and 110,
+and the mean for state 2 to be between 50 and 80, specify
+\begin{Scode}
+hranges = list(mean=list(lower=c(80,50), upper=c(110,80)))
+\end{Scode}
+Maximum likelihood estimation is then performed on the appropriate log
+or logit-transformed scale so that these constraints are
+satisfied. See the \Rfunction{msm} help page for further details.
+Note that initial values should be strictly within the ranges, and not
+on the range boundary.
 

### 2013-07-23 pkg/R/constants.R

@@ -4,7 +4,7 @@
 ### and names of parameters for each distribution
 ### MUST BE KEPT IN THE SAME ORDER as the C variable HMODELS in src/lik.c
 .msm.HMODELPARS <- list(
-                        categorical=c("ncats","basecat","prob"),  # vector
+                        categorical=c("ncats","basecat","p"),
                         identity = NULL,
                         uniform = c("lower", "upper"),
                         normal = c("mean", "sd"),
@@ -48,19 +48,26 @@
 ### Defined ranges for parameters
 .msm.PARRANGES <- list(qbase=c(0, Inf), lower=c(-Inf,Inf), upper=c(-Inf, Inf),
                        mean=c(-Inf, Inf), sd=c(0, Inf),
-                       meanlog=c(-Inf,Inf), sdlog=c(0, Inf), rate=c(0, Inf), shape=c(0, Inf), shape1=c(0,Inf), shape2=c(0,Inf),
+                       meanlog=c(-Inf,Inf), sdlog=c(0, Inf), rate=c(0, Inf), shape=c(0, Inf), scale=c(0, Inf), shape1=c(0,Inf), shape2=c(0,Inf),
                        prob=c(0, 1), meanerr=c(-Inf, Inf), sderr=c(0, Inf), disp=c(0, Inf),
-                       df=c(0, Inf)
+                       p=c(-Inf,Inf), # handled separately using multinomial logit
+                       initp=c(-Inf,Inf), # handled separately using multinomial logit
+                       df=c(0, Inf),
+                       qcov=c(-Inf,Inf),hcov=c(-Inf,Inf),initpcov=c(-Inf,Inf)
                        )
+for (i in .msm.AUXPARS) .msm.PARRANGES[[i]] <- c(-Inf, Inf)
+.msm.PARRANGES <- do.call("rbind",.msm.PARRANGES)
+colnames(.msm.PARRANGES) <- c("lower","upper")
 
 ### Transforms to optimise some parameters on a different scale
 ### Univariate transforms only: doesn't include multinomial logit transform used for misclassification p and initial state probs
+
 .msm.TRANSFORMS <-
-  do.call("rbind",
-          lapply(.msm.PARRANGES,
+    do.call("rbind",
+          apply(.msm.PARRANGES, 1,
                  function(x) {
-                     if (identical(x, c(0, Inf))) c(fn="log",inv="exp")
-                     else if (identical(x, c(0, 1))) c(fn="qlogis",inv="plogis")
[…]

### 2013-07-23 pkg/R/hmm-dists.R

@@ -45,7 +45,6 @@ hmmDIST <- function(label, link, r, call, ...)
     hdist
 }
 
-## Check for initial values outside defined ranges
 hmmCheckInits <- function(pars)
   {
       for (i in names(pars)) {
@@ -55,18 +54,9 @@ hmmCheckInits <- function(pars)
               if (!identical(all.equal(pars[i], round(pars[i])), TRUE))
                 stop("Value of ", i, " should be integer")
           }
-          else if (i %in% names(.msm.PARRANGES))
-            if (!in.range(pars[i], .msm.PARRANGES[[i]]))
-              stop("Initial value ", pars[i], " of parameter ", i, " outside allowed range ",
-                   "[", paste(.msm.PARRANGES[[i]], collapse=","), "]")
+          ## Range check now done in msm.form.hranges
       }
   }
-
-in.range <- function(x, interval) {
-    if (!is.numeric(interval) || length(interval)!=2) stop("interval should be a numeric vector of length 2")
-    if (!is.numeric(x)) stop("x should be numeric")
-    ( (x >= interval[1]) & (x <= interval[2]) )
-}
 
 hmmIdent <- function(x)
 {
@@ -186,7 +176,7 @@ hmmMEUnif <- function(lower, upper, sderr, meanerr=0)
                match.call(),
                meanerr=meanerr)
   }
- 
+
 hmmT <- function(mean, scale, df)
   {
       hmmDIST(label="t",

### 2013-07-23 pkg/R/hmm.R

@@ -324,6 +324,56 @@ msm.form.hcovconstraint <- function(constraint, hmodel)
       match(constr, unique(constr))
   }
 
+## User-supplied range constraints on HMM parameters.
+
+msm.form.hranges <- function(ranges, hmodel)
+{
+    hranges <- .msm.PARRANGES[hmodel$plabs,]
+    if (hmodel$ncoveffs>0) { 
+        hcovranges <- matrix(rep(c(-Inf, Inf), hmodel$ncoveffs), nrow=hmodel$ncoveffs, byrow=TRUE)
+        rownames(hcovranges) <- hmodel$covlabels
+        hranges <- rbind(hranges, hcovranges)
+    }
+    if (!is.null(ranges)) {
+        if (!is.list(ranges)) stop("expected \"hranges\" to be a list")
+        for (i in names(ranges)) {
+            if ( ! (i %in% c(hmodel$plabs, hmodel$covlabels)))
+                stop("parameter \"", i, "\" in \"hranges\" unknown")
+            ran.default <- hranges[rownames(hranges)==i,]
+            ran.user <- do.call("cbind", ranges[[i]])
+            for (j in seq_len(nrow(ran.user))) { 
+                if (ran.user[j,1] < ran.default[j,1]) { 
+                    warning("User-supplied lower bound of ",ran.user[j,1]," for ", i,
+                            " less than theoretical minimum of ", ran.default[j,1], ", ignoring")
+                    ran.user[j,1] <- ran.default[j,1]
+                }
+                if (ran.user[j,2] > ran.default[j,2]) { 
+                    warning("User-supplied upper bound of ",ran.user[j,2]," for ", i,
+                            " less than theoretical maximum of ", ran.default[j,2], ", ignoring")
+                    ran.user[j,2] <- ran.default[j,2]
+                }
+                hranges[rownames(hranges)==i,][j,] <- ran.user[j,]
+            }
+        }
+    }
+    ## ideally should be strict here if estimated, but allow inits on Inf boundary if not estimated.
+    for (i in seq(along=hmodel$pars)){ 
+        if (!in.range(hmodel$pars[i], hranges[i,], strict=FALSE))
+            stop("Initial value ", hmodel$pars[i], " of parameter \"", hmodel$plabs[i], "\" outside allowed range ",
+                 "[", paste(hranges[i,], collapse=","), "]")
+    }
+    hranges
+}
+
[…]

### 2013-07-23 pkg/R/msm.R

@@ -20,6 +20,7 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
                 hcovariates = NULL, # list of formulae specifying covariates model for each hidden state
                 hcovinits = NULL,      # initial values for covariate effects on hidden emission distribution
                 hconstraint = NULL, # constraints on hidden Markov model parameters
+                hranges = NULL, # range constraints for HMM parameters
                 qconstraint = NULL, # constraints on equality of baseline intensities
                 econstraint = NULL, # constraints on equality of baseline misc probs
                 initprobs = NULL,  # initial state occupancy probabilities
@@ -169,6 +170,7 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
         hmodel$constr <- msm.form.hconstraint(hconstraint, hmodel)
         hmodel$covconstr <- msm.form.hcovconstraint(hconstraint, hmodel)
     }
+    if (hmodel$hidden) hmodel$ranges <- msm.form.hranges(hranges, hmodel)
 ### INITIAL STATE OCCUPANCY PROBABILITIES IN HMMS
     if (hmodel$hidden) hmodel <- msm.form.initprobs(hmodel, msmdata)
 
@@ -289,8 +291,10 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
             p$ci <- cbind(p$params - qnorm(1 - 0.5*(1-cl))*sqrt(diag(p$covmat)),
                           p$params + qnorm(1 - 0.5*(1-cl))*sqrt(diag(p$covmat)))
             p$ci[p$fixedpars,] <- NA
-            for (lab in rownames(.msm.TRANSFORMS))
-                p$ci[p$plabs==lab] <- get(.msm.TRANSFORMS[lab,"inv"])(p$ci[p$plabs==lab, ])
+#            for (lab in rownames(.msm.TRANSFORMS))
+#                p$ci[p$plabs==lab] <- get(.msm.TRANSFORMS[lab,"inv"])(p$ci[p$plabs==lab, ])
+            for (i in 1:2)
+                p$ci[,i] <- gexpit(p$ci[,i], p$ranges[,"lower"], p$ranges[,"upper"])
         }
         else {
             p$foundse <- FALSE
[…]

### 2013-07-23 pkg/R/utils.R

@@ -369,3 +369,37 @@ qgeneric <- function(pdist, p, ...)
     if (any(is.nan(ret))) warning("NaNs produced")
     ret
 }
+
+
+## Transform vector of parameters constrained on [a, b] to real line.
+## Vectorised.  a=-Inf or b=Inf represent unbounded below or above.
+
+glogit <- function(x, a, b) {
+    ret <- numeric(length(x))
+    nn <- is.infinite(a) & is.infinite(b)
+    nb <- is.infinite(a) & is.finite(b)
+    an <- is.finite(a) & is.infinite(b)
+    ab <- is.finite(a) & is.finite(b)
+    ret[nn] <- x[nn]
+    ret[nb] <- log(b[nb] - x[nb])
+    ret[an] <- log(x[an] - a[an])
+    ret[ab] <- log((x[ab] - a[ab]) / (b[ab] - x[ab]))
+    ret
+}
+
+## Inverse transform vector of parameters constrained on [a, b]: back
+## from real line to constrained scale.  Vectorised.  a=-Inf or b=Inf
+## represent unbounded below or above.
+
+gexpit <- function(x, a, b) {
+    ret <- numeric(length(x))
+    nn <- is.infinite(a) & is.infinite(b)
+    nb <- is.infinite(a) & is.finite(b)
+    an <- is.finite(a) & is.infinite(b)
+    ab <- is.finite(a) & is.finite(b)
+    ret[nn] <- x[nn]
+    ret[nb] <- b[nb] - exp(x[nb])
+    ret[an] <- exp(x[an]) + a[an]
+    ret[ab] <- (b[ab]*exp(x[ab]) + a[ab]) / (1 + exp(x[ab]))
+    ret
+}

### 2013-08-04 pkg/inst/doc/msm-manual.Rnw
Changes for 1.2.5 including expm method, new totlos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@23 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 The \Rpackage{msm} package calculates $P(t)$ analytically for selected
-2, 3, 4 and 5-state models. These models are illustrated in
-Figures \ref{fig:anp2}--\ref{fig:anp5}.  For other models, $P(t)$ is
-determined from the matrix exponential, calculated using eigensystem
+2, 3, 4 and 5-state models, illustrated in
+Figures \ref{fig:anp2}--\ref{fig:anp5}.  For other models, which can have any transition 
+structure on any number of states in principle, $P(t)$ is
+determined from the matrix exponential.  This is calculated using eigensystem
 decomposition (if eigenvalues are distinct) or a method based on

### 2013-08-04 pkg/R/msm.R

@@ -38,6 +38,7 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
                 opt.method = c("optim","nlm","fisher"),
                 hessian = NULL,
                 use.deriv = TRUE,
+                use.expm = TRUE,
                 analyticp=TRUE,
                 ... # options to optim or nlm
                 )
@@ -51,7 +52,7 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
 ### MODEL FOR TRANSITION INTENSITIES
     if (gen.inits)
         qmatrix <- crudeinits.msm(formula, subject, qmatrix, data, censor, censor.states)
-    qmodel <- msm.form.qmodel(qmatrix, qconstraint, exacttimes, analyticp)
+    qmodel <- msm.form.qmodel(qmatrix, qconstraint, exacttimes, analyticp, use.expm)
 
 ### MISCLASSIFICATION MODEL
     if (!missing(ematrix)) {
@@ -415,7 +416,7 @@ msm.fixdiag.ematrix <- function(ematrix)
     ematrix
 }
 
-msm.form.qmodel <- function(qmatrix, qconstraint=NULL, exacttimes=FALSE, analyticp=TRUE)
+msm.form.qmodel <- function(qmatrix, qconstraint=NULL, exacttimes=FALSE, analyticp=TRUE, use.expm=FALSE)
 {
     msm.check.qmatrix(qmatrix)
     nstates <- dim(qmatrix)[1]
@@ -454,7 +455,7 @@ msm.form.qmodel <- function(qmatrix, qconstraint=NULL, exacttimes=FALSE, analyti
     }
     qmodel <- list(nstates=nstates, iso=iso, perm=perm, qperm=qperm,
                    npars=npars, imatrix=imatrix, qmatrix=qmatrix, inits=inits,
-                   constr=constr, ndpars=ndpars, exacttimes=exacttimes)
+                   constr=constr, ndpars=ndpars, exacttimes=exacttimes, expm=as.numeric(use.expm))
     class(qmodel) <- "msmqmodel"
     qmodel
 }
@@ -497,7 +498,7 @@ msm.form.emodel <- function(ematrix, econstraint=NULL, initprobs=NULL, est.initp
             est.initprobs <- FALSE
         }
         else {
-            if (length(initprobs) != qmodel$nstates) stop("initprobs of length ", length(initprobs), ", should be ", qmodel$nstates)
[…]

### 2013-08-04 pkg/src/lik.c

@@ -168,7 +168,7 @@ void update_likhidden(double *curr, int nc, int obsno, msmdata *d, qmodel *qm,
 #endif
     /* calculate the transition probability (P) matrix for the time interval dt */
     Pmat(pmat, d->time[obsno] - d->time[obsno-1], qmat, qm->nst,
-	 (d->obstype[obsno] == OBS_EXACT), qm->iso, qm->perm, qm->qperm);
+	 (d->obstype[obsno] == OBS_EXACT), qm->iso, qm->perm, qm->qperm, qm->expm);
     for(j = 0; j < qm->nst; ++j)
 	{
 #ifdef DEBUG2
@@ -285,7 +285,7 @@ void update_likcensor(int obsno, double *prev, double *curr, int np, int nc,
     double contrib;
     int i, j, k;
     Pmat(pmat, d->time[obsno] - d->time[obsno-1], qmat, qm->nst,
-	 (d->obstype[obsno] == OBS_EXACT), qm->iso, qm->perm,  qm->qperm);
+	 (d->obstype[obsno] == OBS_EXACT), qm->iso, qm->perm,  qm->qperm, qm->expm);
     for(i = 0; i < nc; ++i)
 	{
 	    newp[i] = 0.0;
@@ -356,7 +356,7 @@ double liksimple(msmdata *d, qmodel *qm, cmodel *cm, hmodel *hm)
 		   P matrix for this */
 		/* pointer to Q matrix for ith datapoint */
 		qmat = &(qm->intens[MI3(0, 0, i, qm->nst, qm->nst)]);
-		Pmat(pmat, d->timelag[i], qmat, qm->nst, (d->obstype[i] == OBS_EXACT), qm->iso, qm->perm,  qm->qperm);
+		Pmat(pmat, d->timelag[i], qmat, qm->nst, (d->obstype[i] == OBS_EXACT), qm->iso, qm->perm,  qm->qperm, qm->expm);
 	    }
 	    if (d->obstype[i] == OBS_DEATH)
 		contrib = pijdeath(d->fromstate[i], d->tostate[i], pmat, qmat, qm->nst);
@@ -386,7 +386,7 @@ double liksimple_subj(int pt, /* ordinal subject ID */
 	from = fprec(d->obs[i-1] - 1, 0); /* convert state outcome to integer */
 	to = fprec(d->obs[i] - 1, 0);
 	qmat = &(qm->intens[MI3(0, 0, i, qm->nst, qm->nst)]);
[…]

### 2013-08-04 pkg/src/pijt.c

@@ -25,8 +25,17 @@
 #define MEXP_PADE 1
 #define MEXP_SERIES 2
 #include "R_ext/Lapack.h"
+#include "R_ext/Rdynload.h"
 
 #define NODERIVDEBUG
+
+/* Interface to expm package. */
+typedef enum {Ward_2, Ward_1, Ward_buggy_octave} precond_type;
+void (*expm)(double *x, int n, double *z, precond_type precond_kind);
+void R_init_msm(DllInfo *dll)
+{
+    expm = (void (*) (double, int, double, precond_type)) R_GetCCallable("expm", "expm");
+}
 
 /* Set A to be an n x n identity matrix */
 
@@ -323,7 +332,7 @@ void Eigen(Matrix mat, int n, vector revals, vector ievals, Matrix evecs, int *e
    invertibility by calculating determinant, thus reducing number of
    cases where it is necessary to use power series or Pade? */
 
-void MatrixExp(Matrix mat, int n, Matrix expmat, double t,
+void MatrixExpMSM(Matrix mat, int n, Matrix expmat, double t,
 	       int degen, /* currently not used */
 	       int method)
 {
@@ -343,10 +352,10 @@ void MatrixExp(Matrix mat, int n, Matrix expmat, double t,
 	break;
       }
     if (repeated_entries (revals, n) || (err != 0) || degen || complex_evals){
-	if (method == MEXP_PADE)
-	    MatrixExpPade(expmat, mat, n, t);
-	else if (method == MEXP_SERIES)
+	if (method == MEXP_SERIES)
 	    MatrixExpSeries(mat, n, expmat, t);
+	else 
+	    MatrixExpPade(expmat, mat, n, t);
     }
     else {
 	for (i=0; i<n; ++i)
@@ -368,15 +377,31 @@ void MatrixExpR(double *mat, int *n, double *expmat, double *t,
     if (*iso > 0)
 	AnalyticP(expmat, *t, *n, *iso, perm, qperm, mat, degen);
     else
-	MatrixExp(mat, *n, expmat, *t, *degen, *method);
+	MatrixExpMSM(mat, *n, expmat, *t, *degen, *method);
 }
 
+void MatrixExpEXPM(double *mat, int *n, double *expmat, double *t,
+		int *method, int *iso, int *perm, int *qperm,
+		int *degen){
+    int i;
+    int nsq = (*n)*(*n);
[…]

### 2013-09-13 pkg/inst/doc/msm-manual.Rnw
Version 1.2.6
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@25 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 package provides a function \Rfunction{crudeinits.msm} for calculating
-initial values in this way.
+initial values in this way (for non-hidden Markov models).
 
@@
 times of transition, ``exact death times'', or a mixture of these. See the
-\Rfunarg{obstype} argument to \Rfunction{msm}.
+\Rfunarg{obstype} argument to \Rfunction{msm}.  Note you do not always have to specify \Rfunarg{death}, 
+even if there are states which correspond to death.  This is
+only used for states whose entry time is observed in this specific way.
 

### 2013-09-13 pkg/R/msm.R

@@ -50,25 +50,28 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
     if (missing(data)) data <- environment(formula)
 
 ### MODEL FOR TRANSITION INTENSITIES
-    if (gen.inits)
-        qmatrix <- crudeinits.msm(formula, subject, qmatrix, data, censor, censor.states)
+    if (gen.inits) {
+        if (is.null(hmodel) && is.null(ematrix))
+            qmatrix <- crudeinits.msm(formula, subject, qmatrix, data, censor, censor.states)
+        else warning("gen.inits not supported for hidden Markov models, ignoring")
+    }
     qmodel <- msm.form.qmodel(qmatrix, qconstraint, exacttimes, analyticp, use.expm)
 
 ### MISCLASSIFICATION MODEL
-    if (!missing(ematrix)) {
+    if (!is.null(ematrix)) {
         emodel <- msm.form.emodel(ematrix, econstraint, initprobs, est.initprobs, qmodel)
     }
     else emodel <- list(misc=FALSE, npars=0, ndpars=0)
     if (emodel$npars==0) emodel <- list(misc=FALSE, npars=0, ndpars=0) # user supplied degenerate ematrix with no misclassification
 
 ### GENERAL HIDDEN MARKOV MODEL
-    if (!missing(hmodel)) {
+    if (!is.null(hmodel)) {
         msm.check.hmodel(hmodel, qmodel$nstates)
-        if (!missing(hcovariates)) msm.check.hcovariates(hcovariates, qmodel)
+        if (!is.null(hcovariates)) msm.check.hcovariates(hcovariates, qmodel)
         hmodel <- msm.form.hmodel(hmodel, hconstraint, initprobs, est.initprobs, qmodel)
     }
     else {
-        if (!missing(hcovariates)) stop("hcovariates have been specified, but no hmodel")
+        if (!is.null(hcovariates)) stop("hcovariates have been specified, but no hmodel")
         hmodel <- list(hidden=FALSE, models=rep(0, qmodel$nstates), nipars=0, nicoveffs=0, totpars=0, ncoveffs=0) # might change later if misc
     }
 
@@ -611,7 +614,7 @@ msm.form.data <- function(formula, subject=NULL, obstype=NULL, obstrue=NULL, cov
     subject <- subject[final.rows]
     time <- subset(time, statetimerows.kept %in% final.rows)
     state <- subset(state, statetimerows.kept %in% final.rows)
[…]

### 2013-09-13 pkg/R/outputs.R

@@ -48,6 +48,75 @@ print.msm <- function(x, ...)
 
 }
 
+## Experimental: more helpful and tidier print output
+## Arrange covariates in columns aligned with baselines as option?
+
+printnew.msm <- function(x, ...)
+{
+    cat("\nCall:\n", deparse(x$call), "\n\n", sep = "")
+    if (!attr(x,"fixed")) {
+        cat ("Maximum likelihood estimates: \n\n")
+        covmessage <-
+            if (x$qcmodel$ncovs == 0) ""
+            else paste("with covariates set to", (if (x$center) "their means" else "0"))
+        cat(paste("Transition intensities ",covmessage,":\n\n",sep=""))
+        y <- mattotrans(x, x$Qmatrices$baseline, x$QmatricesL$baseline, x$QmatricesU$baseline, keep.diag=TRUE)
+        print(y)
+        fres <- matrix("", nrow=nrow(y), ncol=x$qcmodel$ncovs+1)
+        colnames(fres) <- c("baseline", x$qcmodel$covlabels)
+        rownames(fres) <- rownames(y)
+        fres[,1] <- format.ci(y[,1],y[,2],y[,3]) # is this worth it?
+        ## HRs(ORs) not log HRs for covariates and misccovariates.
+        cat("\nHazard ratios for covariates on transition intensities:\n\n")
+        hrs <- hazard.msm(x)
+        print(hrs)
+        im <- t(x$qmodel$imatrix); diag(im) <- -1; nd <- which(im[im!=0]==1)
+        for (i in seq(length=x$qcmodel$ncovs)){
+            nm <- x$qcmodel$covlabels[[i]]
+            hrs <- exp(mattotrans(x, x$Qmatrices[[nm]], x$QmatricesL[[nm]], x$QmatricesU[[nm]], keep.diag=FALSE))
+            fres[nd,1+i] <- format.ci(hrs[,1], hrs[,2], hrs[,3])
+        }
+        cat ("Maximum likelihood estimates: \n\n")
+        cat ("Transition intensities with hazard ratios for covariates\n")
+        cat(paste("Baseline intensities are ",covmessage,"\n\n",sep=""))
+        print(fres, quote=FALSE)
+        ## initprobs also presented as ORs, note multinomial scale
+        ## Is it also worth presenting transition rates as columns?
+        ## Keep old print method
+        if (x$emodel$misc) {
[…]

### 2013-12-03 pkg/src/doc/msm-manual.Rnw
Changes for 1.3, intended for CRAN.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@28 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2013-09-13 pkg/inst/doc/msm-manual.Rnw
@@
 <<>>=
-hmodel1 <- list(hmmNorm(mean=100, sd=16), hmmNorm(mean=54, sd=18), hmmIdent(999))
-
-fev1.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q, death=3, hmodel=hmodel1,
-                hcovariates=list(~acute, ~acute, NULL), hcovinits = list(-8, -8, NULL),
+hmodel1 <- list(hmmNorm(mean=100, sd=16), hmmNorm(mean=54, sd=18),
+                hmmIdent(999))
+
+fev1.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q,
+                death=3, hmodel=hmodel1,
+                hcovariates=list(~acute, ~acute, NULL),
+                hcovinits = list(-8, -8, NULL),
                 hconstraint = list(acute = c(1,1)), method="BFGS")
@@
 
-Our initial values are again 100 and 54 for the means of the
+Our initial values are 90 and 54 for the means of the
 within-state distribution of \emph{underlying} FEV$_1$, and 16 and 18
@@
 
-<<>>=
-hmodel2 <- list(hmmMETNorm(mean=100, sd=16, sderr=8, lower=80, upper=Inf, meanerr=0),
-                hmmMETNorm(mean=54, sd=18, sderr=8, lower=0, upper=80, meanerr=0),
+The convergence of maximum likelihood estimation in this example is
+particularly sensitive to the optimisation method and options, initial values,
+the unit of the time variable and whether covariates are centered,
+probably because the likelihood surface is irregular near to the
+true maximum.
+
+<<>>=
+hmodel2 <- list(hmmMETNorm(mean=90, sd=16, sderr=8,
+                           lower=80, upper=Inf, meanerr=0),
+                hmmMETNorm(mean=54, sd=18, sderr=8,
+                           lower=0, upper=80, meanerr=0),
                 hmmIdent(999))
 
-fev2.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q, death=3, hmodel=hmodel2,
-                hcovariates=list(~acute, ~acute, NULL), hcovinits = list(-8, -8, NULL),
-                hconstraint = list(sderr = c(1,1), acute = c(1,1)), method="BFGS")
+fev2.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q,
+                death=3, hmodel=hmodel2,
+                hcovariates=list(~acute, ~acute, NULL),
+                hcovinits = list(-8, -8, NULL),
+                hconstraint = list(sderr = c(1,1), acute = c(1,1)),
+                control=list(maxit=10000), center=TRUE)
 
@@
 
+Development versions of \Rpackage{msm} are to be found on R-Forge,
+\url{http://msm.r-forge.r-project.org/}.  These are often more recent
+than the version released on CRAN, so if you think you have found a
+bug, then please check first to see whether it has been fixed on the
+R-Forge version.
+
 \section{Get in touch}

### 2013-12-03 pkg/R/boot.R

@@ -144,8 +144,8 @@ totlos.ci.msm <- function(x, start=1, end=NULL, fromt=0, tot=Inf, covariates="me
     apply(t.array, 2, function(x)(quantile(x, c(0.5 - cl/2, 0.5 + cl/2))))
 }
 
-efpt.ci.msm <- function(x, qmatrix=NULL, tostate, covariates="mean", cl=0.95, B=1000) {
-    t.list <- boot.msm(x, function(x)efpt.msm(x, qmatrix, tostate, covariates, ci="none"), B)
+efpt.ci.msm <- function(x, qmatrix=NULL, tostate, start, covariates="mean", cl=0.95, B=1000) {
+    t.list <- boot.msm(x, function(x)efpt.msm(x, qmatrix, start, tostate, covariates, ci="none"), B)
     t.array <- do.call("rbind", t.list)
     apply(t.array, 2, function(x)(quantile(x, c(0.5 - cl/2, 0.5 + cl/2))))
 }
@@ -251,8 +251,8 @@ totlos.normci.msm <- function(x, start=1, end=NULL, fromt=0, tot=Inf, covariates
     apply(t.array, 2, function(x)(quantile(x, c(0.5 - cl/2, 0.5 + cl/2))))
 }
 
-efpt.normci.msm <- function(x, qmatrix=NULL, tostate, covariates="mean", cl=0.95, B=1000, ...) {
-    t.list <- normboot.msm(x, function(x)efpt.msm(x, qmatrix, tostate, covariates, ci="none", ...), B)
+efpt.normci.msm <- function(x, qmatrix=NULL, tostate, start, covariates="mean", cl=0.95, B=1000, ...) {
+    t.list <- normboot.msm(x, function(x)efpt.msm(x, qmatrix, tostate, start, covariates, ci="none", ...), B)
     t.array <- do.call("rbind", t.list)
     apply(t.array, 2, function(x)(quantile(x, c(0.5 - cl/2, 0.5 + cl/2))))
 }

### 2013-12-03 pkg/R/msm.R

@@ -32,7 +32,7 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
                 censor = NULL,
                 censor.states = NULL,
                 pci = NULL,
-                cl = 0.95, # width of confidence intervals
+                cl = 0.95, # width of confidence intervals in saved result components
                 fixedpars = NULL, # specify which parameters to fix. TRUE for all parameters
                 center = TRUE, # center covariates at their means during optimisation
                 opt.method = c("optim","nlm","fisher"),
@@ -105,6 +105,8 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
             pci <- tdmodel$tcut
         }
     }
+
+### AGGREGATE DATA
     if (hmodel$hidden || (cmodel$ncens > 0)) {
         msmdata <- msm.aggregate.hmmdata(msmdata.obs)
         msmdata$fromstate <- msmdata$tostate <- msmdata$timelag <- numeric(0)
@@ -298,7 +300,7 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
 #            for (lab in rownames(.msm.TRANSFORMS))
 #                p$ci[p$plabs==lab] <- get(.msm.TRANSFORMS[lab,"inv"])(p$ci[p$plabs==lab, ])
             for (i in 1:2)
-                p$ci[,i] <- gexpit(p$ci[,i], p$ranges[,"lower"], p$ranges[,"upper"])
+                p$ci[,i] <- gexpit(p$ci[,i], p$ranges[,"lower",drop=FALSE], p$ranges[,"upper",drop=FALSE])
         }
         else {
             p$foundse <- FALSE
@@ -368,7 +370,8 @@ msm <- function(formula,   # formula with  observed Markov states   ~  observati
                        hmodel = hmodel,
                        cmodel = cmodel,
                        pci = pci,
-                       paramdata=p
+                       paramdata=p,
+                       cl=cl
                        )
     attr(msmobject, "fixed") <- p$fixed
     class(msmobject) <- "msm"
@@ -445,12 +448,7 @@ msm.form.qmodel <- function(qmatrix, qconstraint=NULL, exacttimes=FALSE, analyti
         ## analytic P matrix is implemented for this particular intensity matrix
[…]

### 2013-12-03 pkg/R/outputs.R

@@ -5,6 +5,7 @@ print.msm <- function(x, ...)
     cat("\nCall:\n", deparse(x$call), "\n\n", sep = "")
 
     if (!attr(x,"fixed")) {
+        cat("[Note: a cleaner summary is available from \"printnew.msm\", which will be the default in future versions.]\n")
         cat ("Maximum likelihood estimates: \n")
         covmessage <-
             if (x$qcmodel$ncovs == 0) ""
@@ -49,82 +50,97 @@ print.msm <- function(x, ...)
 }
 
 ## Experimental: more helpful and tidier print output
-## TODO allow covariates=? option?
 
-printnew.msm <- function(x, rect=TRUE, ...)
+printnew.msm <- function(x, covariates=NULL, digits=4, ...)
 {
     cat("\nCall:\n", deparse(x$call), "\n\n", sep = "")
-    if (!attr(x,"fixed")) {
-        cat ("Maximum likelihood estimates: \n\n")
-        y <- mattotrans(x, x$Qmatrices$baseline, x$QmatricesL$baseline, x$QmatricesU$baseline, keep.diag=TRUE)
-        covmessage <-
-            if (x$qcmodel$ncovs == 0) ""
-            else paste(" with covariates set to", (if (x$center) "their means" else "0"))
-
-        if (rect) {
-            ## ALTERNATIVE RECTANGULAR FORMAT
-            ## COVARIATES IN COLUMNS ALIGNED WITH BASELINES
+    if (!x$foundse) {
+        cat("Optimisation probably not converged to the maximum likelihood.\noptim() reported convergence but estimated Hessian not positive-definite.\n")
+#        cat("See $opt component of fitted model object for current parameter values,\nwith positive parameters log-transformed.\n")
+    }
+    else { 
+        if (is.null(x$cl)) {
+            cl <- 0.95
+            warning("Found msm object saved before version 1.3. Assuming this was run with the default cl=0.95")
+        } else cl <- x$cl
+        if (!attr(x,"fixed")) {
+                                        #        cat("[New print format for \"msm\" objects since version 1.3. See help(print.msm) for more details. To obtain old format, use \"printold.msm\"]\n\n",sep="")
+            cat ("Maximum likelihood estimates: \n\n")
+            if (is.null(covariates)) {
[…]

### 2013-12-03 pkg/src/analyticp.c

@@ -17,6 +17,7 @@ the matrix exponential.
 */
 
 #include "msm.h"
+#include <stdio.h>
 
 void p2q1(Matrix pmat, double t, Matrix qmat, int *degen);
 void p2q12(Matrix pmat, double t, Matrix qmat, int *degen);
@@ -56,15 +57,21 @@ pfn P5FNS[] = {
 void AnalyticP(Matrix pmat, double t, int nstates, int iso, int *perm, int *qperm, Matrix qmat, int *degen)
 {
     int i, j;
+    Matrix qmat_base = (Matrix) Calloc( (nstates)*(nstates), double);
     Matrix pmat_base = (Matrix) Calloc( (nstates)*(nstates), double);
+    for (i=0; i<nstates; ++i) {
+	for (j=0; j<nstates; ++j) {
+	    qmat_base[MI(i,j,nstates)] = qmat[MI(qperm[i]-1,qperm[j]-1,nstates)];
+	}
+    }
     if (nstates==2)
-	(P2FNS[iso-1])(pmat_base, t, qmat, degen);
+	(P2FNS[iso-1])(pmat_base, t, qmat_base, degen);
     else if (nstates==3)
-	(P3FNS[iso-1])(pmat_base, t, qmat, degen);
+	(P3FNS[iso-1])(pmat_base, t, qmat_base, degen);
     else if (nstates==4)
-	(P4FNS[iso-1])(pmat_base, t, qmat, degen);
+	(P4FNS[iso-1])(pmat_base, t, qmat_base, degen);
     else if (nstates==5)
-	(P5FNS[iso-1])(pmat_base, t, qmat, degen);
+	(P5FNS[iso-1])(pmat_base, t, qmat_base, degen);
     else error("internal error in GetAnalyticP. Send a bug report to the package maintainer.");
     if (*degen) return;
     for (i=0; i<nstates; ++i)
@@ -95,10 +102,16 @@ void p2q12(Matrix pmat, double t, Matrix qmat, int *degen)
 {
[…]

### 2013-12-03 pkg/src/pijt.c

@@ -34,7 +34,7 @@ typedef enum {Ward_2, Ward_1, Ward_buggy_octave} precond_type;
 void (*expm)(double *x, int n, double *z, precond_type precond_kind);
 void R_init_msm(DllInfo *dll)
 {
-    expm = (void (*) (double, int, double, precond_type)) R_GetCCallable("expm", "expm");
+    expm = (void (*) (double*, int, double*, precond_type)) R_GetCCallable("expm", "expm");
 }
 
 /* Set A to be an n x n identity matrix */
@@ -232,7 +232,7 @@ MatrixExpPade(double *ExpAt, double *A, int n, double t)
   double l1 = F77_CALL(dlange)("1", &n, &n, At, &n, 0); /* L-1 norm */
   double linf = F77_CALL(dlange)("i", &n, &n, At, &n, Temp); /* L-Infinity norm */
   double K = (log(l1) + log(linf))/log(4);
-  int npower =  (int)(K) + 4;
+  int npower = (R_FINITE(K) ? (int)(K)+4 : NA_INTEGER);
   double scale = 1;
 
   /* Multiply by t */
@@ -354,7 +354,7 @@ void MatrixExpMSM(Matrix mat, int n, Matrix expmat, double t,
     if (repeated_entries (revals, n) || (err != 0) || degen || complex_evals){
 	if (method == MEXP_SERIES)
 	    MatrixExpSeries(mat, n, expmat, t);
-	else 
+	else
 	    MatrixExpPade(expmat, mat, n, t);
     }
     else {
@@ -388,8 +388,8 @@ void MatrixExpEXPM(double *mat, int *n, double *expmat, double *t,
     double *matt = Calloc(nsq, double);
     if (*iso > 0)
 	AnalyticP(expmat, *t, *n, *iso, perm, qperm, mat, degen);
-    else { 
-	for (i=0; i<((*n)*(*n)); ++i) { 
+    else {
+	for (i=0; i<((*n)*(*n)); ++i) {
 	    matt[i] = (*t) * mat[i];
 	}
 	expm(matt, *n, expmat, Ward_2);
@@ -401,7 +401,7 @@ void MatrixExpEXPM(double *mat, int *n, double *expmat, double *t,
 
 /* Calculates the whole transition matrix in time t given an intensity matrix */
 
[…]

### 2013-12-03 pkg/tests/misc.R

@@ -3,11 +3,12 @@
 source("local.R")
 
 library(msm)
-# library(msm, lib.loc="../../lib/sun/0.7.5")
-data(cav)
+# library(msm, lib.loc="~/msm/src/1.2.7")
+# data(cav) # not needed since 1.3
+
 oneway4.q <- rbind(c(0, 0.148, 0, 0.0171), c(0, 0, 0.202, 0.081), c(0, 0, 0, 0.126), c(0, 0, 0, 0))
-rownames(oneway4.q) <- colnames(oneway4.q) <- c("Well","Mild","Severe","Death")
 ematrix <- rbind(c(0, 0.1, 0, 0),c(0.1, 0, 0.1, 0),c(0, 0.1, 0, 0),c(0, 0, 0, 0))
+rownames(oneway4.q) <- colnames(oneway4.q) <- rownames(ematrix) <- colnames(ematrix) <- c("Well","Mild","Severe","Death")
 
 ## Plain misc model with no covs
 
@@ -49,7 +50,7 @@ if (developer.local) {
     ##    stopifnot(isTRUE(all.equal(3929.39438312539, misccov.msm$minus2loglik, tol=1e-06))) ## 0.7.1 and earlier
 ##    stopifnot(isTRUE(all.equal(3929.59504496140, misccov.msm$minus2loglik, tol=1e-06))) # 1.2.2 and earlier
     stopifnot(isTRUE(all.equal(3929.60136975058, misccov.msm$minus2loglik, tol=1e-06)))
-    if(interactive()) save(misccov.msm, file="~/msm/devel/models/misccov.msm.rda")
+    if(interactive()) save(misccov.msm, file="~/work/msm/devel/models/misccov.msm.rda")
 
     system.time(misccovboth.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                                        qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars=FALSE,
@@ -60,15 +61,15 @@ if (developer.local) {
     ##    stopifnot(isTRUE(all.equal(3921.40046811911, misccovboth.msm$minus2loglik, tol=1e-06))) # 0.7.1 and earlier
 ##     stopifnot(isTRUE(all.equal(3921.42240883417, misccovboth.msm$minus2loglik, tol=1e-06))) 1.2.2 and earlier
     stopifnot(isTRUE(all.equal(3921.42117997675, misccovboth.msm$minus2loglik, tol=1e-06)))
[…]

### 2013-12-03 pkg/tests/simple.R

@@ -178,7 +178,7 @@ if (0) {
 
 ## Multiple death states (Jean-Luc's data)
 if (developer.local) {
-    c2.df <- read.table("~/msm/tests/jeanluc/donneesaveccancerPT.txt", header=TRUE)
+    c2.df <- read.table("~/work/msm/tests/jeanluc/donneesaveccancerPT.txt", header=TRUE)
     qx <- rbind( c(0, 0.005, 0, 0, 0), c(0, 0, 0.01, 0.02,0), c(0, 0, 0, 0.04, 0.03), c(0, 0, 0, 0, 0), c(0, 0, 0, 0, 0))
     c2.msm <- msm(state~years, subject=PTNUM, data=c2.df,
                   qmatrix=qx, death=c(4, 5), method="BFGS", fixedpars = 1:5,
@@ -202,7 +202,7 @@ if (developer.local) {
     stopifnot(isTRUE(all.equal(62915.1638036017, c2.msm$minus2loglik, tol=1e-06)))
 
 ### G Marshall's diabetic retinopathy data
-    marsh.df <- read.table("~/msm/tests/markov/test.dat", col.names=c("subject","eyes","time","duration","hba1"))
+    marsh.df <- read.table("~/work/msm/tests/markov/test.dat", col.names=c("subject","eyes","time","duration","hba1"))
     marsh.df$hba1 <- marsh.df$hba1 - mean(marsh.df$hba1)
     marsh.msm <-
       msm(eyes ~ time, subject=subject, qmatrix = rbind(c(0,0.02039,0,0), c(0.007874,0,0.01012,0), c(0,0.01393,0,0.01045), c(0,0,0,0)),
@@ -211,8 +211,8 @@ if (developer.local) {
     system.time(marsh.msm <-
       msm(eyes ~ time, subject=subject, qmatrix = rbind(c(0,0.02039,0,0), c(0.007874,0,0.01012,0), c(0,0.01393,0,0.01045), c(0,0,0,0)),
           covariates = ~ hba1, data = marsh.df))
-    stopifnot(isTRUE(all.equal(310.989863258621, marsh.msm$minus2loglik, tol=1e-06)))
-    stopifnot(isTRUE(all.equal(-0.0496235211442196, qmatrix.msm(marsh.msm, covariates=0)$estimates[1,1], tol=1e-06)))
[…]

### 2013-12-05 pkg/inst/doc/msm-manual.Rnw
Doc update about lazy loading data
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@30 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 An example data set, taken from monitoring a set of heart transplant
-recipients, is provided with \Rpackage{msm}.  Sharples
+recipients, is provided with \Rpackage{msm}.  (Note: since \Rpackage{msm} 
+version 1.3, the command \Rfunction{data(cav)} is no longer needed to 
+load the data --- it is now ``lazy-loaded'' when required). Sharples
 \etal \cite{my:cav} studied the progression of coronary allograft
@@
 Markov models.
-
-This data set can be made available to the current R session by
-issuing the command
-
-<<>>=
-data(cav)
-@
 
@@
 the \Rpackage{msm} package in a dataset called \Robject{fev}.
-
-<<>>=
-data(fev)
-@
-
 We fit models Models 1 and 2, each with three states and common $Q$

### 2013-12-05 pkg/inst/doc/msm-manual.Rnw
Don't evaluate odds.msm call in manual, see if this fixes build
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@31 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Don't evaluate fev2.msm, see if this fixes the build
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@32 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \item[Odds ratios for misclassification] The function
-  \Rfunction{odds.msm} gives the estimated odds ratios corresponding to
+  \Rfunction{odds.msm} would give the estimated odds ratios corresponding to
   each covariate effect on the misclassification probabilities.
 
-<<>>=
+\begin{Scode}
 odds.msm(cavmiscsex.msm)
-@
+\end{Scode}
 
@@
 
-<<>>=
+\begin{Scode}
 hmodel2 <- list(hmmMETNorm(mean=90, sd=16, sderr=8,
@@
 sojourn.msm(fev2.msm)
-@
+\end{Scode}
 
@@
 
-The following plot illustrates a trajectory of declining FEV$_1$ from
-the first lung transplant recipient in this dataset.  This is produced
-by the following R code.  The Viterbi algorithm is used to locate the
+The following code will create a plot that illustrates a trajectory of declining FEV$_1$ from
+the first lung transplant recipient in this dataset.  The Viterbi algorithm is used to locate the
 most likely point at which this individual moved from BOS state 1 to
 BOS state 2, according to the fitted Model 2.  This is illustrated by
-the vertical dotted line.  This is the point at which the individual's
+a vertical dotted line.  This is the point at which the individual's
 lung function started to remain consistently below 80\% baseline
@@
 
-<<fig=TRUE>>=
+\begin{Scode}
 keep <- fev$ptnum==1 & fev$fev<999
@@
 text(min2 + 500, 50, "STATE 2")
-@
+\end{Scode}
 

### 2013-12-30 pkg/inst/doc/msm-manual.Rnw
Also remove version detection code, see if that fixes build
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@33 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Trigger build to test build breakage, and doc typos
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@34 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+% version <- gsub("Version: +", "",
+%                packageDescription("msm", lib.loc="../../..")$Version)
+
 <<echo=FALSE>>=
-version <- gsub("Version: +", "",
-                packageDescription("msm", lib.loc="../../..")$Version)
+version <- 1.3
 @
@@
 }}
-\author{Christopher Jackson\\Medical Research Council Biostatistics Unit\\Cambridge, U.K.\\ \texttt{chris.jackson@mrc-bsu.cam.ac.uk}}
+\author{Christopher Jackson\\MRC Biostatistics Unit\\Cambridge, U.K.\\ \texttt{chris.jackson@mrc-bsu.cam.ac.uk}}
 \date{}

### 2013-12-31 pkg/inst/doc/msm-manual.Rnw
Make trivial vignette to test build
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@35 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \documentclass{article}
-
-%% Need to modify Sweave.sty to pass pdftex option to graphicx.
-\usepackage{Sweave-local}
-\usepackage{times}
-\usepackage{url}
-\addtolength{\textwidth}{2cm}
-
-\newcommand{\Exp}{\mathop{\mathrm{Exp}}}
-\newcommand{\etal}{{\textit{et al.}}}
-\newcommand{\Rfunction}[1]{{\texttt{#1}}}
-\newcommand{\Robject}[1]{{\texttt{#1}}}
-\newcommand{\Rpackage}[1]{{\textit{#1}}}
-\newcommand{\Rmethod}[1]{{\texttt{#1}}}
-\newcommand{\Rfunarg}[1]{{\texttt{#1}}}
-\newcommand{\Rclass}[1]{{\textit{#1}}}
-
-% version <- gsub("Version: +", "",
-%                packageDescription("msm", lib.loc="../../..")$Version)
-
-<<echo=FALSE>>=
-version <- 1.3
-@
-
-\title{Multi-state modelling with R: the {\tt msm} package \vskip 0.2in
-  \large{Version %
-<<echo=FALSE,results=tex>>=
-cat(version)
-@
-\vskip 0.1in
-<<echo=FALSE,results=tex>>=
-cat(format(Sys.time(), "%d %B, %Y"))
-@
-}}
-\author{Christopher Jackson\\MRC Biostatistics Unit\\Cambridge, U.K.\\ \texttt{chris.jackson@mrc-bsu.cam.ac.uk}}
-\date{}
-%% label with date when Rnw is compiled, not when tex is compiled. should be release date of package
-
-\bibliographystyle{unsrt}
 
@@
 
-  The multi-state Markov model is a useful way of describing a process
-  in which an individual moves through a series of states in
-  continuous time.  The \Rpackage{msm} package for R allows a general
-  multi-state model to be fitted to longitudinal data. Data often
-  consist of observations of the process at arbitrary times, so that
-  the exact times when the state changes are unobserved.  For example,
-  the progression of chronic diseases is often described by stages of
-  severity, and the state of the patient may only be known at doctor
-  or hospital visits.  Features of \Rpackage{msm} include the ability
-  to model transition rates and hidden Markov output models in terms
-  of covariates, and the ability to model data with a variety of
-  observation schemes, including censored states.
-
-  Hidden Markov models, in which the true path through states is only
-  observed through some error-prone marker, can also be fitted.  The
-  observation is generated, conditionally on the underlying states,
-  via some distribution.  An example is a screening misclassification
-  model in which states are observed with error.  More generally,
-  hidden Markov models can have a continuous response, with some
-  arbitrary distribution, conditionally on the underlying state.
-
-  This manual introduces the theory behind multi-state Markov and
-  hidden Markov models, and gives a tutorial in the typical use of the
-  \Rpackage{msm} package, illustrated by some typical applications to
-  modelling chronic diseases.
-
-  Much of the material in this manual is published, in a more concise
-  form, in Journal of Statistical Software (2011) 38(8):1-29,
-  \url{http://www.jstatsoft.org/v38/i08/}
+Test
 
@@
 \label{sec:multistate}
-
-\subsection{Introduction}
-\label{sec:intro}
-
-Figure~\ref{fig:multi} illustrates a multi-state model in continuous
-time. Its four states are labelled {\bf 1, 2, 3, 4}.  At a time $t$
-the individual is in state $S(t)$.  The arrows show which transitions
-are possible between states. The next state to which the individual
-moves, and the time of the change, are governed by a set of {\em
-  transition intensities} $q_{rs}(t, z(t))$ for each pair of states
-$r$ and $s$.  The intensities may also depend on the time of the
-process $t$, or more generally a set of individual-specific or
-time-varying explanatory variables $z(t)$.  The intensity represents
-the instantaneous risk of moving from state $r$ to state $s$:
-\begin{equation}
-  \label{eq:multi:intensity}
-  q_{rs}(t, z(t))  =  \lim_{\delta t \rightarrow 0} P (S(t+\delta t) = s | S(t) = r) / \delta t
-\end{equation}
-The intensities form a matrix $Q$ whose rows sum to zero, so that the
-diagonal entries are defined by $q_{rr} = - \sum_{s \neq r} q_{rs}$.
-
-To fit a multi-state model to data, we estimate this transition
-intensity matrix.  We concentrate on {\em Markov} models here. The
-Markov assumption is that future evolution only depends on the current
-state.  That is, $q_{rs}(t, z(t), \mathcal{F}_t)$ is independent of
-$\mathcal{F}_t$, the observation history $\mathcal{F}_t$ of the
-process up to the time preceding $t$.  See, for example, Cox and
-Miller\cite{cox:miller} for a thorough
-introduction to the theory of continuous-time Markov chains.
-
-\begin{figure}[p]
-  \begin{center}
-    \scalebox{0.4}{\includegraphics{figures/multistate}}
-    \[
-    Q = \left(
-      \begin{array}{llll}
-        q_{11} & q_{12} & q_{13} & q_{14}\\
-        q_{21} & q_{22} & q_{23} & q_{24}\\
-        q_{31} & q_{32} & q_{33} & q_{34}\\
-        q_{41} & q_{42} & q_{43} & q_{44}\\
-      \end{array}
-    \right )
-    \]
-    \caption{\label{fig:multi}General multi-state model.}
-  \end{center}
-\end{figure}
-
-In a time-homogeneous continuous-time Markov model, a single period of occupancy (or
-\emph{sojourn time}) in state $r$ has an exponential distribution,
-with rate given by $-q_{rr}$, (or
-mean $-1 / q_{rr}$).  The remaining elements of the $r$th row of $Q$
-are \emph{proportional to} the probabilities governing the next state
-after $r$ to which the individual makes a transition.  The probability
-that the individual's next move from state $r$ is to state $s$ is
-$-q_{rs} / q_{rr}$.
-
-\subsection{Disease progression models}
-
-The development of the \Rpackage{msm} package was motivated by
-applications to disease modelling.  Many chronic diseases have a
-natural interpretation in terms of staged progression. Multi-state
-Markov models in continuous time are often used to model the course of
-diseases. A commonly-used model is illustrated in
-Figure \ref{fig:disease}.  This represents a series of successively
-more severe disease stages, and an `absorbing' state, often death.
-The patient may advance into or recover from adjacent disease stages,
-or die at any disease stage.  Observations of the state $S_i(t)$ are
-made on a number of individuals $i$ at arbitrary times $t$, which may
-vary between individuals.  The stages of disease may be modelled as a
-homogeneous continuous-time Markov process, with a transition matrix
-$Q$, pictured below Figure \ref{fig:disease}.
-
-A commonly-used model is the \emph{illness-death} model, with three
-states representing health, illness and death
-(Figure \ref{fig:multi:illdeath}).  Transitions are permitted from
-health to illness, illness to death and health to death.  Recovery
-from illness to health is sometimes also considered.
-
-A wide range of medical situations have been modelled using
-multi-state methods, for example, screening for abdominal aortic
-aneurysms (Jackson \etal\cite{jackson:sharples:2003}),
-problems following lung transplantation (Jackson and
-Sharples\cite{jackson:sharples:2002}), problems following heart
-transplantation (Sharples\cite{sharp:gibbs}, Klotz and
-Sharples\cite{klotz:est}), hepatic cancer (Kay\cite{kay:mark}), HIV
-infection and AIDS (Longini \etal\cite{long1}, Satten and
-Longini\cite{sattlong}, Guihenneuc-Jouyaux \etal\cite{rich},
-Gentleman \etal\cite{gentlaw}), diabetic complications
-(Marshall and Jones\cite{retino2}, Andersen\cite{andersen88}), breast
-cancer screening (Duffy and Chen\cite{duffy95}, Chen \emph{et
-  al.}\cite{duffy96}), cervical cancer screening (Kirby and
-Spiegelhalter\cite{kirby:spiegelhalter}) and liver cirrhosis (Andersen
-\etal\cite{ander:proth}).  Many of these references also
-describe the mathematical theory, which will be reviewed in the
-following sections.
-
-\begin{figure}[p]
-  \centering
-  \vskip 1cm
-  \scalebox{1.0}{\includegraphics{figures/general}}
-    \[
-    Q = \left(
-      \begin{array}{llllll}
-        q_{11} & q_{12} & 0 & 0 & \ldots & q_{1n}\\
-        q_{21} & q_{22} & q_{23} & 0 & \ldots & q_{2n}\\
-        0 & q_{32} & q_{33} & q_{34} & \ddots & q_{3n}\\
-        0 & 0 & q_{43} & q_{44} & \ddots & q_{4n}\\
-        \vdots & \vdots & \ddots & \ddots & \ddots & \vdots\\
-        0 & 0 & 0 & 0 & \ldots & 0\\
-      \end{array}
-    \right )
-    \]
-  \caption{\label{fig:disease}General model for disease progression.}
-  \end{figure}
-
-\begin{figure}[p]
-  \begin{center}
-    \scalebox{0.4}{\includegraphics{figures/illdeath}}
-    \caption{Illness-death model.}
-    \label{fig:multi:illdeath}
-  \end{center}
-\end{figure}
-
-\subsection{Arbitrary observation times}
-\label{sec:arbitr-observ-times}
-
-Longitudinal data from monitoring disease progression are often
-incomplete in some way.  Usually patients are seen at intermittent
-follow-up visits, at which monitoring information is collected, but
-information from the periods between visits is not available.  Often
-the exact time of disease onset is unknown. Thus, the changes of state
-in a multi-state model usually occur at unknown times.  Also a subject
-may only be followed up for a portion of their disease history.  A
-fixed observation schedule may be specified in advance, but in
-practice times of visits may vary due to patient and hospital
-pressures. The states of disease progression models often include
-death. Death times are commonly recorded to within a day.  Also
-observations may be censored.  For example, at the end of a study, an
-individual may be known only to be alive, and in an unknown state.
-
-A typical sampling situation is illustrated in
-Figure \ref{fig:multi:sampling}. The individual is observed at four
-occasions through 10 months. The final occasion is the death date
-which is recorded to within a day. The only other information
-available is the occupation of states 2, 2, and 1 at respective times
-1.5, 3.5 and 5. The times of movement between states and the state
-occupancy in between the observation times are unknown.  Although the
-patient was in state 3 between 7 and 9 months this was not observed at
-all.
-
-\paragraph{Informative sampling times}
-To fit a model to longitudinal data with arbitrary sampling times we
-must consider the reasons why observations were made at the given
-times.  This is analogous to the problem of missing data, where the
-fact that a particular observation is missing may implicitly give
-information about the value of that observation. Possible observation
-schemes include:
-\begin{itemize}
-\item {\em fixed}. Each patient is observed at fixed intervals
-  specified in advance.
-\item {\em random}. The sampling times vary randomly, independently
-  of the current state of the disease.
-\item {\em doctor's care}. More severely ill patients are monitored
-  more closely. The next sampling time is chosen on the basis of the
-  current disease state.
-\item patient {\em self-selection}. A patient may decide to visit the
-  doctor on occasions when they are in a poor condition.
-\end{itemize}
-Gr\"uger \etal \cite{gruger:valid} discussed conditions under
-which sampling times are \emph{informative}. If a multi-state model is
-fitted, ignoring the information available in the sampling times, then
-inference may be biased.  Mathematically, because the sampling
-times are often themselves random, they should be modelled along with
-the observation process $X_t$. However the ideal situation is where
-the joint likelihood for the times and the process is proportional to
-the likelihood obtained if the sampling times were fixed in advance.
-Then the parameters of the process can be estimated independently of
-the parameters of the sampling scheme.
-
-In particular, they showed that fixed, random and doctor's care
-observation policies are not informative, whereas patient
-self-selection is informative.   Note that \Rpackage{msm} does not
-deal with informative sampling times. See, e.g. \cite{sweeting:inform:msm}
-for some methods in this case, which require specialised programming.
-
-\begin{figure}[p]
-  \begin{center}
-    \scalebox{0.6}{\rotatebox{270}{\includegraphics{figures/sampling}}}
-    \caption{Evolution of a multi-state model. The process is observed
-      on four occasions.}
-    \label{fig:multi:sampling}
-  \end{center}
-\end{figure}
-
-\subsection{Likelihood for the multi-state model}
-\label{sec:multi:likelihood}
-
-Kalbfleisch and Lawless\cite{kalbfleisch:lawless} and later
-Kay \cite{kay:mark} described a general method for evaluating the
-likelihood for a general multi-state model in continuous time,
-applicable to any form of transition matrix. The only available
-information is the observed state at a set of times, as in
-Figure \ref{fig:multi:sampling}.  The sampling times are assumed to be
-non-informative.
-
-\paragraph{Transition probability matrix}
-The likelihood is calculated from the \emph{transition probability matrix}
-$P(t)$.  For a time-homogeneous process, the $(r,s)$ entry of $P(t)$,
-$p_{rs}(t)$, is the probability of being in state $s$ at a time $t+u$
-in the future, given the state at time $u$ is $r$.  It does not say
-anything about the time of transition from $r$ to $s$, indeed the
-process may have entered other states between times $u$ and $t+u$.
-$P(t)$ can be calculated by taking the matrix exponential of the
-scaled transition intensity matrix (see, for example, Cox and
-Miller \cite{cox:miller}).
-\begin{equation}
-  \label{eq:exptq}
-  P(t) = \Exp(tQ)
-\end{equation}
-
-The matrix exponential $\Exp$ is different from a scalar exponential.
-The exponential of a matrix is defined by the same "power series"
-$\Exp(X) = 1 + X^2/2! + X^3/3! + ...$ as the scalar exponential,
-except that each term $X^k$ in the series is defined by matrix products,
-not element-wise scalar multiplication.  It is notoriously difficult
-to calculate reliably, as discussed by Moler and van Loan \cite{matrixexp}.
-For simpler models, it is feasible to calculate an analytic expression
-for each element of $P(t)$ in terms of $Q$.  This is generally faster
-and avoids the potential numerical instability of calculating the
-matrix exponential.  Symbolic algebra sofware, such as Mathematica,
-can be helpful for obtaining these expressions.  For example, the
-three-state illness-death model with no recovery has a transition
-intensity matrix of
-\[
-Q = \left(
-  \begin{array}{llllll}
-    -(q_{12} + q_{13}) & q_{12} & q_{13}\\
-    0 & - q_{23} & q_{23}\\
-    0 & 0 & 0\\
-  \end{array}
-\right )
-\]
-The corresponding time $t$ transition probabilities are
-\begin{eqnarray*}
-  p_{11}(t) & = & e^{-(q_{12} + q_{13})t}\\
-  p_{12}(t) & = &
-\left\{
-\begin{array}{ll}
-\frac{ q_{12}(-1 + e^{(q_{12}+q_{13}-q_{23})t})e^{-(q_{12}+q_{13})t} } { (q_{12} + q_{13} - q_{23}) } & (q_{12} + q_{13} \neq q_{23})\\
-q_{12}te^{(-(q_{12} + q_{13})t} &  (q_{12} + q_{13} = q_{23})
-\end{array}
-\right.
-\\
-  p_{13}(t) & = &
-\left\{
-\begin{array}{ll}
-\frac{1 + (-q_{13} + q_{23})e^{-(q_{12}+q_{13})t}} {q_{12} + q_{13} - q_{23}} - \frac { q_{12}e^{-q_{23}t}} {q_{12} + q_{13} - q_{23} } & (q_{12} + q_{13} \neq q_{23})\\
- (-1 + e^{(q_{12} + q_{13})t} - q_{12}t)e^{-(q_{12} + q_{13})t} & (q_{12} + q_{13} = q_{23})
-\end{array}
-\right.
-\\
-  p_{21}(t) & = & 0\\
-  p_{22}(t) & = & e^{-q_{23}t}\\
-  p_{23}(t) & = & 1 - e^{-q_{23}t}\\
-  p_{31}(t) & = & 0\\
-  p_{32}(t) & = & 0\\
-  p_{33}(t) & = & 1\\
-\end{eqnarray*}
-The \Rpackage{msm} package calculates $P(t)$ analytically for selected
-2, 3, 4 and 5-state models, illustrated in
-Figures \ref{fig:anp2}--\ref{fig:anp5}.  For other models, which can have any transition
-structure on any number of states in principle, $P(t)$ is
-determined from the matrix exponential.  This is calculated using eigensystem
-decomposition (if eigenvalues are distinct) or a method based on
-Pad\'e approximants with scaling and squaring \cite{matrixexp} (if
-there are repeated eigenvalues). Notice that the states are not
-labelled in these figures.  Each graph can correspond to several
-different $Q$ matrices, depending on how the states are labelled.  For
-example, Figure 1 a) illustrates the model defined by either \( Q =
-\left(
-  \begin{array}{ll}
-    - q_{12} & q_{12} \\
-    0 & 0
-\end{array}
-\right)
-\)
-or
-\(
-Q = \left( \begin{array}{ll}
-0 & 0\\
-q_{21} & - q_{21}
-\end{array}
-\right)
-\).
-
-\begin{figure}
-\begin{center}
-a) \includegraphics[width=3cm]{figures/p2q1} \hskip 3cm
-b) \includegraphics[width=3cm]{figures/p2q12}
-\end{center}
-\caption{\label{fig:anp2}Two-state models fitted using analytic $P(t)$ matrices in \Rpackage{msm}. Implemented for all permutations of state labels 1, 2.}
-\end{figure}
-\begin{figure}
-\begin{center}
-a) \includegraphics[width=3cm]{figures/p3q12} \hskip 3cm
-b) \includegraphics[width=5cm]{figures/p3q14} \vskip 1cm
-c) \includegraphics[width=3cm]{figures/p3q16}\hskip 3cm
-d) \includegraphics[width=3cm]{figures/p3q124}\vskip 1cm
-e) \includegraphics[width=3cm]{figures/p3q135}\hskip 3cm
-f) \includegraphics[width=3cm]{figures/p3q1246}
-\end{center}
-\caption{\label{fig:anp3}Three-state models fitted using analytic $P(t)$ matrices in \Rpackage{msm}. Implemented for all permutations of state labels 1, 2, 3.}
-\end{figure}
-\begin{figure}
-\begin{center}
-a) \includegraphics[width=7cm]{figures/p4q159} \vskip 1cm
-b) \includegraphics[width=7cm]{figures/p4q13569}
-\end{center}
-\caption{\label{fig:anp4}Four-state models fitted using analytic $P(t)$ matrices in \Rpackage{msm}. Implemented for all permutations of state labels 1, 2, 3, 4.}
-\end{figure}
-\begin{figure}
-\begin{center}
-a) \includegraphics[width=9cm]{figures/p5q1_6_11_16}\vskip 1cm
-b) \includegraphics[width=9cm]{figures/p5q1_4_6_8_11_12_16}\vskip 1cm
-c) \includegraphics[width=5cm]{figures/p5q1_6_7_11_12}
-\end{center}
-\caption{\label{fig:anp5}Five-state models fitted using analytic $P(t)$ matrices in \Rpackage{msm}. Implemented for all permutations of state labels 1, 2, 3, 4, 5.}
-\end{figure}
-
-\paragraph{Likelihood for intermittently-observed processes}
-Suppose $i$ indexes $M$ individuals.  The data for individual $i$
-consist of a series of times $(t_{i1}, \ldots, t_{in_i})$ and
-corresponding states $(S(t_{i1}), \ldots, S(t_{in_i}))$.  Consider a
-general multi state model, with a pair of successive observed disease
-states $S(t_j), S(t_{j+1})$ at times $t_j, t_{j+1}$. The contribution
-to the likelihood from this pair of states is
-
-\begin{equation}
-  \label{eq:multi:lik:contrib}
-  L_{i, j} =  p_{S(t_j)S(t_{j+1})}(t_{j+1} - t_j)
-\end{equation}
-This is the entry of the transition matrix $P(t)$ at the $S(t_j)$th
-row and $S(t_{j+1})$th column, evaluated at $t = t_{j+1} - t_j$.
-
-The full likelihood $L(Q)$ is the product of all such terms $L_{i,j}$
-over all individuals and all transitions. It depends on the unknown
-transition matrix $Q$, which was used to determine $P(t)$.
-
-\paragraph{Exactly-observed death times}
-In observational studies of chronic diseases, it is common that the
-time of death is known, but the state on the previous instant before
-death is unknown. If $S(t_{j+1}) = D $ is such a death state, then the
-contribution to the likelihood is summed over the unknown state $m$ on
-the instant before death:
-
-\begin{equation}
-  \label{eq:multi:lik:death}
-  L_{i, j} =  \sum_{m \neq D} p_{S(t_j),m}(t_{j+1} - t_j) q_{m, D}
-\end{equation}
-The sum is taken over all possible states $m$ which can be visited
-between $S(t_j)$ and $D$.
-
-\paragraph{Exactly observed transition times}
-If the times $(t_{i1}, \ldots, t_{in_i})$ had been the {\em exact}
-transition times between the states, with no transitions between the
-observation times, then the contributions would be of the form
-
-\begin{equation}
-  \label{eq:multi:lik:exact}
-  L_{i, j} =  \exp(q_{S(t_j)S(t_{j})}(t_{j+1} - t_j)) q_{S(t_j)S(t_{j+1})}
-\end{equation}
-since the state is assumed to be $S(t_j)$ throughout the interval
-between $t_j$ and $t_{j+1}$ with a known transition to state
-$S(t_{j+1})$ at $t_{j+1}$.  \Rpackage{msm} is restricted to Markov
-models, but much richer models are possible for this type of data. For
-example, Putter \etal \cite{putter:mstate} discussed the
-\Rpackage{mstate} software for semi-parametric multi-state models with
-non-parametric baseline hazards and Cox regression.  The Markov
-assumption is restrictive but necessary in general to compute a
-likelihood for intermittently-observed processes.
-
-\paragraph{Censored states}
-A censored quantity is one whose exact value is unknown, but known to
-be in a certain interval.  For example, in survival analysis, a death
-time is \emph{right-censored} if the study ends and the patient is
-still alive, since the death time is known to be greater than the end
-time.  In multi-state models for intermittently-observed processes,
-the times of changes of state are usually \emph{interval censored},
-known to be within bounded intervals.  This leads to a likelihood
-based on equation \ref{eq:multi:lik:contrib}.
-
-In some circumstances, \emph{states} may be censored as well as
-\emph{event times}.  For example, at the end of some chronic disease
-studies, patients are known to be alive but in an \emph{unknown
-  state}.  For such a censored observation $S(t_{j+1})$, known only to
-be a state in the set $C$, the equivalent contribution to the
-likelihood is
-\begin{equation}
-  \label{eq:multi:lik:deathcens}
-  L_{i, j} =  \sum_{m \in C} p_{S(t_j),m}(t_{j+1} - t_j)
-\end{equation}
-Note that this special likelihood is not needed if the state is known
-at the end of the study.  In this case,
-likelihood \ref{eq:multi:lik:contrib} applies.  Although the
-\emph{survival time} is censored, the \emph{state} at the end of the
-study is not censored.
-
-More generally, suppose every observation from a particular individual
-is censored.  Observations $1, 2, \ldots n_i$ are known only to be in
-the sets $C_1, C_2, \ldots, C_{n_i}$ respectively. The likelihood for
-this individual is a sum of the likelihoods of all possible paths
-through the unobserved states.
-\begin{equation}
-  \label{eq:multi:lik:cens}
-  L_i = \sum_{s_{n_i} \in C_{n_i}} \ldots \sum_{s_2 \in C_2} \sum_{s_1 \in C_1} p_{s_1 s_2}(t_2 - t_1) p_{s_2 s_3} (t_3 - t_2) \ldots p_{s_{n_i-1} s_{n_i}} (t_{n_i} - t_{n_i-1})
-\end{equation}
-Suppose the states comprising the set $C_j$ are $c^{(j)}_1, \ldots,
-c^{(j)}_{m_j}$. This likelihood can also be written as a matrix
-product, say,
-\begin{equation}
-  \label{eq:multi:lik:cens:matrix}
-  L_i = \mathbf{1}^T P^{1,2} P^{2,3} \ldots P^{n_i-1, n_i} \mathbf{1}
-\end{equation}
-where $P^{j-1, j}$ is a $m_{j-1} \times m_j$ matrix with $(r,s)$ entry
-$p_{c^{(j-1)}_r c^{(j)}_s}(t_j - t_{j-1})$, and $\mathbf{1}$ is the
-vector of ones.
-
-The \Rpackage{msm} package allows multi-state models to be fitted to
-data from processes with arbitrary observation times (panel data), exactly-observed
-transition times, exact death times and censored states, or a mixture
-of these schemes.
-
-\subsection{Covariates}
-\label{sec:multi:covariates}
-
-The relation of constant or time-varying characteristics of
-individuals to their transition rates is often of interest in a
-multi-state model.  Explanatory variables for a particular transition
-intensity can be investigated by modelling the intensity as a function
-of these variables.  Marshall and Jones \cite{retino2} described a
-form of a {\em proportional hazards} model, where the transition
-intensity matrix elements $q_{rs}$ which are of interest can be
-replaced by
-\[
-q_{rs}(z(t)) = q_{rs}^{(0)} \exp(\beta_{rs}^T z(t))
-\]
-The new $Q$ is then used to determine the likelihood. If the
-covariates $z(t)$ are time dependent, the contributions to the
-likelihood of the form $p_{rs} (t - u)$ are replaced by
-\[
-p_{rs}(t - u, z(u))
-\]
-although this requires that the value of the covariate is known at
-every observation time $u$. Sometimes covariates are observed at
-different times to the main response, for example recurrent disease
-events or other biological markers. In some of these cases it could be
-assumed that the covariate is a step function which remains constant
-between its observation times.  If the main response (the state of the
-Markov process) is not observed at the times when the covariate
-changes, it could be considered as a "censored" state (as in
-Section \ref{sec:multi:likelihood}).
-
-The \Rpackage{msm} package allows individual-specific or time
-dependent covariates to be fitted to transition intensities.  In order
-to calculate transition probabilities $P(t)$ on which the likelihood
-depends, time-dependent covariates are assumed to be
-piecewise-constant.  Models whose intensities change with time are
-called \emph{time-inhomogeneous}. An important special case handled by
-\Rpackage{msm} is the model in which intensities change at a series of
-times common to each individual.
-
-Marshall and Jones \cite{retino2} described likelihood ratio and Wald
-tests for covariate selection and testing hypotheses, for example
-whether the effect of a covariate is the same for all forward
-transitions in a disease progression model, or whether the effect on
-backward transitions is equal to minus the effect on forward
-transitions.
-
-\subsection{Hidden Markov models}
-\label{sec:hmm}
-
-In a {\em hidden Markov model} (HMM) the states of the Markov chain
-are not observed. The observed data are governed by some probability
-distribution (the \emph{emission} distribution) conditionally on the
-unobserved state.  The evolution of the underlying Markov chain is
-governed by a transition intensity matrix $Q$ as before.
-(Figure \ref{fig:multi:hidden}).  Hidden Markov models are mixture
-models, where observations are generated from a certain number of
-unknown distributions.  However the distribution changes through time
-according to states of a hidden Markov chain.  This class of model is
-commonly used in areas such as speech and signal processing
-\cite{juang:rabiner} and the analysis of biological sequence data
-\cite{biolog:seq}. In engineering and biological sequencing
-applications, the Markov process usually evolves over an
-equally-spaced, discrete `time' space.  Therefore most of the theory
-of HMM estimation was developed for discrete-time models.
-
-HMMs have less frequently been used in medicine, where continuous time
-processes are often more suitable.  A disease process evolves in
-continuous time, and patients are often monitored at irregular and
-differing intervals.  These models are suitable for estimating
-population quantities for chronic diseases which have a natural staged
-interpretation, but which can only be diagnosed by an error-prone
-marker.  The \Rpackage{msm} package can fit continuous-time hidden
-Markov models with a variety of emission distributions.
-
-\begin{figure}[htbp]
-  \begin{center}
-    \scalebox{0.6}{\rotatebox{270}{\includegraphics{figures/hidden}}}
-    \caption{A hidden Markov model in continuous time. Observed states
-      are generated conditionally on an underlying Markov process. }
-    \label{fig:multi:hidden}
-  \end{center}
-\end{figure}
-
-\subsubsection{Misclassification models}
-
-An example of a hidden Markov model is a multi-state model with
-misclassification.  Here the observed data are states, assumed to be
-misclassifications of the true, underlying states.
-
-For example, consider a disease progression model with at least a
-disease-free and a disease state. When screening for the presence of
-the disease, the screening process can sometimes be subject to error.
-Then the Markov disease process $S_i(t)$ for individual $i$ is not
-observed directly, but through realisations $O_i(t)$.  The quality of
-a diagnostic test is often measured by the probabilities that the true
-and observed states are equal, $Pr(O_i(t) = r | S_i(t) = r)$. Where
-$r$ represents a `positive' disease state, this is the {\em
-  sensitivity}, or the probability that a true positive is detected by
-the test.  Where $r$ represents a `negative' or disease-free state,
-this represents the {\em specificity}, or the probability that, given
-the condition of interest is absent, the test produces a negative
-result.
-
-As an extension to the simple multi-state model described in
-section \ref{sec:multistate}, the \Rpackage{msm} package can fit a
-general multi-state model with misclassification.  For patient $i$,
-observation time $t_{ij}$, observed states $O_{ij}$ are generated
-conditionally on true states $S_{ij}$ according to a {\em
-  misclassification matrix} $E$. This is a $n \times n$ matrix, whose
-$(r,s)$ entry is
-\begin{equation}
-  \label{eq:misc}
-  e_{rs} = Pr(O(t_{ij}) = s | S(t_{ij}) = r),
-\end{equation}
-which we first assume to be independent of time $t$. Analogously to
-the entries of $Q$, some of the $e_{rs}$ may be fixed to reflect
-knowledge of the diagnosis process. For example, the probability of
-misclassification may be negligibly small for non-adjacent states of
-disease.  Thus the progression through underlying states is governed
-by the transition intensity matrix $Q$, while the observation process
-of the underlying states is governed by the misclassification matrix
-$E$.
-
-To investigate explanatory variables $w(t)$ for the probability $e_{rs}$ of
-misclassification as state $s$ given underlying state $r$, a
-multinomial logistic regression model can be used:
-\begin{equation}
-  \label{eq:misccovs}
-  \log \frac{e_{rs}(t)}{e_{rs_0}(t)}  =   \gamma_{rs}^T w(t).
-\end{equation}
-where $s_0$ is some baseline state, usually chosen as the underlying
-state, or the state with the highest probability (for numerical
-stability).
-
-\subsubsection{General hidden Markov model}
-\label{sec:hmm:general}
-
-Consider now a general hidden Markov model in continuous time.  The
-true state of the model $S_{ij}$ evolves as an unobserved Markov
-process.  Observed data $y_{ij}$ are generated conditionally true
-states $S_{ij} = 1, 2, \ldots, n$ according to a set of distributions
-$f_1(y | \theta_1, \gamma_1)$, $f_2(y | \theta_2, \gamma_2)$,
-$\ldots$, $f_n(y | \theta_n, \gamma_n)$, respectively.  $\theta_r$ is
-a vector of parameters for the state $r$ distribution.  One or more of
-these parameters may depend on explanatory variables through a
-link-transformed linear model with coefficients $\gamma_r$.
-
-\subsubsection{Likelihood for general hidden Markov models}
-
-A type of EM algorithm known as the {\em Baum-Welch} or {\em
-  forward-backward} algorithm \cite{baum:petrie66, baum:petrie70}, is
-commonly used for hidden Markov model estimation in discrete-time
-applications.  See, for example, Durbin \etal
-\cite{biolog:seq}, Albert \cite{albert99}.  A generalisation of this
-algorithm to continuous time was described by Bureau \etal
-\cite{bureau:hughes:shiboski:00}.
-
-The \Rpackage{msm} package uses a direct method of calculating
-likelihoods in discrete or continuous time based on matrix products.
-This type of method has been described by Macdonald and Zucchini
-\cite[pp.  77--79]{macdonald:zucchini}, Lindsey
-\cite[p.73]{lindsey:rm} and Guttorp \cite{guttorp}.  Satten and
-Longini \cite{sattlong} used this method to calculate likelihoods for
-a hidden Markov model in continuous time with observations of a
-continuous marker generated conditionally on underlying discrete
-states.
-
-Patient $i$'s contribution to the likelihood is
-\begin{eqnarray}
-  \label{eq:multi:hiddencontrib}
-  L_i     & = & Pr(y_{i1}, \ldots, y_{in_i})\\
-  & = & \sum Pr(y_{i1}, \ldots, y_{in_i} | S_{i1}, \ldots, S_{in_i})
-  Pr(S_{i1}, \ldots, S_{in_i}) \nonumber
-\end{eqnarray}
-where the sum is taken over all possible paths of underlying states
-$S_{i1}, \ldots, S_{in_i}$.  Assume that the observed states are
-conditionally independent given the values of the underlying states.
-Also assume the Markov property, $Pr(S_{ij}|S_{i,j-1}, \ldots, S_{i1})
-= Pr(S_{ij}|S_{i,j-1})$.  Then the contribution $L_i$ can be written
-as a product of matrices, as follows. To derive this matrix product,
-decompose the overall sum in equation \ref{eq:multi:hiddencontrib}
-into sums over each underlying state. The sum is accumulated over the
-unknown first state, the unknown second state, and so on until the
-unknown final state:
-\begin{eqnarray}
-  \label{eq:multi:hiddenlik}
-  L_i & = & \sum_{S_{i1}} Pr(y_{i1}|S_{i1})Pr(S_{i1}) \sum_{S_{i2}} Pr(y_{i2}|S_{i2})Pr(S_{i2}|S_{i1}) \sum_{S_{i3}} Pr(y_{i3}|S_{i3}) Pr(S_{i3}|S_{i2})
-  \nonumber \\
-  &   & \ldots \sum_{S_{in_i}} Pr(y_{in_i}|S_{in_i})
-  Pr(S_{in_i}|S_{in_{i-1}})
-\end{eqnarray}
-where $Pr(y_{ij}|S_{ij})$ is the emission probability density.  For
-misclassification models, this is the misclassification probability
-$e_{S_{ij} O_{ij}}$.  For general hidden Markov models, this is the
-probability density
-$f_{S_{ij}}(y_{ij}|\theta_{S_{ij}},\gamma_{S_{ij}})$.
-$Pr(S_{i,j+1}|S_{ij})$ is the $(S_{ij}, S_{i,j+1})$ entry of the
-Markov chain transition matrix $P(t)$ evaluated at $t = t_{i,j+1} -
-t_{ij}$.  Let $f$ be the vector with $r$ element the product of the
-initial state occupation probability $Pr(S_{i1}=r)$ and $Pr(y_{i1}|
-r)$, and let $\mathbf 1$ be a column vector consisting of ones. For $j
-= 2, \ldots, n_i$ let $T_{ij}$ be the $R \times R$ matrix (where $R$
-is the number of states) with $(r,s)$ entry
-\[
-Pr(y_{ij}| s) p_{rs} (t_{ij} - t_{i,j-1})
-\]
-Then subject $i$'s likelihood contribution is
-\begin{equation}
-  L_i =  f T_{i2} T_{i3}, \ldots T_{in_i} \mathbf 1
-  \label{eq:multi:hidden:matprod}
-\end{equation}
-
-If $S(t_{j}) = D$ is an absorbing state such as death, measured
-without error, whose entry time is known exactly, then the
-contribution to the likelihood is summed over the unknown state at the
-previous instant before death.  The $(r,s)$ entry of $T_{ij}$ is then
-\[
-p_{rs}(t_{j} - t_{j-1}) q_{s, D}
-\]
-Section \ref{sec:fitting:hmm:misc} describes how to fit multi-state
-models with misclassification using the \Rpackage{msm} package, and
-Section \ref{sec:fitting:hmm:general} describes how to apply general
-hidden Markov models.
-
-\subsubsection{Example of a general hidden Markov model}
-\label{sec:hmm:example:fev}
-
-Jackson and Sharples \cite{jackson:sharples:2002} described a model
-for FEV$_1$ (forced expiratory volume in 1 second) in recipients of
-lung transplants.  These patients were at risk of BOS (bronchiolitis
-obliterans syndrome), a progressive, chronic deterioration in lung
-function.  In this example, BOS was modelled as a discrete, staged
-process, a model of the form illustrated in Figure \ref{fig:disease},
-with 4 states.  State 1 represents absence of BOS. State 1 BOS is
-roughly defined as a sustained drop below 80\% below baseline FEV$_1$,
-while state 2 BOS is a sustained drop below 65\% baseline.  FEV$_1$ is
-measured as a percentage of a baseline value for each individual,
-determined in the first six months after transplant, and assumed to be
-100\% baseline at six months.
-
-As FEV$_1$ is subject to high short-term variability due to acute
-events and natural fluctuations, the exact BOS state at each
-observation time is difficult to determine.  Therefore, a hidden
-Markov model for FEV$_1$, conditionally on underlying BOS states, was
-used to model the natural history of the disease.  Discrete states are
-appropriate as onset is often sudden.
-
-\paragraph{Model 1}
-Jackson \cite{my:phd} considered models for these data where
-FEV$_1$ were Normally distributed, with an unknown mean and variance
-conditionally each state (\ref{eq:fev:normal}).  This model seeks the
-most likely location for the within-state FEV$_1$ means.
-
-\begin{equation}
-  \label{eq:fev:normal}
-  y_{ij} | \{ S_{ij} = k\}  \sim N(\mu_k + \beta x_{ij}, \sigma^2_k)
-\end{equation}
-
-\paragraph{Model 2}
-Jackson and Sharples \cite{jackson:sharples:2002} used a more complex
-two-level model for FEV$_1$ measurements. Level
-1 (\ref{eq:fev:level1}) represents the short-term fluctuation error of
-the marker around its underlying continuous value $y^{hid}_{ij}$.
-Level 2 (\ref{eq:fev:level2}) represents the distribution of
-$y^{hid}_{ij}$ conditionally on each underlying state, as follows.
-\begin{equation}
-  \label{eq:fev:level1}
-  y_{ij} | y^{hid}_{ij} \qquad \sim  N ( y^{hid}_{ij} + \beta x_{ij} ,
-  \sigma^2_\epsilon)
-\end{equation}
-\begin{equation}
-  \label{eq:fev:level2}
-  y^{hid}_{ij} | S_{ij} \quad \sim \quad
-  \left\{
-    \begin{array}{cll}
-      \mbox{State}& \mbox{Three state model}               & \mbox{Four state model} \\
-      S_{ij} = 0  & N(\mu_0, \sigma^2_0)I_{[80, \infty)} & N(\mu_0, \sigma^2_0)I_{[80, \infty)} \\
-      S_{ij} = 1  & N(\mu_1, \sigma^2_1)I_{(0, 80)}      & Uniform(65, 80)                        \\
-      S_{ij} = 2  & \mbox{(death)}                         & N(\mu_2, \sigma^2_2)I_{(0, 65)}      \\
-      S_{ij} = 3  &                                        & \mbox{(death)}
-    \end{array}
-  \right .
-\end{equation}
-
-Integrating over $y^{hid}_{ij}$ gives an explicit distribution for
-$y_{ij}$ conditionally on each underlying state (given in
-Section \ref{sec:fitting:hmm:general}, Table \ref{tab:hmm:dists}).
-Similar distributions were originally applied by Satten and
-Longini \cite{sattlong} to modelling the progression through discrete,
-unobserved HIV disease states using continuous CD4 cell counts.  The
-\Rpackage{msm} package includes density, quantile, cumulative density
-and random number generation functions for these distributions.
-
-In both models 1 and 2, the term $\beta x_{ij}$ models the short-term
-fluctuation of the marker in terms of acute events. $x_{ij}$ is an
-indicator for the occurrence of an acute rejection or infection
-episode within 14 days of the observation of FEV$_1$.
-
-Section \ref{sec:fitting:hmm:general} describes how these and more
-general hidden Markov models can be fitted with the \Rpackage{msm}
-package.
-
-\clearpage
-
-\section{Fitting multi-state models with {\tt msm}}
-
-<<echo=FALSE>>=
-options(width = 60)
-@
-
-\Rpackage{msm} is a package of functions for multi-state modelling using
-the R statistical software.  The \Rfunction{msm} function itself
-implements maximum-likelihood estimation for general multi-state
-Markov or hidden Markov models in continuous time. We illustrate
-its use with a set of data from monitoring heart transplant patients.
-Throughout this section ``\textsl{\texttt{>}}'' indicates the R
-command prompt, \textsl{\texttt{slanted typewriter}} text indicates R
-commands, and \texttt{typewriter} text R output.
-
-\subsection{Installing \tt{msm}}
-\label{sec:installing}
-
-The easiest way to install the \Rpackage{msm} package on a computer
-connected to the Internet is to run the R command:
-
-\begin{Scode}
-  install.packages("msm")
-\end{Scode}
-
-This downloads \Rpackage{msm} from the CRAN archive of contributed R
-packages (\texttt{cran.r-project.org} or one of its mirrors) and
-installs it to the default R system library.  To install to a
-different location, for example if you are a normal user with no
-administrative privileges, create a directory in which R packages are
-to be stored, say, \texttt{/your/library/dir}, and run
-
-\begin{Scode}
-  install.packages("msm", lib='/your/library/dir')
-\end{Scode}
-
-After \Rpackage{msm} has been installed, its functions can be made
-visible in an R session by
-<<>>=
-library(msm)
-@
-or, if it has been installed into a non-default library,
-
-\begin{Scode}
-  library(msm, lib.loc='/your/library/dir')
-\end{Scode}
-
-\subsection{Getting the data in}
-\label{sec:datain}
-
-The data are specified as a series of observations, grouped by
-patient. At minimum there should be a data frame with variables
-indicating
-\begin{itemize}
-\item the time of the observation,
-\item the observed state of the process.
-\end{itemize}
-If the data do not also contain
-\begin{itemize}
-\item the subject identification number,
-\end{itemize}
-then all the observations are assumed to be from the same subject.
-The subject ID does not need to be numeric, but data must be grouped
-by subject, and observations must be ordered by time within subjects.
-If the model includes variables with missing values, then the corresponding
-observations  are omitted by \Rfunction{msm} with a warning.  If you have missing data,
-as in any statistical model, it is recommended to ensure these do not
-result in biases.
-
-An example data set, taken from monitoring a set of heart transplant
-recipients, is provided with \Rpackage{msm}.  (Note: since \Rpackage{msm}
-version 1.3, the command \Rfunction{data(cav)} is no longer needed to
-load the data --- it is now ``lazy-loaded'' when required). Sharples
-\etal \cite{my:cav} studied the progression of coronary allograft
-vasculopathy (CAV), a post-transplant deterioration of the arterial
-walls, using these data.  Risk factors and the accuracy of the
-screening test were investigated using multi-state Markov and hidden
-Markov models.
-
-The first three patient histories are shown below. There are 622
-patients in all.  \Robject{PTNUM} is the subject identifier.
-Approximately each year after transplant, each patient has an
-angiogram, at which CAV can be diagnosed.  The result of the test is
-in the variable \Robject{state}, with possible values 1, 2, 3
-representing CAV-free, mild CAV and moderate or severe CAV
-respectively.  A value of 4 is recorded at the date of death.
-\Robject{years} gives the time of the test in years since the heart
-transplant. Other variables include \Robject{age} (age at screen),
-\Robject{dage} (donor age), \Robject{sex} (0=male, 1=female),
-\Robject{pdiag} (primary diagnosis, or reason for transplant - IHD
-represents ischaemic heart disease, IDC represents idiopathic dilated
-cardiomyopathy), \Robject{cumrej} (cumulative number of rejection
-episodes), and \Robject{firstobs}, an indicator which is 1 when the
-observation corresponds to the patient's transplant (the first
-observation), and 0 when the observation corresponds to a later
-angiogram.
-
-<<>>=
-cav[1:21,]
-@
-
-A useful way to summarise multi-state data is as a frequency table of
-pairs of consecutive states.  This counts over all individuals, for
-each state $r$ and $s$, the number of times an individual had an
-observation of state $r$ followed by an observation of state $s$. The
-function \Rfunction{statetable.msm} can be used to produce such a table,
-as follows,
-
-<<>>=
-statetable.msm(state, PTNUM, data=cav)
-@
-Thus there were 148 CAV-free deaths, 48 deaths from state 2, and
-55 deaths from state 3.  On only four occasions was there an
-observation of severe CAV followed by an observation of no CAV.
-
-\subsection{Specifying a model}
-\label{sec:specifying:model}
-
-We now specify the multi-state model to be fitted to the data.  A
-model is governed by a transition intensity matrix $Q$. For the heart
-transplant example, there are four possible states through which the
-patient can move, corresponding to CAV-free, mild/moderate CAV, severe
-CAV and death.  We assume that the patient can advance or recover from
-consecutive states while alive, and die from any state. Thus the model
-is illustrated by Figure \ref{fig:disease} with four states, and we
-have
-
-\[
-Q = \left(
-  \begin{array}{llll}
-    -(q_{12} + q_{14}) & q_{12} &  0     & q_{14}\\
-    q_{21} & -(q_{21}+q_{23}+q_{24}) & q_{23} & q_{24}\\
-      0    & q_{32} & -(q_{32}+q_{34}) & q_{34}\\
-      0    &   0    &   0    &   0   \\
-  \end{array}
-\right )
-\]
-
-It is important to remember that this defines which
-\emph{instantaneous} transitions can occur in the Markov process, and
-that the data are \emph{snapshots} of the process (see
-section \ref{sec:arbitr-observ-times}).  Although there were 44
-occasions on which a patient was observed in state 1 followed by state
-3, the underlying model specifies that the patient must have passed
-through state 2 in between.  If your data represent the exact and
-complete transition times of the process, then you must specify
-\Rfunarg{exacttimes=TRUE} or \Rfunarg{obstype=2} in the call to
-\Rfunction{msm}.
-
-To tell \Rfunction{msm} what the allowed transitions of our model are,
-we define a matrix of the same size as $Q$, containing zeroes in the
-positions where the entries of $Q$ are zero.  All other positions
-contain an initial value for the corresponding transition intensity.
-The diagonal entries supplied in this matrix do not matter, as the
-diagonal entries of $Q$ are defined as minus the sum of all the other
-entries in the row.  This matrix will eventually be used as the
-\Rfunarg{qmatrix} argument to the \Rfunction{msm} function. For
-example,
-
-<<>>=
-twoway4.q  <-  rbind ( c(0, 0.25, 0, 0.25),
-                       c(0.166, 0, 0.166, 0.166),
-                       c(0, 0.25, 0, 0.25),
-                       c(0, 0, 0, 0) )
-@
-
-Fitting the model is a process of finding values of the seven unknown
-transition intensities: $q_{12}$, $q_{14}$, $q_{21}$, $q_{23}$,
-$q_{24}$, $q_{32}$, $q_{34}$, which maximise the likelihood.
-
-\subsection{Specifying initial values}
-\label{sec:inits}
-
-The likelihood is maximised by numerical methods, which need a set of
-initial values to start the search for the maximum.  For reassurance
-that the true maximum likelihood estimates have been found, models
-should be run repeatedly starting from different initial
-values. However a sensible choice of initial values can be important
-for unstable models with flat or multi-modal likelihoods.  For
-example, the transition rates for a model with misclassification could
-be initialised at the corresponding estimates for an approximating
-model without misclassification. Initial values for a model without
-misclassification could be set by supposing that transitions between
-states take place only at the observation times.  If we observe
-$n_{rs}$ transitions from state $r$ to state $s$, and a total of $n_r$
-transitions from state $r$, then $q_{rs} / q_{rr}$ can be estimated by
-$n_{rs} / n_r$. Then, given a total of $T_r$ years spent in state $r$,
-the mean sojourn time $1 / q_{rr}$ can be estimated as $T_r / n_r$.
-Thus, $n_{rs} / T_r$ is a crude estimate of $q_{rs}$. The \Rpackage{msm}
-package provides a function \Rfunction{crudeinits.msm} for calculating
-initial values in this way (for non-hidden Markov models).
-
-<<>>=
-twoway4.crude.q  <- crudeinits.msm(state ~ years, PTNUM, data=cav, qmatrix=twoway4.q)
-@
-
-However, if there are are many changes of state in between the
-observation times, then this crude approach may fail to give sensible
-initial values.  For the heart transplant example we could also guess
-that the mean period in each state before moving to the next is about
-2 years, and there is an equal probability of progression, recovery or
-death.  This gives $q_{rr} = - 0.5$ for $r = 1, 2, 3$, and $q_{12} =
-q_{14} = 0.25$, $q_{21} = q_{23} = q_{24} = 0.166$, $q_{32} = q_{34} =
-0.25$, and the initial value matrix \Robject{twoway4.q} shown above,
-which we now use to fit the model.
-
-%% TODO explain about gen.inits=TRUE here
-
-\subsection{Running \Rfunction{msm}}
-\label{sec:running}
-
-To fit the model, call the \Rfunction{msm} function with the appropriate
-arguments.  For our running example, we have defined a data set
-\Robject{cav}, a matrix \Robject{twoway4.q} indicating the allowed
-transitions, and initial values. We are ready to run \Rfunction{msm}.
-
-\paragraph{Model 1: simple bidirectional model}
-
-<<>>=
-cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                    qmatrix = twoway4.q, death = 4 )
-@
-
-In this example the day of death is assumed to be recorded exactly, as
-is usual in studies of chronic diseases.  At the previous instant
-before death the state of the patient is unknown.  Thus we specify
-\Rfunarg{death = 4}, to indicate to \Rfunction{msm} that the
-entry times into state 4 are observed in this manner.
-If the model had
-five states, and states 4 and 5 were two competing causes of death with
-times recorded exactly in this way, then we would specify \Rfunarg{death =
-  c(4,5)}.
-
-By default, the data are assumed to represent snapshots of the process
-at arbitrary times.  However, observations can also represent exact
-times of transition, ``exact death times'', or a mixture of these. See the
-\Rfunarg{obstype} argument to \Rfunction{msm}.  Note you do not always have to specify \Rfunarg{death},
-even if there are states which correspond to death.  This is
-only used for states whose entry time is observed in this specific way.
-
-While the \Rfunction{msm} function runs, it searches for the maximum
-of the likelihood of the unknown parameters.  Internally, it uses the
-R function \Rfunction{optim} to minimise the minus log-likelihood.
-When the data set, the model, or both, are large, then this may take a
-long time.  It can then be useful to see the progress of the
-optimisation algorithm. To do this, we can specify a \Rfunarg{control}
-argument to \Rfunction{msm}, which is passed internally to the
-\Rfunction{optim} function.  For example \texttt{control = list(trace=1,
-  REPORT=1)}. See the help page for \Rfunction{optim},
-
-<<eval=FALSE>>=
-help(optim)
-@
-for more options to control the optimisation.  When completed, the
-\Rfunction{msm} function returns a value.  This value is a list of the
-important results of the model fitting, including the parameter
-estimates and their covariances.  To keep these results for
-post-processing, we store them in an R object, here called
-\Robject{cav.msm}.  When running several similar \Rfunction{msm} models,
-it is recommended to store the respective results in
-informatively-named objects.
-
-\subsection{Showing results}
-
-To show the maximum likelihood estimates and 95\% confidence
-intervals, type the name of the fitted model object at the R command
-prompt.  \footnote{This is equivalent to typing
-  \texttt{print.msm(cav.msm)}.  The function \Rfunction{print.msm}
-  formats the important information in the model object for printing
-  on the screen.}  The confidence level can be changed using the
-\Rfunarg{cl} argument to \Rfunction{msm}.
-
-<<>>=
-cav.msm
-@
-
-From the estimated intensity matrix, we see patients are three times
-as likely to develop symptoms than die without symptoms (first row).
-After disease onset (state 2), progression to severe symptoms (state
-3) is 50\% more likely than recovery.  Once in the severe state, death
-is more likely than recovery, and a mean of 1 / -0.44 = 2.3 years is
-spent in state 3 before death or recovery.
-
-Section \ref{sec:extractor} describes various functions that can be
-used to obtain summary information from the fitted model.
-
-\subsection{Covariates on the transition rates}
-\label{sec:msm:covariates}
-
-We now model the effect of explanatory variables on the rates of
-transition, using a proportional intensities model. Now we have an
-intensity matrix $Q(z)$ which depends on a covariate vector $z$. For
-each entry of $Q(z)$, the transition intensity for patient $i$ at
-observation time $j$ is $q_{rs}(z_{ij}) = q_{rs}^{(0)}
-\exp(\beta_{rs}^T z_{ij})$.  The covariates $z$ are specified through
-the \Rfunarg{covariates} argument to \Rfunction{msm}. If $z_{ij}$ is
-time-dependent, we assume it is constant in between the observation
-times of the Markov process.  \Rfunction{msm} calculates the
-probability for a state transition from times $t_{i,j-1}$ to $t_{ij}$
-using the covariate value at time $t_{i,j-1}$.
-
-We consider a model with just one covariate, female sex.  Out of the
-622 transplant recipients, 535 are male and 87 are female.  By
-default, all linear covariate effects $\beta_{rs}$ are initialised to
-zero. To specify different initial values, use a \Rfunarg{covinits}
-argument, as described in \Rfunction{help(msm)}.  Initial values given
-in the \Rfunarg{qmatrix} represent the intensities with covariate values
-set to their means in the data.  In the following model, all transition
-intensities are modelled in terms of sex.
-
-\paragraph{Model 2: sex as a covariate}
-<<>>=
-cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                     qmatrix = twoway4.q, death = 4, covariates = ~ sex)
-@
-
-The \Robject{msm} object will now include the estimated covariate
-effects and their confidence intervals.
-
-<<>>=
-cavsex.msm
-@
-Comparing the estimated log-linear effects of sex and their standard
-errors, we see that the disease onset rate is smaller for females,
-whereas none of the other effects are significant. The first matrix shown in the output of printing
-\Robject{cavsex.msm} is the estimated transition intensity matrix
-$q_{rs}(z) = q_{rs}^{(0)} \exp(\beta_{rs}^T z)$ with the covariate $z$
-set to its mean value in the data.  This represents an average
-intensity matrix for the population of 535 male and 87 female
-patients. To extract separate intensity matrices for male and female
-patients ($z = 0$ and $1$ respectively), use the function
-\Rfunction{qmatrix.msm}, as shown below.  This and similar summary
-functions will be described in more detail in
-section \ref{sec:extractor}.
-
-<<>>=
-qmatrix.msm(cavsex.msm, covariates=list(sex=0)) # Male
-qmatrix.msm(cavsex.msm, covariates=list(sex=1)) # Female
-@
-
-Different transition rates may be easily modelled on different
-covariates by specifying a named list of formulae as the
-\Rfunarg{covariates} argument.  Each element of the list has a name
-identifying the transition.  In the model below, the transition rate
-from state 1 to state 2 and the rate from state 3 to state 4 are each
-modelled on sex as a covariate, but no other intensities have
-covariates on them.
-
-\paragraph{Model 2a: transition-specific covariates}
-<<eval=FALSE>>=
-cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                     qmatrix = twoway4.q, death = 4,
-                     covariates = list("1-2" = ~ sex, "3-4" = ~sex) )
-@
-
-We may also want to constrain the effect of a covariate to be equal for
-certain transition rates, to reduce the number of parameters in the
-model, or to investigate hypotheses on the covariate effects.  A
-\Rfunarg{constraint} argument can be used to indicate which of the
-transition rates have common covariate effects.
-
-\paragraph{Model 3: constrained covariate effects}
-
-<<eval=FALSE>>=
-cav3.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                   qmatrix = twoway4.q, death = 4,
-                   covariates = ~ sex, constraint = list(sex=c(1,2,3,1,2,3,2)) )
-@
-
-This constrains the effect of sex to be equal for the progression
-rates $q_{12}, q_{23}$, equal for the death rates $q_{14}, q_{24},
-q_{34}$, and equal for the recovery rates $q_{21}, q_{32}$. The
-intensity parameters are assumed to be ordered by reading across the
-rows of the transition matrix, starting at the first row: ($q_{12},
-q_{14}, q_{21}, q_{23}, q_{24}, q_{32}, q_{34}$), giving constraint
-indicators \Rfunarg{(1,2,3,1,2,3,2)}.  Any vector of increasing
-numbers can be used for the indicators.  Negative entries can be used
-to indicate that some effects are equal to minus others:
-\Rfunarg{(1,2,3,-1,2,3,2)} sets the fourth effect to be minus the
-first.
-
-In a similar manner, we can constrain some of the baseline transition
-intensities to be equal to one another, using the
-\Rfunarg{qconstraint} argument.  For example, to constrain the rates
-$q_{12}$ and $q_{23}$ to be equal, and $q_{24}$ and $q_{34}$ to be
-equal, specify \Rfunarg{qconstraint = c(1,2,3,1,4,5,4)}.
-
-\subsection{Fixing parameters at their initial values}
-
-For exploratory purposes we may want to fit a model assuming that some
-parameters are fixed, and estimate the remaining parameters.  This may
-be necessary in cases where there is not enough information in the
-data to be able to estimate a proposed model, and we have strong prior
-information about a certain transition rate. To do this, use the
-\Rfunarg{fixedpars} argument to \Rfunction{msm}.  For model 1, the
-following statement fixes the parameters numbered 6, 7, that is,
-$q_{32}$, $q_{34}$, to their initial values (0.25 and 0.25,
-respectively).
-
-\paragraph{Model 4: fixed parameters}
-<<eval=FALSE>>=
-cav4.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                qmatrix = twoway4.q, death = 4,
-                control = list(trace=2, REPORT=1),
-                fixedpars = c(6, 7) )
-@
-
-A \Rfunarg{fixedpars} statement can also be useful for fixing covariate
-effect parameters to zero, that is to assume no effect of a covariate
-on a certain transition rate.
-
-\subsection{Extractor functions}
-\label{sec:extractor}
-
-We may want to extract some of the information from the \Rfunction{msm}
-model fit for post-processing, for example for plotting graphs or
-generating summary tables.  A set of functions is provided for
-extracting interesting features of the fitted model.
-
-\begin{description}
-
-\item[Intensity matrices] The function \Rfunction{qmatrix.msm}
-  extracts the estimated transition intensity matrix and its confidence intervals
-  for a given set of covariate values, as shown in
-  section \ref{sec:msm:covariates}.  Confidence intervals are
-  calculated from the covariance matrix of the estimates by assuming
-  the distribution is symmetric on the log scale.  Standard errors for
-  the intensities are also available from the object returned by
-  \Rfunction{qmatrix.msm}.  These are calculated by the delta method.
-  The \Rpackage{msm} package provides a general-purpose function
-  \Rfunction{deltamethod} for estimating the variance of a function of
-  a random variable $X$ given the expectation and variance of $X$. See
-  \texttt{help(deltamethod)} for further details.  Bootstrap
-  confidence intervals are also available for \Rfunction{qmatrix.msm}
-  and for most output functions; these are often more accurate, at the
-  cost of computational time.  For more about bootstrapping in
-  \Rpackage{msm}, see Section \ref{sec:boot}.
-
-\item[Transition probability matrices] The function
-  \Rfunction{pmatrix.msm} extracts the estimated transition probability
-  matrix $P(t)$ within a given time. For example, for model 1, the 10
-  year transition probabilities are given by:
-
-<<>>=
-pmatrix.msm(cav.msm, t=10)
-@
-
-  Thus, a typical person in state 1, disease-free, has a probability
-  of 0.5 of being dead ten years from now, a probability of 0.3 being
-  still disease-free, and probabilities of 0.1 of being alive with
-  mild/moderate or severe disease, respectively.
-
-  This assumes $Q$ is constant within the desired time interval.  For
-  non-homogeneous processes, where $Q$ varies with time-dependent
-  covariates but can be approximated as piecewise constant, there is
-  an equivalent function \Rfunction{pmatrix.piecewise.msm}.  Consult its
-  help page for further details.
-
-  If \Rfunarg{ci=''norm''} is specified, then a confidence interval is
-  calculated based on drawing a random sample (default size 1000) from
-  the assumed multivariate normal distribution of the maximum
-  likelihood estimates and covariance matrix, and transforming.  If
-  \Rfunarg{ci=''boot''} is specified, then a bootstrap confidence
-  interval for the transition probability matrix is calculated (see Section \ref{sec:boot}) .
-  However, both of these are computationally intensive, particularly
-  the bootstrap method, so no confidence interval is calculated by default.
-
-\item[Mean sojourn times] The function \Rfunction{sojourn.msm}
-  extracts the estimated mean sojourn times in each transient state
-  $r$, for a given set of covariate values. This is calculated as $-1
-  / \hat q_{rr}$, where $\hat q_{rr}$ is the $r$th diagonal entry of
-  the estimated transition intensity matrix.
-
-<<>>=
-sojourn.msm(cav.msm)
-@
-
-\item[Probability that each state is next] The function
-  \Rfunction{pnext.msm} extracts the matrix of probabilities $-q_{rs}
-  / q_{rr}$ that the next state after state $r$ is state $s$, for each
-  $r$ and $s$.  Together with the mean sojourn times, this gives a more
-  intuitive parameterisation of a continuous-time Markov model than
-  the raw transition intensities $q_{rs}$.  Note these are different from
-  the transition probabilities in a given time $t$ returned by
-  \Rfunction{pmatrix.msm}.
-
-<<>>=
-pnext.msm(cav.msm)
-@
-
-\item[Total length of stay] Mean sojourn times describe the
-  average period in a single stay in a state. For processes with
-  successive periods of recovery and relapse, we may want to
-  forecast the total time spent healthy or diseased, before
-        death.  The function \Rfunction{totlos.msm} estimates the
-        forecasted total length of time spent in each transient state
-        $s$ between two future time points $t_1$ and $t_2$, for a
-        given set of covariate values. This defaults to the expected
-        amount of time spent in each state between the start of the
-        process (time 0, the present time) and death or a specified
-        future time. This is obtained as
-\[
-L_s = \int_{t_1}^{t_2} P(t)_{r,s} dt
-\]
-where $r$ is the state at the start of the process, which defaults to
-1. This is calculated using numerical integration.  For model 1, each
-patient is forecasted to spend 8.8 years disease free, 2.2 years with
-mild or moderate disease and 1.8 years with severe disease.
-
-Bootstrap and asymptotic confidence intervals are available, as for
-\Rfunction{pmatrix.msm}, but are not calculated by default.
-
-<<>>=
-totlos.msm(cav.msm)
-@
-
-\item[Expected first passage times] The function \Rfunction{efpt.msm} estimates
-  the expected time until the process first enters a given state or set of states,
-  also called the ``hitting time''.  See its help page for further details.
-
-\item[Ratio of transition intensities] The function
-  \Rfunction{qratio.msm} estimates a ratio of two entries of the
-  transition intensity matrix at a given set of covariate values,
-  together with a confidence interval estimated assuming normality on
-  the log scale and using the delta method.  For example, we may want
-  to estimate the ratio of the progression rate $q_{12}$ into the
-  first state of disease to the corresponding recovery rate $q_{21}$.
-  For example in model 1, recovery is 1.8 times as likely as
-  progression.
-
-<<>>=
-qratio.msm(cav.msm, ind1=c(2,1), ind2=c(1,2))
-@
-
-\item[Hazard ratios for transition] The function \Rfunction{hazard.msm}
-  gives the estimated hazard ratios corresponding to each covariate
-  effect on the transition intensities. 95\% confidence limits are
-  computed by assuming normality of the log-effect. For example, for
-  model 2 with female sex as a covariate, the following hazard ratios
-  show more clearly that the only transition on which the effect of
-  sex is significant at the 5\% level is the 1-2 transition.
-
-<<>>=
-hazard.msm(cavsex.msm)
-@
-
-\end{description}
-
-\paragraph{Setting covariate values}
-All of these extractor functions take an argument called
-\Rfunarg{covariates}.  If this argument is omitted, for example,
-
-<<eval=FALSE>>=
-qmatrix.msm(cav.msm)
-@
-then the intensity matrix is evaluated as $Q(\bar x)$ with all
-covariates set to their mean values $\bar x$ in the data.
-Alternatively, set \Rfunarg{covariates} to 0 to return the result
-$Q(0)$ with covariates set to zero. This will usually be
-preferable for categorical covariates, where we wish to see the
-result for the baseline category.
-
-<<eval=FALSE>>=
-qmatrix.msm(cavsex.msm, covariates = 0)
-@
-Alternatively, the desired covariate values can be specified
-explicitly as a list,
-
-<<eval=FALSE>>=
-qmatrix.msm(cavsex.msm, covariates = list(sex = 1))
-@
-
-Values of categorical covariates must be quoted. For example, consider
-a covariate \texttt{smoke}, representing tobacco smoking status, with
-three levels, \texttt{NON, CURRENT, EX}, representing a non-smoker,
-current smoker or ex-smoker.
-
-\begin{Scode}
-qmatrix.msm(example.msm, covariates = list(age = 60, smoke=''CURRENT''))
-\end{Scode}
-
-\subsection{Survival plots}
-
-In studies of chronic disease, an important use of multi-state models
-is in predicting the probability of survival for patients in
-increasingly severe states of disease, for some time $t$ in the
-future.  This can be obtained directly from the transition probability
-matrix $P(t)$.
-
-The \Rfunction{plot} method for \Robject{msm} objects produces a plot
-of the expected probability of survival against time, from each
-transient state. Survival is defined as not entering the final
-absorbing state.
-
-<<fig=TRUE>>=
-plot(cav.msm, legend.pos=c(8, 1))
-@
-
-This shows that the 10-year survival probability with severe CAV is
-approximately 0.1, as opposed to 0.3 with mild CAV and 0.5 without
-CAV.  With severe CAV the survival probability diminishes very quickly
-to around 0.3 in the first five years after transplant.  The
-\Rfunarg{legend.pos} argument adjusts the position of the legend in
-case of clashes with the plot lines.  A \Rfunarg{times} argument can
-be supplied to indicate the time interval to forecast survival for.
-
-A more sophisticated analysis of these data might explore competing
-causes of death from causes related or unrelated to the disease under
-study.
-
-\subsection{Bootstrapping}
-\label{sec:boot}
-Most of \Rpackage{msm}'s output functions present confidence intervals
-based on asymptotic standard errors calculated from the Hessian, or
-transformations of these using the delta method.  The asymptotic
-standard errors are expected to be underestimates of the true standard
-errors (Cramer-Rao lower bound).  For some output functions, such as
-\Rfunction{pmatrix.msm}, and functions based on
-\Rfunction{pmatrix.msm} such as \Rfunction{totlos.msm} and
-\Rfunction{prevalence.msm}, the delta method cannot be used at all to
-obtain standard errors.  In these cases, confidence intervals can be
-calculated by drawing a random sample from the assumed multivariate
-normal distribution of the maximum likelihood estimates and covariance
-matrix, and transforming.  However, this is still based on potentially
-inaccurate asymptotic theory.  The \Rpackage{msm} package provides the
-function \Rfunction{boot.msm} to enable bootstrap refitting of
-\Rfunction{msm} models, an alternative way to estimate uncertainty.
-
-For non-hidden Markov models, a bootstrap dataset is drawn by
-resampling pairs of consecutive states from the full data, i.e.
-\emph{transitions}.  These are assumed to be independent when
-calculating the likelihood (Section \ref{sec:multi:likelihood}).  For
-hidden Markov models and models with censoring, a bootstrap dataset is
-drawn by resampling complete series from independent subjects.  The
-bootstrap datasets have the same number of transitions, or subjects,
-respectively, as the original data.
-
-For most output extractor functions provided with \Rpackage{msm}, the
-option \Rfunarg{ci=''boot''} is available, as a wrapper around
-\Rfunction{boot.msm}, to enable bootstrap confidence intervals to be calculated.
-But any user-defined output statistic can be bootstrapped, as follows.
-The function \Rfunction{boot.msm} is called with the fitted
-\Rfunction{msm} model as first argument, and an R function specifying
-the statistic to be bootstrapped as the second argument \texttt{stat}.
-The return value from \Rfunction{boot.msm} is a list of \texttt{B}
-replicates (by default, \texttt{B=1000}) of the desired statistic.  For
-example, to bootstrap the transition intensity matrix of the heart
-transplantation model \Robject{cav.msm}:
-\begin{Scode}
-  q.list <- boot.msm(cav.msm, stat=function(x){qmatrix.msm(x)$estimates})
-\end{Scode}
-Note that for \Rfunction{boot.msm} to be able to refit the original
-model that produced \Robject{cav.msm}, all objects used in the
-original model fit (for example, in this case, \Robject{twoway4.q})
-must be in the working environment.  Otherwise, \Rfunction{boot.msm}
-will give an ``object not found'' error.
-
-The user can then summarise these replicates by calculating empirical
-standard deviations or quantile-based intervals.  In this example,
-\Robject{q.list} is a list of 1000 4$\times$4 matrices.  The following
-code calculates the bootstrap standard error as the empirical standard
-deviation of the 1000 replicates, and a similar 95\% bootstrap
-confidence interval.
-\begin{Scode}
-  q.array <- array(unlist(q.list), dim=c(4,4,1000))
-  apply(q.array, c(1,2), sd)
-  apply(q.array, c(1,2), function(x)quantile(x, c(0.025, 0.975)))
-\end{Scode}
-
-Note that when bootstrapping, the refits of the model to the resampled
-datasets may occasionally fail to converge (as discussed in
-Section \ref{sec:failure}) even if the original model fit did
-converge.  In these cases, a warning is given, but
-\Rfunction{boot.msm} simply discards the failed dataset and moves on
-to the next bootstrap iteration. Unless convergence failure occurs for
-a large proportion of iterations, this should not affect the accuracy
-of the final bootstrap confidence intervals.
-
-\subsection{Convergence failure}
-\label{sec:failure}
-
-Inevitably if over-complex models are applied with insufficient data
-then the parameters of the model will not be identifiable.  This will
-result in the optimisation algorithm failing to find the maximum of
-the log-likelihood, or even failing to evaluate the likelihood. For
-example, it will commonly be inadvisable to include several covariates
-in a model simultaneously.
-
-In some circumstances, the optimisation may report convergence, but
-fail to calculate any standard errors.  In these cases, the Hessian of
-the log-likelihood at the reported solution is not positive definite.
-Thus the reported solution may be a saddle point rather than the
-maximum likelihood, or it may be close to the maximum.
-
-\begin{description}
-\item[Initial values] Make sure that a sensible set of initial values
-  have been chosen. The optimisation may only converge within a
-  limited range of `informative' initial values.  It is also sensible
-  to run the model for several different initial values to ensure that
-  the estimation has converged to a global rather than a local optimum.
-
-\item[Scaling] It is often necessary to apply a scaling factor to
-  normalise the likelihood (\Rfunarg{fnscale}), or certain individual
-  parameters \Rfunarg{(parscale)}.  This may prevent overflow or
-  underflow problems within the optimisation.  For example, if the value
-  of the -2 $\times$ log-likelihood is around 5000, then the following
-  option leads to an minimisation of the -2 $\times$ log-likelihood on
-  an approximate unit scale: \Rfunarg{control = list(fnscale = 5000)}
-
-  It is also advisable to analyse all variables, including covariates
-  and the time unit, on a roughly normalised scale. For example,
-  working in terms of a time unit of months or years instead of days,
-  when the data range over thousands of days.
-
-\item[Convergence criteria] ``False convergence'', in which
-  \Rfunction{optim()} reports convergence of the optimisation but the
-  Hessian is not positive definite, can sometimes be solved by
-  tightening the criteria (\Rfunarg{reltol}, defaults to
-  \texttt{1e-08}) for reporting convergence. For example,
-  \Rfunarg{control = list(reltol = 1e-16)}.
-
-  Alternatively consider using smaller step sizes for the numerical
-  approximation to the gradient, used in calculating the Hessian.
-  This is given by the control parameter \Rfunarg{ndeps}. For example,
-  for a model with 5 parameters, \Rfunarg{control = list(ndeps =
-    rep(1e-6, 5))}
-
-\item[Choice of algorithm] By default, \Rfunction{msm} uses a
-  Nelder-Mead algorithm which does not use analytic derivatives of the
-  objective function.   For potentially greater speed and accuracy,
-  consider using a quasi-Newton algorithm such as
-  the ``BFGS'' method of \Rfunction{optim}, which can make use of
-  analytic derivatives, for example, \Rfunarg{method = ``BFGS'', use.deriv = TRUE}.
-  Or use the \Rfunction{nlm} R function via \Rfunarg{msm(..., opt.method = "nlm" , ...)}
-  Note also the Fisher scoring method available for non-hidden Markov
-  models for panel data, via  \Rfunarg{msm(..., opt.method = "fisher", ...)},
-  is expected to be faster than the generic methods, but less robust to
-  bad initial values.
-
-\item[\Rfunction{optim} "function cannot be evaluated at initial parameters"]
-  To diagnose this problem, run \Rfunction{msm} again with
-  \Rfunarg{fixedpars=TRUE} set, to calculate the -2 log-likelihood at
-  the initial values.  This will probably be \Robject{Inf}. To show
-  the contributions of individual subjects to the overall log
-  likelihood, call \Rfunction{logLik.msm(x, by.subject=TRUE)}, where
-  \Robject{x} is the fitted model object.  If only a few subjects give
-  an infinite log-likelihood, then you can check whether their state
-  histories are particularly unusual and conflict with the model.  For
-  example, they might appear to make unusually large jumps between
-  states in short periods of time.  For models with misclassification,
-  note that the default true initial state distribution
-  \Rfunarg{initprobs} puts all individuals in true state 1 at their
-  first observation.  If someone starts in a much higher state, this
-  may result in an infinite log-likelihood, and changing
-  \Rfunarg{initprobs} would be sensible.
-
-\item[Model simplification]
-
-  If none of these numerical adjustments lead to convergence, then the
-  model is probably over-complicated.  There may not be enough
-  information in the data on a certain transition rate.  It is always
-  recommended to count all the pairs of transitions between states in
-  successive observation times, making a frequency table of previous
-  state against current state (function \Rfunction{statetable.msm}).
-  Although the data are a series of snapshots of a continuous-time
-  process, and the actual transitions take place in between the
-  observation times, this type of table may still be helpful.  If
-  there are not many observed `transitions' from state 2 to state 4,
-  for example, then the data may be insufficient to estimate $q_{24}$.
-
-  For a staged disease model (Figure \ref{fig:disease}), the number of
-  disease states should be low enough that all transition rates can be
-  estimated.  Consecutive states of disease severity should be merged
-  if necessary.  If it is realistic, consider applying constraints on
-  the intensities or the covariate effects so that the parameters are
-  equal for certain transitions, or zero for certain transitions.
-
-  Be careful to use a observation scheme and transition matrix
-  appropriate to your data (see
-  Section \ref{sec:arbitr-observ-times}). By default, \Rfunction{msm}
-  assumes that the data represent snapshots of the process, and the
-  true state is unknown between observation times.  In such
-  circumstances, it is rarely feasible to estimate an intensity matrix
-  with \emph{instantaneous} transitions allowed between every pair of
-  states.  This would be easier if the complete course of the process
-  is known \Rfunarg{(exacttimes = TRUE)} in the call to
-  \Rfunction{msm}.  Understand the difference between
-  \emph{instantaneous} and \emph{interval} transitions - although
-  individuals may be in state 1 at time $t_r$, and state 3 at time
-  $t_{r+1}$, that doesn't mean that instantaneous transitions from 1
-  to 3 should be permitted.
-
-\end{description}
-
-\subsection{Model assessment}
-\label{sec:model-assessment}
-
-\paragraph{Observed and expected prevalence}
-
-To compare the relative fit of two nested models, it is easy to
-compare their likelihoods.  However it is not always easy to determine
-how well a fitted multi-state model describes an irregularly-observed
-process.  Ideally we would like to compare observed data with fitted
-or expected data under the model.  If there were times at which all
-individuals were observed then the fit of the expected numbers in each
-state or {\em prevalences} can be assessed directly at those times.
-Otherwise, some approximations are necessary.  We could assume that an
-individual's state at an arbitrary time $t$ was the same as the state
-at their previous observation time.  This might be fairly accurate if
-observation times are close together.  This approach is taken by the
-function \Rfunction{prevalence.msm}, which constructs a table of observed
-and expected numbers and percentages of individuals in each state at a
-set of times.
-
-A set of expected counts can be produced if the process begins at a
-common time for all individuals.  Suppose at this time, each
-individual is in state 0.  Then given $n(t)$ individuals are under
-observation at time $t$, the expected number of individuals in state
-$r$ at time $t$ is $n(t) P(t)_{0,r}$.   If the covariates on which
-$P(t)$ depends vary between individuals, then this can be
-averaged over the covariates observed in the data.
-
-For example, we calculate the observed and expected numbers and
-percentages at two-yearly intervals up to 20 years after transplant,
-for the heart transplant model \Rfunction{cav.msm}.  The number of
-individuals still alive and under observation decreases from 622 to
-251 at year 20.  The observed and expected percentages are plotted
-against time.
-
-<<>>=
-options(digits=3)
-prevalence.msm(cav.msm, times=seq(0,20,2))
-@
-<<fig=TRUE>>=
-plot.prevalence.msm(cav.msm, mintime=0, maxtime=20)
-@
-
-Comparing the observed and expected percentages in each state,
-we see that the predicted number of individuals who die (State 4) is
-under-estimated by the model from about year 8 onwards.  Similarly the
-number of individuals sill alive and free of CAV (State 1) is
-over-estimated by the model from about year 8 onwards.
-
-Such discrepancies could have many causes.  One possibility is that
-the transition rates vary with the time since the beginning of the
-process, the age of the patient, or some other omitted covariate, so
-that the Markov model is {\em non-homogeneous}.  This could be
-accounted for by modelling the intensity as a function of age, for
-example, such as a piecewise-constant function.  The \Rfunarg{pci}
-argument to \Rfunction{msm} can be used to automatically construct
-models with transition intensities which are piecewise-constant in
-time.
-
-In this example, the hazard of death may increase with age, so that
-the model underestimates the number of deaths when forecasting far
-into the future.  Another cause of poor model fit may sometimes be the
-failure of the Markov assumption.  That is, the transition intensities
-may depend on the time spent in the current state (a semi-Markov
-process) or other characteristics of the process history.  Accounting
-for the process history is difficult as the process is only observed
-through a series of snapshots.  For a multi-state model with one-way
-progression through states, and frequent observations, we may be able
-to estimate the time spent in each state by each individual.
-
-However, if it is known that individuals who died would not have been
-followed up after a certain time, had they survived to that time, then
-they should not be included in the observed prevalence of the death
-state after that time.  This can be accounted for by passing a vector
-of maximum potential follow-up times, one for each individual in the
-same order as the original data, in the \Rfunarg{censtime} argument to
-\Rfunction{prevalence.msm}.  Ignoring the potential follow-up times is
-likely to have resulted in overestimates of the number of deaths
-at later times in the CAV example, though these times are not
-available in the data supplied with \Rpackage{msm}.
-
-\paragraph{Pearson-type goodness-of-fit test}
-\label{sec:pearson}
-
-Suppose that the true transition times are unknown, and data consist
-of observations of the process at arbitrary times which differ between
-individuals (panel data).  Assessing goodness of fit by prevalence
-counts then involves estimating the observed prevalence at a series of
-points by some form of interpolation. This is only advisable if
-observation times are close together.  An alternative method of
-assessing goodness-of-fit is to construct tables of observed and
-expected numbers of transitions, as described by Aguirre-Hernandez and
-Farewell \cite{ahf}.  This leads to a formal test of goodness-of-fit,
-analogous to the classical Pearson $\chi^2$ test for contingency
-tables. The tables are constructed as follows.  Each pair of
-successive observations in the data (\emph{transition}) is classified
-by
-\begin{itemize}
-\item the starting state $r$ and finishing state $s$,
-\item time between the start of the process and the first of the pair
-  of observations (indexed by $h$),
-\item time interval between the observations (indexed by $l_h$, within
-  categories $h$),
-\item (if there are fitted covariates) the impact of covariates, as
-  summarised by $q_{rr}$ (indexed by $c$),
-\item any other grouping of interest for diagnosing lack of fit
-  (indexed by $g$).
-\end{itemize}
-Groupings of continuous quantities are normally defined by quantiles,
-so that there are a similar number of observed transitions in each
-(one-dimensional) category.  The observed and expected numbers of
-transitions in each group are then defined by
-\[
-o_{hl_h rscg} = \sum I(S(t_{i,j+1}) = s, S(t_{ij}) = r)
-\]
-\[
-e_{hl_h rscg} = \sum P(S(t_{i,j+1}) = s | S(t_{ij}) = r)
-\]
-where $I(A)$ is the indicator function for an event $A$ and the
-summation is over the set of transitions in the category defined by
-$h,l_h,c,g$, over all individuals $i$. The Pearson-type test statistic
-is then
-\[
-T = \sum_{hl_h rscg} \frac{(o_{hl_h rscg} - e_{hl_h rscg})^2}{e_{hl_h rscg}}
-\]
-The classical Pearson test statistic is distributed as $\chi^2_{n-p}$,
-where $n$ is the number of independent cells in the table and $p$ is
-the number of estimated parameters $p$. But the null distribution of
-$T$ is not exactly $\chi^2$, since the time intervals are
-non-identical, therefore the observed transitions are realizations
-from a set of independent but non-identical multinomial distributions.
-Titman \cite{titman:asympnull} showed that the null distribution
-of $T$ is asymptotically approximated by a weighted sum of $\chi^2_1$
-random variables.   Aguirre-Hernandez and Farewell \cite{ahf} also showed that $\chi^2_{n-p}$
-is a good approximation if there are no fitted covariates. For models
-with covariates, the null mean of $T$ is higher than $n - p$, but
-lower than $n$.  Therefore, upper and lower bounds for the true
-$p$-value of the statistic can be obtained from the $\chi^2_{n-p}$ and
-$\chi^2_n$ distributions.  Aguirre-Hernandez and Farewell \cite{ahf}
-also described a bootstrap procedure for obtaining an accurate $p$-value.
-
-Titman and Sharples \cite{titman:sharples} described modifications to
-the test to correct for the biases introduced where in addition to the
-panel-data observation scheme:
-\begin{itemize}
-\item Times of death are known exactly.  In this case, transitions
-  ending in death are classified according to the next scheduled
-  observation time after the death, which is estimated by multiple
-  imputation from a Kaplan-Meier estimate of the distribution of time
-  intervals between observations.
-
-\item An individual's final observation is censored, so that they are
-  only known to be alive at that point.
-
-\item States are misclassified.
-\end{itemize}
-
-The \Rpackage{msm} package provides the function
-\Rfunction{pearson.msm} to perform the Pearson-type test.  By default,
-three groups are used for each of $h$, $l_h$ and $c$.  Often the
-number of groups will need to be reduced in cases where the resulting
-contingency tables are sparse (thus there are several low expected
-counts and the variance of $T$ is inflated).
-
-The test is now performed on the model \Robject{cav.msm} for the heart
-transplant dataset (a version of which was also analysed by
-Titman and Sharples \cite{titman:sharples}). The default three interval groups are used,
-and two groups of the time since the start of the process. The
-\Rfunarg{transitions} argument groups the transitions from state 3 to
-each of states 1, 2 and 3 (the 9th, 10th and 11th transitions)
-together in the table, since these transitions are infrequent.
-
-<<>>=
-options(digits=2)
-pearson.msm(cav.msm, timegroups=2, transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
-@
-
-The first two tables in the output show the contingency tables of
-observed and expected numbers of transitions.  Note that the observed
-number of transitions in certain categories is not a whole number,
-since these are averaged over multiple imputations of the next
-scheduled observation time following deaths.  The column \texttt{Time}
-is the group defined by time since the start of the process, and the
-column \texttt{Interval} is the group defined by intervals between
-observations.  The columns indicate the allowed transitions, or pairs
-of states which can be observed in successive observations.
-
-The third table presents the ``deviance'', the value of
-$\frac{(o_{hl_h rscg} - e_{hl_h rscg})^2}{e_{hl_h rscg}}$ for each
-cell, multipled by the sign of $o_{hl_h rscg} - e_{hl_h rscg}$ to
-indicate whether there were more or fewer transitions than expected in
-each cell.  These can indicate areas of bad fit.  For example,
-systematic changes in deviance by time or time interval between
-observations can indicate that a model with time-varying transition
-intensities is more suitable.  Changes in deviance by covariate impact
-may indicate heterogeneity between individuals which is unexplained
-by the fitted covariates.  Changes in deviance with the length of the
-interval between observations may also indicate failure of the Markov
-assumption, and that a semi-Markov model (in which intensities depend
-on the time spent in the current state) may fit better.
-
-In this example, the test statistic is 100.  \Robject{p.upper} is an
-upper bound for the $p$-value of the statistic based on an asymptotic
-$\chi^2_{42}$ distribution, therefore the model does not fit well.  It
-is not clear from the table of deviances which aspects of the fit are
-most influental to the test statistic.  However, the two-way Markov
-model itself is not biologically plausible, as discussed in
-Section \ref{sec:fitting:hmm:misc}.
-
-For non-hidden Markov models for panel data, \Rfunction{pearson.msm}
-also presents the accurate analytic p-value of
-Titman \cite{titman:asympnull}.  For all models,
-\Rfunction{pearson.msm} provides an option for parametric
-bootstrapping to obtain an accurate p-value.
-
-\subsection{Fitting misclassification models with \Rpackage{msm}}
-\label{sec:fitting:hmm:misc}
-
-In fact, in the heart transplant example from
-section \ref{sec:datain}, it is not medically realistic for patients
-to recover from a diseased state to a healthy state.  Progression of
-coronary artery vasculopathy is thought to be an irreversible process.
-The angiography scan for CAV is actually subject to error, which leads
-to some false measurements of CAV states and apparent recoveries.
-Thus we account for misclassification by fitting a \emph{hidden Markov
-model} using \Rpackage{msm}.  Firstly we replace the two-way multi-state
-model by a one-way model with transition intensity matrix
-\[
-Q = \left(
-  \begin{array}{llll}
-    -(q_{12} + q_{14}) & q_{12} &  0     & q_{14}\\
-      0    & -(q_{23}+q_{24}) & q_{23} & q_{24}\\
-      0    &   0    & -q_{34} & q_{34}\\
-      0    &   0    &   0    &   0   \\
-  \end{array}
-\right )
-\]
-We also assume that true state 1 (CAV-free) can be classified as state
-1 or 2, state 2 (mild/moderate CAV) can be classified as state 1, 2 or
-3, while state 3 (severe CAV) can be classified as state 2 or 3.
-Recall that state 4 represents death.  Thus our matrix of misclassification
-probabilities is
-\[
-E = \left(
-  \begin{array}{llll}
-    1 - e_{12} & e_{12} & 0 & 0 \\
-    e_{21} & 1 - e_{21} - e_{23} & e_{23} & 0 \\
-    0 & e_{32} & 1 - e_{32} & 0 \\
-    0 & 0 & 0 & 0\\
-  \end{array}
-  \right)
-\]
-with underlying states as rows, and observed states as columns.
-
-To model observed states with misclassification, we define an matrix
-\Rfunarg{ematrix} indicating the states that can be misclassified.
-Rows of this matrix correspond to true states, columns to observed
-states.  It should contains zeroes in the positions where
-misclassification is not permitted.  Non-zero entries are initial
-values for the corresponding misclassification probabilities. We then
-call \Rfunction{msm} as before, but with this matrix as the
-\Rfunarg{ematrix} argument.  Initial values of 0.1 are assumed for
-each of the four misclassification probabilities $e_{12}, e_{21},
-e_{23}, e_{32}$.  Zeroes are given where the elements of $E$ are zero.
-The diagonal elements supplied in \Rfunarg{ematrix} are ignored, as
-rows must sum to one. The matrix \Rfunarg{qmatrix}, specifying
-permitted transition intensities and their initial values, also
-changes to correspond to the new $Q$ representing the progression-only
-model for the underlying states.
-
-The true state for every patient at the date of transplant is known to
-be ``CAV-free'', not misclassified.  To indicate this we use the
-argument \Rfunarg{obstrue} to \Rfunction{msm}.  This is set to be a
-variable in the dataset, \Rfunarg{firstobs}, indicating where the
-observed state equals the true state.  This takes the value of 1 at
-the patient's first observation, at the transplant date, and 0
-elsewhere.
-
-We use an alternative quasi-Newton optimisation algorithm
-\Rfunarg{(method="BFGS")} which can often be faster or more robust
-than the default Nelder-Mead simplex-based algorithm.  An optional
-argument \Rfunarg{initprobs} could also have been given here,
-representing the vector of the probabilities of occupying each true
-state at the initial observation
-(equation \ref{eq:multi:hidden:matprod}).  This can also be a matrix
-with number of rows equal to the number of subjects, if these
-probabilities are subject-dependent and known. If not given, all
-individuals are assumed to be in true state 1 at their initial
-observation.  If \Rfunarg{est.initprobs=TRUE} is specified, then these
-probabilites are estimated as part of the model fit, using
-a vector \Rfunarg{initprobs} as initial values.  Covariate effects on these
-probabilities can also be estimated using a multinomial logistic
-regression model, if an \Rfunarg{initcovariates} argument is
-specified.  See \Rfunction{help(msm)} for further details.
-
-\paragraph{Model 5: multi-state model with misclassification}
-
-<<>>=
-oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
-                     c(0, 0, 0.202, 0.081),
-                     c(0, 0, 0, 0.126),
-                     c(0, 0, 0, 0))
-ematrix <- rbind(c(0, 0.1, 0, 0),
-                 c(0.1, 0, 0.1, 0),
-                 c(0, 0.1, 0, 0),
-                 c(0, 0, 0, 0))
-cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                   qmatrix = oneway4.q, ematrix = ematrix, death = 4,
-                   obstrue = firstobs, method="BFGS")
-cavmisc.msm
-@
-
-Thus there is an estimated probability of about 0.03 that
-mild/moderate CAV will be diagnosed erroneously, but a rather higher
-probability of 0.17 that underlying mild/moderate CAV will be
-diagnosed as CAV-free.  Between the two CAV states, the mild state
-will be misdiagnosed as severe with a probability of 0.06, and the
-severe state will be misdiagnosed as mild with a probability of 0.12.
-
-The model also estimates the progression rates through underlying
-states.  An average of 8 years is spent disease-free, an average of
-about 3 years is spent with mild/moderate disease, and periods of severe
-disease also last about 3 years on average before death.
-
-\subsection{Effects of covariates on misclassification rates}
-
-We can investigate how the probabilities of misclassification depend
-on covariates in a similar way to the transition intensities, using a
-\Rfunarg{misccovariates} argument to \Rfunction{msm}. For example, we
-now include female sex as a covariate for the misclassification
-probabilities.  The linear effects on the log odds of each misclassified
-state relative to the true state are initialised to zero by default (but
-this can be changed with the \Rfunarg{misccovinits} argument).
-
-\paragraph{Model 6: misclassification model with misclassification probabilities modelled on sex}
-
-<<>>=
-cavmiscsex.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                          qmatrix = oneway4.q, ematrix = ematrix, death = 4,
-                          misccovariates = ~sex, obstrue=firstobs, method="BFGS")
-@
-<<>>=
-cavmiscsex.msm
-@
-The large confidence interval for the 1/2 misclassification suggests
-that women are no more or less likely than men to have a false
-positive angiography scan. On the other hand, women have slightly more
-false negatives.
-
-\subsection{Extractor functions}
-
-As well as the functions described in section \ref{sec:extractor} for
-extracting useful information from fitted models, there are a number
-of extractor functions specific to models with misclassification.
-
-\begin{description}
-\item[Misclassification matrix] The function \Rfunction{ematrix.msm}
-gives the estimated misclassification probability matrix at the given
-covariate values.  For illustration, the fitted misclassification
-probabilities for men and women in model 6 are given by
-
-<<>>=
-ematrix.msm(cavmiscsex.msm, covariates=list(sex=0))
-ematrix.msm(cavmiscsex.msm, covariates=list(sex=1))
-@
-The confidence intervals for the estimates for women are higher, since
-there are only 87 women in this set of 622 patients.
-
-\item[Odds ratios for misclassification] The function
-  \Rfunction{odds.msm} would give the estimated odds ratios corresponding to
-  each covariate effect on the misclassification probabilities.
-
-\begin{Scode}
-odds.msm(cavmiscsex.msm)
-\end{Scode}
-
-\item[Observed and expected prevalences]
-
-  The function \Rfunction{prevalence.msm} is intended to assess the
-  goodness of fit of the hidden Markov model for the \emph{observed}
-  states to the data.  Tables of observed prevalences of observed
-  states are calculated as described in
-  section \ref{sec:model-assessment}, by assuming that observed states
-  are retained between observation times.
-
-  The expected numbers of individuals in each observed state are
-  calculated similarly. Suppose the process begins at a common time
-  for all individuals, and at this time, the probability of occupying
-  \emph{true} state $r$ is $f_r$.  Then given $n(t)$ individuals under
-  observation at time $t$, the expected number of individuals in true
-  state $r$ at time $t$ is the $r$th element of the vector $n(t) f
-  P(t)$. Thus the expected number of individuals in \emph{observed}
-  state $r$ is the $r$th element of the vector $n(t) f P(t) E$, where
-  $E$ is the misclassification probability matrix.
-
-  The expected prevalences (not shown) for this example are similar to
-  those forecasted by the model without misclassification, with
-  underestimates of the rates of death from 8 years onwards.  To
-  improve this model's long-term prediction ability, it is probably
-  necessary to account for the natural increase in the hazard of death
-  from any cause as people become older.
-
-\item[Goodness-of-fit test] The Pearson-type goodness-of-fit test is
-  performed, as in Section \ref{sec:pearson}.  The table of deviances
-  indicates that there are more 1-3 and 1-4 transitions than expected
-  in short intervals, and fewer in long intervals.  This may indicate
-  some time-dependence in the transition rates. Indeed, Titman
-  \cite{titman:phd} found that a model with piecewise-constant
-  transition intensities gave a greatly improved fit to these data.
-<<>>=
-pearson.msm(cavmisc.msm, timegroups=2, transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
-@
-
-\end{description}
-
-\subsection{Recreating the path through underlying states}
-
-In speech recognition and signal processing, {\em decoding} is the
-procedure of determining the underlying states that are most likely to
-have given rise to the observations. The most common method of
-reconstructing the most likely state path is the {\em Viterbi}
-algorithm. Originally proposed by Viterbi \cite{viterbi}, it is also
-described by Durbin \etal \cite{biolog:seq} and Macdonald and
-Zucchini \cite{macdonald:zucchini} for discrete-time hidden Markov
-chains.  For continuous-time models it proceeds as follows. Suppose
-that a hidden Markov model has been fitted and a Markov transition
-matrix $P(t)$ and misclassification matrix $E$ are known.  Let
-$v_k(t_i)$ be the probability of the most probable path ending in state
-$k$ at time $t_i$.
-\begin{enumerate}
-\item Estimate $v_k(t_1)$ using known or estimated initial-state
-  occupation probabilities.
-\item For $i = 1 \ldots N$, calculate $v_l(t_i) = e_{l,O_{t_i}} \max_k
-  v_k(t_{i-1}) P_{kl}(t_{i} - t_{i-1})$. Let $K_i(l)$ be the
-  maximising value of $k$.
-\item At the final time point $t_N$, the most likely underlying
-  state $S^*_N$ is the value of $k$ which maximises $v_k(t_N)$.
-\item Retrace back through the time points, setting $S^*_{i-1} =
-  K_i(S^*_i)$.
-\end{enumerate}
-
-The computations should be done in log space to prevent underflow.
-The \Rpackage{msm} package provides the function \Rfunction{viterbi.msm} to
-implement this method.  For example, the following is an extract from
-a result of calling \Rfunction{viterbi.msm} to determine the most likely
-underlying states for all patients. The results for patient 100103 are
-shown, who appeared to `recover' to a less severe state of disease
-while in state 3.  We assume this is not biologically possible for the
-true states, so we expect that either the observation of state 3 at
-time 4.98 was an erroneous observation of state 2, or their apparent
-state 2 at time 5.94 was actually state 3.  According to the expected
-path constructed using the Viterbi algorithm, it is the observation at
-time 5.94 which is most probably misclassified.
-
-<<>>=
-vit <- viterbi.msm(cavmisc.msm)
-vit[vit$subject==100103,]
-@
-
-\subsection{Fitting general hidden Markov models with \Rpackage{msm}}
-\label{sec:fitting:hmm:general}
-
-The \Rpackage{msm} package provides a framework for fitting
-continuous-time hidden Markov models with general, continuous
-outcomes.  As before, we use the \Rfunction{msm} function itself.
-
-\paragraph{Specifying the hidden Markov model}
-
-A hidden Markov model consists of two related components:
-\begin{itemize}
-\item the model for the evolution of the underlying Markov chain,
-\item the set of models for the observed data conditionally on each
-  underlying state.
-\end{itemize}
-
-The model for the transitions between underlying states is specified
-as before, by supplying a \Rfunarg{qmatrix}.  The model for the
-outcomes is specified using the argument \Rfunarg{hmodel} to
-\Rfunction{msm}.  This is a list, with one element for each underlying
-state, in order.  Each element of the list should be an object
-returned by a hidden Markov model \emph{constructor function}.  The
-HMM constructor functions provided with \Rpackage{msm} are listed in
-Table \ref{tab:hmm:dists}.  There is a separate constructor function
-for each class of outcome distribution, such as uniform, normal or
-gamma.
-
-Consider a three-state hidden Markov model, with a transition
-intensity matrix of
-\[
-Q = \left(
-  \begin{array}{llll}
-    -q_{12} & q_{12} &  0  \\
-    0  & -q_{23} & q_{23}\\
-    0    &   0    &   0   \\
-  \end{array}
-\right )
-\]
-Suppose the outcome distribution for state 1 is Normal$(\mu_1,
-\sigma^2_1)$, the distribution for state 2 is Normal$(\mu_2,
-\sigma^2_2)$, and state 3 is exactly observed. Observations of state 3
-are given a label of -9 in the data.  Here our \Rfunarg{hmodel}
-argument should be a list of objects returned by \Rfunction{hmmNorm} and
-\Rfunction{hmmIdent} constructor functions.
-
-We must specify initial values for the parameters as the arguments to
-the constructor functions.  For example, we take initial values of
-$\mu_1 = 90, \sigma_1 = 8, \mu_2 = 70, \sigma_2 = 8$.  Initial values
-for $q_{12}$ and $q_{23}$ are 0.25 and 0.2. Finally suppose the
-observed data are in a variable called \Robject{y}, the measurement
-times are in \Robject{time}, and subject identifiers are in
-\Robject{ptnum}.  The call to \Rfunction{msm} to estimate the
-parameters of this hidden Markov model would then be
-\begin{Scode}
-  msm( y ~ time, subject=ptnum, data = example.df,
-  qmatrix = rbind( c(0, 0.25, 0), c(0, 0, 0.2), c(0, 0, 0)),
-  hmodel = list (hmmNorm(mean=90, sd=8), hmmNorm(mean=70, sd=8),
-                 hmmIdent(-9)) )
-\end{Scode}
-
-\begin{table}[htbp]
-\scriptsize
-  \centering
-  \begin{tabular}{lp{0.8in}p{0.6in}p{0.6in}p{0.7in}l}
-    \hline
-    Function & Distribution & Parameters &  &  Location (link) & Density for an observation $x$  \\
-    \hline
-    \Rfunction{hmmCat}   & Categorical   & \Rfunarg{prob, basecat}  & $p,c_0$ & $p$ (logit) & $p_x$, $x = 1, \ldots, n$  \\
-    \Rfunction{hmmIdent}   & Identity  & \Rfunarg{x}  & $x_0$  & & $I_{x = x_0}$  \\
-    \Rfunction{hmmUnif}   & Uniform  & \Rfunarg{lower, upper}  & $l,u$  & & $1 / (u - l)$, $u \leq x \leq l$  \\
-    \Rfunction{hmmNorm}   & Normal  & \Rfunarg{mean, sd}  & $\mu,\sigma$  & $\mu$ (identity) & $\phi(x, \mu, \sigma) = \frac{1}{\sqrt{2 \pi \sigma^2}} \exp(-(x - \mu)^2/(2 \sigma^2) )$  \\
-    \Rfunction{hmmLNorm}   & Log-normal  & \Rfunarg{meanlog, sdlog}  & $\mu,\sigma$  & $\mu$ (identity) & $\frac{1}{x \sqrt{2 \pi \sigma^2}} \exp(-(\log x - \mu)^2 / (2 \sigma^2))$   \\
-    \Rfunction{hmmExp}   & Exponential  & \Rfunarg{rate}  & $\lambda$  & $\lambda$ (log) & $\lambda e^{- \lambda x}$, $x > 0$  \\
-    \Rfunction{hmmGamma}   & Gamma  & \Rfunarg{shape, rate}  & $n,\lambda$  &  $\lambda$ (log) & $\frac{\lambda^n}{\Gamma(n)}x^{n-1} \exp(-\lambda x)$, $x > 0, n > 0, \lambda > 0$   \\
-    \Rfunction{hmmWeibull}   & Weibull  & \Rfunarg{shape, scale}  & $a, b$  & $b$ (log) & $\frac{a}{b} (\frac{x}{b})^{a-1} \exp{(-(\frac{x}{b})^a)}$, $x > 0$  \\
-    \Rfunction{hmmPois}   & Poisson  & \Rfunarg{rate}  & $\lambda$  & $\lambda$ (log) & $\lambda^x \exp(-\lambda)/x!$, $x = 0, 1, 2, \ldots$  \\
-    \Rfunction{hmmBinom}   & Binomial  & \Rfunarg{size, prob}  & $n,p$  & $p$ (logit) & ${n \choose x} p^x (1-p)^{n-x}$  \\
-    \Rfunction{hmmNBinom}   & Negative binomial  & \Rfunarg{disp, prob}  & $n,p$  & $p$ (logit) & $\Gamma(x+n)/(\Gamma(n)x!) p^n (1-p)^x$ \\
-    \Rfunction{hmmBeta}   & Beta  & \Rfunarg{shape1,shape2}  & $a,b$  & & $ \Gamma(a+b) / (\Gamma(a)\Gamma(b))x^{a-1}(1-x)^{b-1}$ \\
-    \Rfunction{hmmT}   & Student $t$  & \Rfunarg{mean, scale, df}  & $\mu,\sigma,k$  &  $\mu$ (identity) & $\frac{\Gamma\left((k+1)/2\right)}{\Gamma(k/2)}{\sqrt{\frac{1}{k\pi\sigma^2}}}\left\{1 + \frac{1}{k\sigma^2}(x - \mu)^{2} \right\}^{-(k + 1)/2}$ \\
-    \Rfunction{hmmTNorm}   & Truncated normal  & \Rfunarg{mean, sd, lower, upper}  & $\mu,\sigma,l,u$  & $\mu$ (identity) &
-\parbox[t]{2in}{$\phi(x, \mu, \sigma) / \\ (\Phi(u, \mu, \sigma) - \Phi(l, \mu, \sigma))$, \\ where $\Phi(x,\mu,\sigma) = \int_{-\infty}^x \phi(u,\mu,\sigma) du$}
- \\
-    \Rfunction{hmmMETNorm}   & Normal with truncation and measurement error  & \Rfunarg{mean, sd, lower, upper, sderr, meanerr}  & \parbox[t]{1in} {$\mu_0,\sigma_0,l,u$, \\ $\sigma_\epsilon,\mu_\epsilon$}  &  $\mu_\epsilon$ (identity) &
-\parbox[t]{2in}{$( \Phi(u, \mu_2, \sigma_3) - \Phi(l, \mu_2, \sigma_3)) / $ \\
-$(\Phi(u, \mu_0, \sigma_0) - \Phi(l, \mu_0, \sigma_0)) $ \\
-$\times \phi(x, \mu_0 + \mu_\epsilon, \sigma_2)$, \\
-$\sigma_2^2 = \sigma_0^2 + \sigma_\epsilon^2$, \\
-$\sigma_3 = \sigma_0 \sigma_\epsilon / \sigma_2$, \\
-$\mu_2 = (x - \mu_\epsilon) \sigma_0^2 + \mu_0 \sigma_\epsilon^2$}
- \\
-    \Rfunction{hmmMEUnif}   & Uniform with measurement error  & \Rfunarg{lower, upper, sderr, meanerr}  & $l,u,\mu_\epsilon,\sigma_\epsilon$  & $\mu_\epsilon$ (identity) &
-    \parbox[t]{2in}{$(\Phi(x, \mu_\epsilon+l, \sigma_\epsilon) - \Phi(x, \mu_\epsilon+u, \sigma_\epsilon)) / \\ (u - l)$} \\
-    \hline
-  \end{tabular}
-  \caption{Hidden Markov model distributions in \Rpackage{msm}.}
-  \label{tab:hmm:dists}
-\end{table}
-
-\paragraph{Covariates on hidden Markov model parameters}
-
-Most of the outcome distributions can be parameterised by covariates,
-using a link-transformed linear model.  For example, an observation
-$y_{ij}$ may have distribution $f_1$ conditionally on underlying state
-1. The link-transformed parameter $\theta_1$ is a linear function of
-the covariate vector $x_{ij}$ at the same observation time.
-\begin{eqnarray*}
-  \label{eq:hmm:covs}
-  y_{ij} | S_{ij} &  \sim &   f_1 (y | \theta_1, \gamma_1)\\
-  g(\theta_1) & = & \alpha + \beta^T x_{ij}
-\end{eqnarray*}
-Specifically, parameters named as the ``Location'' parameter in
-Table \ref{tab:hmm:dists} can be modelled in terms of covariates, with
-the given link function.
-
-The \Rfunarg{hcovariates} argument to \Rfunction{msm} specifies the
-model for covariates on the hidden Markov outcome distributions.  This
-is a list of the same length as the number of underlying states, and
-the same length as the \Rfunarg{hmodel} list.  Each element of the
-list is a formula, in standard R linear model notation, defining the
-covariates on the distribution for the corresponding state.  If there
-are no covariates for a certain hidden state, then insert a \texttt{NULL} in
-the corresponding place in the list.  For example, in the three-state
-normal-outcome example above, suppose that the normal means on states
-1 and 2 are parameterised by a single covariate $x$.
-\[
-\mu_1 = \alpha_1 + \beta_1 x_{ij}, \qquad \mu_2 = \alpha_2 + \beta_2 x_{ij}.
-\]
-The equivalent call to \Rfunction{msm} would be
-\begin{Scode}
-  msm( state ~ time, subject=ptnum, data = example.df,
-  qmatrix = rbind( c(0, 0.25, 0), c(0, 0, 0.2), c(0, 0, 0)),
-  hmodel = list (hmmNorm(mean=90, sd=8), hmmNorm(mean=70, sd=8),
-                 hmmIdent(-9)),
-  hcovariates = list ( ~ x, ~ x, NULL)
- ).
-\end{Scode}
-
-\paragraph{Constraints on hidden Markov model parameters}
-
-Sometimes it is realistic that parameters are shared between some of
-the state-specific outcome distributions.  For example, the
-Normally-distributed outcome in the previous example could have a
-common variance $\sigma^2_1 = \sigma^2_2 = \sigma^2$ between states 1
-and 2, but differing means.  It would also be realistic for any
-covariates on the mean to have a common effect $\beta_1 = \beta_2 =
-\beta$ on the state 1 and 2 outcome distributions.
-
-The argument \Rfunarg{hconstraint} to \Rfunction{msm} specifies which
-hidden Markov model parameters are constrained to be equal.  This is a
-named list.  Each element is a vector of constraints on the named
-hidden Markov model parameter. The vector has length equal to the
-number of times that class of parameter appears in the whole model.
-As for the other constraint arguments such as \Rfunarg{qconstraint},
-identical values of this vector indicate parameters constrained to be
-equal.
-
-For example, consider the three-state hidden Markov model described
-above, with normally-distributed outcomes for states 1 and 2.
-To constrain the outcome variance to be equal for states 1 and 2,
-and to also constrain the effect of \Robject{x} on the outcome mean
-to be equal for states 1 and 2, specify
-\begin{Scode}
-hconstraint = list(sd = c(1,1), x=c(1,1))
-\end{Scode}
-
-Parameters of the
-outcome distributions may also be constrained within specific ranges.
-If chosen carefully, this may improve identifiability of hidden Markov states.
-For example to constrain the mean for state 1 to be between 80 and 110,
-and the mean for state 2 to be between 50 and 80, specify
-\begin{Scode}
-hranges = list(mean=list(lower=c(80,50), upper=c(110,80)))
-\end{Scode}
-Maximum likelihood estimation is then performed on the appropriate log
-or logit-transformed scale so that these constraints are
-satisfied. See the \Rfunction{msm} help page for further details.
-Note that initial values should be strictly within the ranges, and not
-on the range boundary.
-
-\paragraph{FEV$_1$ after lung transplants}
-
-Now we give an example of fitting a hidden Markov model to a real
-dataset.  The data on FEV$_1$ measurements from lung transplant
-recipients, described in \ref{sec:hmm:example:fev}, are provided with
-the \Rpackage{msm} package in a dataset called \Robject{fev}.
-We fit models Models 1 and 2, each with three states and common $Q$
-matrix.
-<<>>=
-three.q <- rbind(c(0, exp(-6), exp(-9)), c(0, 0, exp(-6)), c(0, 0, 0))
-@
-
-The simpler Model 1 is specified as follows. Under this model the
-FEV$_1$ outcome is Normal with unknown mean and variance, and the mean
-and variance are different between BOS state 1 and state 2.
-\Rfunarg{hcovariates} specifies that the mean of the Normal outcome
-depends linearly on acute events.  Specifically, this covariate is an
-indicator for the occurrence of an acute event within 14 days of the
-observation, denoted \texttt{acute} in the data.  As an initial guess,
-we suppose the mean FEV$_1$ is 100\% baseline in state 1, and 54\%
-baseline in state 2, with corresponding standard deviations 16 and 18,
-and FEV$_1$ observations coinciding with acute events are on average
-8\% baseline lower. \Rfunarg{hconstraint} specifies that the acute
-event effect is equal between state 1 and state 2.
-
-Days of death are coded as 999 in the \texttt{fev} outcome variable.
-
-<<>>=
-hmodel1 <- list(hmmNorm(mean=100, sd=16), hmmNorm(mean=54, sd=18),
-                hmmIdent(999))
-
-fev1.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q,
-                death=3, hmodel=hmodel1,
-                hcovariates=list(~acute, ~acute, NULL),
-                hcovinits = list(-8, -8, NULL),
-                hconstraint = list(acute = c(1,1)), method="BFGS")
-
-fev1.msm
-
-sojourn.msm(fev1.msm)
-@
-
-Printing the \Rclass{msm} object \Rfunarg{fev1.msm} shows estimates
-and confidence intervals for the transition intensity matrix and the
-hidden Markov model parameters.  The estimated within-state means of
-FEV$_1$ are around 98\% and 52\% baseline respectively.  From the
-estimated transition intensities, individuals spend around 1421 days
-(3.9 years) before getting BOS, after which they live for an average
-of 1248 days (3.4 years).  FEV$_1$ is lower by an average of 8\%
-baseline within 14 days of acute events.
-
-Model 2, where the outcome distribution is a more complex two-level
-model, is specified as follows.  We use the distribution defined by
-equations \ref{eq:fev:level1}--\ref{eq:fev:level2}. The
-\Rfunction{hmmMETNorm} constructor defines the truncated normal
-outcome with an additional normal measurement error.  The explicit
-probability density for this distribution is given in
-Table \ref{tab:hmm:dists}.
-
-Our initial values are 90 and 54 for the means of the
-within-state distribution of \emph{underlying} FEV$_1$, and 16 and 18
-for the standard errors.  This time, underlying FEV$_1$ is truncated
-normal. The truncation limits \Rfunarg{lower} and \Rfunarg{upper} are
-not estimated.  We take an initial measurement error standard
-deviation of \Rfunarg{sderr=8}. The extra shift \Rfunarg{meanerr} in
-the measurement error model is fixed to zero and not estimated.
-
-The \Rfunarg{hconstraint} specifies that the measurement error
-variance $\sigma^2_\epsilon$ is equal between responses in states 1
-and 2, as is the effect of short-term acute events on the FEV$_1$
-response.
-
-The convergence of maximum likelihood estimation in this example is
-particularly sensitive to the optimisation method and options, initial values,
-the unit of the time variable and whether covariates are centered,
-probably because the likelihood surface is irregular near to the
-true maximum.
-
-\begin{Scode}
-hmodel2 <- list(hmmMETNorm(mean=90, sd=16, sderr=8,
-                           lower=80, upper=Inf, meanerr=0),
-                hmmMETNorm(mean=54, sd=18, sderr=8,
-                           lower=0, upper=80, meanerr=0),
-                hmmIdent(999))
-
-fev2.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q,
-                death=3, hmodel=hmodel2,
-                hcovariates=list(~acute, ~acute, NULL),
-                hcovinits = list(-8, -8, NULL),
-                hconstraint = list(sderr = c(1,1), acute = c(1,1)),
-                control=list(maxit=10000), center=TRUE)
-
-fev2.msm
-
-sojourn.msm(fev2.msm)
-\end{Scode}
-
-Under this model the standard deviation of FEV$_1$ measurements caused
-by measurement error (more realistically, natural short-term
-fluctuation) is around 9\% baseline.  The estimated effect of acute
-events on FEV$_1$ and sojourn times in the BOS-free state and in BOS
-before death are similar to Model 1.
-
-The following code will create a plot that illustrates a trajectory of declining FEV$_1$ from
-the first lung transplant recipient in this dataset.  The Viterbi algorithm is used to locate the
-most likely point at which this individual moved from BOS state 1 to
-BOS state 2, according to the fitted Model 2.  This is illustrated by
-a vertical dotted line.  This is the point at which the individual's
-lung function started to remain consistently below 80\% baseline
-FEV$_1$.
-
-\begin{Scode}
-keep <- fev$ptnum==1 & fev$fev<999
-plot(fev$days[keep], fev$fev[keep], type="l",
-ylab=expression(paste("% baseline ", FEV[1])), xlab="Days after transplant")
-vit <- viterbi.msm(fev2.msm)[keep,]
-(max1 <- max(vit$time[vit$fitted==1]))
-(min2 <- min(vit$time[vit$fitted==2]))
-abline(v = mean(max1,min2), lty=2)
-text(max1 - 500, 50, "STATE 1")
-text(min2 + 500, 50, "STATE 2")
-\end{Scode}
-
-\paragraph{An alternative way of specifying a misclassification model}
-
-This general framework for specifying hidden Markov models can also be
-used to specify multi-state models with misclassification.  A
-misclassification model is a hidden Markov model with a categorical
-outcome distribution.  So instead of an \Rfunarg{ematrix} argument to
-\Rfunction{msm}, we can use a \Rfunarg{hmodel} argument with
-\Rfunction{hmmCat} constructor functions.
-
-\Rfunction{hmmCat} takes at least one argument \Rfunarg{prob}, a
-vector of probabilities of observing outcomes of $1, 2, \ldots, n$
-respectively, where $n$ is the length of \Rfunarg{prob}.  All outcome
-probabilities with an initial value of zero are assumed to be fixed at
-zero.  \Rfunarg{prob} is scaled if necessary to sum to one.
-
-The model in section \ref{sec:fitting:hmm:misc} specifies that an
-individual occupying underlying state 1 can be observed as states 2
-(and 1), underlying state 2 can be observed as states 1, 2 or 3, and
-state 3 can be observed as states 2 or 3, and underlying state 4
-(death) cannot be misclassified. Initial values of 0.1 are given for
-the 1-2, 2-1, 2-3 and 3-2 misclassification probabilities.
-
-This is equivalent to the model below, specified using a
-\Rfunarg{hmodel} argument to \Rfunction{msm}.  The maximum likelihood
-estimates should be the same as before (Model 5).
-
-<<>>=
-oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
-                   c(0, 0, 0.202, 0.081),
-                   c(0, 0, 0, 0.126),
-                   c(0, 0, 0, 0))
-cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                     hmodel = list (hmmCat(c(0.9, 0.1, 0, 0)), hmmCat(c(0.1, 0.8, 0.1, 0)), hmmCat(c(0, 0.1, 0.9, 0)), hmmIdent(4)),
-                     qmatrix = oneway4.q, obstrue=firstobs, death = 4, method="BFGS")
-cavmisc.msm
-@
-
-\subsubsection{Defining a new hidden Markov model distribution}
-
-Suppose the hidden Markov model outcome distributions supplied with
-\Rpackage{msm} (Table \ref{tab:hmm:dists}) are insufficient.  We want
-to define our own univariate distribution, called
-\Rfunction{hmmNewDist}, taking two parameters \Robject{location} and
-\Robject{scale}.  Download the source package, for example
-\texttt{msm-0.7.2.tar.gz} for version 0.7.2, from CRAN and edit the
-files in there, as follows.
-
-\begin{enumerate}
-\item Add an element to \Robject{.msm.HMODELPARS} in the file
-  \texttt{R/constants.R}, naming the parameters of the distribution.
-  For example
-\begin{Scode}
-  newdist = c('location', 'scale')
-\end{Scode}
-
-\item Add a corresponding element to the C variable \texttt{HMODELS}
-  in the file \texttt{src/lik.c}.  This MUST be in the same position
-  as in the \Robject{.msm.HMODELPARS} list.  For example,
-\begin{Scode}
-hmmfn HMODELS[] = {
-  ...,
-  hmmNewDist
-};.
-\end{Scode}
-
-\item The new distribution is allowed to have one parameter which can
-  be modelled in terms of covariates.  Add the name of this parameter
-  to the named vector \Robject{.msm.LOCPARS} in
-  \texttt{R/constants.R}.  For example \texttt{newdist = 'location'}.
-  Specify \texttt{newdist = NA} if there is no such parameter.
-
-\item Supposed we have specified a parameter with a non-standard name,
-  that is, one which doesn't already appear in
-  \Robject{.msm.HMODELPARS}. Standard names include, for example,
-  \texttt{'mean'}, \texttt{'sd'}, \texttt{'shape'} or
-  \texttt{'scale'}.  Then we should add the allowed range of the
-  parameter to \Robject{.msm.PARRANGES}. In this example, we add
-  \texttt{meanpars = c(-Inf, Inf)} to \Robject{.msm.PARRANGES}.  This
-  ensures that the optimisation to estimate the parameter takes place
-  on a suitable scale, for example, a log scale for a parameter
-  constrained to be positive.  If the parameter should be fixed during
-  maximum likelihood estimation (for example, the denominator of a
-  binomial distribution) then add its name to \Robject{.msm.AUXPARS}.
-
-\item Add an R constructor function for the distribution to
-  \texttt{R/hmm-dists.R}.  For a simple univariate distribution, this is of the
-  form
-\begin{Scode}
-  hmmNewDist <- function(location, scale)
-  {
-    hmmDIST(label = "newdist",
-    link = "identity",
-    r = function(n) rnewdist(n, location, scale),
-    match.call())
-  }
-\end{Scode}
-
-  \begin{itemize}
-  \item The \texttt{'label'} must be the same as the name you
-    supplied for the new element of \Robject{.msm.HMODELPARS}
-
-  \item \texttt{link} is the link function for modelling the location
-    parameter of the distribution as a linear function of covariates. This
-    should be the quoted name of an R function.  A log link is
-    \texttt{'log'} and a logit link is \texttt{'qlogis'}.  If
-    using a new link function other than \texttt{'identity'},
-    \texttt{'log'}, or \texttt{'qlogis'}, you should add its name
-    to the vector \Robject{.msm.LINKFNS} in \texttt{R/constants.R},
-    and add the name of the corresponding inverse link to
-    \Robject{.msm.INVLINK}.  You should also add the names of these
-    functions to the C array \texttt{LINKFNS} in \texttt{src/lik.c},
-    and write the functions if they do not already exist.
-
-  \item \texttt{r} is an R function, of the above format, returning a
-    vector of \texttt{n} random values from the distribution.
-    You should write this if it doesn't already exist.
-  \end{itemize}
-
-\item Add the name of the new constructor function to the NAMESPACE in
-  the top-level directory of the source package.
-
-\item Write a C function to compute the probability density of the
-  distribution, and put this in \texttt{src/hmm.c}, with a declaration
-  in \texttt{src/hmm.h}. This must be of the form
-\begin{Scode}
-  double hmmNewDist(double x, double *pars)
-\end{Scode}
-
-  where \texttt{*pars} is a vector of the parameters of the
-  distribution, and the density is evaluated at \texttt{x}.
-
-\item Update the documentation (\texttt{man/hmm-dists.Rd}) and the
-  distribution table in\\ \texttt{inst/doc/msm-manual.Rnw}) if you like.
-
-\item Recompile the package (see the ``Writing R Extensions'' manual)
-
-\end{enumerate}
-
-Your new distribution will be available to use in the \Rfunarg{hmodel}
-argument to \Rfunction{msm}, as, for example
-\begin{Scode}
-hmodel = list(..., hmmNewDist(location = 0, scale = 1), ...)
-\end{Scode}
-If your distribution may be of interest to others, ask me nicely
-(\texttt{chris.jackson@mrc-bsu.cam.ac.uk}) to include it in a future
-release.
-
-\clearpage
-
-\section{\Rpackage{msm} reference guide}
-
-The R help page for \Rfunction{msm} gives details of all the allowed
-arguments and options to the \Rfunction{msm} function.  To view this
-online in R, type:
-
-<<eval=FALSE>>=
-help(msm)
-@
-
-Similarly all the other functions in the package have help pages,
-which should always be consulted in case of doubt about how to
-call them.  The web-browser based help interface may often be
-convenient - type
-
-<<eval=FALSE>>=
-help.start()
-@
-and navigate to \textsf{Packages} $\ldots$ \textsf{msm}, which brings
-up a list of all the functions in the package with links to their
-documentation, and a link to this manual in PDF format.
-
-\appendix
-
-\section{Changes in the msm package}
-
-For a detailed list of the changes in recent versions of
-\Rpackage{msm}, see the \texttt{NEWS} file in the top-level directory
-of the installed package.  The \texttt{Changelog} file in the
-top-level directory of the source package contains more technical
-detail about these changes.
-
-Development versions of \Rpackage{msm} are to be found on R-Forge,
-\url{http://msm.r-forge.r-project.org/}.  These are often more recent
-than the version released on CRAN, so if you think you have found a
-bug, then please check first to see whether it has been fixed on the
-R-Forge version.
-
-\section{Get in touch}
-
-If you use \Rpackage{msm} in your work, whatever your field of
-application, please let me know, for my own interest!  Suggestions for
-improvement are welcome.
-
-\clearpage
-
-\bibliography{msm}
+Test
 

### 2013-12-31 pkg/inst/doc/msm-manual.Rnw
Another build test
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@36 edb9625f-4e0d-4859-8d74-9fd3b1da38cb
Test build again: just preamble, title and abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@37 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \documentclass{article}
+
+%% Need to modify Sweave.sty to pass pdftex option to graphicx.
+\usepackage{Sweave-local}
+\usepackage{times}
+\usepackage{url}
+\addtolength{\textwidth}{2cm}
+
+\newcommand{\Exp}{\mathop{\mathrm{Exp}}}
+\newcommand{\etal}{{\textit{et al.}}}
+\newcommand{\Rfunction}[1]{{\texttt{#1}}}
+\newcommand{\Robject}[1]{{\texttt{#1}}}
+\newcommand{\Rpackage}[1]{{\textit{#1}}}
+\newcommand{\Rmethod}[1]{{\texttt{#1}}}
+\newcommand{\Rfunarg}[1]{{\texttt{#1}}}
+\newcommand{\Rclass}[1]{{\textit{#1}}}
+
+%% version <- 1.3
+
+<<echo=FALSE>>=
+version <- gsub("Version: +", "",
+               packageDescription("msm", lib.loc="../../..")$Version)
+@
+
+\title{Multi-state modelling with R: the {\tt msm} package \vskip 0.2in
+  \large{Version %
+<<echo=FALSE,results=tex>>=
+cat(version)
+@
+\vskip 0.1in
+<<echo=FALSE,results=tex>>=
+cat(format(Sys.time(), "%d %B, %Y"))
+@
+}}
+\author{Christopher Jackson\\MRC Biostatistics Unit\\Cambridge, U.K.\\ \texttt{chris.jackson@mrc-bsu.cam.ac.uk}}
+\date{}
+%% label with date when Rnw is compiled, not when tex is compiled. should be release date of package
+
+\bibliographystyle{unsrt}
 
@@
 
-Test
+  The multi-state Markov model is a useful way of describing a process
+  in which an individual moves through a series of states in
+  continuous time.  The \Rpackage{msm} package for R allows a general
+  multi-state model to be fitted to longitudinal data. Data often
+  consist of observations of the process at arbitrary times, so that
+  the exact times when the state changes are unobserved.  For example,
+  the progression of chronic diseases is often described by stages of
+  severity, and the state of the patient may only be known at doctor
+  or hospital visits.  Features of \Rpackage{msm} include the ability
+  to model transition rates and hidden Markov output models in terms
+  of covariates, and the ability to model data with a variety of
+  observation schemes, including censored states.
+
+  Hidden Markov models, in which the true path through states is only
+  observed through some error-prone marker, can also be fitted.  The
+  observation is generated, conditionally on the underlying states,
+  via some distribution.  An example is a screening misclassification
+  model in which states are observed with error.  More generally,
+  hidden Markov models can have a continuous response, with some
+  arbitrary distribution, conditionally on the underlying state.
+
+  This manual introduces the theory behind multi-state Markov and
+  hidden Markov models, and gives a tutorial in the typical use of the
+  \Rpackage{msm} package, illustrated by some typical applications to
+  modelling chronic diseases.
+
+  Much of the material in this manual is published, in a more concise
+  form, in Journal of Statistical Software (2011) 38(8):1-29,
+  \url{http://www.jstatsoft.org/v38/i08/}
 
 \end{abstract}
-
-\tableofcontents
-\clearpage
 

### 2013-12-31 pkg/inst/doc/msm-manual.Rnw
Another build test, up to running first model
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@38 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \label{sec:multistate}
-Test
+
+\subsection{Introduction}
+\label{sec:intro}
+
+Figure~\ref{fig:multi} illustrates a multi-state model in continuous
+time. Its four states are labelled {\bf 1, 2, 3, 4}.  At a time $t$
+the individual is in state $S(t)$.  The arrows show which transitions
+are possible between states. The next state to which the individual
+moves, and the time of the change, are governed by a set of {\em
+  transition intensities} $q_{rs}(t, z(t))$ for each pair of states
+$r$ and $s$.  The intensities may also depend on the time of the
+process $t$, or more generally a set of individual-specific or
+time-varying explanatory variables $z(t)$.  The intensity represents
+the instantaneous risk of moving from state $r$ to state $s$:
+\begin{equation}
+  \label{eq:multi:intensity}
+  q_{rs}(t, z(t))  =  \lim_{\delta t \rightarrow 0} P (S(t+\delta t) = s | S(t) = r) / \delta t
+\end{equation}
+The intensities form a matrix $Q$ whose rows sum to zero, so that the
+diagonal entries are defined by $q_{rr} = - \sum_{s \neq r} q_{rs}$.
+
+To fit a multi-state model to data, we estimate this transition
+intensity matrix.  We concentrate on {\em Markov} models here. The
+Markov assumption is that future evolution only depends on the current
+state.  That is, $q_{rs}(t, z(t), \mathcal{F}_t)$ is independent of
+$\mathcal{F}_t$, the observation history $\mathcal{F}_t$ of the
+process up to the time preceding $t$.  See, for example, Cox and
+Miller\cite{cox:miller} for a thorough
+introduction to the theory of continuous-time Markov chains.
+
+\begin{figure}[p]
+  \begin{center}
+    \scalebox{0.4}{\includegraphics{figures/multistate}}
+    \[
+    Q = \left(
+      \begin{array}{llll}
+        q_{11} & q_{12} & q_{13} & q_{14}\\
+        q_{21} & q_{22} & q_{23} & q_{24}\\
+        q_{31} & q_{32} & q_{33} & q_{34}\\
+        q_{41} & q_{42} & q_{43} & q_{44}\\
+      \end{array}
+    \right )
+    \]
+    \caption{\label{fig:multi}General multi-state model.}
+  \end{center}
+\end{figure}
+
+In a time-homogeneous continuous-time Markov model, a single period of occupancy (or
+\emph{sojourn time}) in state $r$ has an exponential distribution,
+with rate given by $-q_{rr}$, (or
+mean $-1 / q_{rr}$).  The remaining elements of the $r$th row of $Q$
+are \emph{proportional to} the probabilities governing the next state
+after $r$ to which the individual makes a transition.  The probability
+that the individual's next move from state $r$ is to state $s$ is
+$-q_{rs} / q_{rr}$.
+
+\subsection{Disease progression models}
+
+The development of the \Rpackage{msm} package was motivated by
+applications to disease modelling.  Many chronic diseases have a
+natural interpretation in terms of staged progression. Multi-state
+Markov models in continuous time are often used to model the course of
+diseases. A commonly-used model is illustrated in
+Figure \ref{fig:disease}.  This represents a series of successively
+more severe disease stages, and an `absorbing' state, often death.
+The patient may advance into or recover from adjacent disease stages,
+or die at any disease stage.  Observations of the state $S_i(t)$ are
+made on a number of individuals $i$ at arbitrary times $t$, which may
+vary between individuals.  The stages of disease may be modelled as a
+homogeneous continuous-time Markov process, with a transition matrix
+$Q$, pictured below Figure \ref{fig:disease}.
+
+A commonly-used model is the \emph{illness-death} model, with three
+states representing health, illness and death
+(Figure \ref{fig:multi:illdeath}).  Transitions are permitted from
+health to illness, illness to death and health to death.  Recovery
+from illness to health is sometimes also considered.
+
+A wide range of medical situations have been modelled using
+multi-state methods, for example, screening for abdominal aortic
+aneurysms (Jackson \etal\cite{jackson:sharples:2003}),
+problems following lung transplantation (Jackson and
+Sharples\cite{jackson:sharples:2002}), problems following heart
+transplantation (Sharples\cite{sharp:gibbs}, Klotz and
+Sharples\cite{klotz:est}), hepatic cancer (Kay\cite{kay:mark}), HIV
+infection and AIDS (Longini \etal\cite{long1}, Satten and
+Longini\cite{sattlong}, Guihenneuc-Jouyaux \etal\cite{rich},
+Gentleman \etal\cite{gentlaw}), diabetic complications
+(Marshall and Jones\cite{retino2}, Andersen\cite{andersen88}), breast
+cancer screening (Duffy and Chen\cite{duffy95}, Chen \emph{et
+  al.}\cite{duffy96}), cervical cancer screening (Kirby and
+Spiegelhalter\cite{kirby:spiegelhalter}) and liver cirrhosis (Andersen
+\etal\cite{ander:proth}).  Many of these references also
+describe the mathematical theory, which will be reviewed in the
+following sections.
+
+\begin{figure}[p]
+  \centering
+  \vskip 1cm
+  \scalebox{1.0}{\includegraphics{figures/general}}
+    \[
+    Q = \left(
+      \begin{array}{llllll}
+        q_{11} & q_{12} & 0 & 0 & \ldots & q_{1n}\\
+        q_{21} & q_{22} & q_{23} & 0 & \ldots & q_{2n}\\
+        0 & q_{32} & q_{33} & q_{34} & \ddots & q_{3n}\\
+        0 & 0 & q_{43} & q_{44} & \ddots & q_{4n}\\
+        \vdots & \vdots & \ddots & \ddots & \ddots & \vdots\\
+        0 & 0 & 0 & 0 & \ldots & 0\\
+      \end{array}
+    \right )
+    \]
+  \caption{\label{fig:disease}General model for disease progression.}
+  \end{figure}
+
+\begin{figure}[p]
+  \begin{center}
+    \scalebox{0.4}{\includegraphics{figures/illdeath}}
+    \caption{Illness-death model.}
+    \label{fig:multi:illdeath}
+  \end{center}
+\end{figure}
+
+\subsection{Arbitrary observation times}
+\label{sec:arbitr-observ-times}
+
+Longitudinal data from monitoring disease progression are often
+incomplete in some way.  Usually patients are seen at intermittent
+follow-up visits, at which monitoring information is collected, but
+information from the periods between visits is not available.  Often
+the exact time of disease onset is unknown. Thus, the changes of state
+in a multi-state model usually occur at unknown times.  Also a subject
+may only be followed up for a portion of their disease history.  A
+fixed observation schedule may be specified in advance, but in
+practice times of visits may vary due to patient and hospital
+pressures. The states of disease progression models often include
+death. Death times are commonly recorded to within a day.  Also
+observations may be censored.  For example, at the end of a study, an
+individual may be known only to be alive, and in an unknown state.
+
+A typical sampling situation is illustrated in
+Figure \ref{fig:multi:sampling}. The individual is observed at four
+occasions through 10 months. The final occasion is the death date
+which is recorded to within a day. The only other information
+available is the occupation of states 2, 2, and 1 at respective times
+1.5, 3.5 and 5. The times of movement between states and the state
+occupancy in between the observation times are unknown.  Although the
+patient was in state 3 between 7 and 9 months this was not observed at
+all.
+
+\paragraph{Informative sampling times}
+To fit a model to longitudinal data with arbitrary sampling times we
+must consider the reasons why observations were made at the given
+times.  This is analogous to the problem of missing data, where the
+fact that a particular observation is missing may implicitly give
+information about the value of that observation. Possible observation
+schemes include:
+\begin{itemize}
+\item {\em fixed}. Each patient is observed at fixed intervals
+  specified in advance.
+\item {\em random}. The sampling times vary randomly, independently
+  of the current state of the disease.
+\item {\em doctor's care}. More severely ill patients are monitored
+  more closely. The next sampling time is chosen on the basis of the
+  current disease state.
+\item patient {\em self-selection}. A patient may decide to visit the
+  doctor on occasions when they are in a poor condition.
+\end{itemize}
+Gr\"uger \etal \cite{gruger:valid} discussed conditions under
+which sampling times are \emph{informative}. If a multi-state model is
+fitted, ignoring the information available in the sampling times, then
+inference may be biased.  Mathematically, because the sampling
+times are often themselves random, they should be modelled along with
+the observation process $X_t$. However the ideal situation is where
+the joint likelihood for the times and the process is proportional to
+the likelihood obtained if the sampling times were fixed in advance.
+Then the parameters of the process can be estimated independently of
+the parameters of the sampling scheme.
+
+In particular, they showed that fixed, random and doctor's care
+observation policies are not informative, whereas patient
+self-selection is informative.   Note that \Rpackage{msm} does not
+deal with informative sampling times. See, e.g. \cite{sweeting:inform:msm}
+for some methods in this case, which require specialised programming.
+
+\begin{figure}[p]
+  \begin{center}
+    \scalebox{0.6}{\rotatebox{270}{\includegraphics{figures/sampling}}}
+    \caption{Evolution of a multi-state model. The process is observed
+      on four occasions.}
+    \label{fig:multi:sampling}
+  \end{center}
+\end{figure}
+
+\subsection{Likelihood for the multi-state model}
+\label{sec:multi:likelihood}
+
+Kalbfleisch and Lawless\cite{kalbfleisch:lawless} and later
+Kay \cite{kay:mark} described a general method for evaluating the
+likelihood for a general multi-state model in continuous time,
+applicable to any form of transition matrix. The only available
+information is the observed state at a set of times, as in
+Figure \ref{fig:multi:sampling}.  The sampling times are assumed to be
+non-informative.
+
+\paragraph{Transition probability matrix}
+The likelihood is calculated from the \emph{transition probability matrix}
+$P(t)$.  For a time-homogeneous process, the $(r,s)$ entry of $P(t)$,
+$p_{rs}(t)$, is the probability of being in state $s$ at a time $t+u$
+in the future, given the state at time $u$ is $r$.  It does not say
+anything about the time of transition from $r$ to $s$, indeed the
+process may have entered other states between times $u$ and $t+u$.
+$P(t)$ can be calculated by taking the matrix exponential of the
+scaled transition intensity matrix (see, for example, Cox and
+Miller \cite{cox:miller}).
+\begin{equation}
+  \label{eq:exptq}
+  P(t) = \Exp(tQ)
+\end{equation}
+
+The matrix exponential $\Exp$ is different from a scalar exponential.
+The exponential of a matrix is defined by the same "power series"
+$\Exp(X) = 1 + X^2/2! + X^3/3! + ...$ as the scalar exponential,
+except that each term $X^k$ in the series is defined by matrix products,
+not element-wise scalar multiplication.  It is notoriously difficult
+to calculate reliably, as discussed by Moler and van Loan \cite{matrixexp}.
+For simpler models, it is feasible to calculate an analytic expression
+for each element of $P(t)$ in terms of $Q$.  This is generally faster
+and avoids the potential numerical instability of calculating the
+matrix exponential.  Symbolic algebra sofware, such as Mathematica,
+can be helpful for obtaining these expressions.  For example, the
+three-state illness-death model with no recovery has a transition
+intensity matrix of
+\[
+Q = \left(
+  \begin{array}{llllll}
+    -(q_{12} + q_{13}) & q_{12} & q_{13}\\
+    0 & - q_{23} & q_{23}\\
+    0 & 0 & 0\\
+  \end{array}
+\right )
+\]
+The corresponding time $t$ transition probabilities are
+\begin{eqnarray*}
+  p_{11}(t) & = & e^{-(q_{12} + q_{13})t}\\
+  p_{12}(t) & = &
+\left\{
+\begin{array}{ll}
+\frac{ q_{12}(-1 + e^{(q_{12}+q_{13}-q_{23})t})e^{-(q_{12}+q_{13})t} } { (q_{12} + q_{13} - q_{23}) } & (q_{12} + q_{13} \neq q_{23})\\
+q_{12}te^{(-(q_{12} + q_{13})t} &  (q_{12} + q_{13} = q_{23})
+\end{array}
+\right.
+\\
+  p_{13}(t) & = &
+\left\{
+\begin{array}{ll}
+\frac{1 + (-q_{13} + q_{23})e^{-(q_{12}+q_{13})t}} {q_{12} + q_{13} - q_{23}} - \frac { q_{12}e^{-q_{23}t}} {q_{12} + q_{13} - q_{23} } & (q_{12} + q_{13} \neq q_{23})\\
+ (-1 + e^{(q_{12} + q_{13})t} - q_{12}t)e^{-(q_{12} + q_{13})t} & (q_{12} + q_{13} = q_{23})
+\end{array}
+\right.
+\\
+  p_{21}(t) & = & 0\\
+  p_{22}(t) & = & e^{-q_{23}t}\\
+  p_{23}(t) & = & 1 - e^{-q_{23}t}\\
+  p_{31}(t) & = & 0\\
+  p_{32}(t) & = & 0\\
+  p_{33}(t) & = & 1\\
+\end{eqnarray*}
+The \Rpackage{msm} package calculates $P(t)$ analytically for selected
+2, 3, 4 and 5-state models, illustrated in
+Figures \ref{fig:anp2}--\ref{fig:anp5}.  For other models, which can have any transition
+structure on any number of states in principle, $P(t)$ is
+determined from the matrix exponential.  This is calculated using eigensystem
+decomposition (if eigenvalues are distinct) or a method based on
+Pad\'e approximants with scaling and squaring \cite{matrixexp} (if
+there are repeated eigenvalues). Notice that the states are not
+labelled in these figures.  Each graph can correspond to several
+different $Q$ matrices, depending on how the states are labelled.  For
+example, Figure 1 a) illustrates the model defined by either \( Q =
+\left(
+  \begin{array}{ll}
+    - q_{12} & q_{12} \\
+    0 & 0
+\end{array}
+\right)
+\)
+or
+\(
+Q = \left( \begin{array}{ll}
+0 & 0\\
+q_{21} & - q_{21}
+\end{array}
+\right)
+\).
+
+\begin{figure}
+\begin{center}
+a) \includegraphics[width=3cm]{figures/p2q1} \hskip 3cm
+b) \includegraphics[width=3cm]{figures/p2q12}
+\end{center}
+\caption{\label{fig:anp2}Two-state models fitted using analytic $P(t)$ matrices in \Rpackage{msm}. Implemented for all permutations of state labels 1, 2.}
+\end{figure}
+\begin{figure}
+\begin{center}
+a) \includegraphics[width=3cm]{figures/p3q12} \hskip 3cm
+b) \includegraphics[width=5cm]{figures/p3q14} \vskip 1cm
+c) \includegraphics[width=3cm]{figures/p3q16}\hskip 3cm
+d) \includegraphics[width=3cm]{figures/p3q124}\vskip 1cm
+e) \includegraphics[width=3cm]{figures/p3q135}\hskip 3cm
+f) \includegraphics[width=3cm]{figures/p3q1246}
+\end{center}
+\caption{\label{fig:anp3}Three-state models fitted using analytic $P(t)$ matrices in \Rpackage{msm}. Implemented for all permutations of state labels 1, 2, 3.}
+\end{figure}
+\begin{figure}
+\begin{center}
+a) \includegraphics[width=7cm]{figures/p4q159} \vskip 1cm
+b) \includegraphics[width=7cm]{figures/p4q13569}
+\end{center}
+\caption{\label{fig:anp4}Four-state models fitted using analytic $P(t)$ matrices in \Rpackage{msm}. Implemented for all permutations of state labels 1, 2, 3, 4.}
+\end{figure}
+\begin{figure}
+\begin{center}
+a) \includegraphics[width=9cm]{figures/p5q1_6_11_16}\vskip 1cm
+b) \includegraphics[width=9cm]{figures/p5q1_4_6_8_11_12_16}\vskip 1cm
+c) \includegraphics[width=5cm]{figures/p5q1_6_7_11_12}
+\end{center}
+\caption{\label{fig:anp5}Five-state models fitted using analytic $P(t)$ matrices in \Rpackage{msm}. Implemented for all permutations of state labels 1, 2, 3, 4, 5.}
+\end{figure}
+
+\paragraph{Likelihood for intermittently-observed processes}
+Suppose $i$ indexes $M$ individuals.  The data for individual $i$
+consist of a series of times $(t_{i1}, \ldots, t_{in_i})$ and
+corresponding states $(S(t_{i1}), \ldots, S(t_{in_i}))$.  Consider a
+general multi state model, with a pair of successive observed disease
+states $S(t_j), S(t_{j+1})$ at times $t_j, t_{j+1}$. The contribution
+to the likelihood from this pair of states is
+
+\begin{equation}
+  \label{eq:multi:lik:contrib}
+  L_{i, j} =  p_{S(t_j)S(t_{j+1})}(t_{j+1} - t_j)
+\end{equation}
+This is the entry of the transition matrix $P(t)$ at the $S(t_j)$th
+row and $S(t_{j+1})$th column, evaluated at $t = t_{j+1} - t_j$.
+
+The full likelihood $L(Q)$ is the product of all such terms $L_{i,j}$
+over all individuals and all transitions. It depends on the unknown
+transition matrix $Q$, which was used to determine $P(t)$.
+
+\paragraph{Exactly-observed death times}
+In observational studies of chronic diseases, it is common that the
+time of death is known, but the state on the previous instant before
+death is unknown. If $S(t_{j+1}) = D $ is such a death state, then the
+contribution to the likelihood is summed over the unknown state $m$ on
+the instant before death:
+
+\begin{equation}
+  \label{eq:multi:lik:death}
+  L_{i, j} =  \sum_{m \neq D} p_{S(t_j),m}(t_{j+1} - t_j) q_{m, D}
+\end{equation}
+The sum is taken over all possible states $m$ which can be visited
+between $S(t_j)$ and $D$.
+
+\paragraph{Exactly observed transition times}
+If the times $(t_{i1}, \ldots, t_{in_i})$ had been the {\em exact}
+transition times between the states, with no transitions between the
+observation times, then the contributions would be of the form
+
+\begin{equation}
+  \label{eq:multi:lik:exact}
+  L_{i, j} =  \exp(q_{S(t_j)S(t_{j})}(t_{j+1} - t_j)) q_{S(t_j)S(t_{j+1})}
+\end{equation}
+since the state is assumed to be $S(t_j)$ throughout the interval
+between $t_j$ and $t_{j+1}$ with a known transition to state
+$S(t_{j+1})$ at $t_{j+1}$.  \Rpackage{msm} is restricted to Markov
+models, but much richer models are possible for this type of data. For
+example, Putter \etal \cite{putter:mstate} discussed the
+\Rpackage{mstate} software for semi-parametric multi-state models with
+non-parametric baseline hazards and Cox regression.  The Markov
+assumption is restrictive but necessary in general to compute a
+likelihood for intermittently-observed processes.
+
+\paragraph{Censored states}
+A censored quantity is one whose exact value is unknown, but known to
+be in a certain interval.  For example, in survival analysis, a death
+time is \emph{right-censored} if the study ends and the patient is
+still alive, since the death time is known to be greater than the end
+time.  In multi-state models for intermittently-observed processes,
+the times of changes of state are usually \emph{interval censored},
+known to be within bounded intervals.  This leads to a likelihood
+based on equation \ref{eq:multi:lik:contrib}.
+
+In some circumstances, \emph{states} may be censored as well as
+\emph{event times}.  For example, at the end of some chronic disease
+studies, patients are known to be alive but in an \emph{unknown
+  state}.  For such a censored observation $S(t_{j+1})$, known only to
+be a state in the set $C$, the equivalent contribution to the
+likelihood is
+\begin{equation}
+  \label{eq:multi:lik:deathcens}
+  L_{i, j} =  \sum_{m \in C} p_{S(t_j),m}(t_{j+1} - t_j)
+\end{equation}
+Note that this special likelihood is not needed if the state is known
+at the end of the study.  In this case,
+likelihood \ref{eq:multi:lik:contrib} applies.  Although the
+\emph{survival time} is censored, the \emph{state} at the end of the
+study is not censored.
+
+More generally, suppose every observation from a particular individual
+is censored.  Observations $1, 2, \ldots n_i$ are known only to be in
+the sets $C_1, C_2, \ldots, C_{n_i}$ respectively. The likelihood for
+this individual is a sum of the likelihoods of all possible paths
+through the unobserved states.
+\begin{equation}
+  \label{eq:multi:lik:cens}
+  L_i = \sum_{s_{n_i} \in C_{n_i}} \ldots \sum_{s_2 \in C_2} \sum_{s_1 \in C_1} p_{s_1 s_2}(t_2 - t_1) p_{s_2 s_3} (t_3 - t_2) \ldots p_{s_{n_i-1} s_{n_i}} (t_{n_i} - t_{n_i-1})
+\end{equation}
+Suppose the states comprising the set $C_j$ are $c^{(j)}_1, \ldots,
+c^{(j)}_{m_j}$. This likelihood can also be written as a matrix
+product, say,
+\begin{equation}
+  \label{eq:multi:lik:cens:matrix}
+  L_i = \mathbf{1}^T P^{1,2} P^{2,3} \ldots P^{n_i-1, n_i} \mathbf{1}
+\end{equation}
+where $P^{j-1, j}$ is a $m_{j-1} \times m_j$ matrix with $(r,s)$ entry
+$p_{c^{(j-1)}_r c^{(j)}_s}(t_j - t_{j-1})$, and $\mathbf{1}$ is the
+vector of ones.
+
+The \Rpackage{msm} package allows multi-state models to be fitted to
+data from processes with arbitrary observation times (panel data), exactly-observed
+transition times, exact death times and censored states, or a mixture
+of these schemes.
+
+\subsection{Covariates}
+\label{sec:multi:covariates}
+
+The relation of constant or time-varying characteristics of
+individuals to their transition rates is often of interest in a
+multi-state model.  Explanatory variables for a particular transition
+intensity can be investigated by modelling the intensity as a function
+of these variables.  Marshall and Jones \cite{retino2} described a
+form of a {\em proportional hazards} model, where the transition
+intensity matrix elements $q_{rs}$ which are of interest can be
+replaced by
+\[
+q_{rs}(z(t)) = q_{rs}^{(0)} \exp(\beta_{rs}^T z(t))
+\]
+The new $Q$ is then used to determine the likelihood. If the
+covariates $z(t)$ are time dependent, the contributions to the
+likelihood of the form $p_{rs} (t - u)$ are replaced by
+\[
+p_{rs}(t - u, z(u))
+\]
+although this requires that the value of the covariate is known at
+every observation time $u$. Sometimes covariates are observed at
+different times to the main response, for example recurrent disease
+events or other biological markers. In some of these cases it could be
+assumed that the covariate is a step function which remains constant
+between its observation times.  If the main response (the state of the
+Markov process) is not observed at the times when the covariate
+changes, it could be considered as a "censored" state (as in
+Section \ref{sec:multi:likelihood}).
+
+The \Rpackage{msm} package allows individual-specific or time
+dependent covariates to be fitted to transition intensities.  In order
+to calculate transition probabilities $P(t)$ on which the likelihood
+depends, time-dependent covariates are assumed to be
+piecewise-constant.  Models whose intensities change with time are
+called \emph{time-inhomogeneous}. An important special case handled by
+\Rpackage{msm} is the model in which intensities change at a series of
+times common to each individual.
+
+Marshall and Jones \cite{retino2} described likelihood ratio and Wald
+tests for covariate selection and testing hypotheses, for example
+whether the effect of a covariate is the same for all forward
+transitions in a disease progression model, or whether the effect on
+backward transitions is equal to minus the effect on forward
+transitions.
+
+\subsection{Hidden Markov models}
+\label{sec:hmm}
+
+In a {\em hidden Markov model} (HMM) the states of the Markov chain
+are not observed. The observed data are governed by some probability
+distribution (the \emph{emission} distribution) conditionally on the
+unobserved state.  The evolution of the underlying Markov chain is
+governed by a transition intensity matrix $Q$ as before.
+(Figure \ref{fig:multi:hidden}).  Hidden Markov models are mixture
+models, where observations are generated from a certain number of
+unknown distributions.  However the distribution changes through time
+according to states of a hidden Markov chain.  This class of model is
+commonly used in areas such as speech and signal processing
+\cite{juang:rabiner} and the analysis of biological sequence data
+\cite{biolog:seq}. In engineering and biological sequencing
+applications, the Markov process usually evolves over an
+equally-spaced, discrete `time' space.  Therefore most of the theory
+of HMM estimation was developed for discrete-time models.
+
+HMMs have less frequently been used in medicine, where continuous time
+processes are often more suitable.  A disease process evolves in
+continuous time, and patients are often monitored at irregular and
+differing intervals.  These models are suitable for estimating
+population quantities for chronic diseases which have a natural staged
+interpretation, but which can only be diagnosed by an error-prone
+marker.  The \Rpackage{msm} package can fit continuous-time hidden
+Markov models with a variety of emission distributions.
+
+\begin{figure}[htbp]
+  \begin{center}
+    \scalebox{0.6}{\rotatebox{270}{\includegraphics{figures/hidden}}}
+    \caption{A hidden Markov model in continuous time. Observed states
+      are generated conditionally on an underlying Markov process. }
+    \label{fig:multi:hidden}
+  \end{center}
+\end{figure}
+
+\subsubsection{Misclassification models}
+
+An example of a hidden Markov model is a multi-state model with
+misclassification.  Here the observed data are states, assumed to be
+misclassifications of the true, underlying states.
+
+For example, consider a disease progression model with at least a
+disease-free and a disease state. When screening for the presence of
+the disease, the screening process can sometimes be subject to error.
+Then the Markov disease process $S_i(t)$ for individual $i$ is not
+observed directly, but through realisations $O_i(t)$.  The quality of
+a diagnostic test is often measured by the probabilities that the true
+and observed states are equal, $Pr(O_i(t) = r | S_i(t) = r)$. Where
+$r$ represents a `positive' disease state, this is the {\em
+  sensitivity}, or the probability that a true positive is detected by
+the test.  Where $r$ represents a `negative' or disease-free state,
+this represents the {\em specificity}, or the probability that, given
+the condition of interest is absent, the test produces a negative
+result.
+
+As an extension to the simple multi-state model described in
+section \ref{sec:multistate}, the \Rpackage{msm} package can fit a
+general multi-state model with misclassification.  For patient $i$,
+observation time $t_{ij}$, observed states $O_{ij}$ are generated
+conditionally on true states $S_{ij}$ according to a {\em
+  misclassification matrix} $E$. This is a $n \times n$ matrix, whose
+$(r,s)$ entry is
+\begin{equation}
+  \label{eq:misc}
+  e_{rs} = Pr(O(t_{ij}) = s | S(t_{ij}) = r),
+\end{equation}
+which we first assume to be independent of time $t$. Analogously to
+the entries of $Q$, some of the $e_{rs}$ may be fixed to reflect
+knowledge of the diagnosis process. For example, the probability of
+misclassification may be negligibly small for non-adjacent states of
+disease.  Thus the progression through underlying states is governed
+by the transition intensity matrix $Q$, while the observation process
+of the underlying states is governed by the misclassification matrix
+$E$.
+
+To investigate explanatory variables $w(t)$ for the probability $e_{rs}$ of
+misclassification as state $s$ given underlying state $r$, a
+multinomial logistic regression model can be used:
+\begin{equation}
+  \label{eq:misccovs}
+  \log \frac{e_{rs}(t)}{e_{rs_0}(t)}  =   \gamma_{rs}^T w(t).
+\end{equation}
+where $s_0$ is some baseline state, usually chosen as the underlying
+state, or the state with the highest probability (for numerical
+stability).
+
+\subsubsection{General hidden Markov model}
+\label{sec:hmm:general}
+
+Consider now a general hidden Markov model in continuous time.  The
+true state of the model $S_{ij}$ evolves as an unobserved Markov
+process.  Observed data $y_{ij}$ are generated conditionally true
+states $S_{ij} = 1, 2, \ldots, n$ according to a set of distributions
+$f_1(y | \theta_1, \gamma_1)$, $f_2(y | \theta_2, \gamma_2)$,
+$\ldots$, $f_n(y | \theta_n, \gamma_n)$, respectively.  $\theta_r$ is
+a vector of parameters for the state $r$ distribution.  One or more of
+these parameters may depend on explanatory variables through a
+link-transformed linear model with coefficients $\gamma_r$.
+
+\subsubsection{Likelihood for general hidden Markov models}
+
+A type of EM algorithm known as the {\em Baum-Welch} or {\em
+  forward-backward} algorithm \cite{baum:petrie66, baum:petrie70}, is
+commonly used for hidden Markov model estimation in discrete-time
+applications.  See, for example, Durbin \etal
+\cite{biolog:seq}, Albert \cite{albert99}.  A generalisation of this
+algorithm to continuous time was described by Bureau \etal
+\cite{bureau:hughes:shiboski:00}.
+
+The \Rpackage{msm} package uses a direct method of calculating
+likelihoods in discrete or continuous time based on matrix products.
+This type of method has been described by Macdonald and Zucchini
+\cite[pp.  77--79]{macdonald:zucchini}, Lindsey
+\cite[p.73]{lindsey:rm} and Guttorp \cite{guttorp}.  Satten and
+Longini \cite{sattlong} used this method to calculate likelihoods for
+a hidden Markov model in continuous time with observations of a
+continuous marker generated conditionally on underlying discrete
+states.
+
+Patient $i$'s contribution to the likelihood is
+\begin{eqnarray}
+  \label{eq:multi:hiddencontrib}
+  L_i     & = & Pr(y_{i1}, \ldots, y_{in_i})\\
+  & = & \sum Pr(y_{i1}, \ldots, y_{in_i} | S_{i1}, \ldots, S_{in_i})
+  Pr(S_{i1}, \ldots, S_{in_i}) \nonumber
+\end{eqnarray}
+where the sum is taken over all possible paths of underlying states
+$S_{i1}, \ldots, S_{in_i}$.  Assume that the observed states are
+conditionally independent given the values of the underlying states.
+Also assume the Markov property, $Pr(S_{ij}|S_{i,j-1}, \ldots, S_{i1})
+= Pr(S_{ij}|S_{i,j-1})$.  Then the contribution $L_i$ can be written
+as a product of matrices, as follows. To derive this matrix product,
+decompose the overall sum in equation \ref{eq:multi:hiddencontrib}
+into sums over each underlying state. The sum is accumulated over the
+unknown first state, the unknown second state, and so on until the
+unknown final state:
+\begin{eqnarray}
+  \label{eq:multi:hiddenlik}
+  L_i & = & \sum_{S_{i1}} Pr(y_{i1}|S_{i1})Pr(S_{i1}) \sum_{S_{i2}} Pr(y_{i2}|S_{i2})Pr(S_{i2}|S_{i1}) \sum_{S_{i3}} Pr(y_{i3}|S_{i3}) Pr(S_{i3}|S_{i2})
+  \nonumber \\
+  &   & \ldots \sum_{S_{in_i}} Pr(y_{in_i}|S_{in_i})
+  Pr(S_{in_i}|S_{in_{i-1}})
+\end{eqnarray}
+where $Pr(y_{ij}|S_{ij})$ is the emission probability density.  For
+misclassification models, this is the misclassification probability
+$e_{S_{ij} O_{ij}}$.  For general hidden Markov models, this is the
+probability density
+$f_{S_{ij}}(y_{ij}|\theta_{S_{ij}},\gamma_{S_{ij}})$.
+$Pr(S_{i,j+1}|S_{ij})$ is the $(S_{ij}, S_{i,j+1})$ entry of the
+Markov chain transition matrix $P(t)$ evaluated at $t = t_{i,j+1} -
+t_{ij}$.  Let $f$ be the vector with $r$ element the product of the
+initial state occupation probability $Pr(S_{i1}=r)$ and $Pr(y_{i1}|
+r)$, and let $\mathbf 1$ be a column vector consisting of ones. For $j
+= 2, \ldots, n_i$ let $T_{ij}$ be the $R \times R$ matrix (where $R$
+is the number of states) with $(r,s)$ entry
+\[
+Pr(y_{ij}| s) p_{rs} (t_{ij} - t_{i,j-1})
+\]
+Then subject $i$'s likelihood contribution is
+\begin{equation}
+  L_i =  f T_{i2} T_{i3}, \ldots T_{in_i} \mathbf 1
+  \label{eq:multi:hidden:matprod}
+\end{equation}
+
+If $S(t_{j}) = D$ is an absorbing state such as death, measured
+without error, whose entry time is known exactly, then the
+contribution to the likelihood is summed over the unknown state at the
+previous instant before death.  The $(r,s)$ entry of $T_{ij}$ is then
+\[
+p_{rs}(t_{j} - t_{j-1}) q_{s, D}
+\]
+Section \ref{sec:fitting:hmm:misc} describes how to fit multi-state
+models with misclassification using the \Rpackage{msm} package, and
+Section \ref{sec:fitting:hmm:general} describes how to apply general
+hidden Markov models.
+
+\subsubsection{Example of a general hidden Markov model}
+\label{sec:hmm:example:fev}
+
+Jackson and Sharples \cite{jackson:sharples:2002} described a model
+for FEV$_1$ (forced expiratory volume in 1 second) in recipients of
+lung transplants.  These patients were at risk of BOS (bronchiolitis
+obliterans syndrome), a progressive, chronic deterioration in lung
+function.  In this example, BOS was modelled as a discrete, staged
+process, a model of the form illustrated in Figure \ref{fig:disease},
+with 4 states.  State 1 represents absence of BOS. State 1 BOS is
+roughly defined as a sustained drop below 80\% below baseline FEV$_1$,
+while state 2 BOS is a sustained drop below 65\% baseline.  FEV$_1$ is
+measured as a percentage of a baseline value for each individual,
+determined in the first six months after transplant, and assumed to be
+100\% baseline at six months.
+
+As FEV$_1$ is subject to high short-term variability due to acute
+events and natural fluctuations, the exact BOS state at each
+observation time is difficult to determine.  Therefore, a hidden
+Markov model for FEV$_1$, conditionally on underlying BOS states, was
+used to model the natural history of the disease.  Discrete states are
+appropriate as onset is often sudden.
+
+\paragraph{Model 1}
+Jackson \cite{my:phd} considered models for these data where
+FEV$_1$ were Normally distributed, with an unknown mean and variance
+conditionally each state (\ref{eq:fev:normal}).  This model seeks the
+most likely location for the within-state FEV$_1$ means.
+
+\begin{equation}
+  \label{eq:fev:normal}
+  y_{ij} | \{ S_{ij} = k\}  \sim N(\mu_k + \beta x_{ij}, \sigma^2_k)
+\end{equation}
+
+\paragraph{Model 2}
+Jackson and Sharples \cite{jackson:sharples:2002} used a more complex
+two-level model for FEV$_1$ measurements. Level
+1 (\ref{eq:fev:level1}) represents the short-term fluctuation error of
+the marker around its underlying continuous value $y^{hid}_{ij}$.
+Level 2 (\ref{eq:fev:level2}) represents the distribution of
+$y^{hid}_{ij}$ conditionally on each underlying state, as follows.
+\begin{equation}
+  \label{eq:fev:level1}
+  y_{ij} | y^{hid}_{ij} \qquad \sim  N ( y^{hid}_{ij} + \beta x_{ij} ,
+  \sigma^2_\epsilon)
+\end{equation}
+\begin{equation}
+  \label{eq:fev:level2}
+  y^{hid}_{ij} | S_{ij} \quad \sim \quad
+  \left\{
+    \begin{array}{cll}
+      \mbox{State}& \mbox{Three state model}               & \mbox{Four state model} \\
+      S_{ij} = 0  & N(\mu_0, \sigma^2_0)I_{[80, \infty)} & N(\mu_0, \sigma^2_0)I_{[80, \infty)} \\
+      S_{ij} = 1  & N(\mu_1, \sigma^2_1)I_{(0, 80)}      & Uniform(65, 80)                        \\
+      S_{ij} = 2  & \mbox{(death)}                         & N(\mu_2, \sigma^2_2)I_{(0, 65)}      \\
+      S_{ij} = 3  &                                        & \mbox{(death)}
+    \end{array}
+  \right .
+\end{equation}
+
+Integrating over $y^{hid}_{ij}$ gives an explicit distribution for
+$y_{ij}$ conditionally on each underlying state (given in
+Section \ref{sec:fitting:hmm:general}, Table \ref{tab:hmm:dists}).
+Similar distributions were originally applied by Satten and
+Longini \cite{sattlong} to modelling the progression through discrete,
+unobserved HIV disease states using continuous CD4 cell counts.  The
+\Rpackage{msm} package includes density, quantile, cumulative density
+and random number generation functions for these distributions.
+
+In both models 1 and 2, the term $\beta x_{ij}$ models the short-term
+fluctuation of the marker in terms of acute events. $x_{ij}$ is an
+indicator for the occurrence of an acute rejection or infection
+episode within 14 days of the observation of FEV$_1$.
+
+Section \ref{sec:fitting:hmm:general} describes how these and more
+general hidden Markov models can be fitted with the \Rpackage{msm}
+package.
+
+\clearpage
+
+\section{Fitting multi-state models with {\tt msm}}
+
+<<echo=FALSE>>=
+options(width = 60)
+@
+
+\Rpackage{msm} is a package of functions for multi-state modelling using
+the R statistical software.  The \Rfunction{msm} function itself
+implements maximum-likelihood estimation for general multi-state
+Markov or hidden Markov models in continuous time. We illustrate
+its use with a set of data from monitoring heart transplant patients.
+Throughout this section ``\textsl{\texttt{>}}'' indicates the R
+command prompt, \textsl{\texttt{slanted typewriter}} text indicates R
+commands, and \texttt{typewriter} text R output.
+
+\subsection{Installing \tt{msm}}
+\label{sec:installing}
+
+The easiest way to install the \Rpackage{msm} package on a computer
+connected to the Internet is to run the R command:
+
+\begin{Scode}
+  install.packages("msm")
+\end{Scode}
+
+This downloads \Rpackage{msm} from the CRAN archive of contributed R
+packages (\texttt{cran.r-project.org} or one of its mirrors) and
+installs it to the default R system library.  To install to a
+different location, for example if you are a normal user with no
+administrative privileges, create a directory in which R packages are
+to be stored, say, \texttt{/your/library/dir}, and run
+
+\begin{Scode}
+  install.packages("msm", lib='/your/library/dir')
+\end{Scode}
+
+After \Rpackage{msm} has been installed, its functions can be made
+visible in an R session by
+<<>>=
+library(msm)
+@
+or, if it has been installed into a non-default library,
+
+\begin{Scode}
+  library(msm, lib.loc='/your/library/dir')
+\end{Scode}
+
+\subsection{Getting the data in}
+\label{sec:datain}
+
+The data are specified as a series of observations, grouped by
+patient. At minimum there should be a data frame with variables
+indicating
+\begin{itemize}
+\item the time of the observation,
+\item the observed state of the process.
+\end{itemize}
+If the data do not also contain
+\begin{itemize}
+\item the subject identification number,
+\end{itemize}
+then all the observations are assumed to be from the same subject.
+The subject ID does not need to be numeric, but data must be grouped
+by subject, and observations must be ordered by time within subjects.
+If the model includes variables with missing values, then the corresponding
+observations  are omitted by \Rfunction{msm} with a warning.  If you have missing data,
+as in any statistical model, it is recommended to ensure these do not
+result in biases.
+
+An example data set, taken from monitoring a set of heart transplant
+recipients, is provided with \Rpackage{msm}.  (Note: since \Rpackage{msm}
+version 1.3, the command \Rfunction{data(cav)} is no longer needed to
+load the data --- it is now ``lazy-loaded'' when required). Sharples
+\etal \cite{my:cav} studied the progression of coronary allograft
+vasculopathy (CAV), a post-transplant deterioration of the arterial
+walls, using these data.  Risk factors and the accuracy of the
+screening test were investigated using multi-state Markov and hidden
+Markov models.
+
+The first three patient histories are shown below. There are 622
+patients in all.  \Robject{PTNUM} is the subject identifier.
+Approximately each year after transplant, each patient has an
+angiogram, at which CAV can be diagnosed.  The result of the test is
+in the variable \Robject{state}, with possible values 1, 2, 3
+representing CAV-free, mild CAV and moderate or severe CAV
+respectively.  A value of 4 is recorded at the date of death.
+\Robject{years} gives the time of the test in years since the heart
+transplant. Other variables include \Robject{age} (age at screen),
+\Robject{dage} (donor age), \Robject{sex} (0=male, 1=female),
+\Robject{pdiag} (primary diagnosis, or reason for transplant - IHD
+represents ischaemic heart disease, IDC represents idiopathic dilated
+cardiomyopathy), \Robject{cumrej} (cumulative number of rejection
+episodes), and \Robject{firstobs}, an indicator which is 1 when the
+observation corresponds to the patient's transplant (the first
+observation), and 0 when the observation corresponds to a later
+angiogram.
+
+<<>>=
+cav[1:21,]
+@
+
+A useful way to summarise multi-state data is as a frequency table of
+pairs of consecutive states.  This counts over all individuals, for
+each state $r$ and $s$, the number of times an individual had an
+observation of state $r$ followed by an observation of state $s$. The
+function \Rfunction{statetable.msm} can be used to produce such a table,
+as follows,
+
+<<>>=
+statetable.msm(state, PTNUM, data=cav)
+@
+Thus there were 148 CAV-free deaths, 48 deaths from state 2, and
+55 deaths from state 3.  On only four occasions was there an
+observation of severe CAV followed by an observation of no CAV.
+
+\subsection{Specifying a model}
+\label{sec:specifying:model}
+
+We now specify the multi-state model to be fitted to the data.  A
+model is governed by a transition intensity matrix $Q$. For the heart
+transplant example, there are four possible states through which the
+patient can move, corresponding to CAV-free, mild/moderate CAV, severe
+CAV and death.  We assume that the patient can advance or recover from
+consecutive states while alive, and die from any state. Thus the model
+is illustrated by Figure \ref{fig:disease} with four states, and we
+have
+
+\[
+Q = \left(
+  \begin{array}{llll}
+    -(q_{12} + q_{14}) & q_{12} &  0     & q_{14}\\
+    q_{21} & -(q_{21}+q_{23}+q_{24}) & q_{23} & q_{24}\\
+      0    & q_{32} & -(q_{32}+q_{34}) & q_{34}\\
+      0    &   0    &   0    &   0   \\
+  \end{array}
+\right )
+\]
+
+It is important to remember that this defines which
+\emph{instantaneous} transitions can occur in the Markov process, and
+that the data are \emph{snapshots} of the process (see
+section \ref{sec:arbitr-observ-times}).  Although there were 44
+occasions on which a patient was observed in state 1 followed by state
+3, the underlying model specifies that the patient must have passed
+through state 2 in between.  If your data represent the exact and
+complete transition times of the process, then you must specify
+\Rfunarg{exacttimes=TRUE} or \Rfunarg{obstype=2} in the call to
+\Rfunction{msm}.
+
+To tell \Rfunction{msm} what the allowed transitions of our model are,
+we define a matrix of the same size as $Q$, containing zeroes in the
+positions where the entries of $Q$ are zero.  All other positions
+contain an initial value for the corresponding transition intensity.
+The diagonal entries supplied in this matrix do not matter, as the
+diagonal entries of $Q$ are defined as minus the sum of all the other
+entries in the row.  This matrix will eventually be used as the
+\Rfunarg{qmatrix} argument to the \Rfunction{msm} function. For
+example,
+
+<<>>=
+twoway4.q  <-  rbind ( c(0, 0.25, 0, 0.25),
+                       c(0.166, 0, 0.166, 0.166),
+                       c(0, 0.25, 0, 0.25),
+                       c(0, 0, 0, 0) )
+@
+
+Fitting the model is a process of finding values of the seven unknown
+transition intensities: $q_{12}$, $q_{14}$, $q_{21}$, $q_{23}$,
+$q_{24}$, $q_{32}$, $q_{34}$, which maximise the likelihood.
+
+\subsection{Specifying initial values}
+\label{sec:inits}
+
+The likelihood is maximised by numerical methods, which need a set of
+initial values to start the search for the maximum.  For reassurance
+that the true maximum likelihood estimates have been found, models
+should be run repeatedly starting from different initial
+values. However a sensible choice of initial values can be important
+for unstable models with flat or multi-modal likelihoods.  For
+example, the transition rates for a model with misclassification could
+be initialised at the corresponding estimates for an approximating
+model without misclassification. Initial values for a model without
+misclassification could be set by supposing that transitions between
+states take place only at the observation times.  If we observe
+$n_{rs}$ transitions from state $r$ to state $s$, and a total of $n_r$
+transitions from state $r$, then $q_{rs} / q_{rr}$ can be estimated by
+$n_{rs} / n_r$. Then, given a total of $T_r$ years spent in state $r$,
+the mean sojourn time $1 / q_{rr}$ can be estimated as $T_r / n_r$.
+Thus, $n_{rs} / T_r$ is a crude estimate of $q_{rs}$. The \Rpackage{msm}
+package provides a function \Rfunction{crudeinits.msm} for calculating
+initial values in this way (for non-hidden Markov models).
+
+<<>>=
+twoway4.crude.q  <- crudeinits.msm(state ~ years, PTNUM, data=cav, qmatrix=twoway4.q)
+@
+
+However, if there are are many changes of state in between the
+observation times, then this crude approach may fail to give sensible
+initial values.  For the heart transplant example we could also guess
+that the mean period in each state before moving to the next is about
+2 years, and there is an equal probability of progression, recovery or
+death.  This gives $q_{rr} = - 0.5$ for $r = 1, 2, 3$, and $q_{12} =
+q_{14} = 0.25$, $q_{21} = q_{23} = q_{24} = 0.166$, $q_{32} = q_{34} =
+0.25$, and the initial value matrix \Robject{twoway4.q} shown above,
+which we now use to fit the model.
+
+%% TODO explain about gen.inits=TRUE here
+
+\subsection{Running \Rfunction{msm}}
+\label{sec:running}
+
+To fit the model, call the \Rfunction{msm} function with the appropriate
+arguments.  For our running example, we have defined a data set
+\Robject{cav}, a matrix \Robject{twoway4.q} indicating the allowed
+transitions, and initial values. We are ready to run \Rfunction{msm}.
+
+\paragraph{Model 1: simple bidirectional model}
 

### 2014-01-01 pkg/inst/doc/msm-manual.Rnw
Build test again. Includes all non-hidden models.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@39 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+<<>>=
+cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
+                    qmatrix = twoway4.q, death = 4 )
+@
+
+In this example the day of death is assumed to be recorded exactly, as
+is usual in studies of chronic diseases.  At the previous instant
+before death the state of the patient is unknown.  Thus we specify
+\Rfunarg{death = 4}, to indicate to \Rfunction{msm} that the
+entry times into state 4 are observed in this manner.
+If the model had
+five states, and states 4 and 5 were two competing causes of death with
+times recorded exactly in this way, then we would specify \Rfunarg{death =
+  c(4,5)}.
+
+By default, the data are assumed to represent snapshots of the process
+at arbitrary times.  However, observations can also represent exact
+times of transition, ``exact death times'', or a mixture of these. See the
+\Rfunarg{obstype} argument to \Rfunction{msm}.  Note you do not always have to specify \Rfunarg{death},
+even if there are states which correspond to death.  This is
+only used for states whose entry time is observed in this specific way.
+
+While the \Rfunction{msm} function runs, it searches for the maximum
+of the likelihood of the unknown parameters.  Internally, it uses the
+R function \Rfunction{optim} to minimise the minus log-likelihood.
+When the data set, the model, or both, are large, then this may take a
+long time.  It can then be useful to see the progress of the
+optimisation algorithm. To do this, we can specify a \Rfunarg{control}
+argument to \Rfunction{msm}, which is passed internally to the
+\Rfunction{optim} function.  For example \texttt{control = list(trace=1,
+  REPORT=1)}. See the help page for \Rfunction{optim},
+
+<<eval=FALSE>>=
+help(optim)
+@
+for more options to control the optimisation.  When completed, the
+\Rfunction{msm} function returns a value.  This value is a list of the
+important results of the model fitting, including the parameter
+estimates and their covariances.  To keep these results for
+post-processing, we store them in an R object, here called
+\Robject{cav.msm}.  When running several similar \Rfunction{msm} models,
+it is recommended to store the respective results in
+informatively-named objects.
+
+\subsection{Showing results}
+
+To show the maximum likelihood estimates and 95\% confidence
+intervals, type the name of the fitted model object at the R command
+prompt.  \footnote{This is equivalent to typing
+  \texttt{print.msm(cav.msm)}.  The function \Rfunction{print.msm}
+  formats the important information in the model object for printing
+  on the screen.}  The confidence level can be changed using the
+\Rfunarg{cl} argument to \Rfunction{msm}.
+
+<<>>=
+cav.msm
+@
+
+From the estimated intensity matrix, we see patients are three times
+as likely to develop symptoms than die without symptoms (first row).
+After disease onset (state 2), progression to severe symptoms (state
+3) is 50\% more likely than recovery.  Once in the severe state, death
+is more likely than recovery, and a mean of 1 / -0.44 = 2.3 years is
+spent in state 3 before death or recovery.
+
+Section \ref{sec:extractor} describes various functions that can be
+used to obtain summary information from the fitted model.
+
+\subsection{Covariates on the transition rates}
+\label{sec:msm:covariates}
+
+We now model the effect of explanatory variables on the rates of
+transition, using a proportional intensities model. Now we have an
+intensity matrix $Q(z)$ which depends on a covariate vector $z$. For
+each entry of $Q(z)$, the transition intensity for patient $i$ at
+observation time $j$ is $q_{rs}(z_{ij}) = q_{rs}^{(0)}
+\exp(\beta_{rs}^T z_{ij})$.  The covariates $z$ are specified through
+the \Rfunarg{covariates} argument to \Rfunction{msm}. If $z_{ij}$ is
+time-dependent, we assume it is constant in between the observation
+times of the Markov process.  \Rfunction{msm} calculates the
+probability for a state transition from times $t_{i,j-1}$ to $t_{ij}$
+using the covariate value at time $t_{i,j-1}$.
+
+We consider a model with just one covariate, female sex.  Out of the
+622 transplant recipients, 535 are male and 87 are female.  By
+default, all linear covariate effects $\beta_{rs}$ are initialised to
+zero. To specify different initial values, use a \Rfunarg{covinits}
+argument, as described in \Rfunction{help(msm)}.  Initial values given
+in the \Rfunarg{qmatrix} represent the intensities with covariate values
+set to their means in the data.  In the following model, all transition
+intensities are modelled in terms of sex.
+
+\paragraph{Model 2: sex as a covariate}
+<<>>=
+cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
+                     qmatrix = twoway4.q, death = 4, covariates = ~ sex)
+@
+
+The \Robject{msm} object will now include the estimated covariate
+effects and their confidence intervals.
+
+<<>>=
+cavsex.msm
+@
+Comparing the estimated log-linear effects of sex and their standard
+errors, we see that the disease onset rate is smaller for females,
+whereas none of the other effects are significant. The first matrix shown in the output of printing
+\Robject{cavsex.msm} is the estimated transition intensity matrix
+$q_{rs}(z) = q_{rs}^{(0)} \exp(\beta_{rs}^T z)$ with the covariate $z$
+set to its mean value in the data.  This represents an average
+intensity matrix for the population of 535 male and 87 female
+patients. To extract separate intensity matrices for male and female
+patients ($z = 0$ and $1$ respectively), use the function
+\Rfunction{qmatrix.msm}, as shown below.  This and similar summary
+functions will be described in more detail in
+section \ref{sec:extractor}.
+
+<<>>=
+qmatrix.msm(cavsex.msm, covariates=list(sex=0)) # Male
+qmatrix.msm(cavsex.msm, covariates=list(sex=1)) # Female
+@
+
+Different transition rates may be easily modelled on different
+covariates by specifying a named list of formulae as the
+\Rfunarg{covariates} argument.  Each element of the list has a name
+identifying the transition.  In the model below, the transition rate
+from state 1 to state 2 and the rate from state 3 to state 4 are each
+modelled on sex as a covariate, but no other intensities have
+covariates on them.
+
+\paragraph{Model 2a: transition-specific covariates}
+<<eval=FALSE>>=
+cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
+                     qmatrix = twoway4.q, death = 4,
+                     covariates = list("1-2" = ~ sex, "3-4" = ~sex) )
+@
+
+We may also want to constrain the effect of a covariate to be equal for
+certain transition rates, to reduce the number of parameters in the
+model, or to investigate hypotheses on the covariate effects.  A
+\Rfunarg{constraint} argument can be used to indicate which of the
+transition rates have common covariate effects.
+
+\paragraph{Model 3: constrained covariate effects}
+
+<<eval=FALSE>>=
+cav3.msm <- msm( state ~ years, subject=PTNUM, data = cav,
+                   qmatrix = twoway4.q, death = 4,
+                   covariates = ~ sex, constraint = list(sex=c(1,2,3,1,2,3,2)) )
+@
+
+This constrains the effect of sex to be equal for the progression
+rates $q_{12}, q_{23}$, equal for the death rates $q_{14}, q_{24},
+q_{34}$, and equal for the recovery rates $q_{21}, q_{32}$. The
+intensity parameters are assumed to be ordered by reading across the
+rows of the transition matrix, starting at the first row: ($q_{12},
+q_{14}, q_{21}, q_{23}, q_{24}, q_{32}, q_{34}$), giving constraint
+indicators \Rfunarg{(1,2,3,1,2,3,2)}.  Any vector of increasing
+numbers can be used for the indicators.  Negative entries can be used
+to indicate that some effects are equal to minus others:
+\Rfunarg{(1,2,3,-1,2,3,2)} sets the fourth effect to be minus the
+first.
+
+In a similar manner, we can constrain some of the baseline transition
+intensities to be equal to one another, using the
+\Rfunarg{qconstraint} argument.  For example, to constrain the rates
+$q_{12}$ and $q_{23}$ to be equal, and $q_{24}$ and $q_{34}$ to be
+equal, specify \Rfunarg{qconstraint = c(1,2,3,1,4,5,4)}.
+
+\subsection{Fixing parameters at their initial values}
+
+For exploratory purposes we may want to fit a model assuming that some
+parameters are fixed, and estimate the remaining parameters.  This may
+be necessary in cases where there is not enough information in the
+data to be able to estimate a proposed model, and we have strong prior
+information about a certain transition rate. To do this, use the
+\Rfunarg{fixedpars} argument to \Rfunction{msm}.  For model 1, the
+following statement fixes the parameters numbered 6, 7, that is,
+$q_{32}$, $q_{34}$, to their initial values (0.25 and 0.25,
+respectively).
+
+\paragraph{Model 4: fixed parameters}
+<<eval=FALSE>>=
+cav4.msm <- msm( state ~ years, subject=PTNUM, data = cav,
+                qmatrix = twoway4.q, death = 4,
+                control = list(trace=2, REPORT=1),
+                fixedpars = c(6, 7) )
+@
+
+A \Rfunarg{fixedpars} statement can also be useful for fixing covariate
+effect parameters to zero, that is to assume no effect of a covariate
+on a certain transition rate.
+
+\subsection{Extractor functions}
+\label{sec:extractor}
+
+We may want to extract some of the information from the \Rfunction{msm}
+model fit for post-processing, for example for plotting graphs or
+generating summary tables.  A set of functions is provided for
+extracting interesting features of the fitted model.
+
+\begin{description}
+
+\item[Intensity matrices] The function \Rfunction{qmatrix.msm}
+  extracts the estimated transition intensity matrix and its confidence intervals
+  for a given set of covariate values, as shown in
+  section \ref{sec:msm:covariates}.  Confidence intervals are
+  calculated from the covariance matrix of the estimates by assuming
+  the distribution is symmetric on the log scale.  Standard errors for
+  the intensities are also available from the object returned by
+  \Rfunction{qmatrix.msm}.  These are calculated by the delta method.
+  The \Rpackage{msm} package provides a general-purpose function
+  \Rfunction{deltamethod} for estimating the variance of a function of
+  a random variable $X$ given the expectation and variance of $X$. See
+  \texttt{help(deltamethod)} for further details.  Bootstrap
+  confidence intervals are also available for \Rfunction{qmatrix.msm}
+  and for most output functions; these are often more accurate, at the
+  cost of computational time.  For more about bootstrapping in
+  \Rpackage{msm}, see Section \ref{sec:boot}.
+
+\item[Transition probability matrices] The function
+  \Rfunction{pmatrix.msm} extracts the estimated transition probability
+  matrix $P(t)$ within a given time. For example, for model 1, the 10
+  year transition probabilities are given by:
+
+<<>>=
+pmatrix.msm(cav.msm, t=10)
+@
+
+  Thus, a typical person in state 1, disease-free, has a probability
+  of 0.5 of being dead ten years from now, a probability of 0.3 being
+  still disease-free, and probabilities of 0.1 of being alive with
+  mild/moderate or severe disease, respectively.
+
+  This assumes $Q$ is constant within the desired time interval.  For
+  non-homogeneous processes, where $Q$ varies with time-dependent
+  covariates but can be approximated as piecewise constant, there is
+  an equivalent function \Rfunction{pmatrix.piecewise.msm}.  Consult its
+  help page for further details.
+
+  If \Rfunarg{ci=''norm''} is specified, then a confidence interval is
+  calculated based on drawing a random sample (default size 1000) from
+  the assumed multivariate normal distribution of the maximum
+  likelihood estimates and covariance matrix, and transforming.  If
+  \Rfunarg{ci=''boot''} is specified, then a bootstrap confidence
+  interval for the transition probability matrix is calculated (see Section \ref{sec:boot}) .
+  However, both of these are computationally intensive, particularly
+  the bootstrap method, so no confidence interval is calculated by default.
+
+\item[Mean sojourn times] The function \Rfunction{sojourn.msm}
+  extracts the estimated mean sojourn times in each transient state
+  $r$, for a given set of covariate values. This is calculated as $-1
+  / \hat q_{rr}$, where $\hat q_{rr}$ is the $r$th diagonal entry of
+  the estimated transition intensity matrix.
+
+<<>>=
+sojourn.msm(cav.msm)
+@
+
+\item[Probability that each state is next] The function
+  \Rfunction{pnext.msm} extracts the matrix of probabilities $-q_{rs}
+  / q_{rr}$ that the next state after state $r$ is state $s$, for each
+  $r$ and $s$.  Together with the mean sojourn times, this gives a more
+  intuitive parameterisation of a continuous-time Markov model than
+  the raw transition intensities $q_{rs}$.  Note these are different from
+  the transition probabilities in a given time $t$ returned by
+  \Rfunction{pmatrix.msm}.
+
+<<>>=
+pnext.msm(cav.msm)
+@
+
+\item[Total length of stay] Mean sojourn times describe the
+  average period in a single stay in a state. For processes with
+  successive periods of recovery and relapse, we may want to
+  forecast the total time spent healthy or diseased, before
+        death.  The function \Rfunction{totlos.msm} estimates the
+        forecasted total length of time spent in each transient state
+        $s$ between two future time points $t_1$ and $t_2$, for a
+        given set of covariate values. This defaults to the expected
+        amount of time spent in each state between the start of the
+        process (time 0, the present time) and death or a specified
+        future time. This is obtained as
+\[
+L_s = \int_{t_1}^{t_2} P(t)_{r,s} dt
+\]
+where $r$ is the state at the start of the process, which defaults to
+1. This is calculated using numerical integration.  For model 1, each
+patient is forecasted to spend 8.8 years disease free, 2.2 years with
+mild or moderate disease and 1.8 years with severe disease.
+
+Bootstrap and asymptotic confidence intervals are available, as for
+\Rfunction{pmatrix.msm}, but are not calculated by default.
+
+<<>>=
+totlos.msm(cav.msm)
+@
+
+\item[Expected first passage times] The function \Rfunction{efpt.msm} estimates
+  the expected time until the process first enters a given state or set of states,
+  also called the ``hitting time''.  See its help page for further details.
+
+\item[Ratio of transition intensities] The function
+  \Rfunction{qratio.msm} estimates a ratio of two entries of the
+  transition intensity matrix at a given set of covariate values,
+  together with a confidence interval estimated assuming normality on
+  the log scale and using the delta method.  For example, we may want
+  to estimate the ratio of the progression rate $q_{12}$ into the
+  first state of disease to the corresponding recovery rate $q_{21}$.
+  For example in model 1, recovery is 1.8 times as likely as
+  progression.
+
+<<>>=
+qratio.msm(cav.msm, ind1=c(2,1), ind2=c(1,2))
+@
+
+\item[Hazard ratios for transition] The function \Rfunction{hazard.msm}
+  gives the estimated hazard ratios corresponding to each covariate
+  effect on the transition intensities. 95\% confidence limits are
+  computed by assuming normality of the log-effect. For example, for
+  model 2 with female sex as a covariate, the following hazard ratios
+  show more clearly that the only transition on which the effect of
+  sex is significant at the 5\% level is the 1-2 transition.
+
+<<>>=
+hazard.msm(cavsex.msm)
+@
+
+\end{description}
+
+\paragraph{Setting covariate values}
+All of these extractor functions take an argument called
+\Rfunarg{covariates}.  If this argument is omitted, for example,
+
+<<eval=FALSE>>=
+qmatrix.msm(cav.msm)
+@
+then the intensity matrix is evaluated as $Q(\bar x)$ with all
+covariates set to their mean values $\bar x$ in the data.
+Alternatively, set \Rfunarg{covariates} to 0 to return the result
+$Q(0)$ with covariates set to zero. This will usually be
+preferable for categorical covariates, where we wish to see the
+result for the baseline category.
+
+<<eval=FALSE>>=
+qmatrix.msm(cavsex.msm, covariates = 0)
+@
+Alternatively, the desired covariate values can be specified
+explicitly as a list,
+
+<<eval=FALSE>>=
+qmatrix.msm(cavsex.msm, covariates = list(sex = 1))
+@
+
+Values of categorical covariates must be quoted. For example, consider
+a covariate \texttt{smoke}, representing tobacco smoking status, with
+three levels, \texttt{NON, CURRENT, EX}, representing a non-smoker,
+current smoker or ex-smoker.
+
+\begin{Scode}
+qmatrix.msm(example.msm, covariates = list(age = 60, smoke=''CURRENT''))
+\end{Scode}
+
+\subsection{Survival plots}
+
+In studies of chronic disease, an important use of multi-state models
+is in predicting the probability of survival for patients in
+increasingly severe states of disease, for some time $t$ in the
+future.  This can be obtained directly from the transition probability
+matrix $P(t)$.
+
+The \Rfunction{plot} method for \Robject{msm} objects produces a plot
+of the expected probability of survival against time, from each
+transient state. Survival is defined as not entering the final
+absorbing state.
+
+<<fig=TRUE>>=
+plot(cav.msm, legend.pos=c(8, 1))
+@
+
+This shows that the 10-year survival probability with severe CAV is
+approximately 0.1, as opposed to 0.3 with mild CAV and 0.5 without
+CAV.  With severe CAV the survival probability diminishes very quickly
+to around 0.3 in the first five years after transplant.  The
+\Rfunarg{legend.pos} argument adjusts the position of the legend in
+case of clashes with the plot lines.  A \Rfunarg{times} argument can
+be supplied to indicate the time interval to forecast survival for.
+
+A more sophisticated analysis of these data might explore competing
+causes of death from causes related or unrelated to the disease under
+study.
+
+\subsection{Bootstrapping}
+\label{sec:boot}
+Most of \Rpackage{msm}'s output functions present confidence intervals
+based on asymptotic standard errors calculated from the Hessian, or
+transformations of these using the delta method.  The asymptotic
+standard errors are expected to be underestimates of the true standard
+errors (Cramer-Rao lower bound).  For some output functions, such as
+\Rfunction{pmatrix.msm}, and functions based on
+\Rfunction{pmatrix.msm} such as \Rfunction{totlos.msm} and
+\Rfunction{prevalence.msm}, the delta method cannot be used at all to
+obtain standard errors.  In these cases, confidence intervals can be
+calculated by drawing a random sample from the assumed multivariate
+normal distribution of the maximum likelihood estimates and covariance
+matrix, and transforming.  However, this is still based on potentially
+inaccurate asymptotic theory.  The \Rpackage{msm} package provides the
+function \Rfunction{boot.msm} to enable bootstrap refitting of
+\Rfunction{msm} models, an alternative way to estimate uncertainty.
+
+For non-hidden Markov models, a bootstrap dataset is drawn by
+resampling pairs of consecutive states from the full data, i.e.
+\emph{transitions}.  These are assumed to be independent when
+calculating the likelihood (Section \ref{sec:multi:likelihood}).  For
+hidden Markov models and models with censoring, a bootstrap dataset is
+drawn by resampling complete series from independent subjects.  The
+bootstrap datasets have the same number of transitions, or subjects,
+respectively, as the original data.
+
+For most output extractor functions provided with \Rpackage{msm}, the
+option \Rfunarg{ci=''boot''} is available, as a wrapper around
+\Rfunction{boot.msm}, to enable bootstrap confidence intervals to be calculated.
+But any user-defined output statistic can be bootstrapped, as follows.
+The function \Rfunction{boot.msm} is called with the fitted
+\Rfunction{msm} model as first argument, and an R function specifying
+the statistic to be bootstrapped as the second argument \texttt{stat}.
+The return value from \Rfunction{boot.msm} is a list of \texttt{B}
+replicates (by default, \texttt{B=1000}) of the desired statistic.  For
+example, to bootstrap the transition intensity matrix of the heart
+transplantation model \Robject{cav.msm}:
+\begin{Scode}
+  q.list <- boot.msm(cav.msm, stat=function(x){qmatrix.msm(x)$estimates})
+\end{Scode}
+Note that for \Rfunction{boot.msm} to be able to refit the original
+model that produced \Robject{cav.msm}, all objects used in the
+original model fit (for example, in this case, \Robject{twoway4.q})
+must be in the working environment.  Otherwise, \Rfunction{boot.msm}
+will give an ``object not found'' error.
+
+The user can then summarise these replicates by calculating empirical
+standard deviations or quantile-based intervals.  In this example,
+\Robject{q.list} is a list of 1000 4$\times$4 matrices.  The following
+code calculates the bootstrap standard error as the empirical standard
+deviation of the 1000 replicates, and a similar 95\% bootstrap
+confidence interval.
+\begin{Scode}
+  q.array <- array(unlist(q.list), dim=c(4,4,1000))
+  apply(q.array, c(1,2), sd)
+  apply(q.array, c(1,2), function(x)quantile(x, c(0.025, 0.975)))
+\end{Scode}
+
+Note that when bootstrapping, the refits of the model to the resampled
+datasets may occasionally fail to converge (as discussed in
+Section \ref{sec:failure}) even if the original model fit did
+converge.  In these cases, a warning is given, but
+\Rfunction{boot.msm} simply discards the failed dataset and moves on
+to the next bootstrap iteration. Unless convergence failure occurs for
+a large proportion of iterations, this should not affect the accuracy
+of the final bootstrap confidence intervals.
+
+\subsection{Convergence failure}
+\label{sec:failure}
+
+Inevitably if over-complex models are applied with insufficient data
+then the parameters of the model will not be identifiable.  This will
+result in the optimisation algorithm failing to find the maximum of
+the log-likelihood, or even failing to evaluate the likelihood. For
+example, it will commonly be inadvisable to include several covariates
+in a model simultaneously.
+
+In some circumstances, the optimisation may report convergence, but
+fail to calculate any standard errors.  In these cases, the Hessian of
+the log-likelihood at the reported solution is not positive definite.
+Thus the reported solution may be a saddle point rather than the
+maximum likelihood, or it may be close to the maximum.
+
+\begin{description}
+\item[Initial values] Make sure that a sensible set of initial values
+  have been chosen. The optimisation may only converge within a
+  limited range of `informative' initial values.  It is also sensible
+  to run the model for several different initial values to ensure that
+  the estimation has converged to a global rather than a local optimum.
+
+\item[Scaling] It is often necessary to apply a scaling factor to
+  normalise the likelihood (\Rfunarg{fnscale}), or certain individual
+  parameters \Rfunarg{(parscale)}.  This may prevent overflow or
+  underflow problems within the optimisation.  For example, if the value
+  of the -2 $\times$ log-likelihood is around 5000, then the following
+  option leads to an minimisation of the -2 $\times$ log-likelihood on
+  an approximate unit scale: \Rfunarg{control = list(fnscale = 5000)}
+
+  It is also advisable to analyse all variables, including covariates
+  and the time unit, on a roughly normalised scale. For example,
+  working in terms of a time unit of months or years instead of days,
+  when the data range over thousands of days.
+
+\item[Convergence criteria] ``False convergence'', in which
+  \Rfunction{optim()} reports convergence of the optimisation but the
+  Hessian is not positive definite, can sometimes be solved by
+  tightening the criteria (\Rfunarg{reltol}, defaults to
+  \texttt{1e-08}) for reporting convergence. For example,
+  \Rfunarg{control = list(reltol = 1e-16)}.
+
+  Alternatively consider using smaller step sizes for the numerical
+  approximation to the gradient, used in calculating the Hessian.
+  This is given by the control parameter \Rfunarg{ndeps}. For example,
+  for a model with 5 parameters, \Rfunarg{control = list(ndeps =
+    rep(1e-6, 5))}
+
+\item[Choice of algorithm] By default, \Rfunction{msm} uses a
+  Nelder-Mead algorithm which does not use analytic derivatives of the
+  objective function.   For potentially greater speed and accuracy,
+  consider using a quasi-Newton algorithm such as
+  the ``BFGS'' method of \Rfunction{optim}, which can make use of
+  analytic derivatives, for example, \Rfunarg{method = ``BFGS'', use.deriv = TRUE}.
+  Or use the \Rfunction{nlm} R function via \Rfunarg{msm(..., opt.method = "nlm" , ...)}
+  Note also the Fisher scoring method available for non-hidden Markov
+  models for panel data, via  \Rfunarg{msm(..., opt.method = "fisher", ...)},
+  is expected to be faster than the generic methods, but less robust to
+  bad initial values.
+
+\item[\Rfunction{optim} "function cannot be evaluated at initial parameters"]
+  To diagnose this problem, run \Rfunction{msm} again with
+  \Rfunarg{fixedpars=TRUE} set, to calculate the -2 log-likelihood at
+  the initial values.  This will probably be \Robject{Inf}. To show
+  the contributions of individual subjects to the overall log
+  likelihood, call \Rfunction{logLik.msm(x, by.subject=TRUE)}, where
+  \Robject{x} is the fitted model object.  If only a few subjects give
+  an infinite log-likelihood, then you can check whether their state
+  histories are particularly unusual and conflict with the model.  For
+  example, they might appear to make unusually large jumps between
+  states in short periods of time.  For models with misclassification,
+  note that the default true initial state distribution
+  \Rfunarg{initprobs} puts all individuals in true state 1 at their
+  first observation.  If someone starts in a much higher state, this
+  may result in an infinite log-likelihood, and changing
+  \Rfunarg{initprobs} would be sensible.
+
+\item[Model simplification]
+
+  If none of these numerical adjustments lead to convergence, then the
+  model is probably over-complicated.  There may not be enough
+  information in the data on a certain transition rate.  It is always
+  recommended to count all the pairs of transitions between states in
+  successive observation times, making a frequency table of previous
+  state against current state (function \Rfunction{statetable.msm}).
+  Although the data are a series of snapshots of a continuous-time
+  process, and the actual transitions take place in between the
+  observation times, this type of table may still be helpful.  If
+  there are not many observed `transitions' from state 2 to state 4,
+  for example, then the data may be insufficient to estimate $q_{24}$.
+
+  For a staged disease model (Figure \ref{fig:disease}), the number of
+  disease states should be low enough that all transition rates can be
+  estimated.  Consecutive states of disease severity should be merged
+  if necessary.  If it is realistic, consider applying constraints on
+  the intensities or the covariate effects so that the parameters are
+  equal for certain transitions, or zero for certain transitions.
+
+  Be careful to use a observation scheme and transition matrix
+  appropriate to your data (see
+  Section \ref{sec:arbitr-observ-times}). By default, \Rfunction{msm}
+  assumes that the data represent snapshots of the process, and the
+  true state is unknown between observation times.  In such
+  circumstances, it is rarely feasible to estimate an intensity matrix
+  with \emph{instantaneous} transitions allowed between every pair of
+  states.  This would be easier if the complete course of the process
+  is known \Rfunarg{(exacttimes = TRUE)} in the call to
+  \Rfunction{msm}.  Understand the difference between
+  \emph{instantaneous} and \emph{interval} transitions - although
+  individuals may be in state 1 at time $t_r$, and state 3 at time
+  $t_{r+1}$, that doesn't mean that instantaneous transitions from 1
+  to 3 should be permitted.
+
+\end{description}
+
+\subsection{Model assessment}
+\label{sec:model-assessment}
+
+\paragraph{Observed and expected prevalence}
+
+To compare the relative fit of two nested models, it is easy to
+compare their likelihoods.  However it is not always easy to determine
+how well a fitted multi-state model describes an irregularly-observed
+process.  Ideally we would like to compare observed data with fitted
+or expected data under the model.  If there were times at which all
+individuals were observed then the fit of the expected numbers in each
+state or {\em prevalences} can be assessed directly at those times.
+Otherwise, some approximations are necessary.  We could assume that an
+individual's state at an arbitrary time $t$ was the same as the state
+at their previous observation time.  This might be fairly accurate if
+observation times are close together.  This approach is taken by the
+function \Rfunction{prevalence.msm}, which constructs a table of observed
+and expected numbers and percentages of individuals in each state at a
+set of times.
+
+A set of expected counts can be produced if the process begins at a
+common time for all individuals.  Suppose at this time, each
+individual is in state 0.  Then given $n(t)$ individuals are under
+observation at time $t$, the expected number of individuals in state
+$r$ at time $t$ is $n(t) P(t)_{0,r}$.   If the covariates on which
+$P(t)$ depends vary between individuals, then this can be
+averaged over the covariates observed in the data.
+
+For example, we calculate the observed and expected numbers and
+percentages at two-yearly intervals up to 20 years after transplant,
+for the heart transplant model \Rfunction{cav.msm}.  The number of
+individuals still alive and under observation decreases from 622 to
+251 at year 20.  The observed and expected percentages are plotted
+against time.
+
+<<>>=
+options(digits=3)
+prevalence.msm(cav.msm, times=seq(0,20,2))
+@
+<<fig=TRUE>>=
+plot.prevalence.msm(cav.msm, mintime=0, maxtime=20)
+@
+
+Comparing the observed and expected percentages in each state,
+we see that the predicted number of individuals who die (State 4) is
+under-estimated by the model from about year 8 onwards.  Similarly the
+number of individuals sill alive and free of CAV (State 1) is
+over-estimated by the model from about year 8 onwards.
+
+Such discrepancies could have many causes.  One possibility is that
+the transition rates vary with the time since the beginning of the
+process, the age of the patient, or some other omitted covariate, so
+that the Markov model is {\em non-homogeneous}.  This could be
+accounted for by modelling the intensity as a function of age, for
+example, such as a piecewise-constant function.  The \Rfunarg{pci}
+argument to \Rfunction{msm} can be used to automatically construct
+models with transition intensities which are piecewise-constant in
+time.
+
+In this example, the hazard of death may increase with age, so that
+the model underestimates the number of deaths when forecasting far
+into the future.  Another cause of poor model fit may sometimes be the
+failure of the Markov assumption.  That is, the transition intensities
+may depend on the time spent in the current state (a semi-Markov
+process) or other characteristics of the process history.  Accounting
+for the process history is difficult as the process is only observed
+through a series of snapshots.  For a multi-state model with one-way
+progression through states, and frequent observations, we may be able
+to estimate the time spent in each state by each individual.
+
+However, if it is known that individuals who died would not have been
+followed up after a certain time, had they survived to that time, then
+they should not be included in the observed prevalence of the death
+state after that time.  This can be accounted for by passing a vector
+of maximum potential follow-up times, one for each individual in the
+same order as the original data, in the \Rfunarg{censtime} argument to
+\Rfunction{prevalence.msm}.  Ignoring the potential follow-up times is
+likely to have resulted in overestimates of the number of deaths
+at later times in the CAV example, though these times are not
+available in the data supplied with \Rpackage{msm}.
+
+\paragraph{Pearson-type goodness-of-fit test}
+\label{sec:pearson}
+
+Suppose that the true transition times are unknown, and data consist
+of observations of the process at arbitrary times which differ between
+individuals (panel data).  Assessing goodness of fit by prevalence
+counts then involves estimating the observed prevalence at a series of
+points by some form of interpolation. This is only advisable if
+observation times are close together.  An alternative method of
+assessing goodness-of-fit is to construct tables of observed and
+expected numbers of transitions, as described by Aguirre-Hernandez and
+Farewell \cite{ahf}.  This leads to a formal test of goodness-of-fit,
+analogous to the classical Pearson $\chi^2$ test for contingency
+tables. The tables are constructed as follows.  Each pair of
+successive observations in the data (\emph{transition}) is classified
+by
+\begin{itemize}
+\item the starting state $r$ and finishing state $s$,
+\item time between the start of the process and the first of the pair
+  of observations (indexed by $h$),
+\item time interval between the observations (indexed by $l_h$, within
+  categories $h$),
+\item (if there are fitted covariates) the impact of covariates, as
+  summarised by $q_{rr}$ (indexed by $c$),
+\item any other grouping of interest for diagnosing lack of fit
+  (indexed by $g$).
+\end{itemize}
+Groupings of continuous quantities are normally defined by quantiles,
+so that there are a similar number of observed transitions in each
+(one-dimensional) category.  The observed and expected numbers of
+transitions in each group are then defined by
+\[
+o_{hl_h rscg} = \sum I(S(t_{i,j+1}) = s, S(t_{ij}) = r)
+\]
+\[
+e_{hl_h rscg} = \sum P(S(t_{i,j+1}) = s | S(t_{ij}) = r)
+\]
+where $I(A)$ is the indicator function for an event $A$ and the
+summation is over the set of transitions in the category defined by
+$h,l_h,c,g$, over all individuals $i$. The Pearson-type test statistic
+is then
+\[
+T = \sum_{hl_h rscg} \frac{(o_{hl_h rscg} - e_{hl_h rscg})^2}{e_{hl_h rscg}}
+\]
+The classical Pearson test statistic is distributed as $\chi^2_{n-p}$,
+where $n$ is the number of independent cells in the table and $p$ is
+the number of estimated parameters $p$. But the null distribution of
+$T$ is not exactly $\chi^2$, since the time intervals are
+non-identical, therefore the observed transitions are realizations
+from a set of independent but non-identical multinomial distributions.
+Titman \cite{titman:asympnull} showed that the null distribution
+of $T$ is asymptotically approximated by a weighted sum of $\chi^2_1$
+random variables.   Aguirre-Hernandez and Farewell \cite{ahf} also showed that $\chi^2_{n-p}$
+is a good approximation if there are no fitted covariates. For models
+with covariates, the null mean of $T$ is higher than $n - p$, but
+lower than $n$.  Therefore, upper and lower bounds for the true
+$p$-value of the statistic can be obtained from the $\chi^2_{n-p}$ and
+$\chi^2_n$ distributions.  Aguirre-Hernandez and Farewell \cite{ahf}
+also described a bootstrap procedure for obtaining an accurate $p$-value.
+
+Titman and Sharples \cite{titman:sharples} described modifications to
+the test to correct for the biases introduced where in addition to the
+panel-data observation scheme:
+\begin{itemize}
+\item Times of death are known exactly.  In this case, transitions
+  ending in death are classified according to the next scheduled
+  observation time after the death, which is estimated by multiple
+  imputation from a Kaplan-Meier estimate of the distribution of time
+  intervals between observations.
+
+\item An individual's final observation is censored, so that they are
+  only known to be alive at that point.
+
+\item States are misclassified.
+\end{itemize}
+
+The \Rpackage{msm} package provides the function
+\Rfunction{pearson.msm} to perform the Pearson-type test.  By default,
+three groups are used for each of $h$, $l_h$ and $c$.  Often the
+number of groups will need to be reduced in cases where the resulting
+contingency tables are sparse (thus there are several low expected
+counts and the variance of $T$ is inflated).
+
+The test is now performed on the model \Robject{cav.msm} for the heart
+transplant dataset (a version of which was also analysed by
+Titman and Sharples \cite{titman:sharples}). The default three interval groups are used,
+and two groups of the time since the start of the process. The
+\Rfunarg{transitions} argument groups the transitions from state 3 to
+each of states 1, 2 and 3 (the 9th, 10th and 11th transitions)
+together in the table, since these transitions are infrequent.
+
+<<>>=
+options(digits=2)
+pearson.msm(cav.msm, timegroups=2, transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
+@
+
+The first two tables in the output show the contingency tables of
+observed and expected numbers of transitions.  Note that the observed
+number of transitions in certain categories is not a whole number,
+since these are averaged over multiple imputations of the next
+scheduled observation time following deaths.  The column \texttt{Time}
+is the group defined by time since the start of the process, and the
+column \texttt{Interval} is the group defined by intervals between
+observations.  The columns indicate the allowed transitions, or pairs
+of states which can be observed in successive observations.
+
+The third table presents the ``deviance'', the value of
+$\frac{(o_{hl_h rscg} - e_{hl_h rscg})^2}{e_{hl_h rscg}}$ for each
+cell, multipled by the sign of $o_{hl_h rscg} - e_{hl_h rscg}$ to
+indicate whether there were more or fewer transitions than expected in
+each cell.  These can indicate areas of bad fit.  For example,
+systematic changes in deviance by time or time interval between
+observations can indicate that a model with time-varying transition
+intensities is more suitable.  Changes in deviance by covariate impact
+may indicate heterogeneity between individuals which is unexplained
+by the fitted covariates.  Changes in deviance with the length of the
+interval between observations may also indicate failure of the Markov
+assumption, and that a semi-Markov model (in which intensities depend
+on the time spent in the current state) may fit better.
+
+In this example, the test statistic is 100.  \Robject{p.upper} is an
+upper bound for the $p$-value of the statistic based on an asymptotic
+$\chi^2_{42}$ distribution, therefore the model does not fit well.  It
+is not clear from the table of deviances which aspects of the fit are
+most influental to the test statistic.  However, the two-way Markov
+model itself is not biologically plausible, as discussed in
+Section \ref{sec:fitting:hmm:misc}.
+
+For non-hidden Markov models for panel data, \Rfunction{pearson.msm}
+also presents the accurate analytic p-value of
+Titman \cite{titman:asympnull}.  For all models,
+\Rfunction{pearson.msm} provides an option for parametric
+bootstrapping to obtain an accurate p-value.
+
 \end{document}

### 2014-01-01 pkg/inst/doc/msm-manual.Rnw
Build test again, up to misc models with covariate
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@40 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+\subsection{Fitting misclassification models with \Rpackage{msm}}
+\label{sec:fitting:hmm:misc}
+
+In fact, in the heart transplant example from
+section \ref{sec:datain}, it is not medically realistic for patients
+to recover from a diseased state to a healthy state.  Progression of
+coronary artery vasculopathy is thought to be an irreversible process.
+The angiography scan for CAV is actually subject to error, which leads
+to some false measurements of CAV states and apparent recoveries.
+Thus we account for misclassification by fitting a \emph{hidden Markov
+model} using \Rpackage{msm}.  Firstly we replace the two-way multi-state
+model by a one-way model with transition intensity matrix
+\[
+Q = \left(
+  \begin{array}{llll}
+    -(q_{12} + q_{14}) & q_{12} &  0     & q_{14}\\
+      0    & -(q_{23}+q_{24}) & q_{23} & q_{24}\\
+      0    &   0    & -q_{34} & q_{34}\\
+      0    &   0    &   0    &   0   \\
+  \end{array}
+\right )
+\]
+We also assume that true state 1 (CAV-free) can be classified as state
+1 or 2, state 2 (mild/moderate CAV) can be classified as state 1, 2 or
+3, while state 3 (severe CAV) can be classified as state 2 or 3.
+Recall that state 4 represents death.  Thus our matrix of misclassification
+probabilities is
+\[
+E = \left(
+  \begin{array}{llll}
+    1 - e_{12} & e_{12} & 0 & 0 \\
+    e_{21} & 1 - e_{21} - e_{23} & e_{23} & 0 \\
+    0 & e_{32} & 1 - e_{32} & 0 \\
+    0 & 0 & 0 & 0\\
+  \end{array}
+  \right)
+\]
+with underlying states as rows, and observed states as columns.
+
+To model observed states with misclassification, we define an matrix
+\Rfunarg{ematrix} indicating the states that can be misclassified.
+Rows of this matrix correspond to true states, columns to observed
+states.  It should contains zeroes in the positions where
+misclassification is not permitted.  Non-zero entries are initial
+values for the corresponding misclassification probabilities. We then
+call \Rfunction{msm} as before, but with this matrix as the
+\Rfunarg{ematrix} argument.  Initial values of 0.1 are assumed for
+each of the four misclassification probabilities $e_{12}, e_{21},
+e_{23}, e_{32}$.  Zeroes are given where the elements of $E$ are zero.
+The diagonal elements supplied in \Rfunarg{ematrix} are ignored, as
+rows must sum to one. The matrix \Rfunarg{qmatrix}, specifying
+permitted transition intensities and their initial values, also
+changes to correspond to the new $Q$ representing the progression-only
+model for the underlying states.
+
+The true state for every patient at the date of transplant is known to
+be ``CAV-free'', not misclassified.  To indicate this we use the
+argument \Rfunarg{obstrue} to \Rfunction{msm}.  This is set to be a
+variable in the dataset, \Rfunarg{firstobs}, indicating where the
+observed state equals the true state.  This takes the value of 1 at
+the patient's first observation, at the transplant date, and 0
+elsewhere.
+
+We use an alternative quasi-Newton optimisation algorithm
+\Rfunarg{(method="BFGS")} which can often be faster or more robust
+than the default Nelder-Mead simplex-based algorithm.  An optional
+argument \Rfunarg{initprobs} could also have been given here,
+representing the vector of the probabilities of occupying each true
+state at the initial observation
+(equation \ref{eq:multi:hidden:matprod}).  This can also be a matrix
+with number of rows equal to the number of subjects, if these
+probabilities are subject-dependent and known. If not given, all
+individuals are assumed to be in true state 1 at their initial
+observation.  If \Rfunarg{est.initprobs=TRUE} is specified, then these
+probabilites are estimated as part of the model fit, using
+a vector \Rfunarg{initprobs} as initial values.  Covariate effects on these
+probabilities can also be estimated using a multinomial logistic
+regression model, if an \Rfunarg{initcovariates} argument is
+specified.  See \Rfunction{help(msm)} for further details.
+
+\paragraph{Model 5: multi-state model with misclassification}
+
+<<>>=
+oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
+                     c(0, 0, 0.202, 0.081),
+                     c(0, 0, 0, 0.126),
+                     c(0, 0, 0, 0))
+ematrix <- rbind(c(0, 0.1, 0, 0),
+                 c(0.1, 0, 0.1, 0),
+                 c(0, 0.1, 0, 0),
+                 c(0, 0, 0, 0))
+cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
+                   qmatrix = oneway4.q, ematrix = ematrix, death = 4,
+                   obstrue = firstobs, method="BFGS")
+cavmisc.msm
+@
+
+Thus there is an estimated probability of about 0.03 that
+mild/moderate CAV will be diagnosed erroneously, but a rather higher
+probability of 0.17 that underlying mild/moderate CAV will be
+diagnosed as CAV-free.  Between the two CAV states, the mild state
+will be misdiagnosed as severe with a probability of 0.06, and the
+severe state will be misdiagnosed as mild with a probability of 0.12.
+
+The model also estimates the progression rates through underlying
+states.  An average of 8 years is spent disease-free, an average of
+about 3 years is spent with mild/moderate disease, and periods of severe
+disease also last about 3 years on average before death.
+
+\subsection{Effects of covariates on misclassification rates}
+
+We can investigate how the probabilities of misclassification depend
+on covariates in a similar way to the transition intensities, using a
+\Rfunarg{misccovariates} argument to \Rfunction{msm}. For example, we
+now include female sex as a covariate for the misclassification
+probabilities.  The linear effects on the log odds of each misclassified
+state relative to the true state are initialised to zero by default (but
+this can be changed with the \Rfunarg{misccovinits} argument).
+
+\paragraph{Model 6: misclassification model with misclassification probabilities modelled on sex}
+
+<<>>=
+cavmiscsex.msm <- msm(state ~ years, subject = PTNUM, data = cav,
+                          qmatrix = oneway4.q, ematrix = ematrix, death = 4,
+                          misccovariates = ~sex, obstrue=firstobs, method="BFGS")
+@
+<<>>=
+cavmiscsex.msm
+@
+The large confidence interval for the 1/2 misclassification suggests
+that women are no more or less likely than men to have a false
+positive angiography scan. On the other hand, women have slightly more
+false negatives.
+
 \end{document}

### 2014-01-01 pkg/inst/doc/msm-manual.Rnw
Build test again, up to and including first fev fit
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@41 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+\subsection{Extractor functions}
+
+As well as the functions described in section \ref{sec:extractor} for
+extracting useful information from fitted models, there are a number
+of extractor functions specific to models with misclassification.
+
+\begin{description}
+\item[Misclassification matrix] The function \Rfunction{ematrix.msm}
+gives the estimated misclassification probability matrix at the given
+covariate values.  For illustration, the fitted misclassification
+probabilities for men and women in model 6 are given by
+
+<<>>=
+ematrix.msm(cavmiscsex.msm, covariates=list(sex=0))
+ematrix.msm(cavmiscsex.msm, covariates=list(sex=1))
+@
+The confidence intervals for the estimates for women are higher, since
+there are only 87 women in this set of 622 patients.
+
+\item[Odds ratios for misclassification] The function
+  \Rfunction{odds.msm} would give the estimated odds ratios corresponding to
+  each covariate effect on the misclassification probabilities.
+
+\begin{Scode}
+odds.msm(cavmiscsex.msm)
+\end{Scode}
+
+\item[Observed and expected prevalences]
+
+  The function \Rfunction{prevalence.msm} is intended to assess the
+  goodness of fit of the hidden Markov model for the \emph{observed}
+  states to the data.  Tables of observed prevalences of observed
+  states are calculated as described in
+  section \ref{sec:model-assessment}, by assuming that observed states
+  are retained between observation times.
+
+  The expected numbers of individuals in each observed state are
+  calculated similarly. Suppose the process begins at a common time
+  for all individuals, and at this time, the probability of occupying
+  \emph{true} state $r$ is $f_r$.  Then given $n(t)$ individuals under
+  observation at time $t$, the expected number of individuals in true
+  state $r$ at time $t$ is the $r$th element of the vector $n(t) f
+  P(t)$. Thus the expected number of individuals in \emph{observed}
+  state $r$ is the $r$th element of the vector $n(t) f P(t) E$, where
+  $E$ is the misclassification probability matrix.
+
+  The expected prevalences (not shown) for this example are similar to
+  those forecasted by the model without misclassification, with
+  underestimates of the rates of death from 8 years onwards.  To
+  improve this model's long-term prediction ability, it is probably
+  necessary to account for the natural increase in the hazard of death
+  from any cause as people become older.
+
+\item[Goodness-of-fit test] The Pearson-type goodness-of-fit test is
+  performed, as in Section \ref{sec:pearson}.  The table of deviances
+  indicates that there are more 1-3 and 1-4 transitions than expected
+  in short intervals, and fewer in long intervals.  This may indicate
+  some time-dependence in the transition rates. Indeed, Titman
+  \cite{titman:phd} found that a model with piecewise-constant
+  transition intensities gave a greatly improved fit to these data.
+<<>>=
+pearson.msm(cavmisc.msm, timegroups=2, transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
+@
+
+\end{description}
+
+\subsection{Recreating the path through underlying states}
+
+In speech recognition and signal processing, {\em decoding} is the
+procedure of determining the underlying states that are most likely to
+have given rise to the observations. The most common method of
+reconstructing the most likely state path is the {\em Viterbi}
+algorithm. Originally proposed by Viterbi \cite{viterbi}, it is also
+described by Durbin \etal \cite{biolog:seq} and Macdonald and
+Zucchini \cite{macdonald:zucchini} for discrete-time hidden Markov
+chains.  For continuous-time models it proceeds as follows. Suppose
+that a hidden Markov model has been fitted and a Markov transition
+matrix $P(t)$ and misclassification matrix $E$ are known.  Let
+$v_k(t_i)$ be the probability of the most probable path ending in state
+$k$ at time $t_i$.
+\begin{enumerate}
+\item Estimate $v_k(t_1)$ using known or estimated initial-state
+  occupation probabilities.
+\item For $i = 1 \ldots N$, calculate $v_l(t_i) = e_{l,O_{t_i}} \max_k
+  v_k(t_{i-1}) P_{kl}(t_{i} - t_{i-1})$. Let $K_i(l)$ be the
+  maximising value of $k$.
+\item At the final time point $t_N$, the most likely underlying
+  state $S^*_N$ is the value of $k$ which maximises $v_k(t_N)$.
+\item Retrace back through the time points, setting $S^*_{i-1} =
+  K_i(S^*_i)$.
+\end{enumerate}
+
+The computations should be done in log space to prevent underflow.
+The \Rpackage{msm} package provides the function \Rfunction{viterbi.msm} to
+implement this method.  For example, the following is an extract from
+a result of calling \Rfunction{viterbi.msm} to determine the most likely
+underlying states for all patients. The results for patient 100103 are
+shown, who appeared to `recover' to a less severe state of disease
+while in state 3.  We assume this is not biologically possible for the
+true states, so we expect that either the observation of state 3 at
+time 4.98 was an erroneous observation of state 2, or their apparent
+state 2 at time 5.94 was actually state 3.  According to the expected
+path constructed using the Viterbi algorithm, it is the observation at
+time 5.94 which is most probably misclassified.
+
+<<>>=
+vit <- viterbi.msm(cavmisc.msm)
+vit[vit$subject==100103,]
+@
+
+\subsection{Fitting general hidden Markov models with \Rpackage{msm}}
+\label{sec:fitting:hmm:general}
+
+The \Rpackage{msm} package provides a framework for fitting
+continuous-time hidden Markov models with general, continuous
+outcomes.  As before, we use the \Rfunction{msm} function itself.
+
+\paragraph{Specifying the hidden Markov model}
+
+A hidden Markov model consists of two related components:
+\begin{itemize}
+\item the model for the evolution of the underlying Markov chain,
+\item the set of models for the observed data conditionally on each
+  underlying state.
+\end{itemize}
+
+The model for the transitions between underlying states is specified
+as before, by supplying a \Rfunarg{qmatrix}.  The model for the
+outcomes is specified using the argument \Rfunarg{hmodel} to
+\Rfunction{msm}.  This is a list, with one element for each underlying
+state, in order.  Each element of the list should be an object
+returned by a hidden Markov model \emph{constructor function}.  The
+HMM constructor functions provided with \Rpackage{msm} are listed in
+Table \ref{tab:hmm:dists}.  There is a separate constructor function
+for each class of outcome distribution, such as uniform, normal or
+gamma.
+
+Consider a three-state hidden Markov model, with a transition
+intensity matrix of
+\[
+Q = \left(
+  \begin{array}{llll}
+    -q_{12} & q_{12} &  0  \\
+    0  & -q_{23} & q_{23}\\
+    0    &   0    &   0   \\
+  \end{array}
+\right )
+\]
+Suppose the outcome distribution for state 1 is Normal$(\mu_1,
+\sigma^2_1)$, the distribution for state 2 is Normal$(\mu_2,
+\sigma^2_2)$, and state 3 is exactly observed. Observations of state 3
+are given a label of -9 in the data.  Here our \Rfunarg{hmodel}
+argument should be a list of objects returned by \Rfunction{hmmNorm} and
+\Rfunction{hmmIdent} constructor functions.
+
+We must specify initial values for the parameters as the arguments to
+the constructor functions.  For example, we take initial values of
+$\mu_1 = 90, \sigma_1 = 8, \mu_2 = 70, \sigma_2 = 8$.  Initial values
+for $q_{12}$ and $q_{23}$ are 0.25 and 0.2. Finally suppose the
+observed data are in a variable called \Robject{y}, the measurement
+times are in \Robject{time}, and subject identifiers are in
+\Robject{ptnum}.  The call to \Rfunction{msm} to estimate the
+parameters of this hidden Markov model would then be
+\begin{Scode}
+  msm( y ~ time, subject=ptnum, data = example.df,
+  qmatrix = rbind( c(0, 0.25, 0), c(0, 0, 0.2), c(0, 0, 0)),
+  hmodel = list (hmmNorm(mean=90, sd=8), hmmNorm(mean=70, sd=8),
+                 hmmIdent(-9)) )
+\end{Scode}
+
+\begin{table}[htbp]
+\scriptsize
+  \centering
+  \begin{tabular}{lp{0.8in}p{0.6in}p{0.6in}p{0.7in}l}
+    \hline
+    Function & Distribution & Parameters &  &  Location (link) & Density for an observation $x$  \\
+    \hline
+    \Rfunction{hmmCat}   & Categorical   & \Rfunarg{prob, basecat}  & $p,c_0$ & $p$ (logit) & $p_x$, $x = 1, \ldots, n$  \\
+    \Rfunction{hmmIdent}   & Identity  & \Rfunarg{x}  & $x_0$  & & $I_{x = x_0}$  \\
+    \Rfunction{hmmUnif}   & Uniform  & \Rfunarg{lower, upper}  & $l,u$  & & $1 / (u - l)$, $u \leq x \leq l$  \\
+    \Rfunction{hmmNorm}   & Normal  & \Rfunarg{mean, sd}  & $\mu,\sigma$  & $\mu$ (identity) & $\phi(x, \mu, \sigma) = \frac{1}{\sqrt{2 \pi \sigma^2}} \exp(-(x - \mu)^2/(2 \sigma^2) )$  \\
+    \Rfunction{hmmLNorm}   & Log-normal  & \Rfunarg{meanlog, sdlog}  & $\mu,\sigma$  & $\mu$ (identity) & $\frac{1}{x \sqrt{2 \pi \sigma^2}} \exp(-(\log x - \mu)^2 / (2 \sigma^2))$   \\
+    \Rfunction{hmmExp}   & Exponential  & \Rfunarg{rate}  & $\lambda$  & $\lambda$ (log) & $\lambda e^{- \lambda x}$, $x > 0$  \\
+    \Rfunction{hmmGamma}   & Gamma  & \Rfunarg{shape, rate}  & $n,\lambda$  &  $\lambda$ (log) & $\frac{\lambda^n}{\Gamma(n)}x^{n-1} \exp(-\lambda x)$, $x > 0, n > 0, \lambda > 0$   \\
+    \Rfunction{hmmWeibull}   & Weibull  & \Rfunarg{shape, scale}  & $a, b$  & $b$ (log) & $\frac{a}{b} (\frac{x}{b})^{a-1} \exp{(-(\frac{x}{b})^a)}$, $x > 0$  \\
+    \Rfunction{hmmPois}   & Poisson  & \Rfunarg{rate}  & $\lambda$  & $\lambda$ (log) & $\lambda^x \exp(-\lambda)/x!$, $x = 0, 1, 2, \ldots$  \\
+    \Rfunction{hmmBinom}   & Binomial  & \Rfunarg{size, prob}  & $n,p$  & $p$ (logit) & ${n \choose x} p^x (1-p)^{n-x}$  \\
+    \Rfunction{hmmNBinom}   & Negative binomial  & \Rfunarg{disp, prob}  & $n,p$  & $p$ (logit) & $\Gamma(x+n)/(\Gamma(n)x!) p^n (1-p)^x$ \\
+    \Rfunction{hmmBeta}   & Beta  & \Rfunarg{shape1,shape2}  & $a,b$  & & $ \Gamma(a+b) / (\Gamma(a)\Gamma(b))x^{a-1}(1-x)^{b-1}$ \\
+    \Rfunction{hmmT}   & Student $t$  & \Rfunarg{mean, scale, df}  & $\mu,\sigma,k$  &  $\mu$ (identity) & $\frac{\Gamma\left((k+1)/2\right)}{\Gamma(k/2)}{\sqrt{\frac{1}{k\pi\sigma^2}}}\left\{1 + \frac{1}{k\sigma^2}(x - \mu)^{2} \right\}^{-(k + 1)/2}$ \\
+    \Rfunction{hmmTNorm}   & Truncated normal  & \Rfunarg{mean, sd, lower, upper}  & $\mu,\sigma,l,u$  & $\mu$ (identity) &
+\parbox[t]{2in}{$\phi(x, \mu, \sigma) / \\ (\Phi(u, \mu, \sigma) - \Phi(l, \mu, \sigma))$, \\ where $\Phi(x,\mu,\sigma) = \int_{-\infty}^x \phi(u,\mu,\sigma) du$}
+ \\
+    \Rfunction{hmmMETNorm}   & Normal with truncation and measurement error  & \Rfunarg{mean, sd, lower, upper, sderr, meanerr}  & \parbox[t]{1in} {$\mu_0,\sigma_0,l,u$, \\ $\sigma_\epsilon,\mu_\epsilon$}  &  $\mu_\epsilon$ (identity) &
+\parbox[t]{2in}{$( \Phi(u, \mu_2, \sigma_3) - \Phi(l, \mu_2, \sigma_3)) / $ \\
+$(\Phi(u, \mu_0, \sigma_0) - \Phi(l, \mu_0, \sigma_0)) $ \\
+$\times \phi(x, \mu_0 + \mu_\epsilon, \sigma_2)$, \\
+$\sigma_2^2 = \sigma_0^2 + \sigma_\epsilon^2$, \\
+$\sigma_3 = \sigma_0 \sigma_\epsilon / \sigma_2$, \\
+$\mu_2 = (x - \mu_\epsilon) \sigma_0^2 + \mu_0 \sigma_\epsilon^2$}
+ \\
+    \Rfunction{hmmMEUnif}   & Uniform with measurement error  & \Rfunarg{lower, upper, sderr, meanerr}  & $l,u,\mu_\epsilon,\sigma_\epsilon$  & $\mu_\epsilon$ (identity) &
+    \parbox[t]{2in}{$(\Phi(x, \mu_\epsilon+l, \sigma_\epsilon) - \Phi(x, \mu_\epsilon+u, \sigma_\epsilon)) / \\ (u - l)$} \\
+    \hline
+  \end{tabular}
+  \caption{Hidden Markov model distributions in \Rpackage{msm}.}
+  \label{tab:hmm:dists}
+\end{table}
+
+\paragraph{Covariates on hidden Markov model parameters}
+
+Most of the outcome distributions can be parameterised by covariates,
+using a link-transformed linear model.  For example, an observation
+$y_{ij}$ may have distribution $f_1$ conditionally on underlying state
+1. The link-transformed parameter $\theta_1$ is a linear function of
+the covariate vector $x_{ij}$ at the same observation time.
+\begin{eqnarray*}
+  \label{eq:hmm:covs}
+  y_{ij} | S_{ij} &  \sim &   f_1 (y | \theta_1, \gamma_1)\\
+  g(\theta_1) & = & \alpha + \beta^T x_{ij}
+\end{eqnarray*}
+Specifically, parameters named as the ``Location'' parameter in
+Table \ref{tab:hmm:dists} can be modelled in terms of covariates, with
+the given link function.
+
+The \Rfunarg{hcovariates} argument to \Rfunction{msm} specifies the
+model for covariates on the hidden Markov outcome distributions.  This
+is a list of the same length as the number of underlying states, and
+the same length as the \Rfunarg{hmodel} list.  Each element of the
+list is a formula, in standard R linear model notation, defining the
+covariates on the distribution for the corresponding state.  If there
+are no covariates for a certain hidden state, then insert a \texttt{NULL} in
+the corresponding place in the list.  For example, in the three-state
+normal-outcome example above, suppose that the normal means on states
+1 and 2 are parameterised by a single covariate $x$.
+\[
+\mu_1 = \alpha_1 + \beta_1 x_{ij}, \qquad \mu_2 = \alpha_2 + \beta_2 x_{ij}.
+\]
+The equivalent call to \Rfunction{msm} would be
+\begin{Scode}
+  msm( state ~ time, subject=ptnum, data = example.df,
+  qmatrix = rbind( c(0, 0.25, 0), c(0, 0, 0.2), c(0, 0, 0)),
+  hmodel = list (hmmNorm(mean=90, sd=8), hmmNorm(mean=70, sd=8),
+                 hmmIdent(-9)),
+  hcovariates = list ( ~ x, ~ x, NULL)
+ ).
+\end{Scode}
+
+\paragraph{Constraints on hidden Markov model parameters}
+
+Sometimes it is realistic that parameters are shared between some of
+the state-specific outcome distributions.  For example, the
+Normally-distributed outcome in the previous example could have a
+common variance $\sigma^2_1 = \sigma^2_2 = \sigma^2$ between states 1
+and 2, but differing means.  It would also be realistic for any
+covariates on the mean to have a common effect $\beta_1 = \beta_2 =
+\beta$ on the state 1 and 2 outcome distributions.
+
+The argument \Rfunarg{hconstraint} to \Rfunction{msm} specifies which
+hidden Markov model parameters are constrained to be equal.  This is a
+named list.  Each element is a vector of constraints on the named
+hidden Markov model parameter. The vector has length equal to the
+number of times that class of parameter appears in the whole model.
+As for the other constraint arguments such as \Rfunarg{qconstraint},
+identical values of this vector indicate parameters constrained to be
+equal.
+
+For example, consider the three-state hidden Markov model described
+above, with normally-distributed outcomes for states 1 and 2.
+To constrain the outcome variance to be equal for states 1 and 2,
+and to also constrain the effect of \Robject{x} on the outcome mean
+to be equal for states 1 and 2, specify
+\begin{Scode}
+hconstraint = list(sd = c(1,1), x=c(1,1))
+\end{Scode}
+
+Parameters of the
+outcome distributions may also be constrained within specific ranges.
+If chosen carefully, this may improve identifiability of hidden Markov states.
+For example to constrain the mean for state 1 to be between 80 and 110,
+and the mean for state 2 to be between 50 and 80, specify
+\begin{Scode}
+hranges = list(mean=list(lower=c(80,50), upper=c(110,80)))
+\end{Scode}
+Maximum likelihood estimation is then performed on the appropriate log
+or logit-transformed scale so that these constraints are
+satisfied. See the \Rfunction{msm} help page for further details.
+Note that initial values should be strictly within the ranges, and not
+on the range boundary.
+
+\paragraph{FEV$_1$ after lung transplants}
+
+Now we give an example of fitting a hidden Markov model to a real
+dataset.  The data on FEV$_1$ measurements from lung transplant
+recipients, described in \ref{sec:hmm:example:fev}, are provided with
+the \Rpackage{msm} package in a dataset called \Robject{fev}.
+We fit models Models 1 and 2, each with three states and common $Q$
+matrix.
+<<>>=
+three.q <- rbind(c(0, exp(-6), exp(-9)), c(0, 0, exp(-6)), c(0, 0, 0))
+@
+
+The simpler Model 1 is specified as follows. Under this model the
+FEV$_1$ outcome is Normal with unknown mean and variance, and the mean
+and variance are different between BOS state 1 and state 2.
+\Rfunarg{hcovariates} specifies that the mean of the Normal outcome
+depends linearly on acute events.  Specifically, this covariate is an
+indicator for the occurrence of an acute event within 14 days of the
+observation, denoted \texttt{acute} in the data.  As an initial guess,
+we suppose the mean FEV$_1$ is 100\% baseline in state 1, and 54\%
+baseline in state 2, with corresponding standard deviations 16 and 18,
+and FEV$_1$ observations coinciding with acute events are on average
+8\% baseline lower. \Rfunarg{hconstraint} specifies that the acute
+event effect is equal between state 1 and state 2.
+
+Days of death are coded as 999 in the \texttt{fev} outcome variable.
+
+<<>>=
+hmodel1 <- list(hmmNorm(mean=100, sd=16), hmmNorm(mean=54, sd=18),
+                hmmIdent(999))
+
+fev1.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q,
+                death=3, hmodel=hmodel1,
+                hcovariates=list(~acute, ~acute, NULL),
+                hcovinits = list(-8, -8, NULL),
+                hconstraint = list(acute = c(1,1)), method="BFGS")
+
+fev1.msm
+
+sojourn.msm(fev1.msm)
+@
+
 \end{document}

### 2014-01-02 pkg/inst/doc/msm-manual.Rnw
Build test again, up to final new misc model
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@42 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+Printing the \Rclass{msm} object \Rfunarg{fev1.msm} shows estimates
+and confidence intervals for the transition intensity matrix and the
+hidden Markov model parameters.  The estimated within-state means of
+FEV$_1$ are around 98\% and 52\% baseline respectively.  From the
+estimated transition intensities, individuals spend around 1421 days
+(3.9 years) before getting BOS, after which they live for an average
+of 1248 days (3.4 years).  FEV$_1$ is lower by an average of 8\%
+baseline within 14 days of acute events.
+
+Model 2, where the outcome distribution is a more complex two-level
+model, is specified as follows.  We use the distribution defined by
+equations \ref{eq:fev:level1}--\ref{eq:fev:level2}. The
+\Rfunction{hmmMETNorm} constructor defines the truncated normal
+outcome with an additional normal measurement error.  The explicit
+probability density for this distribution is given in
+Table \ref{tab:hmm:dists}.
+
+Our initial values are 90 and 54 for the means of the
+within-state distribution of \emph{underlying} FEV$_1$, and 16 and 18
+for the standard errors.  This time, underlying FEV$_1$ is truncated
+normal. The truncation limits \Rfunarg{lower} and \Rfunarg{upper} are
+not estimated.  We take an initial measurement error standard
+deviation of \Rfunarg{sderr=8}. The extra shift \Rfunarg{meanerr} in
+the measurement error model is fixed to zero and not estimated.
+
+The \Rfunarg{hconstraint} specifies that the measurement error
+variance $\sigma^2_\epsilon$ is equal between responses in states 1
+and 2, as is the effect of short-term acute events on the FEV$_1$
+response.
+
+The convergence of maximum likelihood estimation in this example is
+particularly sensitive to the optimisation method and options, initial values,
+the unit of the time variable and whether covariates are centered,
+probably because the likelihood surface is irregular near to the
+true maximum.
+
+\begin{Scode}
+hmodel2 <- list(hmmMETNorm(mean=90, sd=16, sderr=8,
+                           lower=80, upper=Inf, meanerr=0),
+                hmmMETNorm(mean=54, sd=18, sderr=8,
+                           lower=0, upper=80, meanerr=0),
+                hmmIdent(999))
+
+fev2.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q,
+                death=3, hmodel=hmodel2,
+                hcovariates=list(~acute, ~acute, NULL),
+                hcovinits = list(-8, -8, NULL),
+                hconstraint = list(sderr = c(1,1), acute = c(1,1)),
+                control=list(maxit=10000), center=TRUE)
+
+fev2.msm
+
+sojourn.msm(fev2.msm)
+\end{Scode}
+
+Under this model the standard deviation of FEV$_1$ measurements caused
+by measurement error (more realistically, natural short-term
+fluctuation) is around 9\% baseline.  The estimated effect of acute
+events on FEV$_1$ and sojourn times in the BOS-free state and in BOS
+before death are similar to Model 1.
+
+The following code will create a plot that illustrates a trajectory of declining FEV$_1$ from
+the first lung transplant recipient in this dataset.  The Viterbi algorithm is used to locate the
+most likely point at which this individual moved from BOS state 1 to
+BOS state 2, according to the fitted Model 2.  This is illustrated by
+a vertical dotted line.  This is the point at which the individual's
+lung function started to remain consistently below 80\% baseline
+FEV$_1$.
+
+\begin{Scode}
+keep <- fev$ptnum==1 & fev$fev<999
+plot(fev$days[keep], fev$fev[keep], type="l",
+ylab=expression(paste("% baseline ", FEV[1])), xlab="Days after transplant")
+vit <- viterbi.msm(fev2.msm)[keep,]
+(max1 <- max(vit$time[vit$fitted==1]))
+(min2 <- min(vit$time[vit$fitted==2]))
+abline(v = mean(max1,min2), lty=2)
+text(max1 - 500, 50, "STATE 1")
+text(min2 + 500, 50, "STATE 2")
+\end{Scode}
+
+\paragraph{An alternative way of specifying a misclassification model}
+
+This general framework for specifying hidden Markov models can also be
+used to specify multi-state models with misclassification.  A
+misclassification model is a hidden Markov model with a categorical
+outcome distribution.  So instead of an \Rfunarg{ematrix} argument to
+\Rfunction{msm}, we can use a \Rfunarg{hmodel} argument with
+\Rfunction{hmmCat} constructor functions.
+
+\Rfunction{hmmCat} takes at least one argument \Rfunarg{prob}, a
+vector of probabilities of observing outcomes of $1, 2, \ldots, n$
+respectively, where $n$ is the length of \Rfunarg{prob}.  All outcome
+probabilities with an initial value of zero are assumed to be fixed at
+zero.  \Rfunarg{prob} is scaled if necessary to sum to one.
+
+The model in section \ref{sec:fitting:hmm:misc} specifies that an
+individual occupying underlying state 1 can be observed as states 2
+(and 1), underlying state 2 can be observed as states 1, 2 or 3, and
+state 3 can be observed as states 2 or 3, and underlying state 4
+(death) cannot be misclassified. Initial values of 0.1 are given for
+the 1-2, 2-1, 2-3 and 3-2 misclassification probabilities.
+
+This is equivalent to the model below, specified using a
+\Rfunarg{hmodel} argument to \Rfunction{msm}.  The maximum likelihood
+estimates should be the same as before (Model 5).
+
+<<>>=
+oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
+                   c(0, 0, 0.202, 0.081),
+                   c(0, 0, 0, 0.126),
+                   c(0, 0, 0, 0))
+cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
+                     hmodel = list (hmmCat(c(0.9, 0.1, 0, 0)), hmmCat(c(0.1, 0.8, 0.1, 0)), hmmCat(c(0, 0.1, 0.9, 0)), hmmIdent(4)),
+                     qmatrix = oneway4.q, obstrue=firstobs, death = 4, method="BFGS")
+cavmisc.msm
+@
+
 \end{document}

### 2014-01-02 pkg/inst/doc/msm-manual.Rnw
Build test, excluding new cav misc (we're getting there...)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@43 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
-\paragraph{An alternative way of specifying a misclassification model}
-
-This general framework for specifying hidden Markov models can also be
-used to specify multi-state models with misclassification.  A
-misclassification model is a hidden Markov model with a categorical
-outcome distribution.  So instead of an \Rfunarg{ematrix} argument to
-\Rfunction{msm}, we can use a \Rfunarg{hmodel} argument with
-\Rfunction{hmmCat} constructor functions.
-
-\Rfunction{hmmCat} takes at least one argument \Rfunarg{prob}, a
-vector of probabilities of observing outcomes of $1, 2, \ldots, n$
-respectively, where $n$ is the length of \Rfunarg{prob}.  All outcome
-probabilities with an initial value of zero are assumed to be fixed at
-zero.  \Rfunarg{prob} is scaled if necessary to sum to one.
-
-The model in section \ref{sec:fitting:hmm:misc} specifies that an
-individual occupying underlying state 1 can be observed as states 2
-(and 1), underlying state 2 can be observed as states 1, 2 or 3, and
-state 3 can be observed as states 2 or 3, and underlying state 4
-(death) cannot be misclassified. Initial values of 0.1 are given for
-the 1-2, 2-1, 2-3 and 3-2 misclassification probabilities.
-
-This is equivalent to the model below, specified using a
-\Rfunarg{hmodel} argument to \Rfunction{msm}.  The maximum likelihood
-estimates should be the same as before (Model 5).
-
-<<>>=
-oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
-                   c(0, 0, 0.202, 0.081),
-                   c(0, 0, 0, 0.126),
-                   c(0, 0, 0, 0))
-cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                     hmodel = list (hmmCat(c(0.9, 0.1, 0, 0)), hmmCat(c(0.1, 0.8, 0.1, 0)), hmmCat(c(0, 0.1, 0.9, 0)), hmmIdent(4)),
-                     qmatrix = oneway4.q, obstrue=firstobs, death = 4, method="BFGS")
-cavmisc.msm
-@
-
 \end{document}

### 2014-01-02 pkg/inst/doc/msm-manual.Rnw
Build check, eval fev2, don't eval new misc, leave out appendix
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@44 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2014-01-02 pkg/inst/doc/msm-manual.Rnw
@@
 
-\begin{Scode}
+<<>>=
 hmodel2 <- list(hmmMETNorm(mean=90, sd=16, sderr=8,
@@
 sojourn.msm(fev2.msm)
-\end{Scode}
+@
 
@@
 
-\begin{Scode}
+<<>>=
 keep <- fev$ptnum==1 & fev$fev<999
@@
 text(min2 + 500, 50, "STATE 2")
-\end{Scode}
+@
 
@@
 
-<<>>=
+\begin{Scode}
 oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
@@
 cavmisc.msm
-@
+\end{Scode}
 

### 2014-01-03 pkg/inst/doc/msm-manual.Rnw
Build test again (hopefully final - fev2 and new misc replaced by Scode)
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@45 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

--- 2014-01-02 pkg/inst/doc/msm-manual.Rnw
@@
                 control=list(maxit=10000), center=TRUE)
-
-fev2.msm
-
-sojourn.msm(fev2.msm)
 \end{Scode}
@@
 
+\includegraphics{figures/fev_viterbi}
+
 \paragraph{An alternative way of specifying a misclassification model}
@@
 cavmisc.msm
-@
+\end{Scode}
+
+\subsubsection{Defining a new hidden Markov model distribution}
+
+Suppose the hidden Markov model outcome distributions supplied with
+\Rpackage{msm} (Table \ref{tab:hmm:dists}) are insufficient.  We want
+to define our own univariate distribution, called
+\Rfunction{hmmNewDist}, taking two parameters \Robject{location} and
+\Robject{scale}.  Download the source package, for example
+\texttt{msm-0.7.2.tar.gz} for version 0.7.2, from CRAN and edit the
+files in there, as follows.
+
+\begin{enumerate}
+\item Add an element to \Robject{.msm.HMODELPARS} in the file
+  \texttt{R/constants.R}, naming the parameters of the distribution.
+  For example
+\begin{Scode}
+  newdist = c('location', 'scale')
+\end{Scode}
+
+\item Add a corresponding element to the C variable \texttt{HMODELS}
+  in the file \texttt{src/lik.c}.  This MUST be in the same position
+  as in the \Robject{.msm.HMODELPARS} list.  For example,
+\begin{Scode}
+hmmfn HMODELS[] = {
+  ...,
+  hmmNewDist
+};.
+\end{Scode}
+
+\item The new distribution is allowed to have one parameter which can
+  be modelled in terms of covariates.  Add the name of this parameter
+  to the named vector \Robject{.msm.LOCPARS} in
+  \texttt{R/constants.R}.  For example \texttt{newdist = 'location'}.
+  Specify \texttt{newdist = NA} if there is no such parameter.
+
+\item Supposed we have specified a parameter with a non-standard name,
+  that is, one which doesn't already appear in
+  \Robject{.msm.HMODELPARS}. Standard names include, for example,
+  \texttt{'mean'}, \texttt{'sd'}, \texttt{'shape'} or
+  \texttt{'scale'}.  Then we should add the allowed range of the
+  parameter to \Robject{.msm.PARRANGES}. In this example, we add
+  \texttt{meanpars = c(-Inf, Inf)} to \Robject{.msm.PARRANGES}.  This
+  ensures that the optimisation to estimate the parameter takes place
+  on a suitable scale, for example, a log scale for a parameter
+  constrained to be positive.  If the parameter should be fixed during
+  maximum likelihood estimation (for example, the denominator of a
+  binomial distribution) then add its name to \Robject{.msm.AUXPARS}.
+
+\item Add an R constructor function for the distribution to
+  \texttt{R/hmm-dists.R}.  For a simple univariate distribution, this is of the
+  form
+\begin{Scode}
+  hmmNewDist <- function(location, scale)
+  {
+    hmmDIST(label = "newdist",
+    link = "identity",
+    r = function(n) rnewdist(n, location, scale),
+    match.call())
+  }
+\end{Scode}
+
+  \begin{itemize}
+  \item The \texttt{'label'} must be the same as the name you
+    supplied for the new element of \Robject{.msm.HMODELPARS}
+
+  \item \texttt{link} is the link function for modelling the location
+    parameter of the distribution as a linear function of covariates. This
+    should be the quoted name of an R function.  A log link is
+    \texttt{'log'} and a logit link is \texttt{'qlogis'}.  If
+    using a new link function other than \texttt{'identity'},
+    \texttt{'log'}, or \texttt{'qlogis'}, you should add its name
+    to the vector \Robject{.msm.LINKFNS} in \texttt{R/constants.R},
+    and add the name of the corresponding inverse link to
+    \Robject{.msm.INVLINK}.  You should also add the names of these
+    functions to the C array \texttt{LINKFNS} in \texttt{src/lik.c},
+    and write the functions if they do not already exist.
+
+  \item \texttt{r} is an R function, of the above format, returning a
+    vector of \texttt{n} random values from the distribution.
+    You should write this if it doesn't already exist.
+  \end{itemize}
+
+\item Add the name of the new constructor function to the NAMESPACE in
+  the top-level directory of the source package.
+
+\item Write a C function to compute the probability density of the
+  distribution, and put this in \texttt{src/hmm.c}, with a declaration
+  in \texttt{src/hmm.h}. This must be of the form
+\begin{Scode}
+  double hmmNewDist(double x, double *pars)
+\end{Scode}
+
+  where \texttt{*pars} is a vector of the parameters of the
+  distribution, and the density is evaluated at \texttt{x}.
+
+\item Update the documentation (\texttt{man/hmm-dists.Rd}) and the
+  distribution table in\\ \texttt{inst/doc/msm-manual.Rnw}) if you like.
+
+\item Recompile the package (see the ``Writing R Extensions'' manual)
+
+\end{enumerate}
+
+Your new distribution will be available to use in the \Rfunarg{hmodel}
+argument to \Rfunction{msm}, as, for example
+\begin{Scode}
+hmodel = list(..., hmmNewDist(location = 0, scale = 1), ...)
+\end{Scode}
+If your distribution may be of interest to others, ask me
+(\texttt{chris.jackson@mrc-bsu.cam.ac.uk}) to include it in a future
+release.
+
+\clearpage
+
+\section{\Rpackage{msm} reference guide}
+
+The R help page for \Rfunction{msm} gives details of all the allowed
+arguments and options to the \Rfunction{msm} function.  To view this
+online in R, type:
+
+<<eval=FALSE>>=
+help(msm)
+@
+
+Similarly all the other functions in the package have help pages,
+which should always be consulted in case of doubt about how to
+call them.  The web-browser based help interface may often be
+convenient - type
+
+<<eval=FALSE>>=
+help.start()
+@
+and navigate to \textsf{Packages} $\ldots$ \textsf{msm}, which brings
+up a list of all the functions in the package with links to their
+documentation, and a link to this manual in PDF format.
+
+\appendix
+
+\section{Changes in the msm package}
+
+For a detailed list of the changes in recent versions of
+\Rpackage{msm}, see the \texttt{NEWS} file in the top-level directory
+of the installed package.  The \texttt{Changelog} file in the
+top-level directory of the source package contains more technical
+detail about these changes.
+
+Development versions of \Rpackage{msm} are to be found on R-Forge,
+\url{http://msm.r-forge.r-project.org/}.  These are often more recent
+than the version released on CRAN, so if you think you have found a
+bug, then please check first to see whether it has been fixed on the
+R-Forge version.
+
+\section{Get in touch}
+
+If you use \Rpackage{msm} in your work, whatever your field of
+application, please let me know, for my own interest!  Suggestions for
+improvement are welcome.
+
+\clearpage
+
+\bibliography{msm}
 

### 2014-01-18 pkg/inst/doc/msm-manual.Rnw
0.3 release, adding compiled PDF manual
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@47 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 \begin{array}{ll}
-\frac{ q_{12}(-1 + e^{(q_{12}+q_{13}-q_{23})t})e^{-(q_{12}+q_{13})t} } { (q_{12} + q_{13} - q_{23}) } & (q_{12} + q_{13} \neq q_{23})\\
+\frac{ q_{12} }{q_{12} + q_{13} - q_{23} } (e^{-q_{23} t} - e^{-(q_{12} + q_{13})t})  & (q_{12} + q_{13} \neq q_{23})\\
 q_{12}te^{(-(q_{12} + q_{13})t} &  (q_{12} + q_{13} = q_{23})
@@
 \begin{array}{ll}
-\frac{1 + (-q_{13} + q_{23})e^{-(q_{12}+q_{13})t}} {q_{12} + q_{13} - q_{23}} - \frac { q_{12}e^{-q_{23}t}} {q_{12} + q_{13} - q_{23} } & (q_{12} + q_{13} \neq q_{23})\\
+1 - e^{-(q_{12} + q_{13})t} - \frac{ q_{12} }{q_{12} + q_{13} - q_{23} } (e^{-q_{23} t} - e^{-(q_{12} + q_{13})t}) & (q_{12} + q_{13} \neq q_{23})\\
  (-1 + e^{(q_{12} + q_{13})t} - q_{12}t)e^{-(q_{12} + q_{13})t} & (q_{12} + q_{13} = q_{23})
@@
 the mean sojourn time $1 / q_{rr}$ can be estimated as $T_r / n_r$.
-Thus, $n_{rs} / T_r$ is a crude estimate of $q_{rs}$. The \Rpackage{msm}
-package provides a function \Rfunction{crudeinits.msm} for calculating
-initial values in this way (for non-hidden Markov models).
+Thus, $n_{rs} / T_r$ is a crude estimate of $q_{rs}$.
+
+Such default initial values can be used by supplying
+\Rfunarg{gen.inits=TRUE} in the call to \Rfunction{msm} below, along
+with a \Rfunarg{qmatrix} whose non-zero entries represent the allowed
+transitions of the model.  Alternatively the function \Rfunction{crudeinits.msm} 
+could be used to get this matrix of initial values explicitly as follows.  
+These methods are only available for non-hidden Markov models.
 
 <<>>=
-twoway4.crude.q  <- crudeinits.msm(state ~ years, PTNUM, data=cav, qmatrix=twoway4.q)
+twoway4.crude.q  <- crudeinits.msm(state ~ years, PTNUM, data=cav, 
+                                   qmatrix=twoway4.q)
 @
@@
 which we now use to fit the model.
-
-%% TODO explain about gen.inits=TRUE here
 
@@
                 qmatrix = twoway4.q, death = 4,
-                   covariates = ~ sex, constraint = list(sex=c(1,2,3,1,2,3,2)) )
+                covariates = ~ sex, 
+                constraint = list(sex=c(1,2,3,1,2,3,2)) )
 @
@@
 
-A \Rfunarg{fixedpars} statement can also be useful for fixing covariate
+A \Rfunarg{fixedpars} statement can also be used for fixing covariate
 effect parameters to zero, that is to assume no effect of a covariate
@@
 options(digits=2)
-pearson.msm(cav.msm, timegroups=2, transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
+pearson.msm(cav.msm, timegroups=2, 
+            transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
 @
@@
 cavmiscsex.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                          qmatrix = oneway4.q, ematrix = ematrix, death = 4,
-                          misccovariates = ~sex, obstrue=firstobs, method="BFGS")
+                      qmatrix = oneway4.q, ematrix = ematrix, 
+                      death = 4, misccovariates = ~sex, 
+                      obstrue=firstobs, method="BFGS")
 @
@@
 <<>>=
-pearson.msm(cavmisc.msm, timegroups=2, transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
+pearson.msm(cavmisc.msm, timegroups=2, 
+            transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
 @
@@
 cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                     hmodel = list (hmmCat(c(0.9, 0.1, 0, 0)), hmmCat(c(0.1, 0.8, 0.1, 0)), hmmCat(c(0, 0.1, 0.9, 0)), hmmIdent(4)),
-                     qmatrix = oneway4.q, obstrue=firstobs, death = 4, method="BFGS")
+                     hmodel = list (hmmCat(c(0.9, 0.1, 0, 0)), 
+                                    hmmCat(c(0.1, 0.8, 0.1, 0)), 
+                                    hmmCat(c(0, 0.1, 0.9, 0)), 
+                                    hmmIdent(4)),
+                     qmatrix = oneway4.q, obstrue=firstobs, death = 4, 
+                     method="BFGS")
 cavmisc.msm

### 2014-01-18 pkg/R/outputs.R

@@ -5,7 +5,6 @@ print.msm <- function(x, ...)
     cat("\nCall:\n", deparse(x$call), "\n\n", sep = "")
 
     if (!attr(x,"fixed")) {
-        cat("[Note: a cleaner summary is available from \"printnew.msm\", which will be the default in future versions.]\n")
         cat ("Maximum likelihood estimates: \n")
         covmessage <-
             if (x$qcmodel$ncovs == 0) ""
@@ -46,6 +45,7 @@ print.msm <- function(x, ...)
         }
     }
     cat ("-2 * log-likelihood: ", x$minus2loglik, "\n")
+    cat("[Note: a cleaner summary is available from \"printnew.msm\",\n which will be the default in future versions.]\n")
 
 }
 

### 2014-06-03 pkg/R/constants.R
Changes for 1.3.2
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@50 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -22,11 +22,17 @@
                         t = c("mean","scale","df")
                         )
 
-## TODO - non-central beta, cauchy, chisq, noncentral chisq, F,
-## non-central F, geometric, hypergeometric, logistic, t, noncentral
-## t.
+## TODO - e.g. non-central beta, cauchy, chisq, noncentral chisq, F,
+## non-central F, geometric, hypergeometric, logistic, noncentral t.
 
 .msm.HMODELS <- names(.msm.HMODELPARS)
+
+### Models with analytic derivatives available
+.msm.HMODELS.DERIV <- c("categorical","identity","uniform","normal","lognormal","exponential",
+                        "gamma","weibull","poisson","binomial","nbinom","beta","t")
+
+### Models with expected information matrix available
+.msm.HMODELS.INFO <- c("categorical","identity")
 
 ### Parameter in each distribution that can have covariates on it
 .msm.LOCPARS <- c(categorical="p", identity=NA, uniform=NA, normal="mean", lognormal="meanlog",

### 2014-06-03 pkg/R/hmm.R

@@ -79,10 +79,12 @@ msm.form.hmodel <- function(hmodel, hconstraint=NULL, initprobs=NULL, est.initpr
       hmod
   }
 
-msm.form.hcmodel <- function(hmodel, hcovdata, hcovinits, hconstraint)
+## NOTE removed whichcovh, whichcovh.orig
+
+msm.form.hcmodel <- function(hmodel, mm, hcovinits, hconstraint)
   {
       nst <- hmodel$nstates
-      ncovs <- if (is.null(hcovdata)) rep(0, nst) else sapply(hcovdata, function(x) x$ncovs)
+      ncovs <- if (is.null(mm)) rep(0, nst) else sapply(mm, function(x) {ncol(x)-1})
       ncovs2 <- rep(rep(0, nst), hmodel$npars)
       ncovs2[hmodel$locpars] <- ncovs[hmodel$parstate[hmodel$locpars]]
       coveffstate <- rep(1:nst, tapply(ncovs2, hmodel$parstate, sum))
@@ -93,33 +95,31 @@ msm.form.hcmodel <- function(hmodel, hcovdata, hcovinits, hconstraint)
           if (!(sum(ncovs2) == length(unlist(hcovinits)))) {
               warning("Initial values for hidden covariate effects do not match numbers of covariates, ignoring")
               coveffect <- rep(0, sum(ncovs2))
-          }
-          coveffect <- unlist(hcovinits)
+          } else coveffect <- unlist(hcovinits)
           if (!is.numeric(coveffect)) {
               warning("hcovinits should be numeric")
               coveffect <- rep(0, sum(ncovs2))
           }
       }
-      covlabels <- lapply(hcovdata, function(x) x$covlabels)
+      covlabels <- lapply(mm, function(x) colnames(x)[-1])
       covlabels <- unlist(covlabels[hmodel$parstate[hmodel$locpars]])
       names(coveffect) <- covlabels
       hcmod <- list(ncovs=ncovs2, coveffect=coveffect, covlabels=covlabels,
                     coveffstate=coveffstate, ncoveffs=length(coveffect))
       hmodel <- c(hmodel, hcmod)
       hmodel$plabs[hmodel$plabs=="hcov"] <- paste("hcov.",covlabels,sep="")
-      whichcovh <- lapply(hcovdata, function(x) x$whichcov) # factor contrasts as separate covariates
[…]

### 2014-06-03 pkg/R/msm.R

@@ -1,61 +1,28 @@
-### Function to fit Markov multi-state models in continuous time
-### with either arbitrary observation times or observed exact transition times
-### with or without misclassification between true and underlying states
-
-msm <- function(formula,   # formula with  observed Markov states   ~  observation times (required)
-                subject = NULL, # optional, defaults to all the same if not given
-                data=list(),       # data frame in which to interpret variable names
-                qmatrix,    # matrix of 1s and 0s with indices of allowed transitions (diagonal is ignored) (required)
-                gen.inits = FALSE, # generate initial values for transition intensities using crudeinits.msm
-                ematrix = NULL,    # matrix of 1s and 0s with indices of allowed misclassfications (diagonal is ignored) (required)
-                hmodel = NULL,  # list of constructors for hidden emission distributions
-                obstype = NULL, # optional, defaults to all 1 (snapshots) if not given
-                obstrue = NULL, # for hidden Markov models, which observations represent the true state
-                covariates = NULL, # formula or list of formulae specifying covariates on transition rates.
-                covinits = NULL,      # initial values for covariate effects
-                constraint = NULL, # which intensities have covariates on them (as in Marshall et al.)
-                misccovariates = NULL, # formula or list of formulae specifying covariates on misclassification probs
-                misccovinits = NULL,      # initial values for misclassification covariate effects
-                miscconstraint = NULL, # which misc probs have covariates on them
-                hcovariates = NULL, # list of formulae specifying covariates model for each hidden state
-                hcovinits = NULL,      # initial values for covariate effects on hidden emission distribution
-                hconstraint = NULL, # constraints on hidden Markov model parameters
-                hranges = NULL, # range constraints for HMM parameters
-                qconstraint = NULL, # constraints on equality of baseline intensities
-                econstraint = NULL, # constraints on equality of baseline misc probs
[…]

### 2014-06-03 pkg/R/optim.R

@@ -1,0 +1,221 @@
+deriv.supported <- function(hmodel, cmodel){
+    (!hmodel$hidden || (hmodel$hidden &&
+                        !hmodel$est.initprobs &&
+                        (!any(duplicated(hmodel$constr[hmodel$plabs=="p"]))) &&
+                        (!any(duplicated(hmodel$covconstr[.msm.HMODELS[hmodel$models[hmodel$coveffstate]]=="categorical"]))) &&
+                        all(.msm.HMODELS[hmodel$models %in% .msm.HMODELS.DERIV])
+                        ))
+}
+
+info.supported <- function(msmdata, hmodel, cmodel){
+    ## only panel data and either non-hidden or misclassification models
+    deriv.supported(hmodel, cmodel) &&
+        all(msmdata$mf$"(obstype)"==1) &&
+            (!hmodel$hidden || (hmodel$hidden &&
+                                all(.msm.HMODELS[hmodel$models] %in% .msm.HMODELS.INFO)))
+}
+
+### GENERIC OPTIMISATION FUNCTION
+
+msm.optim <- function(opt.method, p, hessian, use.deriv, msmdata, qmodel, qcmodel, cmodel, hmodel, ...){
+    p$params <- p$allinits
+    gr <- if (deriv.supported(hmodel,cmodel) && use.deriv) deriv.msm else NULL
+    if (isTRUE(getOption("msm.test.analytic.derivatives"))){
+        if (!deriv.supported(hmodel,cmodel)) warning("Analytic derivatives not available for this model")
+        else print(deriv.test(msmdata, qmodel, qcmodel, cmodel, hmodel, msm.unfixallparams(p)))
+    }
+    optfn <- paste("msm.optim", opt.method, sep=".")
+    if (!exists(optfn)) stop("Unknown optimisation method \"", opt.method, "\"")
+
+    args <- c(list(p=p, gr=gr, hessian=hessian,
+                   msmdata=msmdata, qmodel=qmodel, qcmodel=qcmodel,
+                   cmodel=cmodel, hmodel=hmodel), list(...))
+    p <- do.call(optfn, args)
+
+    ## Attach derivative and information matrix at the MLE.
+    ## If all parameters fixed, do this at the initial values for all.
+    pp <- if (opt.method=="fixed") msm.unfixallparams(p) else p
+    pi <- pp$params[pp$optpars]
+    if (deriv.supported(hmodel,cmodel)) {
[…]

### 2014-06-03 pkg/R/pearson.R

@@ -9,11 +9,11 @@ pearson.msm <- function(x, transitions=NULL, timegroups=3, intervalgroups=3, cov
                         pval=TRUE # calculate p-values for test. false if calling from bootstrap
                         ){
 
-    dat <- x$data
+    dat <- x$data$mf
     ## Error handling
     if (!inherits(x, "msm")) stop("expected \"x\" t to be a msm model")
     if (x$hmodel$hidden && !x$emodel$misc) stop("only HMMs handled are misclassification models specified using \"ematrix\"")
-    if (any(dat$obstype==2)) stop("exact transition times are not supported, only panel-observed data")
+    if (any(dat$"(obstype)"==2)) stop("exact transition times are not supported, only panel-observed data")
     if (!is.null(transitions) && !is.numeric(transitions)) stop("expected \"transitions\" to be numeric")
     if (!is.numeric(timegroups) || length(timegroups) != 1) stop ("expected \"timegroups\" to be a single number")
     if (!is.numeric(intervalgroups) || length(intervalgroups) != 1) stop ("expected \"intervalgroups\" to be a single number")
@@ -27,18 +27,19 @@ pearson.msm <- function(x, transitions=NULL, timegroups=3, intervalgroups=3, cov
 
     ## Label various constants
     nst <- x$qmodel$nstates
-    exact.death <- any(dat$obstype == 3)
+    exact.death <- any(dat$"(obstype)" == 3)
     dstates <- if (exact.death) absorbing.msm(x) else NULL
     ndstates <- if (exact.death) transient.msm(x) else 1:nst
     nndstates <- length(ndstates)
 
-    ## Add a few useful variables to the data
-    obstypename <- if (x$hmodel$hidden || x$cmodel$ncens > 0) "obstype" else "obstype.obs"
-    od <- as.data.frame(dat[c("subject","time","state",obstypename)])
+    ## Do minimum needed to port this to msm version 1.4
+    od <- dat[,c("(subject)","(time)","(state)","(obstype)")]
+    names(od) <- c("subject","time","state","obstype")
+    od$cov <- dat[,attr(dat,"covnames.q")]
+    if (x$emodel$misc) od$misccov <- dat[,attr(dat,"covnames.e")]
     ncovs <- x$qcmodel$ncovs
-    if (ncovs > 0) od$cov <- dat$cov.orig
[…]

### 2014-06-03 pkg/R/simul.R

@@ -333,24 +333,21 @@ simhidden.msm <- function(state, hmodel, nstates, beta=NULL, x=NULL)
 ### Used for parametric bootstrap in pearson.msm
 
 simfitted.msm <- function(x, drop.absorb=TRUE, drop.pci.imp=TRUE){
-    dat <- x$data
-    dat$cens <- ifelse(x$data$state %in% 1:x$qmodel$nstates, 0, x$data$state) # 0 if not censored, cens indicator if censored, so that censoring is retained in simulated data
-    sim.df <- as.data.frame(dat[c("subject","time","cens")])
-    sim.df$obstrue <- dat$obstrue # NULL if not HMM
-    sim.df$obstype <- if (!is.null(dat$obstype.obs)) dat$obstype.obs else dat$obstype
-    sim.df$pci.imp <- dat$pci.imp
+    sim.df <- x$data$mf
+    x$data <- expand.data(x)
+    sim.df$"(cens)" <- ifelse(sim.df$"(state)" %in% 1:x$qmodel$nstates, 0, sim.df$"(state)") # 0 if not censored, cens indicator if censored, so that censoring is retained in simulated data.  TODO used in pearson?
     if (x$qcmodel$ncovs > 0) {
-        sim.df <- cbind(sim.df, dat$cov[,x$qcmodel$covlabels,drop=FALSE])
-        sim.df <- cbind(sim.df, dat$cov.orig[,x$qcmodel$covlabels.orig,drop=FALSE])
+        sim.df <- cbind(sim.df, x$data$mm.cov)
         cov.effs <- lapply(x$Qmatrices, function(y)t(y)[t(x$qmodel$imatrix)==1])[x$qcmodel$covlabels]
-    }
-    else cov.effs <- NULL
+    } else cov.effs <- NULL
     if (x$ecmodel$ncovs > 0) {
-        sim.df <- cbind(sim.df, dat$cov[,setdiff(x$ecmodel$covlabels,x$qcmodel$covlabels),drop=FALSE])
-        sim.df <- cbind(sim.df, dat$cov.orig[,setdiff(x$ecmodel$covlabels.orig,x$qcmodel$covlabels.orig),drop=FALSE])
+        sim.df <- cbind(sim.df, x$data$mm.mcov)
         misccov.effs <- lapply(x$Ematrices, function(y)t(y)[t(x$emodel$imatrix)==1])[x$ecmodel$covlabels]
-    }
-    else misccov.effs <- NULL
+    } else misccov.effs <- NULL
[…]

### 2014-06-03 pkg/R/utils.R

@@ -143,7 +143,7 @@ rtnorm <- function (n, mean = 0, sd = 1, lower = -Inf, upper = Inf) {
                           (lower == -Inf & upper > 0) |
                           (is.finite(lower) & is.finite(upper) & (lower < 0) & (upper > 0) & (upper-lower > sqrt(2*pi)))
                           ),
-                         0, # standard "simulate from normal and reject if outside limits" method. Use if bounds are wide.  FIXME HSOULD BE
+                         0, # standard "simulate from normal and reject if outside limits" method. Use if bounds are wide.
                          ifelse(
                                 (lower >= 0 & (upper > lower + 2*sqrt(exp(1)) /
                                  (lower + sqrt(lower^2 + 4)) * exp((lower*2 - lower*sqrt(lower^2 + 4)) / 4))),
@@ -344,6 +344,14 @@ rpexp <- function(n=1, rate=1, t=0)
       ret
   }
 
+mean.pexp <- function(rate=1, t=0){
+    if (length(t) != length(rate)) stop("length of t must be equal to length of rate")
+    if (!isTRUE(all.equal(0, t[1]))) stop("first element of t should be 0")
+    if (is.unsorted(t)) stop("t should be in increasing order")
+    t1 <- c(t[-1], Inf)
+    sum(    (t + 1/rate)*exp(-rate*t) - (t1 - 1/rate)*exp(-rate*t1)  )
+}
+
 qgeneric <- function(pdist, p, ...)
 {
     args <- list(...)
@@ -399,6 +407,26 @@ glogit <- function(x, a, b) {
     ret
 }
 
+dglogit <- function(x, a, b) {
+    if (is.null(a)) a <- -Inf
+    if (is.null(b)) b <- Inf
+    ret <- numeric(length(x))
+    attributes(ret) <- attributes(x)
+    nn <- is.infinite(a) & is.infinite(b)
+    nb <- is.infinite(a) & is.finite(b)
+    an <- is.finite(a) & is.infinite(b)
+    ab <- is.finite(a) & is.finite(b)
+    ret[nn] <- 1
+    ret[nb] <- -1 / (b[nb] - x[nb])
+    ret[an] <- 1 / (x[an] - a[an])
[…]

### 2014-06-03 pkg/src/hmm.c

@@ -47,8 +47,8 @@ double hmmLNorm(double x, double *pars)
 
 double hmmExp(double x, double *pars)
 {
-    double rate = 1 / pars[0];
-    return dexp(x, rate, 0);
+    double mean = 1 / pars[0];
+    return dexp(x, mean, 0);
 }
 
 double hmmGamma(double x, double *pars)

### 2014-06-03 pkg/src/hmm.h

@@ -1,5 +1,6 @@
 
 typedef double (*hmmfn)(double x, double *pars);
+typedef void (*dhmmfn)(double x, double *pars, double *d);
 typedef double (*linkfn)(double x);
 
 double hmmCat(double x, double *pars);
@@ -18,3 +19,20 @@ double hmmMEUnif(double x, double *pars);
 double hmmNBinom(double x, double *pars);
 double hmmBeta(double x, double *pars);
 double hmmT(double x, double *pars);
+
+void DhmmCat(double x, double *pars, double *d);
+void DhmmIdent(double x, double *pars, double *d);
+void DhmmUnif(double x, double *pars, double *d);
+void DhmmNorm(double x, double *pars, double *d);
+void DhmmLNorm(double x, double *pars, double *d);
+void DhmmExp(double x, double *pars, double *d);
+void DhmmGamma(double x, double *pars, double *d);
+void DhmmWeibull(double x, double *pars, double *d);
+void DhmmPois(double x, double *pars, double *d);
+void DhmmBinom(double x, double *pars, double *d);
+void DhmmTNorm(double x, double *pars, double *d);
+void DhmmMETNorm(double x, double *pars, double *d);
+void DhmmMEUnif(double x, double *pars, double *d);
+void DhmmNBinom(double x, double *pars, double *d);
+void DhmmBeta(double x, double *pars, double *d);
+void DhmmT(double x, double *pars, double *d);

### 2014-06-03 pkg/src/hmmderiv.c

@@ -1,0 +1,113 @@
+#include "hmm.h"
+#include <Rmath.h>
+
+/* Derivatives of PDF w.r.t parameters, for each HMM outcome
+ * distribution */
+
+void DhmmCat(double x, double *pars, double *d)
+{
+/* f(x) = p_y if x=y, and 0 otherwise
+   df / dp_y |x   = 1 if x=y, and 0 otherwise */
+    int i;
+    int cat = fprec(x, 0);
+    int ncats = fprec(pars[0], 0);
+    for (i=0; i<ncats+1; ++i)
+	d[i] = 0;
+    if ((cat > ncats) || (cat < 1)) return;
+    d[1+cat] = 1;
+}
+
+void DhmmIdent(double x, double *pars, double *d){
+    d[0] = 0;
+}
+
+void DhmmUnif(double x, double *pars, double *d)
+{
+    d[0]=0; d[1]=0; /* uniform parameters are not estimated */
+}
+
+void DhmmNorm(double x, double *pars, double *d)
+{
+    double mean = pars[0], sd = pars[1], f;
+    f = dnorm(x, mean, sd, 0); 
+    d[0] = f*(x - mean)/R_pow_di(sd,2);
+    d[1] = f*(R_pow_di((x-mean)/sd, 2) - 1)/sd;
+}
+
+void DhmmLNorm(double x, double *pars, double *d)
+{
+    double meanlog = pars[0], sdlog = pars[1], f;
+    f = dlnorm(x, meanlog, sdlog, 0);
+    d[0] = f*(log(x) - meanlog)/R_pow_di(sdlog,2);
+    d[1] = f*(R_pow_di((log(x)-meanlog)/sdlog, 2) - 1)/sdlog;
+}
+
+void DhmmExp(double x, double *pars, double *d)
+{
+    double rate = pars[0];
+    d[0] = (1 - rate*x)*exp(-rate*x);
+}
+
+void DhmmGamma(double x, double *pars, double *d)
+{
+    double shape = pars[0], rate=pars[1], scale = 1 / rate, f;
+    f = dgamma(x, shape, scale, 0);
+    d[0] = f*(log(rate) + log(x) - digamma(shape));
+    d[1] = f*(shape/rate - x);
+}
+
+void DhmmWeibull(double x, double *pars, double *d)
+{
+    double shape = pars[0], scale = pars[1], f, rp;
+    f = dweibull(x, shape, scale, 0);
[…]

### 2014-06-03 pkg/src/lik.c

@@ -13,7 +13,6 @@
 #include <Rmath.h>
 #define NODEBUG
 #define NOVITDEBUG
-#define NOFISHDEBUG
 #define NODERIVDEBUG
 
 linkfn LINKFNS[3][2] = {
@@ -40,6 +39,26 @@ hmmfn HMODELS[] = {
     hmmNBinom,
     hmmBeta,
     hmmT
+};
+
+/* MUST KEEP THIS IN SAME ORDER AS .msm.HMODELPARS IN R/constants.R */
+dhmmfn DHMODELS[] = {
+    DhmmCat,
+    DhmmIdent,
+    DhmmUnif,
+    DhmmNorm,
+    DhmmLNorm,
+    DhmmExp,
+    DhmmGamma,
+    DhmmWeibull,
+    DhmmPois,
+    DhmmBinom,
+    DhmmTNorm,
+    DhmmMETNorm,
+    DhmmMEUnif,
+    DhmmNBinom,
+    DhmmBeta,
+    DhmmT
 };
 
 /* MUST MATCH order of .msm.CTASKS in R/constants.R */
@@ -104,10 +123,10 @@ void GetCensored (double obs, cmodel *cm, int *nc, double **states)
 }
 
 /*
-  Calculate p (obs curr | true i) for hidden Markov models
+  Calculate p (obs curr | true i) for hidden Markov or censoring models
   If observation is not necessarily of the true state (!obstrue),
   then this is just the HMM outcome probability (summed over censor set if necessary)
-  If obstrue, this observation is not misclassified.
+  If obstrue (observation not misclassified) or censored state.
   e.g. censor set 1,2,3,    state set 1,2,3,4,
   pout =   if i in curr 1, else 0
 */
@@ -117,9 +136,11 @@ void GetOutcomeProb(double *pout, double *curr, int nc, double *hpars, hmodel *h
     int i, j;
     for (i=0; i<qm->nst; ++i) {
 	pout[i] = 0;
-	if (!obstrue) {
-	    for (j=0; j<nc; ++j)
+	if (hm->hidden && !obstrue) {
+	    for (j=0; j<nc; ++j){
 		pout[i] += (HMODELS[hm->models[i]])(curr[j], &(hpars[hm->firstpar[i]]));
+		/*		printf("hm=%d,curr=%f,fp=%d\n",hm->models[i],curr[j],hm->firstpar[i]);*/
+	    }
 	}
 	else {
 	    for (j=0; j<nc; ++j)
[…]

### 2014-06-03 pkg/src/pijt.c

@@ -529,7 +529,7 @@ void dpijdeath(int r, int s, Array3 dpmat, Matrix pmat, Array3 dqmat, Matrix qma
     }
 }
 
-/* Derivatives of P matrix */
+/* Derivatives of P matrix (Kalbfleisch & Lawless) */
 
 void DPmat(Array3 dpmat, double t, Array3 dqmat, Matrix qmat, int n, int npars, int exacttimes)
 {

### 2014-06-03 pkg/tests/misc.R

@@ -22,13 +22,12 @@ if (developer.local) {
                                 qmatrix = oneway4.q, ematrix=ematrix, death = 4, # pci=5, # fixedpars=1:5,
                                 control = list(trace=1, REPORT=1), method="BFGS"))
     stopifnot(isTRUE(all.equal(3951.82919869367, misc.msm$minus2loglik, tol=1e-06)))
-    if(interactive()) save(misc.msm, file="~/msm/devel/models/misc.msm.rda")
+    if(interactive()) save(misc.msm, file="~/msm/devel/models/1.4/misc.msm.rda")
 }
 
 ## Covs on transition rates
 misccov.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                 qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars = TRUE,
-                control = list(trace=1, REPORT=1), method="BFGS",
                 covariates = ~ sex, covinits=list(sex=rep(0.1, 5)))
 #stopifnot(isTRUE(all.equal(4299.35653620144, misccov.msm$minus2loglik, tol=1e-06))) # with last obs included in cov means, as pre 1.2.3
 stopifnot(isTRUE(all.equal(4299.38058878142, misccov.msm$minus2loglik, tol=1e-06)))
@@ -37,8 +36,7 @@ stopifnot(isTRUE(all.equal(4299.38058878142, misccov.msm$minus2loglik, tol=1e-06
 ## Covs on misc probs, old way.
 misccov.msm <- msm(state ~ years, subject = PTNUM, data = cav,
                    qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars=TRUE,
-                   misccovariates = ~dage + sex, misccovinits = list(dage=c(0.01,0.02,0.03,0.04), sex=c(-0.013,-0.014,-0.015,-0.016)),
-                   control = list(trace=1, REPORT=1), method="BFGS")
+                   misccovariates = ~dage + sex, misccovinits = list(dage=c(0.01,0.02,0.03,0.04), sex=c(-0.013,-0.014,-0.015,-0.016)))
 #stopifnot(isTRUE(all.equal(4304.90609473048, misccov.msm$minus2loglik, tol=1e-06))) # with last obs included in cov means
[…]

### 2014-06-03 pkg/tests/simple.R

@@ -2,7 +2,6 @@ source("local.R")
 library(msm)
 #library(msm, lib.loc="~/lib/R")
 #library(msm, lib.loc="~/msm/lib/sun/0.7.5")
-data(cav)
 
 ### TESTS FOR SIMPLE NON-HIDDEN MARKOV MODELS
 
@@ -33,18 +32,12 @@ stopifnot(isTRUE(all.equal(4113.16601901957, cav.msm$minus2loglik, tol=1e-06)))
 
 if (developer.local) {
     system.time(cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                               qmatrix = twoway4.q, death = TRUE, fixedpars=FALSE, # opt.method="nlm"
+                               qmatrix = twoway4.q, death = TRUE, fixedpars=FALSE,  
                                method="BFGS",
                                control=list(trace=5, REPORT=1, fnscale=1)
                                )
                 )
     stopifnot(isTRUE(all.equal(3968.7978930519, cav.msm$minus2loglik, tol=1e-06)))
-
-    pnext.msm(cav.msm)
-    pnext.msm(cav.msm, ci="normal")
-    pnext.msm(cav.msm, ci="bootstrap", B=3)
-    if (0)
-        pnext.msm(cav.msm, ci="bootstrap")
 }
 
 ## No death state.
@@ -78,7 +71,7 @@ stopifnot(isTRUE(all.equal(4909.17147259115, cav.msm$minus2loglik, tol=1e-06)))
 if (developer.local) {
     system.time(cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                                  qmatrix = twoway4.q, death = TRUE, fixedpars=FALSE,
-                                 covariates = ~ sex, method="BFGS", control=list(trace=5, REPORT=1))) # 44.13 on new, 260.14 on old
+                                 covariates = ~ sex, method="BFGS", control=list(trace=5, REPORT=1)))
     stopifnot(isTRUE(all.equal(3954.77700438268, cav.msm$minus2loglik, tol=1e-06)))
     qe <- qmatrix.msm(cav.msm)$estimates
     qs <- qmatrix.msm(cav.msm)$SE
@@ -110,7 +103,6 @@ if (developer.local) {
 }
 
[…]

### 2014-06-18 pkg/R/boot.R
testthat tests
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@54 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -170,7 +170,7 @@ totlos.ci.msm <- function(x, start=1, end=NULL, fromt=0, tot=Inf, covariates="me
 }
 
 efpt.ci.msm <- function(x, qmatrix=NULL, tostate, start, covariates="mean", cl=0.95, B=1000, cores=NULL) {
-    t.list <- boot.msm(x, function(x)efpt.msm(x, qmatrix, start, tostate, covariates, ci="none"), B=B, cores=cores)
+    t.list <- boot.msm(x, function(x)efpt.msm(x=x, qmatrix=qmatrix, start=start, tostate=tostate, covariates=covariates, ci="none"), B=B, cores=cores)
     t.array <- do.call("rbind", t.list)
     apply(t.array, 2, function(x)(quantile(x, c(0.5 - cl/2, 0.5 + cl/2))))
 }

### 2014-06-18 pkg/R/msm.R

@@ -84,6 +84,7 @@ msm <- function(formula, subject=NULL, data=list(), qmatrix, gen.inits=FALSE,
     covnames <- unique(unlist(lapply(forms, varnames)))
     temp[["formula"]] <- if (length(covnames) > 0) reformulate(covnames) else ~1
     temp[["na.action"]] <- na.pass # run na.action later so we can pass aux info to it
+    temp[["data"]] <- data
     mf <- eval(temp, parent.frame())
     ## remember user-specified names for later (e.g. bootstrap/cross validation)
     attr(mf, "usernames") <- c(state=all.vars(formula[[2]]), time=all.vars(formula[[3]]), subject=as.character(temp$subject), obstype=as.character(substitute(obstype)), obstrue=as.character(temp$obstrue))
@@ -532,6 +533,7 @@ msm.form.obstype <- function(mf, obstype, dmodel, exacttimes)
     if (!is.null(obstype)) {
         if (!is.numeric(obstype)) stop("obstype should be numeric")
         if (length(obstype) == 1) obstype <- rep(obstype, nrow(mf))
+        else if (length(obstype)!=nrow(mf)) stop("obstype of length ", length(obstype), ", should be length 1 or ",nrow(mf))
         if (any(! obstype[duplicated(mf$"(subject)")] %in% 1:3)) stop("elements of obstype should be 1, 2, or 3") # ignore obstypes at subject's first observation
     }
     else if (!is.null(exacttimes) && exacttimes)

### 2014-06-18 pkg/R/optim.R

@@ -20,10 +20,6 @@ info.supported <- function(msmdata, hmodel, cmodel){
 msm.optim <- function(opt.method, p, hessian, use.deriv, msmdata, qmodel, qcmodel, cmodel, hmodel, ...){
     p$params <- p$allinits
     gr <- if (deriv.supported(hmodel,cmodel) && use.deriv) deriv.msm else NULL
-    if (isTRUE(getOption("msm.test.analytic.derivatives"))){
-        if (!deriv.supported(hmodel,cmodel)) warning("Analytic derivatives not available for this model")
-        else print(deriv.test(msmdata, qmodel, qcmodel, cmodel, hmodel, msm.unfixallparams(p)))
-    }
     optfn <- paste("msm.optim", opt.method, sep=".")
     if (!exists(optfn)) stop("Unknown optimisation method \"", opt.method, "\"")
 
@@ -32,6 +28,10 @@ msm.optim <- function(opt.method, p, hessian, use.deriv, msmdata, qmodel, qcmode
                    cmodel=cmodel, hmodel=hmodel), list(...))
     p <- do.call(optfn, args)
 
+    if (isTRUE(getOption("msm.test.analytic.derivatives"))){
+        if (!deriv.supported(hmodel,cmodel)) warning("Analytic derivatives not available for this model")
+        else p$deriv.test <- deriv.test(msmdata, qmodel, qcmodel, cmodel, hmodel, msm.unfixallparams(p))
+    }
     ## Attach derivative and information matrix at the MLE.
     ## If all parameters fixed, do this at the initial values for all.
     pp <- if (opt.method=="fixed") msm.unfixallparams(p) else p
@@ -196,6 +196,7 @@ msm.optim.bobyqa <- function(p, gr, hessian, msmdata, qmodel, qcmodel, cmodel, h
 deriv.test <- function(msmdata, qmodel, qcmodel, cmodel, hmodel, p){
     an.d <- deriv.msm(p$inits, msmdata, qmodel, qcmodel, cmodel, hmodel, p)
 
+    if (0){ 
     ## fiddly method using stats::numericDeriv
     likwrap <- function(x, ...){
         pars <- list(unlist(list(...)))
@@ -209,7 +210,9 @@ deriv.test <- function(msmdata, qmodel, qcmodel, cmodel, hmodel, p){
[…]

### 2014-06-18 pkg/R/outputs.R

@@ -481,7 +481,7 @@ qematrix.msm <- function(x, covariates="mean", intmisc="intens", sojourn=FALSE, 
         covlabels <- x$ecmodel$covlabels
         covmeans <- x$ecmodel$covmeans
     }
-
+    if ((cl < 0) || (cl > 1)) stop("expected cl in [0,1]")
     se <- lse <- numeric(ni)
     if ((nc == 0) && (is.list(covariates)) && (length(covariates) > 0))
         warning(paste("Ignoring covariates - no covariates in model for",
@@ -830,6 +830,7 @@ qratio.msm <- function(x, ind1, ind2,
         stop("ind1 and ind2 must be numeric vectors of length 2")
     if (any (! (ind1 %in% 1 : x$qmodel$nstates))  |  any (! (ind2 %in% 1 : x$qmodel$nstates) ) )
         stop("ind1 and ind2 must be pairs of states in 1, ..., ", x$qmodel$nstates)
+    if ((cl < 0) || (cl > 1)) stop("expected cl in [0,1]")
     if (q[ind2[1], ind2[2]] == 0)
         stop (paste("Denominator q[",ind2[1],",",ind2[2],"", "] is zero\n", sep=""))
     else if (q[ind1[1], ind1[2]] ==  0) {
@@ -1396,7 +1397,7 @@ get.covhist <- function(x, subset=NULL) {
             mf <- mf[subs,,drop=FALSE]
         }
         n <- length(mf$"(subject)")
-        apaste <- do.call("paste", mf[,attr(mf,"covnames")])
+        apaste <- do.call("paste", mf[,attr(mf,"covnames"),drop=FALSE])
         first <- !duplicated(mf$"(subject)"); last <- rev(!duplicated(rev(mf$"(subject)")))
         keep <- (c(0, apaste[1:(n-1)]) != apaste) | first | last
         ## Keep and tabulate unique covariate series
@@ -1411,7 +1412,7 @@ get.covhist <- function(x, subset=NULL) {
         covseries.t <- paste(covseries, change.times, sep="; ")
         ids <- unique(mf$"(subject)")[!duplicated(covseries.t)] # subj ids, one with each distinct series
         ncombs <- table(covseries.t)[unique(covseries.t)]# how many per series
-        covmat <- cbind(subject=mf$"(subject)", time=mf$"(time)", mf[,attr(mf,"covnames")])
+        covmat <- cbind(subject=mf$"(subject)", time=mf$"(time)", mf[,attr(mf,"covnames"),drop=FALSE])
[…]

### 2014-06-18 pkg/R/simul.R

@@ -356,7 +356,7 @@ simfitted.msm <- function(x, drop.absorb=TRUE, drop.pci.imp=TRUE){
                             misccovariates=misccov.effs,
                             drop.absorb=drop.absorb
                             )
-    if (drop.pci.imp) {
+    if (drop.pci.imp & !is.null(boot.df$"(pci.imp)")) {
         boot.df <- boot.df[!boot.df$"(pci.imp)",]
         boot.df$"(pci.imp)" <- NULL
     }

### 2014-06-18 pkg/src/lik.c

@@ -316,7 +316,7 @@ double likhidden(int pt, /* ordinal subject ID */
     double *newp     = Calloc(qm->nst, double);
     double *pout = Calloc(qm->nst, double);
     double lweight, lik, *hpars;
-    int i, obsno, nc=1;
+    int i, obsno, nc=1, allzero=1;
     if (d->firstobs[pt] + 1 == d->firstobs[pt+1])
       return 0; /* individual has only one observation. Shouldn't happen since 1.3.2 */
     /* Likelihood for individual's first observation */
@@ -329,8 +329,14 @@ double likhidden(int pt, /* ordinal subject ID */
       /* Ignore initprobs if observation is known to be the true state
 	 or TODO, can we set it in R to one for obs state, zero for others? */
       if (!d->obstrue[d->firstobs[pt]]) cump[i] = cump[i]*hm->initp[MI(pt,i,d->npts)];
+      if (!all_equal(cump[i], 0)) allzero = 0;
     }
-    //    printf("cump at 0=%16.12lf\n,", cump[0]);
+    if (allzero) {
+// don't leave this warning in - since extreme values visited by the optimiser break it (e.g. for log odds in misc models, see misccov.msm)
+//	warning("First observation of %f for subject number %d out of %d is impossible for given initial state probabilities and outcome model\n", curr[0], pt+1, d->npts);
+//	printf("initp: "); for (i = 0; i < qm->nst; ++i) printf("%f,",hm->initp[MI(pt,i,d->npts)]); printf("\n");
+//	printf("hpars: "); for (i = 0; i<hm->totpars; ++i) printf("%f,",hm->pars[MI(i, d->firstobs[pt], hm->totpars)]); printf("\n");	
+    }
     lweight=0;
     /* Matrix product loop to accumulate the likelihood for subsequent observations */
     for (obsno = d->firstobs[pt]+1; obsno <= d->firstobs[pt+1] - 1; ++obsno)

### 2014-06-18 pkg/tests/misc.R

@@ -1,241 +1,0 @@
-### TESTS OF MISCLASSIFICATION MODELS
-### SPECIFIED USING THE OLD ematrix SYNTAX
-source("local.R")
-
-library(msm)
-# library(msm, lib.loc="~/msm/src/1.2.7")
-# data(cav) # not needed since 1.3
-
-oneway4.q <- rbind(c(0, 0.148, 0, 0.0171), c(0, 0, 0.202, 0.081), c(0, 0, 0, 0.126), c(0, 0, 0, 0))
-ematrix <- rbind(c(0, 0.1, 0, 0),c(0.1, 0, 0.1, 0),c(0, 0.1, 0, 0),c(0, 0, 0, 0))
-rownames(oneway4.q) <- colnames(oneway4.q) <- rownames(ematrix) <- colnames(ematrix) <- c("Well","Mild","Severe","Death")
-
-## Plain misc model with no covs
-
-misc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars=TRUE)
-misc.msm
-stopifnot(isTRUE(all.equal(4296.9155995778, misc.msm$minus2loglik, tol=1e-06)))
-
-if (developer.local) {
-    system.time(misc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                                qmatrix = oneway4.q, ematrix=ematrix, death = 4, # pci=5, # fixedpars=1:5,
-                                control = list(trace=1, REPORT=1), method="BFGS"))
-    stopifnot(isTRUE(all.equal(3951.82919869367, misc.msm$minus2loglik, tol=1e-06)))
-    if(interactive()) save(misc.msm, file="~/msm/devel/models/1.4/misc.msm.rda")
-}
-
-## Covs on transition rates
-misccov.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars = TRUE,
-                covariates = ~ sex, covinits=list(sex=rep(0.1, 5)))
-#stopifnot(isTRUE(all.equal(4299.35653620144, misccov.msm$minus2loglik, tol=1e-06))) # with last obs included in cov means, as pre 1.2.3
-stopifnot(isTRUE(all.equal(4299.38058878142, misccov.msm$minus2loglik, tol=1e-06)))
-
-
-## Covs on misc probs, old way.
[…]

### 2014-06-18 pkg/tests/simple.R

@@ -1,330 +1,0 @@
-source("local.R")
-library(msm)
-#library(msm, lib.loc="~/lib/R")
-#library(msm, lib.loc="~/msm/lib/sun/0.7.5")
-
-### TESTS FOR SIMPLE NON-HIDDEN MARKOV MODELS
-
-twoway4.q <- rbind(c(-0.5, 0.25, 0, 0.25), c(0.166, -0.498, 0.166, 0.166), c(0, 0.25, -0.5, 0.25), c(0, 0, 0, 0))
-twoway4.q2 <- rbind(c(-0.51, 0.24, 0, 0.25), c(0.162, -0.498, 0.168, 0.166), c(0, 0.26, -0.5, 0.25), c(0, 0, 0, 0))
-twoway3.q <- rbind(c(-0.5, 0.25, 0), c(0.166, -0.498, 0.166), c(0, 0.25, -0.5))
-oneway4.q <- rbind(c(0, 0.148, 0, 0.0171), c(0, 0, 0.202, 0.081), c(0, 0, 0, 0.126), c(0, 0, 0, 0))
-rownames(twoway4.q) <- colnames(twoway4.q) <- c("Well","Mild","Severe","Death")
-rownames(oneway4.q) <- colnames(oneway4.q) <- c("Well","Mild","Severe","Death")
-twoway4.i <- twoway4.q; twoway4.i[twoway4.i!=0] <- 1
-oneway4.i <- oneway4.q; oneway4.i[oneway4.i!=0] <- 1
-
-### CAV DATA
-(stab <- statetable.msm(state, PTNUM, data=cav))
-stopifnot(all(stab == c(1367, 46, 4, 204, 134, 13, 44, 54, 107, 148, 48, 55)))
-(cinits <- crudeinits.msm(state ~ years, PTNUM, data=cav, qmatrix=twoway4.q))
-stopifnot(isTRUE(all.equal(c(-0.117314905786477, 0.116817878212849, 0, 0, 0.067989320398981, -0.375848825554382, 0.049084006577444, 0, 0, 0.137134030945518, -0.256747111328168, 0, 0.049325585387496, 0.121896916396016, 0.207663104750724, 0), as.numeric(cinits))))
-
-## Simple model
-cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                 qmatrix = twoway4.q, death = TRUE, fixedpars=TRUE,
[…]

### 2014-06-18 pkg/tests/testthat/helper.r

@@ -1,0 +1,32 @@
+twoway4.q <- rbind(c(-0.5, 0.25, 0, 0.25), c(0.166, -0.498, 0.166, 0.166), c(0, 0.25, -0.5, 0.25), c(0, 0, 0, 0))
+twoway4.q2 <- rbind(c(-0.51, 0.24, 0, 0.25), c(0.162, -0.498, 0.168, 0.166), c(0, 0.26, -0.5, 0.25), c(0, 0, 0, 0))
+twoway3.q <- rbind(c(-0.5, 0.25, 0), c(0.166, -0.498, 0.166), c(0, 0.25, -0.5))
+oneway4.q <- rbind(c(0, 0.148, 0, 0.0171), c(0, 0, 0.202, 0.081), c(0, 0, 0, 0.126), c(0, 0, 0, 0))
+rownames(twoway4.q) <- colnames(twoway4.q) <- rownames(oneway4.q) <- colnames(oneway4.q) <- c("Well","Mild","Severe","Death")
+twoway4.i <- twoway4.q; twoway4.i[twoway4.i!=0] <- 1
+oneway4.i <- oneway4.q; oneway4.i[oneway4.i!=0] <- 1
+psor.q <- rbind(c(0,0.1,0,0),c(0,0,0.1,0),c(0,0,0,0.1),c(0,0,0,0))
+fiveq <- rbind(c(0,0.01,0,0,0.002), c(0,0,0.07,0,0.01), c(0,0,0,0.07,0.02), c(0,0,0,0,0.03), c(0,0,0,0,0))
+ematrix <- rbind(c(0, 0.1, 0, 0),c(0.1, 0, 0.1, 0),c(0, 0.1, 0, 0),c(0, 0, 0, 0))
+rownames(oneway4.q) <- colnames(oneway4.q) <- rownames(ematrix) <- colnames(ematrix) <- c("Well","Mild","Severe","Death")
+
+cav.cens <- cav
+cav.cens$state[cav$state==4][1:50] <- 99
+cav.cens2 <- cav
+cav.cens2$state[cav$state==4][1:50] <- 99
+cav.cens2$state[cav$state==4][51:100] <- 999
+cav.cens3 <- cav
+ns <- c(cav$state[2:nrow(cav)], 0)
+cav.cens3$state[cav$state==4][1:50] <- 99
+cav.cens3$state[ns==4][1:50] <- 999
+
+
+has_accurate_derivs <- function(){
+    if (!isTRUE(getOption("msm.test.analytic.derivatives")))
[…]

### 2014-06-18 pkg/tests/testthat/test_datasumm.r

@@ -1,0 +1,41 @@
+context("msm data summaries")
+
+test_that("statetable.msm", {
+    stab <- statetable.msm(state, PTNUM, data=cav)
+    expect_that(stab, equals(structure(c(1367L, 46L, 4L, 204L, 134L, 13L, 44L, 54L, 107L, 148L, 48L, 55L), .Dim = 3:4, .Dimnames = structure(list(from = c("1","2", "3"), to = c("1", "2", "3", "4")), .Names = c("from", "to")), class = "table")))
+    expect_equal(as.numeric(stab), c(1367, 46, 4, 204, 134, 13, 44, 54, 107, 148, 48, 55))
+    expect_error(statetable.msm("state","PTNUM"))
+    expect_error(statetable.msm(state,PTNUM), "not found")
+    stabc <- statetable.msm(state, PTNUM, cav.cens)
+    expect_equal(as.numeric(stabc), c(1367, 46, 4, 204, 134, 13, 44, 54, 107, 127, 40, 34, 21, 8, 21))
+})
+
+test_that("crudeinits.msm", {
+    cinits <- crudeinits.msm(state ~ years, PTNUM, data=cav, qmatrix=twoway4.q)
+    expect_equal(as.numeric(cinits), c(-0.117314905786477, 0.116817878212849, 0, 0, 0.067989320398981, -0.375848825554382, 0.049084006577444, 0, 0, 0.137134030945518, -0.256747111328168, 0, 0.049325585387496, 0.121896916396016, 0.207663104750724, 0))
+    expect_error(crudeinits.msm(state ~ years, PTNUM, qmatrix=twoway4.q), "not found")
+})
+
+test_that("crudeinits.msm ignores inconsistent transitions unless exact times", {
+    cav.wrong <- cav; cav.wrong$state[4] <- 1
+    expect_that(crudeinits.msm(state ~ years, PTNUM, oneway4.q, cav), not(throws_error()))
+    expect_warning(msm(state ~ years, subject=PTNUM, data = cav.wrong, qmatrix = twoway4.q, exacttimes=TRUE, fixedpars=TRUE), "inconsistent with intensity")
+    obstype <- rep(2, nrow(cav)); obstype[10] <- 1
+    expect_warning(msm(state ~ years, subject=PTNUM, data = cav.wrong, qmatrix = twoway4.q, obstype=obstype, fixedpars=TRUE), "inconsistent with intensity")
+})
+
[…]

### 2014-06-18 pkg/tests/testthat/test_deriv.r

@@ -1,0 +1,183 @@
+## depends on psor.msm
+
+context("analytic derivatives of likelihood")
+
+test_that("derivatives by subject: sum to overall derivative",{
+    psor.msm <- msm(state ~ months, subject=ptnum, data=psor, qmatrix = psor.q,  covariates = ~ollwsdrt+hieffusn, constraint = list(hieffusn=c(1,1,1),ollwsdrt=c(1,1,2)), fixedpars=FALSE)
+    q.mle <- psor.msm$paramdata$opt$par
+    deriv.overall <- deriv.msm(q.mle, expand.data(psor.msm), psor.msm$qmodel, psor.msm$qcmodel, psor.msm$cmodel, psor.msm$hmodel, psor.msm$paramdata)
+    deriv.subj <- Ccall.msm(q.mle, do.what="deriv.subj", expand.data(psor.msm), psor.msm$qmodel, psor.msm$qcmodel, psor.msm$cmodel, psor.msm$hmodel, psor.msm$paramdata)
+    expect_equal(deriv.overall, colSums(deriv.subj))
+})
+
+options(msm.test.analytic.derivatives=TRUE)
+
+test_that("analytic derivatives match numeric",{
+    cav.msm <- msm(state ~ years, subject=PTNUM, data = cav, qmatrix = twoway4.q, death = TRUE, fixedpars=TRUE)
+    expect_that(cav.msm, has_accurate_derivs())
+    cav.msm <- msm( state ~ years, subject=PTNUM, data = cav, qmatrix = twoway4.q, death = FALSE, fixedpars=TRUE)
+    expect_that(cav.msm, has_accurate_derivs())
+    cav.msm <- msm( state ~ years, subject=PTNUM, data = cav, qconstraint = c(1,1,2,2,2,3,3),  qmatrix = twoway4.q, death = FALSE, fixedpars=TRUE)
+    expect_that(cav.msm, has_accurate_derivs())
+    psor.0.q <- rbind(c(0,0.1,0,0),c(0,0,0.2,0),c(0,0,0,0.3),c(0,0,0,0))
+    psor.msm <- msm(state ~ months, subject=ptnum, data=psor, qmatrix = psor.0.q, fixedpars=TRUE)
+    expect_that(psor.msm, has_accurate_derivs())
[…]

### 2014-06-18 pkg/tests/testthat/test_models.r

@@ -1,0 +1,652 @@
+context("msm simple model likelihoods")
+
+test_that("simple model, death",{
+    cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
+                   qmatrix = twoway4.q, death = TRUE, fixedpars=TRUE,
+                   method="BFGS", control=list(trace=5, REPORT=1))
+    expect_equal(4908.81676837903, cav.msm$minus2loglik)
+})
+
+test_that("simple model, no death",{
+    cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
+                   qmatrix = twoway4.q, death = FALSE, fixedpars=TRUE)
+    expect_equal(4833.00640639644, cav.msm$minus2loglik)
+})
+
+test_that("simple model, covariates",{
+    cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
+                   qmatrix = twoway4.q, death = TRUE, fixedpars=TRUE,
+                   covariates = ~ sex, covinits = list(sex=rep(0.01, 7)))
+    expect_equal(4909.17147259115, cav.msm$minus2loglik)
+})
+
+test_that("autogenerated inits reproduce crudeinits",{
+    cinits <- crudeinits.msm(state ~ years, PTNUM, data=cav, qmatrix=twoway4.q)
+    expect_equal(msm( state ~ years, subject=PTNUM, data = cav, qmatrix = cinits, death = TRUE, fixedpars=TRUE)$minus2loglik,
+                 msm( state ~ years, subject=PTNUM, data = cav, qmatrix=twoway4.i, gen.inits=TRUE, death = TRUE, fixedpars=TRUE)$minus2loglik)
+})
+
+test_that("data as global variables",{
+    state.g <- cav$state; time.g <- cav$years; subj.g <- cav$PTNUM
+    cav.msm <- msm(state.g ~ time.g, subject=subj.g, qmatrix = twoway4.i, gen.inits=TRUE, fixedpars=TRUE)
+    expect_equal(4119.9736299032, cav.msm$minus2loglik)
+})
+
+psor.msm <- msm(state ~ months, subject=ptnum, data=psor, qmatrix = psor.q,  covariates = ~ollwsdrt+hieffusn, constraint = list(hieffusn=c(1,1,1),ollwsdrt=c(1,1,2)))
+test_that("psor model: covariates, constraints",{
[…]

### 2014-06-18 pkg/tests/testthat/test_models_hmm.r

@@ -1,0 +1,143 @@
+context("Hidden Markov model likelihoods")
+
+three.q <- rbind(c(0, exp(-6), exp(-9)), c(0, 0, exp(-6)), c(0, 0, 0))
+
+four.q <-  rbind(c(0, exp(-6), 0, exp(-9)), c(0, 0, exp(-6.01), exp(-9)), c(0, 0, 0, exp(-6.02)), c(0, 0, 0, 0))
+
+five.q <-  rbind(c(0, exp(-6), 0, 0, exp(-9)),
+                 c(0, 0, exp(-6.01), 0, exp(-9)),
+                 c(0, 0, 0, exp(-6.02), exp(-6.03)),
+                 c(0, 0, 0, 0, exp(-6.04)),
+                 c(0, 0, 0, 0, 0))
+
+hmodel3 <- list(hmmNorm(mean=100, sd=16), hmmNorm(mean=54, sd=18), hmmIdent(999))
+hmodel4 <- list(hmmNorm(mean=100, sd=16), hmmNorm(mean=72.5, sd=10), hmmNorm(mean=42.5, sd=18), hmmIdent(999))
+hmodel5 <- list(hmmNorm(mean=100, sd=16), hmmNorm(mean=72.5, sd=10), hmmNorm(mean=62.5, sd=10), hmmNorm(mean=42.5, sd=18), hmmIdent(999))
+
+
+test_that("HMM normal likelihoods: FEV data",{
+    (fev3.hid <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q, death=3, hmodel=hmodel3, fixedpars=TRUE))
+    expect_equal(52388.7381942858, fev3.hid$minus2loglik, tol=1e-06)
+    (fev4.hid <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=four.q, death=4, hmodel=hmodel4, fixedpars=TRUE))
+    expect_equal(50223.9497625937, fev4.hid$minus2loglik, tol=1e-06)
+    (fev5.hid <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=five.q, death=5, hmodel=hmodel5, fixedpars=TRUE))
+    expect_equal(49937.9840668066, fev5.hid$minus2loglik, tol=1e-06)
+})
+
+test_that("HMM with obstrue",{
+    fev$obstrue <- as.numeric(fev$fev==999)
[…]

### 2014-06-18 pkg/tests/testthat/test_models_misc.r

@@ -1,0 +1,102 @@
+context("msm misclassification model likelihoods")
+
+test_that("cav misclassification model with no covariates",{
+    misc.msm <- msm(state ~ years, subject = PTNUM, data = cav, qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars=TRUE)
+    expect_equal(4296.9155995778, misc.msm$minus2loglik, tol=1e-06)
+    miscnew.msm <- msm(state ~ years, subject = PTNUM, data = cav,
+                       qmatrix = oneway4.q, death = 4, fixedpars=TRUE,
+                       hmodel=list(
+                       hmmCat(prob=c(0.9, 0.1, 0, 0)),
+                       hmmCat(prob=c(0.1, 0.8, 0.1, 0)),
+                       hmmCat(prob=c(0, 0.1, 0.9, 0)), hmmIdent())
+                       )
+    expect_equal(miscnew.msm$minus2loglik, misc.msm$minus2loglik)  
+})
+
+test_that("cav misclassification model with covariates on transition rates",{
+    misccov.msm <- msm(state ~ years, subject = PTNUM, data = cav, qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars = TRUE, covariates = ~ sex, covinits=list(sex=rep(0.1, 5)))
+    expect_equal(4299.38058878142, misccov.msm$minus2loglik, tol=1e-06)
+})
+
+test_that("cav misclassification model with covariates on misclassification probabilities",{
+    misccov.msm <- msm(state ~ years, subject = PTNUM, data = cav,
+                       qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars=TRUE,
+                       misccovariates = ~dage + sex, misccovinits = list(dage=c(0.01,0.02,0.03,0.04), sex=c(-0.013,-0.014,-0.015,-0.016)))
+    expect_equal(4306.3077053482, misccov.msm$minus2loglik, tol=1e-06)
+    misccovnew.msm <- msm(state ~ years, subject = PTNUM, data = cav,
+                          qmatrix = oneway4.q, death = 4, fixedpars=TRUE, center=TRUE,
+                          hmodel=list(
+                          hmmCat(prob=c(0.9, 0.1, 0, 0)),
+                          hmmCat(prob=c(0.1, 0.8, 0.1, 0)),
[…]

### 2014-06-24 pkg/R/boot.R
Changes for 1.3.3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@55 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -175,6 +175,17 @@ efpt.ci.msm <- function(x, qmatrix=NULL, tostate, start, covariates="mean", cl=0
     apply(t.array, 2, function(x)(quantile(x, c(0.5 - cl/2, 0.5 + cl/2))))
 }
 
+ppass.ci.msm <- function(x, qmatrix, tot, start, covariates="mean", piecewise.times=NULL, piecewise.covariates=NULL, cl=0.95, B=1000, cores=NULL,...) {
+    t.list <- boot.msm(x, function(x)ppass.msm(x=x, qmatrix=qmatrix, tot=tot, start=start, covariates=covariates,
+                                                piecewise.times=piecewise.times, piecewise.covariates=piecewise.covariates,
+                                                ci="none",...), B=B, cores=cores)
+    nst <- ncol(t.list[[1]])
+    t.array <- do.call("rbind", lapply(t.list, as.vector))
+    ci <- apply(t.array, 2, function(x)(quantile(x, c(0.5 - cl/2, 0.5 + cl/2))))
+    di <- dimnames(t.list[[1]])
+    list(L=matrix(ci[1,],ncol=nst,dimnames=di), U=matrix(ci[2,],ncol=nst, dimnames=di))
+}
+
 expected.ci.msm <- function(x,
                             times=NULL,
                             timezero=NULL,
@@ -282,6 +293,17 @@ efpt.normci.msm <- function(x, qmatrix=NULL, tostate, start, covariates="mean", 
     t.list <- normboot.msm(x, function(x)efpt.msm(x=x, qmatrix=qmatrix, tostate=tostate, start=start, covariates=covariates, ci="none", ...), B)
     t.array <- do.call("rbind", t.list)
     apply(t.array, 2, function(x)(quantile(x, c(0.5 - cl/2, 0.5 + cl/2))))
+}
+
+ppass.normci.msm <- function(x, qmatrix, tot, start, covariates="mean", piecewise.times=NULL, piecewise.covariates=NULL, cl=0.95, B=1000, ...) {
+    t.list <- normboot.msm(x, function(x)ppass.msm(x=x, qmatrix=qmatrix, tot=tot, start=start, covariates=covariates,
+                                                    piecewise.times=piecewise.times, piecewise.covariates=piecewise.covariates,
+                                                    ci="none", ...), B)
+    nst <- ncol(t.list[[1]])
[…]

### 2014-06-24 pkg/R/draic.R

@@ -64,7 +64,7 @@ draic.msm <- function(msm.full,msm.coarse,
     lmm <- logLik.msm(msm.full)  # Complex model on full data
     lss <- logLik.msm(msm.coarse) # Simple model on restricted data
     dfit <- (1/n)*(lss - lsm)
-    res.lik <- c("LL(complex on full)"=lmm,"LL(complex on restricted)"=lsm,"LL(simple on restricted)"=lss)
+    res.lik <- c("complex"=-lsm, "simple"=-lss, "complex-simple"=-(lsm-lss),"(complex-simple)/n"=dfit)
 
     if (!likelihood.only) {
         information <- match.arg(information)
@@ -82,7 +82,7 @@ draic.msm <- function(msm.full,msm.coarse,
             res$paramdata$optpars <- msm.full$paramdata$optpars
             res$paramdata$fixedpars <- c(msm.full$paramdata$fixedpars, res$paramdata$auxpars)
             I <- information.msm(res$paramdata$allinits[res$paramdata$optpars], expand.data(res), res$qmodel,
-                                 res$qcmodel, res$cmodel, res$hmodel, res$paramdata)$info
+                                 res$qcmodel, res$cmodel, res$hmodel, res$paramdata)
             I.restricted <- -0.5*(1/n)*I
         }
         npars.restricted <- sum(diag(I.restricted %*% solve(I.complex)))
@@ -94,17 +94,16 @@ draic.msm <- function(msm.full,msm.coarse,
         dlh <- lh.restricted.obj(mle, by.subject=TRUE) - logLik.msm(msm.coarse, by.subject=TRUE)
         omega.sq <- mean(dlh*dlh) - mean(dlh)^2
         half.width <- qnorm(1 - 0.5*(1 - tl)) * sqrt(omega.sq / n)
-        prob.DRAIC <- pnorm(-DRAIC * sqrt(omega.sq / n))
+        prob.DRAIC <- pnorm(-DRAIC / sqrt(omega.sq / n))
         res.int <- c("2.5%"=DRAIC-half.width,"97.5%"=DRAIC+half.width,"Prob<0"=prob.DRAIC)
 
-        res <- c("DRAIC"=DRAIC, "dtrace"=dcompl,
-                 res.lik,
-                 "npars(complex on restricted)"=npars.restricted,
[…]

### 2014-06-24 pkg/R/msm.R

@@ -566,7 +566,7 @@ msm.form.obstrue <- function(mf, hmodel) {
 
 msm.impute.censored <- function(fromstate, tostate, Pmat, cmodel)
 {
-    ## e.g. cmodel$censor 99,999;  cmodel$states 1,2,1,2,3;  cmodel$index 1, 3, 6
+    ## e.g. cmodel$censor 99,999;  cmodel$states 1,2,1,2,3;  cmodel$index 1, 3, 6
     ## Both from and to are censored
     wb <- which ( fromstate %in% cmodel$censor & tostate %in% cmodel$censor)
     for (i in wb) {
@@ -798,7 +798,7 @@ msm.form.dmodel <- function(death, qmodel, hmodel)
     if (is.logical(death) && death==TRUE)
         states <- nstates
     else if (is.logical(death) && death==FALSE)
-        states <- numeric(0) ## Will be changed to -1 when passing to C
+        states <- numeric(0) ## Will be changed to -1 when passing to C
     else if (!is.numeric(death)) stop("Death states indicator must be numeric")
     else if (length(setdiff(death, 1:nstates)) > 0)
         stop("Death states indicator contains states not in ",statelist)
@@ -1329,109 +1329,61 @@ Ccall.msm <- function(params, do.what="lik", msmdata, qmodel, qcmodel, cmodel, h
     mm.cov <- if (agg) msmdata$mm.cov.agg else msmdata$mm.cov
     Q <- msm.add.qcovs(qmodel, pars, mm.cov)
     DQ <- if (do.what %in% c("deriv","info","deriv.subj","dpmat")) msm.form.dq(qmodel, qcmodel, pars, p, mm.cov) else NULL
-    hpars <- if (hmodel$hidden) msm.add.hmmcovs(hmodel, pars, msmdata$mm.hcov) else NULL
-    DH <- if (hmodel$hidden && (do.what %in% c("deriv","info","deriv.subj"))) msm.form.dh(hmodel, pars, hpars, paramdata, msmdata$mm.hcov) else NULL
+    H <- if (hmodel$hidden) msm.add.hmmcovs(hmodel, pars, msmdata$mm.hcov) else NULL
+    DH <- if (hmodel$hidden && (do.what %in% c("deriv","info","deriv.subj"))) msm.form.dh(hmodel, pars, H, paramdata, msmdata$mm.hcov) else NULL
[…]

### 2014-06-24 pkg/R/outputs.R

@@ -1880,6 +1880,7 @@ scoreresid.msm <- function(x, plot=FALSE){
 # Function to calculate expected first passage times for continuous-time Markov chain with arbitrary Q matrix
 # Returns vector with EFPT for each "from" state in the state space.
 # Could also get CDF simply by making tostate absorbing and calculating pmatrix.
+# TODO time-dependent covariates
 
 efpt.msm <- function(x=NULL, qmatrix=NULL, tostate, start="all", covariates="mean",
                      ci=c("none","normal","bootstrap"), cl = 0.95, B = 1000, cores=NULL, ...)
@@ -1933,35 +1934,46 @@ efpt.msm <- function(x=NULL, qmatrix=NULL, tostate, start="all", covariates="mea
     if (ci=="none") est else rbind(est, e.ci)
 }
 
-## TODO matrix probabilities of passage through each to-state in an interval, given start in each from-state
-## Make every state other than the from-state absorbing,
-## and calculate the P matrix
 
-## ppass.msm <- function(x=NULL, qmatrix=NULL, tostate, covariates="mean",
-##                       ci=c("none","normal","bootstrap"), cl = 0.95, B = 1000)
-## {
-##     ci <- match.arg(ci)
-##     if (!is.null(x)) {
-##         if (!inherits(x, "msm")) stop("expected x to be a msm model")
-##         qmatrix <- qmatrix.msm(x, covariates=covariates, ci="none")
-##     }
-##     else if (!is.null(qmatrix)) {
-##         if (!is.matrix(qmatrix) || (nrow(qmatrix) != ncol(qmatrix)))
-##             stop("expected qmatrix to be a square matrix")
-##         if (ci != "none") {warning("No fitted model supplied: not calculating confidence intervals."); ci <- "none"}
-##     }
-
-##     res <- matrix(dim=dim(qmatrix))
-##     states <- 1:nrow(qmatrix)
-##     for (i in states) {
-##         Qred <- qmatrix; Qred[,states[-i]] <- 0
-##         res[i,] <- MatrixExp(Qred)
-##     }
-
-## TODO handle time dependent covs in x
-
-#    p.ci <- switch(ci,
-#                   bootstrap = ppass.ci.msm(x=x, qmatrix=qmatrix, tostate=tostate, covariates=covariates, cl=cl, B=B),
[…]

### 2014-06-24 pkg/R/simul.R

@@ -113,7 +113,7 @@ getobs.msm <- function(sim, obstimes, death=FALSE, drop.absorb=TRUE)
 
 ### Simulate a multi-state Markov or hidden Markov model dataset using fixed observation times
 
-### Would it be better to make specification of covariate model consistent with model fitting function?
+### Would it be better to make specification of covariate model consistent with model fitting function?
 ### e.g. separate hcovariates and  covariates formulae,
 ### plus covinits and hcovinits?
 
@@ -137,7 +137,7 @@ simmulti.msm <- function(data,           # data frame with subject, times, covar
     msm.check.qmatrix(qmatrix)
     qmatrix <- msm.fixdiag.qmatrix(qmatrix)
 
-### Subject, time and state
+### Subject, time and state
     if (!("subject" %in% names(data)))
         data$subject <- rep(1, nrow(data))
     if (!("time" %in% names(data)))
@@ -198,8 +198,9 @@ simmulti.msm <- function(data,           # data frame with subject, times, covar
 
 ### Starting states
     if (missing(start)) start <- rep(1, n)
+    else if (length(start) == 1) start <- rep(start, n)
     else if (length(start) != n)
-        stop("Supplied ", length(start), " starting states, expected ", n)
+        stop("Supplied ", length(start), " starting states, expected 1 or ", n)
 
     nq <- length(qmatrix[qmatrix > 0])
     misspeccovs <- covnames[sapply(covariates, length) != nq]
@@ -311,7 +312,7 @@ simhidden.msm <- function(state, hmodel, nstates, beta=NULL, x=NULL)
             ## don't change the underlying state if the HMM is the null (identity) model
             if (!(hmodel[[i]]$label=="identity" && (length(hmodel[[i]]$pars) == 0)))  {
                 ## simulate from the sampling function "r" in the HMM object
-                ## transform the location parameter by covariates if necessary
+                ## transform the location parameter by covariates if necessary
                 rcall <- list(n=length(state[state==i]))
                 if (!is.null(beta[[i]])) {
                     link <- get(hmodel[[i]]$link)

### 2014-06-24 pkg/src/lik.c

@@ -11,6 +11,7 @@
 #include "msm.h"
 #include "hmm.h"
 #include <Rmath.h>
+#include <Rdefines.h>
 #define NODEBUG
 #define NOVITDEBUG
 #define NODERIVDEBUG
@@ -111,14 +112,14 @@ void GetCensored (double obs, cmodel *cm, int *nc, double **states)
 	    ++k;
 	if (k < cm->ncens) {
 	    cens = 1;
-	    n =  cm->censstind[k+1] - cm->censstind[k];
+	    n =  cm->index[k+1] - cm->index[k];
 	}
 	else n = 1;
     }
     if (cm->ncens == 0 || !cens)
 	(*states)[0] = obs;
-    else { for (j = cm->censstind[k]; j < cm->censstind[k+1]; ++j)
-	(*states)[j - cm->censstind[k]] = cm->censstates[j]; }
+    else { for (j = cm->index[k]; j < cm->index[k+1]; ++j)
+	(*states)[j - cm->index[k]] = cm->states[j]; }
     *nc = n;
 }
 
@@ -331,11 +332,8 @@ double likhidden(int pt, /* ordinal subject ID */
       if (!d->obstrue[d->firstobs[pt]]) cump[i] = cump[i]*hm->initp[MI(pt,i,d->npts)];
       if (!all_equal(cump[i], 0)) allzero = 0;
     }
-    if (allzero) {
-// don't leave this warning in - since extreme values visited by the optimiser break it (e.g. for log odds in misc models, see misccov.msm)
-//	warning("First observation of %f for subject number %d out of %d is impossible for given initial state probabilities and outcome model\n", curr[0], pt+1, d->npts);
-//	printf("initp: "); for (i = 0; i < qm->nst; ++i) printf("%f,",hm->initp[MI(pt,i,d->npts)]); printf("\n");
-//	printf("hpars: "); for (i = 0; i<hm->totpars; ++i) printf("%f,",hm->pars[MI(i, d->firstobs[pt], hm->totpars)]); printf("\n");	
+    if (allzero && (qm->nliks==1)) {
+	warning("First observation of %f for subject number %d out of %d is impossible for given initial state probabilities and outcome model\n", curr[0], pt+1, d->npts);
     }
     lweight=0;
     /* Matrix product loop to accumulate the likelihood for subsequent observations */
@@ -689,7 +687,7 @@ double liksimple(msmdata *d, qmodel *qm, cmodel *cm, hmodel *hm)
     double lik=0, contrib=0;
[…]

### 2014-06-24 pkg/src/msm.h

@@ -43,12 +43,12 @@ struct msmdata {
     int *subject;
     double *time;
     double *obs; /* observed state or any other HMM observation */
-    int *firstobs;
     int *obstype;
     int *obstrue;
     int *pcomb;
+    int *firstobs;
 
-    int nobs;
+    int nagg;
     int n;
     int npts;
     int ntrans;
@@ -59,30 +59,30 @@ struct qmodel {
     int nst;
     int npars;
     int nopt;
-    int *ivector;
     double *intens;
     double *dintens;
     int iso;
     int *perm;
     int *qperm;
     int expm;
+    int nliks;
 };
 
 struct cmodel {
     int ncens;
     int *censor;
-    int *censstind;
-    int *censstates;
+    int *states;
+    int *index;
 };
 
 struct hmodel {
     int hidden;
     int *models;
+    int totpars;
     int *npars;
     int *firstpar;
     double *pars;
     double *dpars;
-    int totpars;
     int nopt;
     double *initp;
 };

### 2014-06-24 pkg/tests/testthat/test_models.r

@@ -441,6 +441,15 @@ test_that("efpt.msm",{
     expect_equal(efpt.msm(psor.msm, tostate=c(2,3)), c(10.4237243422138, 0, 0, Inf), tol=1e-05)
 })
 
+test_that("ppass.msm",{
+    pp <- ppass.msm(psor.msm, tot=10)
+    pm <- pmatrix.msm(psor.msm, t=10)
+    expect_equal(pp[,4], pm[,4]) # state 4 is absorbing
+    pp <- ppass.msm(qmatrix=twoway4.q, tot=1000)
+    expect_equal(pp[1,2], 0.5)
+    expect_warning(ppass.msm(qmatrix=twoway4.q, tot=100, ci="normal"), "No fitted model supplied: not calculating confidence intervals")
+})
+
 test_that("score residuals",{
     sres <- scoreresid.msm(psor.msm)
     expect_equal(c(0.0608163009112361, 0.187998750251689, 0.0143302186951471), as.numeric(sres[1:3]), tol=1e-05)

### 2014-06-24 pkg/tests/testthat/test_models_misc.r

@@ -89,6 +89,12 @@ test_that("can't mix ematrix and hcovariates",{
                                      hcovariates=list(~dage + sex, ~dage + sex, ~dage + sex, ~1),
                                      hcovinits = list(c(0.01,0.013), c(0.01,0.013,0.01,0.013), c(0.01,0.013), NULL) ), "hcovariates have been specified, but no hmodel")
 })
+
+test_that("data inconsistent with initprobs/ematrix",{
+    cav2 <- cav
+    cav2$state[c(1,8)] <- 3
+    expect_warning(msm(state ~ years, subject = PTNUM, data = cav2, qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars=TRUE), "First observation .+ is impossible")
+})
           
 test_that("various errors",{
     wrong.e <- "foo"

### 2014-11-18 pkg/R/msm.R
Changes for 1.4.1 before applying qematrix.msm rewrite
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@59 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -88,6 +88,11 @@ msm <- function(formula, subject=NULL, data=list(), qmatrix, gen.inits=FALSE,
     mf <- eval(temp, parent.frame())
     ## remember user-specified names for later (e.g. bootstrap/cross validation)
     attr(mf, "usernames") <- c(state=all.vars(formula[[2]]), time=all.vars(formula[[3]]), subject=as.character(temp$subject), obstype=as.character(substitute(obstype)), obstrue=as.character(temp$obstrue))
+    if (is.factor(mf$"(state)")){
+        if (!all(grepl("^[[:digit:]]+$", as.character(mf$"(state)"))))
+            stop("state variable should be numeric or a factor with ordinal numbers as levels")
+        else mf$"(state)" <- as.numeric(as.character(mf$"(state)"))
+    }
     if (!(hmodel$hidden || emodel$misc))
         msm.check.state(qmodel$nstates, mf$"(state)", cmodel$censor)
     if (is.null(mf$"(subject)")) mf$"(subject)" <- rep(1, nrow(mf))

### 2014-11-18 pkg/R/optim.R

@@ -73,8 +73,13 @@ msm.optim.fixed <- function(p, gr, hessian, msmdata, qmodel, qcmodel, cmodel, hm
 msm.optim.optim <- function(p, gr, hessian, msmdata, qmodel, qcmodel, cmodel, hmodel, ...) {
     optim.args <- list(...)
     if (is.null(optim.args$method))
-        optim.args$method <- if (deriv.supported(hmodel, cmodel)) "BFGS" else "Nelder-Mead"
-#        optim.args$method <- if (length(p$inits)==1) "BFGS" else "Nelder-Mead"
+        optim.args$method <- if (deriv.supported(hmodel, cmodel) || (length(p$inits)==1)) "BFGS" else "Nelder-Mead"
+
+    if (is.null(optim.args$control)) optim.args$control <- list()
+    if (is.null(optim.args$control$fnscale))
+        optim.args$control$fnscale <- lik.msm(p$inits, msmdata=msmdata, qmodel=qmodel, qcmodel=qcmodel,
+                                              cmodel=cmodel, hmodel=hmodel, paramdata=p)
+    
     optim.args <- c(optim.args, list(par=p$inits, fn=lik.msm, hessian=hessian, gr=gr,
                                      msmdata=msmdata, qmodel=qmodel, qcmodel=qcmodel,
                                      cmodel=cmodel, hmodel=hmodel, paramdata=p))
@@ -170,7 +175,7 @@ msm.optim.fisher <- function(p, gr, hessian, msmdata, qmodel, qcmodel, cmodel, h
 }
 
 msm.optim.bobyqa <- function(p, gr, hessian, msmdata, qmodel, qcmodel, cmodel, hmodel, ...) {
-    require(minqa)
+    library(minqa)
     optim.args <- list(...)
     optim.args <- c(optim.args, list(par=p$inits, fn=lik.msm,
                                      msmdata=msmdata, qmodel=qmodel, qcmodel=qcmodel,
@@ -198,7 +203,7 @@ msm.optim.bobyqa <- function(p, gr, hessian, msmdata, qmodel, qcmodel, cmodel, h
 deriv.test <- function(msmdata, qmodel, qcmodel, cmodel, hmodel, p){
     an.d <- deriv.msm(p$inits, msmdata, qmodel, qcmodel, cmodel, hmodel, p)
 
-    if (0){
+#    if (0){
[…]

### 2014-11-18 pkg/src/analyticp.c

@@ -78,7 +78,7 @@ void AnalyticP(Matrix pmat, double t, int nstates, int iso, int *perm, int *qper
 	for (j=0; j<nstates; ++j) {
 	    pmat[MI(i,j,nstates)] = pmat_base[MI(perm[i]-1,perm[j]-1,nstates)];
 	}
-    Free(pmat_base);
+    Free(pmat_base); Free(qmat_base);
 }
 
 

### 2014-11-18 pkg/src/lik.c

@@ -108,8 +108,9 @@ void GetCensored (double obs, cmodel *cm, int *nc, double **states)
     if (cm->ncens == 0)
 	n = 1;
     else {
-	while (!all_equal(obs, cm->censor[k]) && k < cm->ncens)
+	while ((k < cm->ncens) && !all_equal(obs, cm->censor[k])){ 
 	    ++k;
+	}
 	if (k < cm->ncens) {
 	    cens = 1;
 	    n =  cm->index[k+1] - cm->index[k];
@@ -535,7 +536,7 @@ void hmm_deriv(int pt, msmdata *d, qmodel *qm, cmodel *cm, hmodel *hm, Array3 pm
     }
     lp *= -2;
     //    printf("lp=%lf\n",lp);
-    Free(curr); Free(aold); Free(anew); Free(phiold); Free(phinew); Free(xiold); Free(xinew); Free(phiold); Free(phinew); Free(dpok);
+    Free(curr); Free(aold); Free(anew); Free(phiold); Free(phinew); Free(xiold); Free(xinew); Free(dxiold); Free(dxinew); Free(dpok);
 }
 
 
@@ -611,7 +612,7 @@ void hmm_info(int pt, msmdata *d, qmodel *qm, cmodel *cm, hmodel *hm, Array3 pma
 	    }
 	}
     }
-    Free(curr); Free(potential); Free(aold); Free(anew); Free(phiold); Free(phinew); Free(xiold); Free(xinew); Free(phiold); Free(phinew); Free(dpok);
+    Free(curr); Free(potential); Free(anew); Free(aold); Free(phiold); Free(phinew); Free(xinew); Free(xiold); Free(dxiold); Free(dxinew); Free(dpok);
 }
 
 

### 2014-11-18 pkg/src/pijt.c

@@ -391,6 +391,11 @@ void MatrixExpEXPM(double *mat, int *n, double *expmat, double *t,
     else {
 	for (i=0; i<((*n)*(*n)); ++i) {
 	    matt[i] = (*t) * mat[i];
+	    /* Check whether any of the elements of Q have overflowed.  If
+	       so, Fortran eigen function will hang in a infinite loop, so
+	       bail out before this happens.  */
+	    if (!R_FINITE(matt[i]))
+		error("numerical overflow in calculating likelihood\n");
 	}
 	expm(matt, *n, expmat, Ward_2);
     }

### 2014-11-18 pkg/tests/testthat/test_analyticp.r

@@ -68,8 +68,8 @@ test_that("3 state analytic P matrices",{
                                         # 1,2,4
     (three.q <- fixq(rbind(c(0, exp(-3), exp(-6)), c(0, 0, exp(-3)), c(0, 0, 0))))
     sim2.df <- simmulti.msm(sim.df[,1:2], qmatrix=three.q)
-    (sim.mod1 <- msm(state ~ time, subject=subject, data=sim2.df, qmatrix = rbind(c(0, exp(-1), exp(-2)), c(0, 0, exp(-1)), c(0, 0, 0)), analyticp=TRUE))
-    (sim.mod2 <- msm(state ~ time, subject=subject, data=sim2.df, qmatrix = rbind(c(0, exp(-1), exp(-2)), c(0, 0, exp(-1)), c(0, 0, 0)), analyticp=FALSE))
+    (sim.mod1 <- msm(state ~ time, subject=subject, data=sim2.df, qmatrix = rbind(c(0, exp(-1), exp(-2)), c(0, 0, exp(-1)), c(0, 0, 0)), analyticp=TRUE, control=list(fnscale=1)))
+    (sim.mod2 <- msm(state ~ time, subject=subject, data=sim2.df, qmatrix = rbind(c(0, exp(-1), exp(-2)), c(0, 0, exp(-1)), c(0, 0, 0)), analyticp=FALSE, control=list(fnscale=1)))
     expect_equal(sim.mod1$minus2loglik, sim.mod2$minus2loglik)
 
                                         # 1,2,4 (=1,2,6)

### 2014-11-18 pkg/tests/testthat/test_models.r

@@ -32,7 +32,8 @@ test_that("data as global variables",{
     expect_equal(4119.9736299032, cav.msm$minus2loglik)
 })
 
-psor.msm <- msm(state ~ months, subject=ptnum, data=psor, qmatrix = psor.q,  covariates = ~ollwsdrt+hieffusn, constraint = list(hieffusn=c(1,1,1),ollwsdrt=c(1,1,2)))
+psor.msm <- msm(state ~ months, subject=ptnum, data=psor, qmatrix = psor.q,  covariates = ~ollwsdrt+hieffusn, constraint = list(hieffusn=c(1,1,1),ollwsdrt=c(1,1,2)), control=list(fnscale=1))
+
 test_that("psor model: covariates, constraints",{
     expect_equal(1114.89946121717, psor.msm$minus2loglik, tol=1e-06)
     expect_equal(0.0959350004999946, psor.msm$Qmatrices$baseline[1,2], tol=1e-06)
@@ -66,6 +67,19 @@ test_that("exact transition times with death, should be same",{
     (msmtest5 <- msm(state ~ time, qmatrix = fiveq,  subject = ptnum, data = bos, death=5, obstype=rep(2, nrow(bos)), fixedpars=TRUE)) # no warning, inconsistently
     expect_equal(3057.85781916437, msmtest5$minus2loglik, tol=1e-06)
 })
+
+
+cav$statefac <- factor(cav$state)
+cav$statefac2 <- factor(cav$state, labels=c("none","mild","severe","death"))
+test_that("factors as states, death",{
+    cav.msm <- msm( statefac ~ years, subject=PTNUM, data = cav,
+                   qmatrix = twoway4.q, death = TRUE, fixedpars=TRUE,
+                   method="BFGS", control=list(trace=5, REPORT=1))
+    expect_equal(4908.81676837903, cav.msm$minus2loglik)
+    expect_error(msm( statefac2 ~ years, subject=PTNUM, data = cav,
+                   qmatrix = twoway4.q, death = TRUE, fixedpars=TRUE), "state variable should be numeric or a factor with ordinal numbers as levels")
+})
+
 
 test_that("factor covariates with factor() in formula",{
     ## Should be no need for users to do this. factors should already be identified as such in the data frame
@@ -322,6 +336,12 @@ test_that("qratio.msm",{
[…]

### 2014-11-18 pkg/tests/testthat/test_models_misc.r

@@ -105,4 +105,5 @@ test_that("various errors",{
     expect_error(misc.msm <- msm(state ~ years, subject = PTNUM, data = cav, qmatrix = oneway4.q, ematrix=wrong.e, death = 4, fixedpars=TRUE),"Number of rows and columns of ematrix should be equal")
     wrong.e <- cbind(c(0,1), c(0,2))
     expect_error(misc.msm <- msm(state ~ years, subject = PTNUM, data = cav, qmatrix = oneway4.q, ematrix=wrong.e, death = 4, fixedpars=TRUE),"Dimensions of qmatrix and ematrix should be the same")
+    expect_warning(msm(state ~ years, subject = PTNUM, data = cav, qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars=TRUE, gen.inits=TRUE), "gen.inits not supported for hidden Markov models, ignoring")
 })

### 2014-12-10 pkg/src/doc/msm-manual.Rnw
Changes for 1.4.1 including phase-type models
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@65 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 
+%% version <- 1.3
+
 <<echo=FALSE>>=
 version <- gsub("Version: +", "",
-                packageDescription("msm", lib.loc="../../..")$Version)
+               packageDescription("msm", lib.loc=c("../..",.libPaths()))$Version)
 @
@@
 
-\tableofcontents
-\clearpage
-
 \section{Multi-state models}
@@
 
-Figure~\ref{fig:multi} illustrates a multi-state model in continuous
+Figure \ref{fig:multi} illustrates a multi-state model in continuous
 time. Its four states are labelled {\bf 1, 2, 3, 4}.  At a time $t$
@@
 An example data set, taken from monitoring a set of heart transplant
-recipients, is provided with \Rpackage{msm}.  Sharples
+recipients, is provided with \Rpackage{msm}.  (Note: since \Rpackage{msm}
+version 1.3, the command \Rfunction{data(cav)} is no longer needed to
+load the data --- it is now ``lazy-loaded'' when required). Sharples
 \etal \cite{my:cav} studied the progression of coronary allograft
@@
 occasions on which a patient was observed in state 1 followed by state
-3, the underlying model specifies that the patient must have passed
-through state 2 in between.  If your data represent the exact and
+3, we can still have $q_{13}=0$. The underlying model specifies that the patient must have passed
+through state 2 in between, rather than jumping straight from 1 to 3.  If your data represent the exact and
 complete transition times of the process, then you must specify
@@
 <<>>=
-twoway4.q  <-  rbind ( c(0, 0.25, 0, 0.25),
+Q  <-  rbind ( c(0, 0.25, 0, 0.25),
                c(0.166, 0, 0.166, 0.166),
@@
 the mean sojourn time $1 / q_{rr}$ can be estimated as $T_r / n_r$.
-Thus, $n_{rs} / T_r$ is a crude estimate of $q_{rs}$. The \Rpackage{msm}
-package provides a function \Rfunction{crudeinits.msm} for calculating
-initial values in this way (for non-hidden Markov models).
+Thus, $n_{rs} / T_r$ is a crude estimate of $q_{rs}$.
+
+Such default initial values can be used by supplying
+\Rfunarg{gen.inits=TRUE} in the call to \Rfunction{msm} below, along
+with a \Rfunarg{qmatrix} whose non-zero entries represent the allowed
+transitions of the model.  Alternatively the function \Rfunction{crudeinits.msm}
+could be used to get this matrix of initial values explicitly as follows.
+These methods are only available for non-hidden Markov models.
 
 <<>>=
-twoway4.crude.q  <- crudeinits.msm(state ~ years, PTNUM, data=cav, qmatrix=twoway4.q)
+Q.crude  <- crudeinits.msm(state ~ years, PTNUM, data=cav,
+                                   qmatrix=Q)
 @
@@
 q_{14} = 0.25$, $q_{21} = q_{23} = q_{24} = 0.166$, $q_{32} = q_{34} =
-0.25$, and the initial value matrix \Robject{twoway4.q} shown above,
+0.25$, and the initial value matrix \Robject{Q} shown above,
 which we now use to fit the model.
-
-%% TODO explain about gen.inits=TRUE here
 
@@
 arguments.  For our running example, we have defined a data set
-\Robject{cav}, a matrix \Robject{twoway4.q} indicating the allowed
+\Robject{cav}, a matrix \Robject{Q} indicating the allowed
 transitions, and initial values. We are ready to run \Rfunction{msm}.
@@
 cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                    qmatrix = twoway4.q, death = 4 )
+                    qmatrix = Q, death = 4)
 @
@@
 @
-for more options to control the optimisation.  When completed, the
-\Rfunction{msm} function returns a value.  This value is a list of the
-important results of the model fitting, including the parameter
-estimates and their covariances.  To keep these results for
-post-processing, we store them in an R object, here called
-\Robject{cav.msm}.  When running several similar \Rfunction{msm} models,
-it is recommended to store the respective results in
-informatively-named objects.
+for more options to control the optimisation.
+\footnote{Note that since version 1.3.2, \Rfunarg{method=''BFGS''}, is the default optimisation algorithm in \Rfunction{msm}, since it can use analytic derivatives, which are available for most models.}
+
+When completed, the \Rfunction{msm} function returns a value.  This
+value is a list of the important results of the model fitting,
+including the parameter estimates and their covariances.  To keep
+these results for post-processing, we store them in an R object, here
+called \Robject{cav.msm}.  When running several similar
+\Rfunction{msm} models, it is recommended to store the respective
+results in informatively-named objects.
 
@@
 
-From the estimated intensity matrix, we see patients are three times
-as likely to develop symptoms than die without symptoms (first row).
+From the estimated intensities, we see patients are three times
+as likely to develop symptoms than die without symptoms (transitions from state 1).
 After disease onset (state 2), progression to severe symptoms (state
@@
 cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                     qmatrix = twoway4.q, death = 4, covariates = ~ sex)
-@
-
-The \Robject{msm} object will now include the estimated covariate
-effects and their confidence intervals.
+                     qmatrix = Q, death = 4, covariates = ~ sex)
+@
+
+Printing the \Robject{msm} object now displays the estimated covariate
+effects and their confidence intervals (note since version 1.3.2 these
+are \emph{hazard ratios} $\exp(\beta_{rs})$, not \emph{log hazard ratios} $\beta_{rs}$ as in previous
+versions).
 
@@
 @
-Comparing the estimated log-linear effects of sex and their standard
-errors, we see that the disease onset rate is smaller for females,
-whereas none of the other effects are significant. The first matrix shown in the output of printing
-\Robject{cavsex.msm} is the estimated transition intensity matrix
+The sizes of the
+confidence intervals for some of the hazard ratios suggests there is no information in the data about 
+the corresponding covariate effects, leading to a likelihood that is a flat function of these parameters, 
+and this model should be simplified. The first column shown in the output
+is the estimated transition intensity matrix
 $q_{rs}(z) = q_{rs}^{(0)} \exp(\beta_{rs}^T z)$ with the covariate $z$
@@
 
-Different transition rates may be easily modelled on different
-covariates by specifying a named list of formulae as the
-\Rfunarg{covariates} argument.  Each element of the list has a name
-identifying the transition.  In the model below, the transition rate
-from state 1 to state 2 and the rate from state 3 to state 4 are each
-modelled on sex as a covariate, but no other intensities have
-covariates on them.
+Since \Rpackage{msm} version 1.2.3, different transition rates may be
+easily modelled on different covariates by specifying a named list of
+formulae as the \Rfunarg{covariates} argument.  Each element of the
+list has a name identifying the transition.  In the model below, the
+transition rate from state 1 to state 2 and the rate from state 1 to
+state 4 are each modelled on sex as a covariate, but no other
+intensities have covariates on them.
 
@@
 cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                     qmatrix = twoway4.q, death = 4,
-                     covariates = list("1-2" = ~ sex, "3-4" = ~sex) )
+                  qmatrix = Q, death = 4,
+                  covariates = list("1-2" = ~ sex, "1-4" = ~sex) )
 @
@@
 cav3.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                   qmatrix = twoway4.q, death = 4,
-                   covariates = ~ sex, constraint = list(sex=c(1,2,3,1,2,3,2)) )
+                qmatrix = Q, death = 4,
+                covariates = ~ sex,
+                constraint = list(sex=c(1,2,3,1,2,3,2)) )
 @
@@
 cav4.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                qmatrix = twoway4.q, death = 4,
+                qmatrix = Q, death = 4,
                 control = list(trace=2, REPORT=1),
@@
 
+\item[Expected number of visits] The function \Rfunction{envisits.msm}
+  estimates the expected number of visits to a state, computed in a
+  similar way to the total length of stay. See its help page for
+  further details.
+
 \item[Ratio of transition intensities] The function
@@
   effect on the transition intensities. 95\% confidence limits are
-  computed by assuming normality of the log-effect. For example, for
-  model 2 with female sex as a covariate, the following hazard ratios
-  show more clearly that the only transition on which the effect of
-  sex is significant at the 5\% level is the 1-2 transition.
+  computed by assuming normality of the log-effect.
 
@@
 model that produced \Robject{cav.msm}, all objects used in the
-original model fit (for example, in this case, \Robject{twoway4.q})
+original model fit (for example, in this case, \Robject{Q})
 must be in the working environment.  Otherwise, \Rfunction{boot.msm}
@@
 \begin{description}
+
+\item[Model simplification]
+
+  Firstly, make sure there are not too many parameters in the model.  There may not be enough
+  information in the data on a certain transition rate.  It is
+  recommended to count all the pairs of transitions between states in
+  successive observation times, making a frequency table of previous
+  state against current state (function \Rfunction{statetable.msm}),
+  and do this for any subgroups defining covariates.
+  Although the data are a series of snapshots of a continuous-time
+  process, and the actual transitions take place in between the
+  observation times, this type of table may still be helpful.  If
+  there are not many observed `transitions' from state 2 to state 4,
+  for example, then the data may be insufficient to estimate $q_{24}$.
+
+  For a staged disease model (Figure \ref{fig:disease}), the number of
+  disease states should be low enough that all transition rates can be
+  estimated.  Consecutive states of disease severity should be merged
+  if necessary.  If it is realistic, consider applying constraints on
+  the intensities or the covariate effects so that the parameters are
+  equal for certain transitions, or zero for certain transitions.
+
+  Be careful to use a observation scheme and transition matrix
+  appropriate to your data (see
+  Section \ref{sec:arbitr-observ-times}). By default, \Rfunction{msm}
+  assumes that the data represent snapshots of the process, and the
+  true state is unknown between observation times.  In such
+  circumstances, it is rarely feasible to estimate an intensity matrix
+  with \emph{instantaneous} transitions allowed between every pair of
+  states.  This would be easier if the complete course of the process
+  is known \Rfunarg{(exacttimes = TRUE)} in the call to
+  \Rfunction{msm}.  Understand the difference between
+  \emph{instantaneous} and \emph{interval} transitions - although
+  individuals may be in state 1 at time $t_r$, and state 3 at time
+  $t_{r+1}$, that doesn't mean that instantaneous transitions from 1
+  to 3 should be permitted.
+
 \item[Initial values] Make sure that a sensible set of initial values
@@
 
-\item[Scaling] It is often necessary to apply a scaling factor to
-  normalise the likelihood (\Rfunarg{fnscale}), or certain individual
-  parameters \Rfunarg{(parscale)}.  This may prevent overflow or
-  underflow problems within the optimisation.  For example, if the value
-  of the -2 $\times$ log-likelihood is around 5000, then the following
-  option leads to an minimisation of the -2 $\times$ log-likelihood on
-  an approximate unit scale: \Rfunarg{control = list(fnscale = 5000)}
+\item[Scaling] It is often necessary to apply a scaling factor to normalise the likelihood (\Rfunarg{fnscale}), or certain individual parameters \Rfunarg{(parscale)}.
+This may prevent overflow or underflow problems within the optimisation.
+For example, if the value of the -2 $\times$ log-likelihood is around 5000, then the following option leads to an minimisation of the -2 $\times$ log-likelihood on an approximate unit scale: \Rfunarg{control = list(fnscale = 5000)}.
+  % Though since version 1.4.1, \Rfunarg{fnscale} is applied automatically using the likelihood at the initial values, unless the user has already supplied it.
+%    If not provided by the user, \code{control=list(fnscale = a)} is
+%    applied automatically to normalise the optimisation, where \code{a}
+%    is the minus twice log likelihood at the initial values.
 
@@
 
-\item[Choice of algorithm] By default, \Rfunction{msm} uses a
-  Nelder-Mead algorithm which does not use analytic derivatives of the
-  objective function.   For potentially greater speed and accuracy,
-  consider using a quasi-Newton algorithm such as
-  the ``BFGS'' method of \Rfunction{optim}, which can make use of
-  analytic derivatives, for example, \Rfunarg{method = ``BFGS'', use.deriv = TRUE}.
-  Or use the \Rfunction{nlm} R function via \Rfunarg{msm(..., opt.method = "nlm" , ...)}
-  Note also the Fisher scoring method available for non-hidden Markov
-  models for panel data, via  \Rfunarg{msm(..., opt.method = "fisher", ...)},
-  is expected to be faster than the generic methods, but less robust to
-  bad initial values.
+\item[Choice of algorithm] By default, since version 1.3.2,
+  \Rfunction{msm} uses the BFGS method of \Rfunction{optim}, which
+  makes use of analytic derivatives.  Analytic derivatives are
+  available for all models in msm, apart from hidden Markov models
+  with unknown initial state probabilities, misclassification models
+  with equality constraints on misclassification probabilities, and
+  truncated or measurement-error outcome distributions.  This speeds
+  up optimisation.  Though alternative algorithms are available such
+  as \Rfunarg{method = ``CG''}.  Or use the \Rfunction{nlm} R function
+  via \Rfunarg{msm(..., opt.method = "nlm" , ...)}  Note also the
+  Fisher scoring method available for non-hidden Markov models for
+  panel data, via \Rfunarg{msm(..., opt.method = "fisher", ...)}, is
+  expected to be faster than the generic methods, but less robust to
+  bad initial values.  Or since version 1.3.2, msm can also use
+  \Rfunarg{method=``bobyqa''} from the \Rpackage{minqa} package, a fast
+  derivative-free method.
 
@@
   \Rfunarg{initprobs} would be sensible.
-
-\item[Model simplification]
-
-  If none of these numerical adjustments lead to convergence, then the
-  model is probably over-complicated.  There may not be enough
-  information in the data on a certain transition rate.  It is always
-  recommended to count all the pairs of transitions between states in
-  successive observation times, making a frequency table of previous
-  state against current state (function \Rfunction{statetable.msm}).
-  Although the data are a series of snapshots of a continuous-time
-  process, and the actual transitions take place in between the
-  observation times, this type of table may still be helpful.  If
-  there are not many observed `transitions' from state 2 to state 4,
-  for example, then the data may be insufficient to estimate $q_{24}$.
-
-  For a staged disease model (Figure \ref{fig:disease}), the number of
-  disease states should be low enough that all transition rates can be
-  estimated.  Consecutive states of disease severity should be merged
-  if necessary.  If it is realistic, consider applying constraints on
-  the intensities or the covariate effects so that the parameters are
-  equal for certain transitions, or zero for certain transitions.
-
-  Be careful to use a observation scheme and transition matrix
-  appropriate to your data (see
-  Section \ref{sec:arbitr-observ-times}). By default, \Rfunction{msm}
-  assumes that the data represent snapshots of the process, and the
-  true state is unknown between observation times.  In such
-  circumstances, it is rarely feasible to estimate an intensity matrix
-  with \emph{instantaneous} transitions allowed between every pair of
-  states.  This would be easier if the complete course of the process
-  is known \Rfunarg{(exacttimes = TRUE)} in the call to
-  \Rfunction{msm}.  Understand the difference between
-  \emph{instantaneous} and \emph{interval} transitions - although
-  individuals may be in state 1 at time $t_r$, and state 3 at time
-  $t_{r+1}$, that doesn't mean that instantaneous transitions from 1
-  to 3 should be permitted.
 
@@
 options(digits=2)
-pearson.msm(cav.msm, timegroups=2, transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
+pearson.msm(cav.msm, timegroups=2,
+            transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
 @
@@
 
-To model observed states with misclassification, we define an matrix
+To model observed states with misclassification, we define a matrix
 \Rfunarg{ematrix} indicating the states that can be misclassified.
@@
 <<>>=
-oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
+Qm <- rbind(c(0, 0.148, 0, 0.0171),
             c(0, 0, 0.202, 0.081),
@@
 cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                   qmatrix = oneway4.q, ematrix = ematrix, death = 4,
-                   obstrue = firstobs, method="BFGS")
+                   qmatrix = Qm, ematrix = ematrix, death = 4,
+                   obstrue = firstobs)
 cavmisc.msm
@@
 cavmiscsex.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                          qmatrix = oneway4.q, ematrix = ematrix, death = 4,
-                          misccovariates = ~sex, obstrue=firstobs, method="BFGS")
+                      qmatrix = Qm, ematrix = ematrix,
+                      death = 4, misccovariates = ~sex,
+                      obstrue=firstobs)
 @
@@
 @
-The large confidence interval for the 1/2 misclassification suggests
-that women are no more or less likely than men to have a false
-positive angiography scan. On the other hand, women have slightly more
+The large confidence interval for the odds ratio for 1/2 misclassification suggests
+there is no information in the data about the difference between genders in 
+the false positive rates for angiography. On the other hand, women have slightly more
 false negatives.
@@
 @
-The confidence intervals for the estimates for women are higher, since
+The confidence intervals for the estimates for women are wider, since
 there are only 87 women in this set of 622 patients.
@@
 <<>>=
-pearson.msm(cavmisc.msm, timegroups=2, transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
+pearson.msm(cavmisc.msm, timegroups=2,
+            transitions=c(1,2,3,4,5,6,7,8,9,9,9,10))
 @
@@
                 hcovinits = list(-8, -8, NULL),
-                hconstraint = list(acute = c(1,1)), method="BFGS")
+                hconstraint = list(acute = c(1,1)))
 
@@
 Printing the \Rclass{msm} object \Rfunarg{fev1.msm} shows estimates
-and confidence intervals for the transition intensity matrix and the
+and confidence intervals for the transition intensities and the
 hidden Markov model parameters.  The estimated within-state means of
@@
                 control=list(maxit=10000), center=TRUE)
-
-fev2.msm
-
-sojourn.msm(fev2.msm)
-@
+\end{Scode}
 
@@
 text(min2 + 500, 50, "STATE 2")
-@
+\end{Scode}
+
+\includegraphics{figures/fev_viterbi}
 
@@
 
-<<>>=
-oneway4.q <- rbind(c(0, 0.148, 0, 0.0171),
+\begin{Scode}
+Qm <- rbind(c(0, 0.148, 0, 0.0171),
             c(0, 0, 0.202, 0.081),
@@
 cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                     hmodel = list (hmmCat(c(0.9, 0.1, 0, 0)), hmmCat(c(0.1, 0.8, 0.1, 0)), hmmCat(c(0, 0.1, 0.9, 0)), hmmIdent(4)),
-                     qmatrix = oneway4.q, obstrue=firstobs, death = 4, method="BFGS")
+                     hmodel = list (hmmCat(c(0.9, 0.1, 0, 0)),
+                                    hmmCat(c(0.1, 0.8, 0.1, 0)),
+                                    hmmCat(c(0, 0.1, 0.9, 0)),
+                                    hmmIdent(4)),
+                     qmatrix = Qm, obstrue=firstobs, death = 4)
 cavmisc.msm
-@
+\end{Scode}
 
@@
 
+\item (Optionally) Write a C function to compute the derivatives of
+  the probability density with respect to the parameters, and put this
+  in \texttt{src/hmmderiv.c}, with a declaration in
+  \texttt{src/hmm.h}.  Then add the model to \texttt{DHMODELS} in
+  \texttt{lik.c} (in the same position as in \texttt{HMODELS}) and
+  \texttt{.msm.HMODELS.DERIV} in \texttt{R/constants.R}.  This will
+  generally double the speed of maximum likelihood estimation, but analytic 
+  derivatives will not be available for all distributions.
+
 \item Update the documentation (\texttt{man/hmm-dists.Rd}) and the
@@
 \end{Scode}
-If your distribution may be of interest to others, ask me nicely
+If your distribution may be of interest to others, ask me
 (\texttt{chris.jackson@mrc-bsu.cam.ac.uk}) to include it in a future

### 2014-12-10 pkg/R/boot.R

@@ -193,6 +193,13 @@ ppass.ci.msm <- function(x, qmatrix, tot, start, covariates="mean", piecewise.ti
     list(L=matrix(ci[1,],ncol=nst,dimnames=di), U=matrix(ci[2,],ncol=nst, dimnames=di))
 }
 
+phasemeans.ci.msm <- function(x, covariates="mean", cl=0.95, B=1000, cores=NULL, ...) {
+    p.list <- boot.msm(x, function(x)phasemeans.msm(x=x, covariates=covariates, ci="none", ...), B=B, cores=cores)
+    p.array <- array(unlist(p.list), dim=c(dim(p.list[[1]]), length(p.list)))
+    p.ci <- apply(p.array, c(1,2), function(x)(quantile(x, c(0.5 - cl/2, 0.5 + cl/2))))
+    aperm(p.ci, c(2,3,1))
+}
+
 expected.ci.msm <- function(x,
                             times=NULL,
                             timezero=NULL,
@@ -236,11 +243,11 @@ normboot.msm <- function(x, stat, B=1000) {
     for (i in 1:B) {
         x.rep <- x
         x.rep$paramdata$params <- params[i,]
-        output <- msm.form.output("intens", x.rep$qmodel, x.rep$qcmodel, x.rep$paramdata)
-        x.rep$Qmatrices <- output$Matrices
+        output <- msm.form.output(x.rep, "intens")
+        x.rep$Qmatrices <- output$Qmatrices
         if (x$emodel$misc) {
-            output <- msm.form.output("misc", x.rep$emodel, x.rep$ecmodel, x.rep$paramdata)
-            x.rep$Ematrices <- output$Matrices
+            output <- msm.form.output(x.rep, "misc")
+            x.rep$Ematrices <- output$Ematrices
             names(x.rep$Ematrices)[1] <- "logitbaseline"
         }
         sim.stat[[i]] <- stat(x.rep)
@@ -341,4 +348,11 @@ expected.normci.msm <- function(x,
     res <- list(aperm(e.tab.ci, c(2,3,1)),  aperm(e.perc.ci, c(2,3,1)))
     names(res) <- c("Expected", "Expected percentages")
     res
+}
+
+phasemeans.normci.msm <- function(x, covariates="mean", cl=0.95, B=1000, ...) {
[…]

### 2014-12-10 pkg/R/draic.R

@@ -110,17 +110,26 @@ draic.msm <- function(msm.full,msm.coarse,
 
 drlcv.msm <- function(msm.full,msm.coarse,tl=0.95,cores=NULL,verbose=TRUE,outfile=NULL)
 {
+    if (!inherits(msm.full, "msm")) stop("Expected \"msm.full\" to be a msm model")
+    if (!inherits(msm.coarse, "msm")) stop("Expected \"msm.coarse\" to be a msm model")
+    if (msm.full$hmodel$hidden || msm.coarse$hmodel$hidden) stop("Hidden Markov models not supported")
+    if (msm.full$cmodel$ncens > 0 || msm.coarse$cmodel$ncens >0) stop("Models with censoring not supported")
     ## Reconstruct datasets used to fit each model as dataframes
-    msm.full.data <- reconstruct.data(msm.full$data$mf)
-    msm.coarse.data <- reconstruct.data(msm.coarse$data$mf)
+    msm.full.data <- reconstruct.data(model.frame(msm.full))
+    msm.coarse.data <- reconstruct.data(model.frame(msm.coarse))
     un <- attr(msm.full.data, "usernames")
     subj <- un["subject"]; st <- un["state"]
     unc <- attr(msm.coarse.data, "usernames")
     subjc <- unc["subject"]; stc <- unc["state"]
-
+    if (nrow(msm.full.data) != nrow(msm.coarse.data)) stop("Full dataset has ",nrow(msm.full.data)," rows, but coarse dataset has ",nrow(msm.coarse.data)," rows, these should be equal.")
+    
     ## Determine map from full to coarsened data
     coarsening.ematrix <- matrix(0,nrow=msm.full$qmodel$nstates, ncol=msm.full$qmodel$nstates)
-    map <- unique(cbind(msm.full.data[,st], msm.coarse.data[,stc]))
+    map <- unique(cbind(full=msm.full.data[,st], coarse=msm.coarse.data[,stc]))
+    if (!all(tapply(map[,"coarse"], map[,"full"], function(x)length(unique(x)))==1)){
+        message("Observed the following state mappings in the data:")
+        print(map); stop("Coarse state space not an aggregation of the the full state space")
+    }
     coarsening.ematrix[map] <- 1
 
     ## Initial state occupancy probabilities assumed equal across possible states
[…]

### 2014-12-10 pkg/R/hmm.R

@@ -373,14 +373,16 @@ in.range <- function(x, interval, strict=FALSE) {
         ( (x >= interval[1]) & (x <= interval[2]) )
 }
 
-## TODO could this be merged with msm.initprobs2mat?
-msm.form.initprobs <- function(hmodel, npts){
-    ## if (!is.null(phase.states)) {
-    ##     hmodel$initpmat <- matrix(0, nrow=msmdata.obs$npts, ncol=qmodel$nstates)
-    ##     initstate <- msmdata.obs$state[!duplicated(msmdata.obs$subject)]
-    ##     hmodel$initpmat[cbind(1:npts, match(initstate, pars))] <- 1
-    ##     hmodel$initpbysubj <- TRUE
-    ## }
+msm.form.initprobs <- function(hmodel, mf){
+    npts <- attr(mf,"npts")
+    if (!is.null(hmodel$phase.states)) {
+        hmodel$initprobs <- matrix(0, nrow=npts, ncol=hmodel$nstates)
+        initstate <- mf$"(state)"[!duplicated(mf$"(subject)")]
+        hmodel$initprobs[cbind(1:npts, match(initstate, hmodel$pars))] <- 1
+        if (hmodel$est.initprobs)
+            warning("Not estimating initial state occupancy probabilities: assuming everyone starts at first phase")
+        hmodel$est.initprobs <- FALSE
+    }
     if (!hmodel$est.initprobs) {
         if (is.matrix(hmodel$initprobs)) {
             if (nrow(hmodel$initprobs) != npts)

### 2014-12-10 pkg/R/msm.R

@@ -6,7 +6,9 @@ msm <- function(formula, subject=NULL, data=list(), qmatrix, gen.inits=FALSE,
                 qconstraint=NULL, econstraint=NULL, initprobs = NULL,
                 est.initprobs=FALSE, initcovariates = NULL, initcovinits = NULL,
                 death = FALSE, exacttimes = FALSE, censor=NULL,
-                censor.states=NULL, pci=NULL, cl = 0.95, fixedpars = NULL, center=TRUE,
+                censor.states=NULL, pci=NULL, phase.states=NULL,
+                phase.inits = NULL, # TODO merge with inits eventually
+                cl = 0.95, fixedpars = NULL, center=TRUE,
                 opt.method="optim", hessian=NULL, use.deriv=TRUE,
                 use.expm=TRUE, analyticp=TRUE, na.action=na.omit, ...)
 {
@@ -22,32 +24,39 @@ msm <- function(formula, subject=NULL, data=list(), qmatrix, gen.inits=FALSE,
         }
         else warning("gen.inits not supported for hidden Markov models, ignoring")
     }
-    qmodel <- msm.form.qmodel(qmatrix, qconstraint, analyticp, use.expm)
+    qmodel <- msm.form.qmodel(qmatrix, qconstraint, analyticp, use.expm, phase.states)
 
-### MISCLASSIFICATION MODEL
-    if (!is.null(ematrix)) {
-        emodel <- msm.form.emodel(ematrix, econstraint, initprobs, est.initprobs, qmodel)
+    if (!is.null(phase.states)) {
+        pm <- msm.form.pmodel(qmodel, phase.states, phase.inits, qconstraint, analyticp, use.expm)
+        qmodel <- pm$qmodel; hmodel <- pm$hmodel
+        emodel <- list(misc=FALSE, npars=0, ndpars=0)
     }
-    else emodel <- list(misc=FALSE, npars=0, ndpars=0)
-    if (emodel$npars==0) emodel <- list(misc=FALSE, npars=0, ndpars=0) # user supplied degenerate ematrix with no misclassification
+    else { 
+### MISCLASSIFICATION MODEL
+        if (!is.null(ematrix)) {
+            emodel <- msm.form.emodel(ematrix, econstraint, initprobs, est.initprobs, qmodel)
+        }
+        else emodel <- list(misc=FALSE, npars=0, ndpars=0)
[…]

### 2014-12-10 pkg/R/outputs.R

@@ -10,6 +10,7 @@ qmatrix.msm <- function(x, covariates="mean", sojourn=FALSE, ci=c("delta","norma
     if ((cl < 0) || (cl > 1)) stop("expected cl in [0,1]")
     se <- lse <- fixed <- numeric(ni)
     logest <- x$Qmatrices$logbaseline
+    if (is.null(x$QmatricesFixed)) x <- msm.form.output(x, "intens") # for back-compat with pre 1.4.1 model objects
     fixed <- x$QmatricesFixed$logbaseline
     for (i in seq_len(nc)) {
         logest <- logest + x$Qmatrices[[i+1]] * covlist[[i]]
@@ -98,6 +99,7 @@ ematrix.msm <- function(x, covariates="mean", ci=c("delta","normal","bootstrap",
     if ((cl < 0) || (cl > 1)) stop("expected cl in [0,1]")
     se <- lse <- numeric(ni)
     logest <- x$Ematrices$logitbaseline
+    if (is.null(x$EmatricesFixed)) x <- msm.form.output(x, "misc") # for back-compat with pre 1.4.1 model objects
     fixed <- x$EmatricesFixed$logitbaseline
     for (i in seq_len(nc)) {
         logest <- logest + x$Ematrices[[i+1]] * covlist[[i]]
@@ -515,6 +517,7 @@ mattotrans <- function(x, matrix, lower, upper, fixed, keep.diag=FALSE, intmisc=
 
 msm.form.qoutput <- function(x, covariates="mean", cl=0.95, digits=4, ...){
     qbase <- qmatrix.msm(x, covariates=covariates, cl=cl)
+    if (is.null(x$QmatricesFixed)) x <- msm.form.output(x, "intens") # for back-compat with pre 1.4.1 model objects
     y <- mattotrans(x, qbase$estimates, qbase$L, qbase$U, qbase$fixed, keep.diag=TRUE)
     ret <- data.frame(base=y)
     fres <- matrix("", nrow=nrow(y), ncol=x$qcmodel$ncovs+1)
@@ -538,6 +541,7 @@ msm.form.qoutput <- function(x, covariates="mean", cl=0.95, digits=4, ...){
 
 msm.form.eoutput <- function(x, covariates="mean", cl=0.95, digits=4, ...){
     ebase <- ematrix.msm(x, covariates=covariates, cl=cl)
+    if (is.null(x$EmatricesFixed)) x <- msm.form.output(x, "misc") # for back-compat with pre 1.4.1 model objects
[…]

### 2014-12-10 pkg/R/phase.R

@@ -1,0 +1,217 @@
+msm.form.pmodel <- function(qmodel, phase.states, inits, qconstraint, analyticp, use.expm){
+    if (any(!(phase.states %in% 1:qmodel$nstates))) stop("phase.states should be in 1,...,",qmodel$nstates)
+    markov.states <- setdiff(1:qmodel$nstates, phase.states)
+    nst <- 2*length(phase.states) + length(markov.states)
+    reps <- rep(1, qmodel$nstates)
+    reps[phase.states] <- 2
+    pars <- rep(1:qmodel$nstates, reps)
+
+    ## user supplied inits: list of list(trans=scalar, exit=(nto x 2) matrix), one for
+    ## each phased state in order, where nto is number of
+    ## destination states from that phased state TODO document that
+    ### TODO document that inits for q are as usual for trans out of non-phased states, and ignored for trans out of phased states
+    if (!is.null(inits)){
+        if (!is.list(inits)) stop("phase.inits should be a list")
+        if (!length(inits)==length(phase.states))
+            stop(sprintf("phase.inits of length %d, but there are %d phased states", length(inits), length(phase.states)))
+        for (i in seq(along=inits)){
+            if (!length(inits[[i]])==2) stop(sprintf("phase.inits[[%d]] list of length %d, should be 2", i, length(inits[[i]])))
+            if (is.null(names(inits[[i]]))) names(inits[[i]]) <- c("trans","exit")
+            if (is.vector(inits[[i]]$trans)) inits[[i]]$trans <- matrix(inits[[i]]$trans, nrow=1)
+            if (is.vector(inits[[i]]$exit)) inits[[i]]$exit <- matrix(inits[[i]]$exit, nrow=1)
+            if (length(inits[[i]]$trans)!=1) stop(sprintf("phase.inits[[%d]]$trans of length %d, should be 1", i, length(inits[[i]]$trans)))
+            if (ncol(inits[[i]]$exit)!=2) stop(sprintf("phase.inits[[%d]]$exit has %d columns, should be 2", i, length(inits[[i]]$exit)))
+        }
+    }
+    ## TODO warn for identical exit rates
+
[…]

### 2014-12-10 pkg/src/lik.c

@@ -1069,18 +1069,9 @@ void pmax(double *x, int n, int *maxi)
 
 /* Calculates the most likely path through underlying states */
 
-void Viterbi(msmdata *d, qmodel *qm, cmodel *cm, hmodel *hm, double *fitted)
+void Viterbi(msmdata *d, qmodel *qm, cmodel *cm, hmodel *hm, double *fitted, double *pstate)
 {
     int i, j, tru, k, kmax, obs, nc = 1;
-/*
-    double *pmat = (double *) S_alloc((qm->nst)*(qm->nst), sizeof(double));
-    int *ptr = (int *) S_alloc((d->n)*(qm->nst), sizeof(int));
-    double *lvold = (double *) S_alloc(qm->nst, sizeof(double));
-    double *lvnew = (double *) S_alloc(qm->nst, sizeof(double));
-    double *lvp = (double *) S_alloc(qm->nst, sizeof(double));
-    double *curr = (double *) S_alloc (qm->nst, sizeof(double));
-    double *pout = (double *) S_alloc(qm->nst, sizeof(double));
-*/
 
     double *pmat = Calloc((qm->nst)*(qm->nst), double);
     int *ptr = Calloc((d->n)*(qm->nst), int);
@@ -1089,8 +1080,9 @@ void Viterbi(msmdata *d, qmodel *qm, cmodel *cm, hmodel *hm, double *fitted)
     double *lvp = Calloc(qm->nst, double);
     double *curr = Calloc (qm->nst, double);
     double *pout = Calloc(qm->nst, double);
+//    double *pstate = Calloc((d->n)*(qm->nst), double);
 
-    double dt;
+    double dt, sump;
     double *qmat, *hpars;
 
     i = 0;
@@ -1115,7 +1107,9 @@ void Viterbi(msmdata *d, qmodel *qm, cmodel *cm, hmodel *hm, double *fitted)
 	    lvold[k] = log(hm->initp[MI(0, k, d->npts)]);
       }
     }
-
+    for (k = 0; k < qm->nst; ++k)
+    	pstate[MI(i,k,d->n)] = exp(lvold[k]); 
+		  
     for (i = 1; i <= d->n; ++i)
 	{
[…]

### 2015-09-14 pkg/R/msm.R
Changes for 1.5.3
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@73 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@ -5,7 +5,7 @@ msm <- function(formula, subject=NULL, data=list(), qmatrix, gen.inits=FALSE,
                 hcovariates = NULL, hcovinits = NULL, hconstraint = NULL, hranges=NULL,
                 qconstraint=NULL, econstraint=NULL, initprobs = NULL,
                 est.initprobs=FALSE, initcovariates = NULL, initcovinits = NULL,
-                death = FALSE, exacttimes = FALSE, censor=NULL,
+                deathexact = NULL, death = NULL, exacttimes = FALSE, censor=NULL,
                 censor.states=NULL, pci=NULL, phase.states=NULL,
                 phase.inits = NULL, # TODO merge with inits eventually
                 cl = 0.95, fixedpars = NULL, center=TRUE,
@@ -63,6 +63,7 @@ msm <- function(formula, subject=NULL, data=list(), qmatrix, gen.inits=FALSE,
     else emodel <- list(misc=FALSE, npars=0, ndpars=0, nipars=0, nicoveffs=0)
 
 ### EXACT DEATH TIMES. Logical values allowed for backwards compatibility (TRUE means final state has exact death time, FALSE means no states with exact death times)
+    if (!is.null(deathexact)) death <- deathexact
     dmodel <- msm.form.dmodel(death, qmodel, hmodel)  # returns death, ndeath,
     if (dmodel$ndeath > 0 && exacttimes) warning("Ignoring death argument, as all states have exact entry times")
 
@@ -416,8 +417,6 @@ msm.check.ematrix <- function(ematrix, nstates)
         stop("Not all elements of ematrix are between 0 and 1")
     invisible()
 }
-
-## FIXME should imatrix be only 
 
 msm.form.emodel <- function(ematrix, econstraint=NULL, initprobs=NULL, est.initprobs, qmodel)
 {
@@ -843,6 +842,7 @@ msm.form.dmodel <- function(death, qmodel, hmodel)
 {
     nstates <- qmodel$nstates
     statelist <- if (nstates==2) "1, 2" else if (nstates==3) "1, 2, 3" else paste("1, 2, ... ,",nstates)
+    if (is.null(death)) death <- FALSE
     if (is.logical(death) && death==TRUE)
         states <- nstates
     else if (is.logical(death) && death==FALSE)
@@ -881,16 +881,8 @@ msm.form.cmodel <- function(censor=NULL, censor.states=NULL, qmatrix)
[…]

### 2015-09-14 pkg/src/lik.c

@@ -14,6 +14,7 @@
 #include <Rdefines.h>
 #define NODEBUG
 #define NOVITDEBUG0
+#define NOVITDEBUG
 #define NODERIVDEBUG
 
 /* MUST KEEP THIS IN SAME ORDER AS .msm.HMODELPARS IN R/constants.R */
@@ -1201,10 +1202,11 @@ void Viterbi(msmdata *d, qmodel *qm, cmodel *cm, hmodel *hm, double *fitted, dou
     double *pfwd = Calloc((d->n)*(qm->nst), double);
     double *pbwd = Calloc((d->n)*(qm->nst), double);
 
-    double dt, pall, psum;
+    double dt, logpall, psum;
     double *qmat, *hpars;
     double *ucfwd = Calloc(d->n, double);
     double *ucbwd = Calloc(d->n, double);
+    double *maxpu = Calloc(d->n, double);
 
     i = 0;
     if (d->obstrue[i]) {
@@ -1305,14 +1307,15 @@ void Viterbi(msmdata *d, qmodel *qm, cmodel *cm, hmodel *hm, double *fitted, dou
 //		    fitted[obs] = (d->obstrue[obs] ? d->obs[MI(0,obs,d->nout)]-1 : kmax);
 		    fitted[obs] = (d->obstrue[obs] ? d->obstrue[obs]-1 : kmax);
 
-		    pall = 0;  // compute full likelihood.  
+		    logpall = 0;  // compute full likelihood.  
 		    ucbwd[obs] = 0;
 		    for (k = 0; k < qm->nst; ++k){
 			pbwd[MI(obs,k,d->n)] = 1;
-			pall += pfwd[MI(obs,k,d->n)]*exp(ucfwd[obs]);
+			logpall += pfwd[MI(obs,k,d->n)];
 		    }
+		    logpall = log(logpall) + ucfwd[obs];
 		    for (k = 0; k < qm->nst; ++k){
-			pstate[MI(obs,k,d->n)] = exp(log(pfwd[MI(obs,k,d->n)]) + log(pbwd[MI(obs,k,d->n)]) - log(pall) + ucfwd[obs]);
[…]

### 2015-09-14 pkg/src/pijt.c

@@ -304,8 +304,9 @@ void Eigen(Matrix mat, int n, vector revals, vector ievals, Matrix evecs, int *e
 	/* Check whether any of the elements of Q have overflowed.  If
 	   so, Fortran eigen function will hang in a infinite loop, so
 	   bail out before this happens.  */
-	if (!R_FINITE(mat[i]))
+	if (!R_FINITE(mat[i])) { 
 	    error("numerical overflow in calculating likelihood\n");
+	}
 	temp[i] = mat[i];
     }
 #ifdef _USE_LINPACK_EIGEN_
@@ -382,7 +383,7 @@ void MatrixExpR(double *mat, int *n, double *expmat, double *t,
 
 void MatrixExpEXPM(double *mat, int *n, double *expmat, double *t,
 		int *method, int *iso, int *perm, int *qperm,
-		int *degen){
+		   int *degen, int *err){
     int i;
     int nsq = (*n)*(*n);
     double *matt = Calloc(nsq, double);
@@ -394,8 +395,15 @@ void MatrixExpEXPM(double *mat, int *n, double *expmat, double *t,
 	    /* Check whether any of the elements of Q have overflowed.  If
 	       so, Fortran eigen function will hang in a infinite loop, so
 	       bail out before this happens.  */
-	    if (!R_FINITE(matt[i]))
+	/* Could we return loglik = -Inf instead of halting by setting
+	   error code?  return all zeros for pmat in that case?
+	   Doesn't help convergence with test case in test/rory.r: lik
+	   stays at zero.
+	 */
+	    if (!R_FINITE(matt[i])){
+//		*err = -1; return;
 		error("numerical overflow in calculating likelihood\n");
+	    }
 	}
 	expm(matt, *n, expmat, Ward_2);
     }
@@ -408,7 +416,7 @@ void MatrixExpEXPM(double *mat, int *n, double *expmat, double *t,
 
 void Pmat(Matrix pmat, double t, Matrix qmat, int nstates, int exacttimes, int iso, ivector perm, ivector qperm, int use_expm)
 {
-    int i,j,method=MEXP_PADE,degen=0;
+    int i,j,method=MEXP_PADE,degen=0,err=0;
     double pii;
     if (exacttimes) {
 	for (i=0; i<nstates; ++i) {
[…]

### 2015-09-14 pkg/tests/testthat/test_datasumm.r

@@ -4,7 +4,6 @@ test_that("statetable.msm", {
     stab <- statetable.msm(state, PTNUM, data=cav)
     expect_that(stab, equals(structure(c(1367L, 46L, 4L, 204L, 134L, 13L, 44L, 54L, 107L, 148L, 48L, 55L), .Dim = 3:4, .Dimnames = structure(list(from = c("1","2", "3"), to = c("1", "2", "3", "4")), .Names = c("from", "to")), class = "table")))
     expect_equal(as.numeric(stab), c(1367, 46, 4, 204, 134, 13, 44, 54, 107, 148, 48, 55))
-    expect_error(statetable.msm("state","PTNUM"))
     expect_error(statetable.msm(state,PTNUM), "not found")
     stabc <- statetable.msm(state, PTNUM, cav.cens)
     expect_equal(as.numeric(stabc), c(1367, 46, 4, 204, 134, 13, 44, 54, 107, 127, 40, 34, 21, 8, 21))

### 2015-09-14 pkg/tests/testthat/test_models.r

@@ -2,28 +2,28 @@ context("msm simple model likelihoods")
 
 test_that("simple model, death",{
     cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                   qmatrix = twoway4.q, death = TRUE, fixedpars=TRUE,
+                   qmatrix = twoway4.q, deathexact = TRUE, fixedpars=TRUE,
                    method="BFGS", control=list(trace=5, REPORT=1))
     expect_equal(4908.81676837903, cav.msm$minus2loglik)
 })
 
 test_that("simple model, no death",{
     cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                   qmatrix = twoway4.q, death = FALSE, fixedpars=TRUE)
+                   qmatrix = twoway4.q, deathexact = FALSE, fixedpars=TRUE)
     expect_equal(4833.00640639644, cav.msm$minus2loglik)
 })
 
 test_that("simple model, covariates",{
     cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                   qmatrix = twoway4.q, death = TRUE, fixedpars=TRUE,
+                   qmatrix = twoway4.q, deathexact = TRUE, fixedpars=TRUE,
                    covariates = ~ sex, covinits = list(sex=rep(0.01, 7)))
     expect_equal(4909.17147259115, cav.msm$minus2loglik)
 })
 
 test_that("autogenerated inits reproduce crudeinits",{
     cinits <- crudeinits.msm(state ~ years, PTNUM, data=cav, qmatrix=twoway4.q)
-    expect_equal(msm( state ~ years, subject=PTNUM, data = cav, qmatrix = cinits, death = TRUE, fixedpars=TRUE)$minus2loglik,
-                 msm( state ~ years, subject=PTNUM, data = cav, qmatrix=twoway4.i, gen.inits=TRUE, death = TRUE, fixedpars=TRUE)$minus2loglik)
+    expect_equal(msm( state ~ years, subject=PTNUM, data = cav, qmatrix = cinits, deathexact = TRUE, fixedpars=TRUE)$minus2loglik,
+                 msm( state ~ years, subject=PTNUM, data = cav, qmatrix=twoway4.i, gen.inits=TRUE, deathexact = TRUE, fixedpars=TRUE)$minus2loglik)
 })
 
 test_that("data as global variables",{
@@ -62,9 +62,9 @@ test_that("exact transition times using obstype vector of all 2",{
 })
 
[…]

### 2015-09-14 pkg/tests/testthat/test_models_misc.r

@@ -1,10 +1,10 @@
 context("msm misclassification model likelihoods")
 
 test_that("cav misclassification model with no covariates",{
-    misc.msm <- msm(state ~ years, subject = PTNUM, data = cav, qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars=TRUE)
+    misc.msm <- msm(state ~ years, subject = PTNUM, data = cav, qmatrix = oneway4.q, ematrix=ematrix, deathexact = 4, fixedpars=TRUE)
     expect_equal(4296.9155995778, misc.msm$minus2loglik, tol=1e-06)
     miscnew.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                       qmatrix = oneway4.q, death = 4, fixedpars=TRUE,
+                       qmatrix = oneway4.q, deathexact = 4, fixedpars=TRUE,
                        hmodel=list(
                        hmmCat(prob=c(0.9, 0.1, 0, 0)),
                        hmmCat(prob=c(0.1, 0.8, 0.1, 0)),
@@ -14,17 +14,17 @@ test_that("cav misclassification model with no covariates",{
 })
 
 test_that("cav misclassification model with covariates on transition rates",{
-    misccov.msm <- msm(state ~ years, subject = PTNUM, data = cav, qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars = TRUE, covariates = ~ sex, covinits=list(sex=rep(0.1, 5)))
+    misccov.msm <- msm(state ~ years, subject = PTNUM, data = cav, qmatrix = oneway4.q, ematrix=ematrix, deathexact = 4, fixedpars = TRUE, covariates = ~ sex, covinits=list(sex=rep(0.1, 5)))
     expect_equal(4299.38058878142, misccov.msm$minus2loglik, tol=1e-06)
 })
 
 test_that("cav misclassification model with covariates on misclassification probabilities",{
     misccov.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                       qmatrix = oneway4.q, ematrix=ematrix, death = 4, fixedpars=TRUE,
+                       qmatrix = oneway4.q, ematrix=ematrix, deathexact = 4, fixedpars=TRUE,
[…]

### 2016-02-19 pkg/src/doc/msm-manual.Rnw
Changes for 1.6.1
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/msm@77 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

@@
 cav.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                    qmatrix = Q, death = 4)
+                    qmatrix = Q, deathexact = 4)
 @
@@
 before death the state of the patient is unknown.  Thus we specify
-\Rfunarg{death = 4}, to indicate to \Rfunction{msm} that the
+\Rfunarg{deathexact = 4}, to indicate to \Rfunction{msm} that the
 entry times into state 4 are observed in this manner.
@@
 five states, and states 4 and 5 were two competing causes of death with
-times recorded exactly in this way, then we would specify \Rfunarg{death =
+times recorded exactly in this way, then we would specify \Rfunarg{deathexact =
   c(4,5)}.
@@
 times of transition, ``exact death times'', or a mixture of these. See the
-\Rfunarg{obstype} argument to \Rfunction{msm}.  Note you do not always have to specify \Rfunarg{death},
-even if there are states which correspond to death.  This is
-only used for states whose entry time is observed in this specific way.
+\Rfunarg{obstype} argument to \Rfunction{msm}.
 
@@
 cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                     qmatrix = Q, death = 4, covariates = ~ sex)
+                     qmatrix = Q, deathexact = 4, covariates = ~ sex)
 @
@@
 cavsex.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                  qmatrix = Q, death = 4,
+                  qmatrix = Q, deathexact = 4,
                   covariates = list("1-2" = ~ sex, "1-4" = ~sex) )
@@
 cav3.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                qmatrix = Q, death = 4,
+                qmatrix = Q, deathexact = 4,
                 covariates = ~ sex,
@@
 cav4.msm <- msm( state ~ years, subject=PTNUM, data = cav,
-                qmatrix = Q, death = 4,
+                qmatrix = Q, deathexact = 4,
                 control = list(trace=2, REPORT=1),
@@
 for the process history is difficult as the process is only observed
-through a series of snapshots.  For a multi-state model with one-way
-progression through states, and frequent observations, we may be able
-to estimate the time spent in each state by each individual.
+through a series of snapshots.  Semi-Markov 
+models may in principle be fitted to this type of data using phase-type distributions.  Since
+version 1.4.1 the \Rfunarg{phase.states} option to \Rfunction{msm} can be used to 
+define some phase-type models.  See \Rfunction{help(msm)} for further details.
 
@@
 cavmisc.msm <- msm(state ~ years, subject = PTNUM, data = cav,
-                   qmatrix = Qm, ematrix = ematrix, death = 4,
+                   qmatrix = Qm, ematrix = ematrix, deathexact = 4,
                    obstrue = firstobs)
@@
                       qmatrix = Qm, ematrix = ematrix,
-                      death = 4, misccovariates = ~sex,
+                      deathexact = 4, misccovariates = ~sex,
                       obstrue=firstobs)
@@
 fev1.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q,
-                death=3, hmodel=hmodel1,
+                deathexact=3, hmodel=hmodel1,
                 hcovariates=list(~acute, ~acute, NULL),
@@
 fev2.msm <- msm(fev ~ days, subject=ptnum, data=fev, qmatrix=three.q,
-                death=3, hmodel=hmodel2,
+                deathexact=3, hmodel=hmodel2,
                 hcovariates=list(~acute, ~acute, NULL),
@@
                                     hmmIdent(4)),
-                     qmatrix = Qm, obstrue=firstobs, death = 4)
+                     qmatrix = Qm, obstrue=firstobs, deathexact = 4)
 cavmisc.msm
 \end{Scode}
+
+\subsubsection{Hidden Markov models with multivariate outcomes}
+
+Since version 1.5.2, \Rpackage{msm} can fit models where at each time
+point, there are multiple outcomes generated conditionally on a single
+hidden Markov state.  The outcomes must be independent conditionally
+on the hidden state, but they may be generated from the same or
+different univariate distributions.
+
+See \Rfunction{help(hmmMV)} for detailed documentation and a worked
+example.
 

### 2016-02-19 pkg/R/hmm.R

@@ -64,7 +64,7 @@ msm.form.hmodel <- function(hmodel, hconstraint=NULL, initprobs=NULL, est.initpr
           hmod <- msm.form.mvhmodel(hmodel)
       } else hmod <- msm.form.univhmodel(hmodel)
       hmod <- c(list(hidden=TRUE, nstates=nst, fitted=FALSE, nipars=nipars,
-                     initprobs=initprobs, est.initprobs=est.initprobs),
+                     initprobs=initprobs, est.initprobs=est.initprobs, ematrix=FALSE),
                 hmod)
       class(hmod) <- "hmodel"
       hmod
@@ -226,13 +226,14 @@ msm.emodel2hmodel <- function(emodel, qmodel)
           labels <- .msm.HMODELS[models]
           locpars <- which(plabs == rep(.msm.LOCPARS[labels], npars))
           hmod <- list(hidden=TRUE, fitted=FALSE, nstates=nst, models=models, labels=labels,
+                       ematrix=TRUE, ## remember if obtained from ematrix, since could change meaning of obstrue
                        nout=rep(1, nst), mv=FALSE,
                        npars=npars, totpars=sum(npars), locpars=locpars,
                        pars=pars, plabs=plabs, parstate=parstate, firstpar=firstpar, nipars=emodel$nipars, initprobs=emodel$initprobs, est.initprobs=emodel$est.initprobs)
           hmod$constr <- msm.econstr2hconstr(emodel$constr, hmod)
       }
       else {
-          hmod <- list(hidden=FALSE, fitted=FALSE, models=rep(0, qmodel$nstates), npars=0, ndpars=0)
+          hmod <- list(hidden=FALSE, fitted=FALSE, ematrix=FALSE, models=rep(0, qmodel$nstates), npars=0, ndpars=0)
       }
       class(hmod) <- "hmodel"
       hmod

### 2016-02-19 pkg/R/msm.R

@@ -38,7 +38,6 @@ msm <- function(formula, subject=NULL, data=list(), qmatrix, gen.inits=FALSE,
         emodel <- msm.form.emodel(ematrix, econstraint, initprobs, est.initprobs, qmodel)
     }
     else emodel <- list(misc=FALSE, npars=0, ndpars=0)
-    if (emodel$npars==0) emodel <- list(misc=FALSE, npars=0, ndpars=0) # user supplied degenerate ematrix with no misclassification
 
 ### GENERAL HIDDEN MARKOV MODEL
     if (!is.null(hmodel)) {
@@ -60,7 +59,10 @@ msm <- function(formula, subject=NULL, data=list(), qmatrix, gen.inits=FALSE,
     if (emodel$misc) {
         hmodel <- msm.emodel2hmodel(emodel, qmodel)
     }
-    else emodel <- list(misc=FALSE, npars=0, ndpars=0, nipars=0, nicoveffs=0)
+    else {
+        emodel <- list(misc=FALSE, npars=0, ndpars=0, nipars=0, nicoveffs=0)
+        hmodel$ematrix <- FALSE
+    }
 
 ### EXACT DEATH TIMES. Logical values allowed for backwards compatibility (TRUE means final state has exact death time, FALSE means no states with exact death times)
     if (!is.null(deathexact)) death <- deathexact
@@ -460,8 +462,13 @@ msm.form.emodel <- function(ematrix, econstraint=NULL, initprobs=NULL, est.initp
     else
         constr <- seq(length=npars)
     ndpars <- if(npars>0) max(constr) else 0
-    emodel <- list(misc=TRUE, npars=npars, nstates=nstates, imatrix=imatrix, ematrix=ematrix, inits=inits,
-                   constr=constr, ndpars=ndpars, nipars=nipars, initprobs=initprobs, est.initprobs=est.initprobs)
+
+    nomisc <- isTRUE(all.equal(as.vector(diag(ematrix)),rep(1,nrow(ematrix)))) # degenerate ematrix with no misclassification: all 1 on diagonal
+    if (nomisc)
+        emodel <- list(misc=FALSE, npars=0, ndpars=0)
+    else                                     
+        emodel <- list(misc=TRUE, npars=npars, nstates=nstates, imatrix=imatrix, ematrix=ematrix, inits=inits,
+                       constr=constr, ndpars=ndpars, nipars=nipars, initprobs=initprobs, est.initprobs=est.initprobs)
[…]

### 2016-02-19 pkg/src/lik.c

@@ -176,12 +176,12 @@ void GetOutcomeProb(double *pout, double *outcome, int nc, int nout, double *hpa
 		}
 	    } else {
 		pout[i] = 0;
-		if (hm->hidden && nc == 1){ /* "state" data contain an actual observation. 
+		if (hm->hidden && nc == 1 && !hm->ematrix){ /* "state" data contain an actual observation. 
 				 get its distribution here conditional on the supplied true state */
 		    if (obstrue == i+1){
 			pout[i] = (HMODELS[hm->models[i]])(outcome[0], &(hpars[hm->firstpar[i]]));
 		    } 
-		} else {  /* "state" data contain a censor indicator */
+		} else {  /* "state" data contain the true state (for hm->ematrix with obstrue) or censor indicator */
 		    for (j=0; j<nc; ++j){
 			if ((int) outcome[j] == i+1)
 			    pout[i] = 1;
@@ -216,7 +216,7 @@ void GetDOutcomeProb(double *dpout, /* qm->nst x hm->nopt */
     for (i=0; i<qm->nst; ++i) {
 	for (l=0; l<hm->nopt; ++l)
 	    dpout[MI(i,l,qm->nst)] = 0;
-	if (hm->hidden && (!obstrue || (obstrue==(i+1)))) {
+	if (hm->hidden && (!obstrue || (obstrue==(i+1) && !hm->ematrix))) {
 	    if (nout > 1) {  /* multivariate outcomes. Censored states not supported			
 				TODO.  This is fiddlier than first thought.
 				not considered what hm->dpars should be, particularly with constraints  */ 
@@ -1245,7 +1245,6 @@ void Viterbi(msmdata *d, qmodel *qm, cmodel *cm, hmodel *hm, double *fitted, dou
     double *qmat, *hpars;
     double *ucfwd = Calloc(d->n, double);
     double *ucbwd = Calloc(d->n, double);
-    double *maxpu = Calloc(d->n, double);
 
     i = 0;
     if (d->obstrue[i]) {
@@ -1537,6 +1536,7 @@ SEXP msmCEntry(SEXP do_what_s, SEXP mf_agg_s, SEXP mf_s, SEXP auxdata_s, SEXP qm
 
     hm.hidden = list_int(hmodel_s,"hidden");
     hm.mv = list_int(hmodel_s,"mv");
+    hm.ematrix = list_int(hmodel_s,"ematrix");
[…]

### 2016-02-19 pkg/src/msm.h

@@ -79,6 +79,7 @@ struct cmodel {
 struct hmodel {
     int hidden;
     int mv;
+    int ematrix;
     int *models;
     int totpars;
     int *npars;

### 2016-02-19 pkg/tests/testthat/test_models.r

@@ -155,11 +155,14 @@ test_that("piecewise constant intensities with pci",{
                     pci = c(5), covinits = list("timeperiod[5,Inf)"=rep(0.01,7)),
                     )
     expect_equal(4906.01423796805, cav5.msm$minus2loglik, tol=1e-06)
+
     cav10.msm <- msm( state ~ years, subject=PTNUM, data = cav,
                      qmatrix = twoway4.q, deathexact = TRUE, pci = c(5,10), fixedpars=TRUE,
                      covinits = list("timeperiod[5,10)"=rep(0.01,7), "timeperiod[10,Inf)"=rep(0.01,7)),
                      )
     expect_equal(4905.61646158639, cav10.msm$minus2loglik, tol=1e-06)
+    
+
 })
 
 test_that("piecewise constant intensities with pci, cut points outside data",{

### 2016-02-19 pkg/tests/testthat/test_models_misc.r

@@ -41,6 +41,11 @@ test_that("misclassification model with no misclassification reduces to simple",
     expect_equal(nomisc.msm$minus2loglik, simple.msm$minus2loglik)
 })
 
+test_that("misclassification model with obstrue",{
+    misc.msm <- msm(state ~ years, subject = PTNUM, data = cav, qmatrix = oneway4.q, ematrix=ematrix, deathexact = 4, fixedpars=TRUE, obstrue=firstobs)
+    expect_equal(4165.84711809003, misc.msm$minus2loglik, tol=1e-06)
+})
+
 test_that("misclassification model with exact times",{
     miscexact.msm <- msm(state ~ years, subject = PTNUM, data = cav, qmatrix = oneway4.q, ematrix=ematrix, exacttimes=TRUE, fixedpars=TRUE)
     expect_equal(4864.14764195147, miscexact.msm$minus2loglik, tol=1e-06)
